# (PART) Introduction to Reproducible R {-} 

# Why be reproducible 
***

It is in your own self-interest.  

## An example workflow

Taking First Steps ...

**Step 1: Some Ideas and Data**

$X_{1} \to Y_{1}$

* You copy some data into a spreadsheet
* do some calculations and tables the same spreadsheet
* some other analysis from here and there, using this software and that.

**Step 2: Persuing the lead for a week or two**

* you beef up the data you got
* add some other types of data 
* copy in a spreadsheet data, manually aggregate
* do some more calculations and tables, same as before




Then, a Little Way Down the Road ...

**1 month later, someone asks about another factor:** $X_{2}$

* You repeat **Step 2** with some data on $X_{2}$.
* The details from your "point and click" method are a bit fuzzy.

It takes a little time, but you successfully redo the analysis.


**4 months later, someone asks about another factor:** $X_{3}\to Y_{1}$

* You again repeat **Step 2** with some data on $X_{3}$.
* You're pretty sure
  * it's the latest version of the spreadsheet.
  * none of tables your tried messed up the order of the rows or columns.

It takes more time -- the data processing was not transparent.



**6 months later, you want to explore:** $X_{2} \to Y_{2}$.

* You found out Excel had some bugs in it's statistical calculations (see e.g., https://biostat.app.vumc.org/wiki/pub/Main/TheresaScott/StatsInExcel.TAScot.handout.pdf).




**2 years later, you want to replicate:** $\{ X_{1}, X_{2}, X_{3} \} \to Y_{1}$


* A rival has proposed an alternative theory. Their idea doesn't actually make any sense, but their visuals are better and statistics are more sophisticated.
* You don't even have that computer anymore.
* A collaborator who handled the data on $X_{2}$ has moved on.



## An alternative workflow

Suppose you decided to code what you did beginning with Step 2.

**It doesn't take much time to update or replicate your results.**

* Your computer runs for 2 hours and reproduces the figures and tables.
(You wrote your big calculations to use multiple cores and this saved 6 hours--each time.)
* You decided to add some more data, and it adds almost no time.
* You see the exact steps you took and found an error
(glad you found it before publication!)


**Your results are transparent and easier to build on.**

* You easily see that not much has changed with the new data.
* You try out a new plot you found in *The Visual Display of Quantitative Information*, by Edward Tufte.
  * It's not a standard plot, but google answers most of your questions.
  * Tutorials help avoid bad practices, such as plotting 2D data as a 3D object (see e.g., https://clauswilke.com/dataviz/no-3d.html).
* You try out an obscure statistical approach that's hot in your field.
  * it doesn't make the paper, but you have some confidence that candidate issue isn't a big problem



## Why R and R-Markdown

Good for complex stats, concise figures, and coherent organization.

* Built and developed by applied statisticians for statistics.
* Used by many in academia and industry.


Good for reproducible research 

* http://www.r-bloggers.com/the-reproducibility-crisis-in-science-and-prospects-for-r/
* http://fmwww.bc.edu/GStat/docs/pointclick.html
* https://github.com/qinwf/awesome-R\#reproducible-research
* A Guide to Reproducible Code in Ecology and Evolution
* https://biostat.app.vumc.org/wiki/pub/Main/TheresaScott/ReproducibleResearch.TAScott.handout.pdf


Good for teaching

* https://doi.org/10.1080/00220485.2019.1618765
* https://doi.org/10.1002/jae.657


Good for getting a job

* As a student, think about labour demand. R skills, unlike other purely academic software, is something future employers use and want. Do more of your own research on this to understand how much to invest.


## Types of Projects

Homework reports are the smallest and probably first document you create. We will create little homework reports using R markdown that are almost entirely self-contained (showing both code and output). To do this, you will need to install [Pandoc](http://pandoc.org) on your computer.

Posters and presentations are another important type of scientific document. R markdown is good at creating both of these, and actually *very* good with some additional packages. So we will also use [flexdashboard](https://pkgs.rstudio.com/flexdashboard/) for posters and [beamer]( https://bookdown.org/yihui/rmarkdown/beamer-presentation.html) for presentions. Since beamer is a pdf output, you will need to install [TinyTex](https://yihui.org/tinytex/), which can be done within R
```{r, eval=F}
install.packages('tinytex')
tinytex::install_tinytex()  # install TinyTeX
```

Install any required packages
```{r, eval=FALSE}
## Packages for Rmarkdown
install.packages("knitr")
install.packages("rmarkdown")
install.packages("bookdown")

## Other packages used in this primer
install.packages("plotly")
install.packages("sf")
```


Note that

* R is our software
* Rstudio is our GUI
* R-markdown is our document

R and R markdown are also both languages: R studio interprets R code to produce statistics, R studio interprets R markdown code to produce pretty documents which contain both writing and statistics. You should already be a bit familiar with R, but not necessarily R Markdown. So to get you started, you should first read and work through this document, and then recreate it yourself as a litte tutorial. Before hopping into programming, however, understand why you will learn to program in both R and R-markdown languages




# Small Scale Projects
***

## An R Code Chunk

Save the following code as `CodeChunk.R`
```{r, eval=F}
sum_squared <- function(x1, x2) {
	y <- (x1 + x2)^2
	return(y)
} 

x <- c(0,1,3,10,6)
sum_squared(x[1], x[3])
sum_squared(x, x[2])
sum_squared(x, x[7])
sum_squared(x, x) 
```

**Clean the workspace** In the right panels, manually cleanup

* save the code as *MyFirstCode.R*
* clear the environment and history (use the broom in top right panel)
* clear unsaved plots (use the broom in bottom right panel)

    
**Replicate** using the grahical user interface (GUI) while in Rstudio either using a "point-click" or "console" method

*GUI: point-click*
click 'Source > Source as a local job' on top right

*GUI: console*
into the console on the bottom left, enter
```{r, eval=FALSE}
source('MyFirstCode.R')
```


**CLI Alternatives** *(Skippable)* There are also alternative ways to replicate via the command line interface (CLI) after opening a terminal
 
*CLI: console*
```{bash, eval=FALSE}
Rscript -e "source('CodeChunk.R')"
```

*CLI: direct*

```{bash, eval=FALSE}
Rscript CodeChunk.R
```

Note that you can open a new terminal in RStudio in the top bar by
clicking 'tools > terminal > new terminal'



## An R Markdown Document

### Copy/Paste Example
<!-- 
**Clean workspace**
Delete any temporary files which you do not want (or start a fresh session).

(for example *summarytable_example.txt* and *plot_example.pdf* and section *Data analysis examples: custom figures*)
-->



**Download** 

See [DataScientism.html](https://jadamso.github.io/Rbooks/DataScientism.html)

Download source file from [DataScientism.Rmd](https://jadamso.github.io/Rbooks/DataScientism.Rmd)


**Replicate**

You can now create the primers by opening the Rstudio GUI and then point-and-click.

Alternatively, you can use the console to run
```{r, eval=FALSE}
rmarkdown::render('DataScientism.Rmd')
```

### A Homework Example {.tabset}
Below is a template of what homework questions (and answers) look like. Create an .Rmd from scratch that produces a similar looking .html file.

**Question 1**
Simulate 100 random observations of the form $y=x\beta+\epsilon$ and plot the relationship. Plot and explore the data interactively via plotly, https://plotly.com/r/line-and-scatter/. Then play around with different styles, https://www.r-graph-gallery.com/13-scatter-plot.html, to best express your point.

*Answer*
I simulate $400$ observations for $\epsilon \sim 2\times N(0,1)$ and $\beta=4$, as seen in this single chunk. Notice an upward trend.
```{r answer1, message=FALSE, warning=FALSE}
n <- 100
E <- rnorm(n)
X <- seq(n)
Y <- 4*X + 2*E

library(plotly)
plot_ly( data=data.frame(X=X,Y=Y), x=~X, y=~Y)
```

**Question 2**
Verify the definition of a line segment for points $A=(0,3), B=(1,5)$ using a $101 \times 101$ grid. Recall a line segment is all points $s$ that have $d(s, A) + d(s, B) = d(A, B)$.

*Answer* 
```{r answer2, message=FALSE, warning=FALSE}
library(sf)
s_1 <- c(0,3)
s_2 <- c(1,5)
Seg1 <- st_linestring( rbind(s_1,s_2) )
grid_pts <- expand.grid(
    x=seq(s_1[1],s_2[1], length.out=101),
    y=seq(s_1[2],s_2[2], length.out=101)
)

Seg1_dist <- dist( Seg1 )
grid_pts$dist <- apply(grid_pts, 1, function(s){
    dist( rbind(s,s_1) ) + dist( rbind(s,s_2) ) })
grid_pts$seg <- grid_pts$dist == Seg1_dist

D_point_seg <- st_multipoint( as.matrix(grid_pts[grid_pts$seg==T,1:2]) ) 
D_point_notseg <- st_multipoint( as.matrix(grid_pts[grid_pts$seg==F,1:2]) ) 

plot(Seg1)
points(D_point_notseg, col=2, pch='.')
points(D_point_seg)
box()
```

## Posters and Slides

See

* [DataScientism_Slides.pdf](https://jadamso.github.io/Rbooks/DataScientism_Slides.pdf)
* [DataScientism_Poster.html](https://jadamso.github.io/Rbooks/DataScientism_Poster.html)

And download source files

* [DataScientism_Slides.Rmd](https://jadamso.github.io/Rbooks/DataScientism_Slides.Rmd)
* [DataScientism_Poster.Rmd](https://jadamso.github.io/Rbooks/DataScientism_Poster.Rmd)



## Getting help w/ R Markdown

For more guidance on how to create Rmarkdown documents, see

* https://github.com/rstudio/cheatsheets/blob/main/rmarkdown.pdf
* https://cran.r-project.org/web/packages/rmarkdown/vignettes/rmarkdown.html
* http://rmarkdown.rstudio.com
* https://bookdown.org/yihui/rmarkdown/
* https://bookdown.org/yihui/rmarkdown-cookbook/
* https://dept.stat.lsa.umich.edu/~jerrick/courses/stat701/notes/rmarkdown.html
* An Introduction to the Advanced Theory and Practice of Nonparametric Econometrics. Raccine 2019. Appendices B \& D.
* https://rmd4sci.njtierney.com/using-rmarkdown.html

If you are still lost, try one of the many online tutorials (such as these)

* https://www.rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf
* https://github.com/adam-p/markdown-here/wiki/Markdown-Cheatsheet
* https://www.neonscience.org/resources/learning-hub/tutorials/rmd-code-intro
* https://m-clark.github.io/Introduction-to-Rmarkdown/
* https://www.stat.cmu.edu/~cshalizi/rmarkdown/
* http://math.wsu.edu/faculty/xchen/stat412/HwWriteUp.Rmd
* http://math.wsu.edu/faculty/xchen/stat412/HwWriteUp.html
* https://holtzy.github.io/Pimp-my-rmd/
* https://ntaback.github.io/UofT_STA130/Rmarkdownforclassreports.html
* https://crd150.github.io/hw_guidelines.html
* https://r4ds.had.co.nz/r-markdown.html
* http://www.stat.cmu.edu/~cshalizi/rmarkdown
* http://www.ssc.wisc.edu/sscc/pubs/RFR/RFR_RMarkdown.html
* http://kbroman.org/knitr_knutshell/pages/Rmarkdown.html




# Medium Scale Projects
***

As you scale up to a medium sized project, however, you will have to be more organized.

Medium sized projects should have their own `Project` folder on your computer with files, subdirectories with files, and subsubdirectories with files. It should look like this
```
Project
    └── README.txt
    └── /Code
        └── MAKEFILE.R
        └── RBLOCK_001_DataClean.R
        └── RBLOCK_002_Figures.R
        └── RBLOCK_003_ModelsTests.R
        └── RBLOCK_004_Robust.R
        └── /Logs
            └── MAKEFILE.Rout
    └── /Data
        └── /Raw
            └── Source1.csv
            └── Source2.shp
            └── Source3.txt
        └── /Clean
            └── AllDatasets.Rdata
            └── MainDataset1.Rds
            └── MainDataset2.csv
    └── /Output
        └── MainFigure.pdf
        └── AppendixFigure.pdf
        └── MainTable.tex
        └── AppendixTable.tex
    └── /Writing
        └── /TermPaper
            └── TermPaper.tex
            └── TermPaper.bib
            └── TermPaper.pdf
        └── /Slides
            └── Slides.Rmd
            └── Slides.html
            └── Slides.pdf
        └── /Poster
            └── Poster.Rmd
            └── Poster.html
            └── Poster.pdf
        └── /Proposal
            └── Proposal.Rmd
            └── Proposal.html
            └── Proposal.pdf
```



## MAKEFILE

There are two main meta-files

* `README.txt` overviews the project structure and what the codes are doing
* `MAKEFILE` explicitly describes and executes all codes. 


If all code is written with the same program, the makefile can be written in that programs code: `MAKEFILE.R`, which looks like
```{r, eval=FALSE}

### Project Structure
home_dir    <- path.expand("~/Desktop/Project/")

data_dir    <- paste0(home_dir, "Data/")
data_dir_r  <- paste0(data_dir, "Raw/")
data_dir_c  <- paste0(data_dir, "Clean/")

out_dir  <- paste0(hdir, "Output/")

code_dir  <- paste0(hdir, "Code/")


### Execute Codes
### libraries are loaded within each RBLOCK
set.wd( code_dir )
source( "RBLOCK_001_DataClean.R" )
source( "RBLOCK_002_Figures.R" )
source( "RBLOCK_003_ModelsTests.R" )
source( "RBLOCK_004_Robust.R" )
```


If some folders or files are not created, you can do this within R
```{r, eval =FALSE}
# create directory called 'Data'
dir.create('Data')

# list the files and directories
list.files(recursive=TRUE, include.dirs=TRUE)
```


## Logging/Sinking

You can then execute the makefile within R and log the output. Either by 

1. Inserting some code that logs/sinks the output 
```{r, eval=FALSE}

### Project Structure
home_dir    <- path.expand("~/Desktop/Project/")

data_dir    <- paste0(home_dir, "Data/")
data_dir_r  <- paste0(data_dir, "Raw/")
data_dir_c  <- paste0(data_dir, "Clean/")

out_dir  <- paste0(hdir, "Output/")

code_dir  <- paste0(hdir, "Code/")

### Log Output
set.wd( code_dir )
sink("MAKEFILE.Rout", append=TRUE, split=TRUE)

### Execute Codes
source( "RBLOCK_001_DataClean.R" )
source( "RBLOCK_002_Figures.R" )
source( "RBLOCK_003_ModelsTests.R" )
source( "RBLOCK_004_Robust.R" )

### Stop Logging Output
sink()
```

2. Starting a session that logs/sinks you sourcing the makefile
```{r, eval=FALSE}
sink("MAKEFILE.Rout", append=TRUE, split=TRUE)
source("MAKEFILE.R")
sink()
```

3. Execute the makefile via the commandline 
```{bash, eval=FALSE}
R CMD BATCH MAKEFILE.R MAKEFILE.Rout
```




## Final Step

Zip your project into a single file that is easy for *others* to identify: `Class_Project_LASTNAME_FIRSTNAME.zip`

Your code should be readable and error free. For code writing guides, see

* https://google.github.io/styleguide/Rguide.html
* https://style.tidyverse.org/
* https://dept.stat.lsa.umich.edu/~jerrick/courses/stat701/notes/codestyle.html
* http://adv-r.had.co.nz/Style.html
* https://www.burns-stat.com/pages/Tutor/R_inferno.pdf


For additional logging capabilities, see https://cran.r-project.org/web/packages/logr/

For very large projects, there are many more tools available at https://cran.r-project.org/web/views/ReproducibleResearch.html



# Large Scale Projects


## Debugging 

We can use the following packages to help deal with various problems that may arise
```{r}
library(profvis)
library(bench)
library(parallel)
library(Rcpp)
library(compiler)
```

(Note that many of the examples are taken from https://adv-r.hadley.nz/).


Problems print to the console
```{r, error=TRUE}
message("This is what a message looks like")

warning("This is what a warning looks like")

stop("This is what an error looks like")
```


Nonproblems also print to the console
```{r, error=TRUE}
cat('cat\n')
print('print')
```


###  Tracing Errors

First, use garbage cleanup to "cleanup" memory leaks
```{r}
gc()
```
Then trace the error process. 

Example of error process
```{r, error=TRUE}
## Let i() check if its argument is numeric
i <- function(i0) {
  if ( !is.numeric(i0) ) {
    stop("`d` must be numeric", call.=FALSE)
  }
  i0 + 10
}

## Let f() call g() call h() call i()
h <- function(i0) i(i0)
g <- function(h0) h(h0)
f <- function(g0) g(g0)

## Observe Error
f("a")
```



Traceback debugging
```{r, error=TRUE}
traceback()
```



Simple print debugging
```{r, error=TRUE}
f2 <- function(g0) {
  cat("f2 calls g2()\n")
  g2(g0)
}
g2 <- function(h0) {
  cat("g2 calls h2() \n")
  cat("b =", h0, "\n")
  h2(h0)
}
h2 <- function(i0) {
  cat("h2 call i() \n")
  i(i0)
}

f2("a")
```


Interactive approach
```{r, error=TRUE}
g3 <- function(h0) {
  browser()
  h(h0)
}
f3 <- function(g0){
  g3(g0)
}
f3("a")
```




###  Checking problems 

To inspect objects
```{r, eval=FALSE}
is.object(f)
is.object(c(1,1))

class(f)
class(c(1,1))

## Storage Mode Type 
typeof(f)
typeof(c(1,1))

storage.mode(f)
storage.mode(c(1,1))
```


To check for valid inputs/outputs
```{r, eval=FALSE}
x <- c(NA, NULL, NaN, Inf, 0)

cat("Vector to inspect: ")
x

cat("NA: ")
is.na(x)

cat("NULL: ")
is.null(x)

cat("NaN: ")
is.nan(x)

cat("Finite: ")
is.finite(x)

cat("Infinite: ")
is.infinite(x)
## Many others
```


To check for values
```{r, eval=FALSE}
all( x > -2 )
any( x > -2 )
## Check Matrix Rows
rowAny <- function(x) rowSums(x) > 0
rowAll <- function(x) rowSums(x) == ncol(x)
```


## Being Proactive

Supressing errors is possible but a bad idea
```{r, error=TRUE}
try(1+2, silent=T)
try(warning('warning'), silent=T)
try(error('error'), silent=T)
```
```{r, error=TRUE}
try(1+2, silent=F)
try(warning('warning'), silent=F)
try(error('error'), silent=F)
```

Try to handle errors
```{r, eval=F}
tryCatch(
  error = function(e) {
    # code to run when error is thrown
  },
  code_to_run_while_handlers_are_active
)
```

Simple Example
```{r}
tryCatch(
    expr = {
        message( log(-Inf) )
        message("Successfully executed the log(x) call.")
    },
    error = function(e){
        message('Caught an error!')
        print(e)
    },
    warning = function(w){
        message('Caught an warning!')
        print(w)
    },
    finally = {
        message('All done, quitting.')
    }
)
```



<!--## Ignore warnings/messages-->
<!--#suppressWarnings()-->
<!--#suppressMessages()-->

Safe Functions
```{r, error=TRUE}
## Define
log_safe <- function(x){
    tryCatch(
        expr = {
            message( log(x) )
            message("Successfully executed the log(x) call.")
        },
        error = function(e){
            message('Caught an error!')
            print(e)
        },
        warning = function(w){
            message('Caught an warning!')
            print(w)
        },
        finally = {
            message('All done, quitting.')
        }
    )
}

## Test 
log_safe( 10)
log_safe(-10)
log_safe(' ')
```











## Optimizing 


In General: Clean code is often faster and less error prone


*Repetitive tasks can be optimized* You end up with code that

* is cleaner, faster, and more general
* can be easily parallelized


*Computers have big memories and are really good at math.*

* First try vectors
* then try `apply` functions


*Don't waste time on code that is not holding you back.*

* Your code may break, be slow, or incorrect.
* Look at what has already done.




###  Benchmarking

The simplest approach

```{r, error=TRUE}
system.time({
    x <- runif(1e5)
    sqrt(x)
})
```




For identifying bottlenecks
```{r, error=TRUE}
## Generate Large Random Dataset
n <- 2e6
x <- runif(n)
y <- runif(n)
z <- runif(n)
XYZ <- cbind(x,y,z)

## Inspect 4 equivalent `row mean` calculations 
profvis::profvis({
    m <- rowSums(XYZ)/ncol(XYZ)
    m <- rowMeans(XYZ)
    m <- apply(XYZ, 1, mean)
    m <- rep(NA, n);  for(i in 1:n){ m[i] <- (x[i] + y[i] + z[i]) / 3 }
})
## rowSums(), colSums(), rowMeans(), and colMeans() are vectorised and fast.
## for loop is not the slowest, but the ugliest.
```



For easily comparing solutions to specific bottlenecks
```{r, error=TRUE}
## 3 Equivalent calculations of the mean of a vector
mean1 <- function(x,p=1) mean(x^p)
mean2 <- function(x,p=1) sum(x^p) / length(x)
mean3 <- function(x,p=1) mean.default(x^p)

## Time them
x <- runif(1e6)
bench::mark(
  mean1(x,.5),
  mean2(x,.5),
  mean3(x,.5)
)
```

Check for easy speed-ups before creating your own program
```{r, error=TRUE}
X <- cbind(1, runif(1e6))
Y <- X %*% c(1,2) + rnorm(1e6)
DAT <- as.data.frame(cbind(Y,X))

system.time({.lm.fit(X, Y) })
system.time({ lm(Y~X, data=DAT) })

## But note that quicker codes 
## tend to have fewer checks
## and return less information
```


**Vectors are generally faster and easier to read than loops**
```{r, error=TRUE}
x <- runif(1e6)

## Compare 2 moving averages

## First Try
ma1 <- function(y){
    z <- y*NA
    for(i in 2:length(y)){
        z[i] <- (y[i]-y[i-1])/2
    }
    return(z)
}

# Optimized using diff
diff( c(2,2,10,9) )

ma2 <- function(y){
    z2 <- diff(y)/2
    z2 <- c(NA, z2) 
    return(z2)
}

all.equal(ma1(y),ma2(y))

bench::mark(
  ma1(y),
  ma2(y)
)
```

###  Bottlenecks

Sometimes there will still be a problematic bottleneck. 



Your next step should be parallelism:

* Write the function as a general vectorized function.
* Apply the same function to every element in a list *at the same time*

```{r, error=TRUE}
## lapply in parallel on {m}ultiple {c}ores
x <- c(10,20,30,40,50)
f <- function(element) { element^element } 
parallel::mclapply( x, mc.cores=2, FUN=f)
```


More power is often not the solution

```{r, error=TRUE}
## vectorize and compile
e_power_e_fun <- compiler::cmpfun( function(vector){ vector^vector} )

## base R
x <- 0:1E6
s_vc <- system.time( e_power_e_vec <- e_power_e_fun(x) )
s_vc

## brute power
x <- 0:1E6
s_bp <- system.time({
  e_power_e_mc <- unlist( parallel::mclapply(x, mc.cores=2, FUN=e_power_e_fun))
})
s_bp

## Same results
all(e_power_e_vec==e_power_e_mc)
```

Parallelism does not go great with a GUI.
For identifying bottlenecks on a cluster without a GUI, try `Rprof()`
```{r, error=TRUE, eval=F}
Rprof( interval = 0.005)
    # Create Data
    x <- runif(2e6)
    y <- sqrt(x)
    ## Loop Format Data
    z <- y*NA
    for(i in 2:length(y)){ z[i] <- (y[i]-y[i-1])/2 }
    ## Regression
    X <- cbind(1,x)[-1,]
    Z <- z[-1]
    reg_fast <- .lm.fit(X, Z)
Rprof(NULL)
summaryRprof()
```


If you still are stuck, you can

* try [Amazon Web Server](https://aws.amazon.com/ec2/) for more brute-power 
* rewrite bottlenecks with a working C++ compiler or Fortran compiler.



##### Compiled Code


To get C++

* On Windows, install Rtools.
* On Mac, install Xcode from the app store.
* On Linux, sudo apt-get install r-base-dev or similar.

To call C++ from R use package `Rcpp`
```{r, error=TRUE}
Rcpp::cppFunction('
  int add(int x, int y, int z) {
    int sum = x + y + z;
    return sum;
  }'
)
add(1, 2, 3)
```

For help getting started with Rcpp, see https://cran.r-project.org/web/packages/Rcpp/vignettes/Rcpp-quickref.pdf


First try to use C++ (or Fortran) code that others have written
```{r, eval=FALSE}
.C
.Fortran
```
For a tutorial, see https://masuday.github.io/fortran_tutorial/r.html


###   Further Programming Guidance

https://rmd4sci.njtierney.com/
https://smac-group.github.io/ds/high-performance-computing.html
https://www.stat.umn.edu/geyer/3701/notes/arithmetic.Rmd

For debugging tips

* https://cran.r-project.org/doc/manuals/R-lang.html#Debugging
* https://cran.r-project.org/doc/manuals/R-exts.html#Debugging
* https://adv-r.hadley.nz/debugging.html
* https://adv-r.hadley.nz/conditions.html
* https://dept.stat.lsa.umich.edu/~jerrick/courses/stat701/notes/debugging.html
* https://dept.stat.lsa.umich.edu/~jerrick/courses/stat701/notes/functions.html


For optimization tips

* https://cran.r-project.org/doc/manuals/R-exts.html#Tidying-and-profiling-R-code
* https://cran.r-project.org/doc/manuals/R-lang.html#Exception-handling
* https://adv-r.hadley.nz/perf-measure.html.libPaths()
* https://adv-r.hadley.nz/perf-improve.html
* https://cran.r-project.org/doc/manuals/R-exts.html#System-and-foreign-language-interfaces
 https://dept.stat.lsa.umich.edu/~jerrick/courses/stat701/notes/profiling.html
* https://adv-r.hadley.nz/rcpp.html


For parallel programming 

* https://dept.stat.lsa.umich.edu/~jerrick/courses/stat701/notes/parallel.html
* https://bookdown.org/rdpeng/rprogdatascience/parallel-computation.html
* https://grantmcdermott.com/ds4e/parallel.html
* https://psu-psychology.github.io/r-bootcamp-2018/talks/parallel_r.html


For general tips

* https://github.com/christophergandrud/Rep-Res-Book
* Efficient R programming. C. Gillespie R. Lovelace. 2021. https://csgillespie.github.io/efficientR/
* Data Science at the Command Line, 1e. Janssens J. 2020. https://www.datascienceatthecommandline.com/1e/
* R Programming for Data Science. Peng R. 2020. https://bookdown.org/rdpeng/rprogdatascience/
* Advanced R. H. Wickham 2019. https://adv-r.hadley.nz/
* Econometrics in R. Grant Farnsworth. 2008. http://cran.r-project.org/doc/contrib/Farnsworth-EconometricsInR.pdf
* The R Inferno. https://www.burns-stat.com/documents/books/the-r-inferno/






# Software {.tabset}


## Latest versions

Make sure your packages are up to date
```{r,  eval=FALSE}
update.packages()
```

After reinstalling, you can update *all* packages stored in *all* `.libPaths()` with the following command
```{r,  eval=FALSE}
pkgs <- installed.packages( .libPaths() )[,'Package']
install.packages(
    pkgs=pkgs,
    lib=.libPaths()[1],
    type='source'
)
```

You can find all old packages
```{r, eval=FALSE}
old.packages(.libPaths())
```

To find all broken packages after an update
```{r,  eval=FALSE}
library(purrr)

set_names(.libPaths()) %>%
  map(function(lib) {
    .packages(all.available = TRUE, lib.loc = lib) %>%
        keep(function(pkg) {
            f <- system.file('Meta', 'package.rds', package = pkg, lib.loc = lib)
            tryCatch({readRDS(f); FALSE}, error = function(e) TRUE)
        })
  })
## https://stackoverflow.com/questions/31935516/installing-r-packages-error-in-readrdsfile-error-reading-from-connection/55997765
```


## Stata

For those transitioning from Stata or replicating others' Stata work, you can work with Stata data and code within R.

Translations of common procedures is provided by https://stata2r.github.io/

Many packages allows you to read data created by different programs. As of right now, `haven` is a particularly useful for reading in Stata files
```{r, eval=FALSE}
library(haven)
read_dta()
## See also foreign::read.dta
```

You can also execute stata commands directly in R via package `Rstata`. (Last time I checked, `Rstata` requires you to have purchased a non-student version of Stata.) Moreover, you can include stata in the markdown reports via package `Statamarkdown`:
```{r, eval=FALSE}
library(Rstata)
library(Statamarkdown)
```

There are many R packages to replicate or otherwise directly copy what Stata does. For example, see the `margins` package  https://cran.r-project.org/web/packages/margins/vignettes/Introduction.html



For more information on R and Stata, see

* https://github.com/lbraglia/RStata
* https://ignacioriveros1.github.io/r/2020/03/22/r_and_stata.html
* https://bookdown.org/yihui/rmarkdown-cookbook/eng-stata.html
* https://rpubs.com/quarcs-lab/stata-from-Rstudio
* https://clanfear.github.io/Stata_R_Equivalency/docs/r_stata_commands.html
* https://libguides.bates.edu/c.php?g=209169&p=7233333

You can also use other software (such as Python) within R. You can also use R within Stata, or both within Python. With R, you can easily import many different data types

* https://cran.r-project.org/doc/manuals/R-data.html
* https://raw.githubusercontent.com/rstudio/cheatsheets/main/data-import.pdf

## General Workflow

If you want to go further down the reproducibility route (recommended, but not required for our class), consider making your entire workflow use Free Open Source Software


**Linux:** An alternative to windows and mac operating systems.
Used in computing clusters, big labs, and phones.
E.g., Ubuntu and Fedora are popular brands

* https://www.r-bloggers.com/linux-data-science-virtual-machine-new-and-upgraded-tools/,
* http://www.howtogeek.com/249966/how-to-install-and-use-the-linux-bash-shell-on-windows-10/


On Fedora, you can open RStudio on the commandline with
```{bash, eval=FALSE}
rstudio
```

Alternatively, you are encouraged to try using R without a GUI. E.g., on Fedora, this document can be created directly via 
```{bash, eval=FALSE}
Rscript -e "rmarkdown::render('RMarkown.Rmd')"
```



**Latex:** An alternative to microsoft word.
Great for writing many equations and typsetting.
Easy to integrate Figures, Tables, and References.
Steep learning curve.

* easiest to get started online with [Overleaf](https://www.overleaf.com/)
* can also download yourself via [Tex Live](https://www.tug.org/texlive/)

To begin programming, see 

* https://biostat.app.vumc.org/wiki/pub/Main/TheresaScott/Intro.to.LaTeX.TAScott.pdf
* https://www.tug.org/begin.html



## Sweave




**Sweave:** is an alternative to Rmarkdown for integrating latex and R. While Rmarkdown "writes R and latex within markdown", Sweave "write R in latex". Sweave files end in ".Rnw" and can be called within R

```{r, eval=FALSE}
Sweave('Sweave_file.Rnw')
```

or directly from the command line

```{bash, eval=FALSE}
R CMD Sweave Sweave_file.Rnw 
```

In both cases, a latex file `Sweave_file.tex` is produced, which can then be converted to `Sweave_file.pdf`.



For more on Sweave,

* https://rpubs.com/YaRrr/SweaveIntro
* https://support.rstudio.com/hc/en-us/articles/200552056-Using-Sweave-and-knitr
* https://www.statistik.lmu.de/~leisch/Sweave/Sweave-manual.pdf


**Knitr:** 
You can also produce a pdf from an .Rnw file via `knitr`

```{bash, eval=FALSE}
Rscript -e "knitr::Sweave2knitr('Sweave_file.Rnw')"
Rscript -e "knitr::knit2pdf('Sweave_file-knitr.Rnw')"
```

For background on knitr

* https://yihui.org/knitr/
* https://kbroman.org/knitr_knutshell/pages/latex.html
* https://sachsmc.github.io/knit-git-markr-guide/knitr/knit.html



**Misc:** 
Some other good packages for posters/presenting you should be aware of

* https://github.com/mathematicalcoffee/beamerposter-rmarkdown-example
* https://github.com/rstudio/pagedown
* https://github.com/brentthorne/posterdown
* https://odeleongt.github.io/postr/
* https://wytham.rbind.io/post/making-a-poster-in-r/
* https://www.animateyour.science/post/How-to-design-an-award-winning-conference-poster





