---
title: "Expanded Report Template"
author: "Jordan Adamson"
date: "`r format(Sys.time(), '%d.%m.%Y')`"
output:
  html_document:
        code_download: true
        self_contained: true
---





# {.tabset}



## Description
<!-- FIRST TAB -->

In this first tab, include the writeup from your previous report. Polish their presentation for others to easily understand your dataset.

### Source

```{r}
```

### Descriptive Statistics

```{r}
```



## Tests 
<!-- SECOND TAB -->

Suppose we are interested in whether people with $\leq 2$ kids have wages that are *typically* the same as those with $>2$ kids. To find out, we will conduct two hypothesis tests. 

```{r, eval=F}
# library(Ecdat)
# Hint: start by creating a dummy variable
LaborSupply$ManyKids <- LaborSupply$kids > 2
```


### Equal Medians

Create a jackknife confidence interval for the medians of both groups.

Then statistically test whether the medians are equal. (Hint: invert a confidence interval for the differences between the means.) 


### Equal Means

Standardize the means for both groups using jackknife standard errors. Then conduct a bootstrap test for equal means. (Hint: impose the null.)


## Replication Info 

```{r, eval=F}
sessionInfo()
```


<!-- ## COMPILE FROM CLI
    Rscript -e "rmarkdown::render('ReportTemplate_2Expanded.Rmd')"
-->


