---
title: "Mr. Helter Skelter's Recipe for All-Natural Experiments"
titlerunning: H. Skelter's Recipe
date: "`r format(Sys.time(), '%d.%m.%Y')`"
thanks: |
  These recipes were inspired by Mrs. Razzle Dazzle's tips on proofs to never be a fop, macaroni, or dandy.
author:
    name: Jordan Adamson
    address: Institute for Empirical Economic Research, Leipzig University
    email: adamson@wifa.uni-leipzig.de
abstract: |
  This letter extends "Cheating with Models" by Eliaz, Kfir, Ran Spiegler, and Yair Weiss (2021, American Economic Review: Insights, 3 (4): 417-34. DOI: 10.1257/aeri.20200635) to indicator variables and the heuristics used to analyze natural experiments. As such, it contributes a missing piece to "Methods Matter: p-Hacking and Publication Bias in Causal Analysis in Economics" by Brodeur, Abel, Nikolai Cook, and Anthony Heyes (2020, American Economic Review, 110 (11): 3634-60. DOI: 10.1257/aer.20190687). The underlying statistical issue is a type of streak selection discussed in "Surprised by the Hot Hand Fallacy? A Truth in the Law of Small Numbers" by Joshua B. Miller and Adam Sanjurjo (2018, Econometrica, 86: 2019-2047. DOI: 10.3982/ECTA14943). The overarching methodological point is to ``estimate, not testimate'' (McCloskey, 2008, p.249 of The Cult of Statistical Significance).
output: rticles::elsevier_article
---

<!--bibliography: bibliography.bib-->
<!--output:-->
<!--    html_document:-->
<!--        code_download: true-->
<!--output: rticles::elsevier_article-->



Time series experiments and causal estimands: exact randomization tests and trading
https://www.tandfonline.com/doi/full/10.1080/01621459.2018.1527225

Econometric analysis of potential outcomes time series: instruments, shocks, linearity and the causal response function
https://arxiv.org/pdf/1903.01637.pdf



In my research, I use only the best. That's why I use the gold standard atop the hierarchy of evidence. My two all-natural experiments practically guarantee three stars of causality in under 10 minutes. Simply substitute the easy-to-find $P( \text{natural experiment} ~|~ \text{big change in time trend} )$ for $P( \text{big change in time trend} ~|~ \text{natural experiment} )$, and you'll be baking-in statistical significance in no time.

We'll use two ready-made datasets today, but you can try multiple datasets or create your own from scratch. You'll get something different every time, and these unique findings will stand out against a confusing literature of mixed results. If your results aren't quite robust enough this time, try again with an eye towards passing the test.




\newpage
# Always start with some good data 

```{r, message=F, warning=F, results='asis', out.width="100%"}
n <- 1000
n_index <- seq(n)

set.seed(1)
random_walk1 <- cumsum(runif(n,-1,1))

set.seed(2)
random_walk2 <- cumsum(runif(n,-1,1))

par(mfrow=c(1,2))
plot(random_walk1, pch=16, col=grey(.5,.5))
plot(random_walk2, pch=16, col=grey(.5,.5))
```



\newpage
# Simply let the data speak with Regression Discontinuity Designs!


```{r, message=F, warning=F, results='asis', out.width="100%"}
## Let the data take shape
## (around the large differences before and after)
n1 <- 290
wind1 <- c(n1-300,n1+300)
dat1 <- data.frame(t=n_index, y=random_walk1, d=1*(n_index > n1))
dat1_sub <- dat1[ n_index>wind1[1] & n_index < wind1[2],]

## Then find your big break
reg0 <- lm(y~t, data=dat1_sub[dat1_sub$d==0,])
reg1 <- lm(y~t, data=dat1_sub[dat1_sub$d==1,])

## The evidence should show openly (it's just science)
plot(random_walk1, pch=16, col=grey(.5,.5), xlim=wind1)
abline(v=n1, lty=2)
lines(reg0$model$t, reg0$fitted.values, col=2)
lines(reg1$model$t, reg1$fitted.values, col=4)

## Dress with some statistics for added credibility
rdd_sub <- lm(y~d+t+d*t, data=dat1_sub)
rdd_full <- lm(y~d+t+d*t, data=dat1)
stargazer::stargazer(rdd_sub, rdd_full, 
    type='latex', table.placement='h!', title='Recipe RDD',
    header=F, omit=c('Constant'),
    notes=c('First column uses a dataset around the discontinuity.',
    'Smaller windows are more causal, and where the effect is bigger.'))
```

\newpage
# For more control, try these Difference-In-Differences! 


```{r, message=F, warning=F, results='asis', out.width="100%"}

## Find a reversal of fortune
## (A good story always goes well with a nice pre-trend)
n2 <- 318
wind2 <- c(n2-20,n2+20)
plot(random_walk2, pch=16, col=4, xlim=wind2, ylim=c(-15,15))
points(random_walk1, pch=16, col=2)
abline(v=n2, lty=2)

## Knead out any effects that are non-causal 
## (or even just correlation)
dat2A <- data.frame(t=n_index, y=random_walk1, d=1*(n_index > n2), RWid=1)
dat2B <- data.frame(t=n_index, y=random_walk2, d=0, RWid=2)
dat2  <- rbind(dat2A, dat2B)
dat2$RWid <- as.factor(dat2$RWid)
dat2$tid <- as.factor(dat2$t)
dat2_sub <- dat2[ dat2$t>wind2[1] & dat2$t < wind2[2],]

## Report the stars for all to enjoy
## (and remind that stable coefficients are the good ones)
did_fe1 <- lm(y~d+tid, data=dat2_sub)
did_fe2 <- lm(y~d+RWid, data=dat2_sub)
did_fe3 <- lm(y~d*RWid+tid, data=dat2_sub)
stargazer::stargazer(did_fe1, did_fe2, did_fe3,
    type='latex', table.placement='h!', title='Recipe DID',
    header=F, omit=c('tid','RWid', 'Constant'),
    notes=c(
     'Fixed effects for time in column 1,
        for id in column 2, and both in column 3.',
     'Fixed effects control for most of your concerns.',
     'Anything else creates a bias in the opposite direction.'))
```


```{r, echo=FALSE}
knitr::knit_exit()
```

<!--^[(I heard someone found $\beta_{1}^{*}$ one time but someone else found $\beta_{2}^{**}$, you might find $\beta_{3}^{***}$!)]-->


<!--bibliography: bibliography.bib
<!--    @article{10.1257/aeri.20200635,-->
<!--    Author = {Eliaz, Kfir and Spiegler, Ran and Weiss, Yair},-->
<!--    Title = {Cheating with Models},-->
<!--    Journal = {American Economic Review: Insights},-->
<!--    Volume = {3},-->
<!--    Number = {4},-->
<!--    Year = {2021},-->
<!--    Month = {December},-->
<!--    Pages = {417-34},-->
<!--    DOI = {10.1257/aeri.20200635},-->
<!--    URL = {https://www.aeaweb.org/articles?id=10.1257/aeri.20200635}}-->


<!--dat2 <- data.frame(t=n_index, y1=random_walk1, y2=random_walk2, d=reversal)-->
<!--dat2_long <- reshape(dat2, varying=list( c('y1','y2') ), direction='long', timevar='RW')-->
<!--sub_id <- dat2_long$t>wind2[1] & dat2_long$t < wind2[2]-->
<!--dat2_sub <- dat2_long[sub_id,]-->


<!-- Practice makes perfect, and you'll be finding causal relationships in no time.-->

<!--www.damianclarke.net/teaching/Microeconometria/microeconometria.pdf-->

<!-- ## COMPILE FROM CLI
    Rscript -e "rmarkdown::render('Template_Paper.Rmd')"
    latexmk -c  && rm *.spl
-->







