---
title: "Poster template"
output:
  flexdashboard::flex_dashboard:
    orientation: rows
    source: embed
    social: menu
---




## Row 1 {data-height=15}
 **Abstract:** 
Template for academic poster preparation with R markdown.

* Point-and-click cumbersome to update, expensive, non-reproducible. 
* $LaTeX$ cumbersome to update, complex (at first).





## Row 2 {data-height=80}

### **Methods**

This poster template depends on the following R packages

```{r, echo=T}
library(flexdashboard)
library(tidyverse)
library(kableExtra)
```


**Preparing your analysis environment**

To make this poster template,
```{r, echo=T, eval=F}
rmarkdown::render('DataScientism_Poster.Rmd')
```

To use data you have saved via `save(YourData, 'YourData.Rdata')`, 
```{r, echo=T, eval=F}
load('YourData.Rdata')
```


### **Results**

Amazing Table *(less preferred)*

```{r}
t(summary(c(volcano))) %>% kbl() %>%  kable_styling()
```


Amazing Figure *(more preferred)*

```{r, out.width='100%', fig.height=4}
filled.contour(volcano)
```




### **Discussion**


Crammed, complex posters are often worse posters.



#### Secondary Result
Only if necessary

```{r, eval=F, echo=T}
## to grab from folder
knitr::include_graphics('../folder/picture.pdf')
```




## Row 3 {data-height=5, .small}

### **Contact:**
Name. email@address. Institution.
  

### **References:**

* https://pkgs.rstudio.com/flexdashboard/articles/examples.html
* https://pagedown.rbind.io/poster-jacobs


<!-- ## COMPILE FROM CLI
    Rscript -e "rmarkdown::render('DataScientism_Poster.Rmd')"
    cp DataScientism_Poster.html ../docs/DataScientism_Poster.html
    
    ## Convert to PDF
    pandoc DataScientism_Poster.html -t latex -o DataScientism_Poster.pdf
    Rscript -e "pagedown::chrome_print('DataScientism_Poster.html')"
--->    


