---
title: "Summary Report Template"
author: "Jordan Adamson"
date: "`r format(Sys.time(), '%d.%m.%Y')`"
output:
  html_document:
        code_download: true
        self_contained: true
---

# Notes
*** 

This is a template that you must update

* Change the name "Jordan Adamson" to your own
* Change the title to just "Summary Report"
* Use the "Ecdat:LaborSupply" dataset instead of "AER:Grunfeld"
* Include your replication info
* Delete this notes section


# Source
***

Describe where and how you got your dataset. E.g.,

> I use the "Grunfeld" dataset from the AER Package by Kleiber and Zeileis (2008).
>
>*Kleiber C, Zeileis A (2008). _Applied Econometrics with R_.
>Springer-Verlag, New York.
><https://doi.org/10.1007/978-0-387-77318-6>,
><https://CRAN.R-project.org/package=AER>.*

Explain what the dataset is. E.g.,

> This is a dataset of 3 variables for 11 large US manufacturing firms between 1935-1954 with annual records.

How the data were collected. E.g.,

> Grunfeld compiled the data largely from firms publicly available reports and financial manuals like Moody’s Industrial Manual.
 

Load in the dataset.
```{r}
data("Grunfeld", package="AER")
# data("LaborSupply", package="plm")
```

Add a table of the main variables of interest from your dataset
```{r}
library(reactable)
reactable(Grunfeld[,c('firm','year','value')],
    groupBy='firm', 
    defaultPageSize=5)
```

# Descriptive Statistics
***

Summarize 2 variables from your dataset using

* Figures for accuracy
* Numbers for precision
* Words for the big-picture interpretation

```{r}
# Simple analysis first
# Then polish figures, making them interactive when helpful
```

# Replication Info 
***

```{r, eval=F}
sessionInfo()
```


<!-- ## COMPILE FROM CLI
    Rscript -e "rmarkdown::render('ReportTemplate_1Descriptive.Rmd')"
-->


