
# First Steps
***

## Why Program in R?

You should program your statistical analysis, and we will cover some of the basics of how to do this in R. You also want your work to be replicable

* *Replicable*: someone collecting new data comes to the same results.
* *Reproducibile*: someone reusing your data comes to the same results.

You can read more about the distinction in many places, including

* <https://www.annualreviews.org/doi/10.1146/annurev-psych-020821-114157>
* <https://nceas.github.io/sasap-training/materials/reproducible_research_in_r_fairbanks/>

We focus on R because it is good for complex stats, concise figures, and coherent organization. It is built and developed by applied statisticians for statistics, and used by many in academia and industry. For students, think about labor demand and what may be good for getting a job. Do some of your own research to best understand how much to invest.


My main sell to you is that being reproducible is in your own self-interest.

#### **An example workflow**. {-}

***First Steps...***

*Step 1*:  Some ideas and data about how variable $X_{1}$ affects variable $Y_{1}$, which we denote as $X_{1}\to Y_{1}$

* You copy some data into a spreadsheet, manually aggregate
* do some calculations and tables the same spreadsheet
* some other analysis from here and there, using this software and that.

*Step 2*:  Pursuing the lead for a week or two

* you extend your dataset with more observations
* copy in a spreadsheet data, manually aggregate
* do some more calculations and tables, same as before

***A Little Way Down the Road ...***

1 month later: someone asks about another factor: $X_{2} \to Y$.

* you download some other type of data
* You repeat <u>Step 2</u> with some data on $X_{2}$.
* The details from your "point and click" method are a bit fuzzy.
* It takes a little time, but you successfully redo the analysis.

4 months later: someone asks about yet another factor: $X_{3}\to Y_{1}$.

* You again repeat <u>Step 2</u> with some data on $X_{3}$.
* You're pretty sure none of tables your tried messed up the order of the rows or columns.
* It takes more time and effort. The data processing was not transparent, but you eventually redo the analysis.

6 months later: you want to explore another outcome: $X_{2} \to Y_{2}$.

* You found out Excel had some bugs in it's statistical calculations (see e.g., <https://biostat.app.vumc.org/wiki/pub/Main/TheresaScott/StatsInExcel.TAScot.handout.pdf>). You now use a new version of the spreadsheet
* You're not sure you merged everything correctly. After much time and effort, most (but not all) of the numbers match exactly.

2 years later: your boss wants you to replicate your work: $X_{1}, X_{2}, X_{3} \to Y_{1}$.

* A rival has proposed something new. Their idea doesn't actually make any sense, but their figures and statistics look better.
* You don't even use that computer anymore and a collaborator who handled the data on $X_{2}$ has moved on.


#### **An alternative workflow**. {-}

Suppose you decided to code what you did beginning with Step 2.

It does not take much time to update or replicate your results.

* Your computer runs for 2 hours and reproduces the figures and tables.
* You also rewrote your big calculations to use multiple cores, this took two hours to do but saved 6 hours *each time* you rerun your code.
* You add some more data. It adds almost no time to see whether much has changed.

Your results are transparent and easier to build on.

* You see the exact steps you took and found an error
  * Google *"worst Excel errors"* and note the frequency they arise from copy/paste via the "point-and-click" approach. E.g., *Fidelity's $2.6 Billion Dividend Error*.
    * Glad you found a problem before sending your research out! See <https://retractionwatch.com/> and <https://econjwatch.org/>. Future economists should also read <https://core.ac.uk/download/pdf/300464894.pdf>.
* You try out a new plot you found in *The Visual Display of Quantitative Information*, by Edward Tufte.
  * It's not a standard plot, but google answers most of your questions.
  * Tutorials help avoid bad practices, such as plotting 2D data as a 3D object (see e.g., <https://clauswilke.com/dataviz/no-3d.html>).
* You try out an obscure statistical approach that's hot in your field.
  * it doesn't make the report, but you have some confidence that candidate issue isn't a big problem


## First Steps

#### **Install R**. {-}

First Install [R](https://cloud.r-project.org/).
Then Install [Rstudio](https://www.rstudio.com/products/rstudio/download/).


For Fedora (linux) users, note that you need to first enable the repo and then install

```{bash, eval=F}
sudo dnf install 'dnf-command(copr)'
sudo dnf copr enable iucar/rstudio
sudo dnf install rstudio-desktop
```


*Make sure you have the latest version of R and Rstudio for class.* If not, then reinstall. 

#### **Interfacing with R Studio**. {-}

Rstudio is perhaps the easiest to get going with. (There are other GUI's.)

In Rstudio, there are 4 panes. (If you do not see 4, click "file > new file > R script" on the top left of the toolbar.)

```{r, echo=F}
knitr::include_graphics("../docs/Figures_Manual/Rstudio.svg")
```

The top left pane is where you write your code. For example, type
```{r, eval=F}
#| code-fold: false
1+1
```

The pane below is where your code is executed. 
Keep you mouse on the same line as your code, and then click "Run". 
You should see
```
> 1+1
[1] 2
```
If you click "Run" again, you should see that same output printed again.

You should add comments to your codes, and you do this with hashtags. For example
```
# This is my first comment!
1+1 # The simplest calculation I could think of
```
You can execute each line one-at-a-time. Or you can highlight them both, to take advantage of how R executes commands line-by-line.

#### **Reading This Textbook**. {-}

As we proceed, you can see both my source code and output like this:
```{r}
1+1
```

There are also special boxes

:::{.callout-note icon=false collapse="true"}
This box contains *need to know* examples. Such as 
```{r}
1+1
```
:::

:::{.callout-tip icon=false collapse="true"}
This box contains *test yourself* examples and questions. Such as 
```{r}
2+7
2/7
```
:::


#### **Assignment**. {-}

You can create "variables" that store values. For example,
```{r}
x <- 1 # Make your first variable
x + 1 # The simplest calculation I could think of
```

```{r}
x <- 23 #Another example
x + 1
```

```{r}
y <- x + 1 #Another example
y
```

Your variables must be defined in order to use them. Otherwise you get an error. For example,
```{r, error=TRUE}
X +   1 # notice that R is sensitive to capitalization 
```

Your variable names do not matter technically, but they should be informative
```{r}
one <- 1 # good variable name
one

one <- 43 # bad variable name
one
```

Good names avoid confusion later
```{r}
x <- 43
x_plus_two <- x + 2 # better
x_plus_two
```

#### **Scripting**. {-}

* Create a folder on your computer to save your scripts
* Save your `R Script` file as *My_First_Script.R* in your folder
* Close Rstudio
* Open your script and re-run it

As you work through the material, make sure to both execute and save your scripts. Add lots of commentary to your scripts. Name your scripts systematically.

There are often many ways to accomplish the same goal. You first scripts will be very basic and rough, but you can edit them later based on what you learn. And you can always ask R for help
```{r, eval=F}
sum(x, 2) # x + 2
?sum
```

We write script in the top left so that we can edit common mistakes.
```{r, eval=F}
# Mistake 1: using undefined objects
Y

#  Mistake 2: spelling and spacing
Y < - 43
Y_plus_z <- Y + z

# Mistake 3: half-completed code
x + y + 
x_plus_y_plus_z <- x + y + z
# Seeing "+" in the bottom console?
# press "Escape" and try again
```


## Mathematical Objects

In R: scalars, vectors, and matrices are different kinds of "objects". 

These objects are used extensively in data analysis

* *scalars*: summary statistics (average household income).
* *vectors*: single variables in data sets (the household income of each family in Vancouver).
* *matrices*: two variables in data sets (the age and education level of every person in class).


Vectors are probably your most common object in R, but we will start with scalars.


#### **Scalars**. {-}

Make your first scalar
```{r}
xs <- 2 # Make your first scalar
xs  # Print the scalar
```

Perform simple calculations and see how R is doing the math for you
```{r}
xs + 2
xs*2 # Perform and print a simple calculation
(xs+1)^2 # Perform and print a simple calculation
xs + NA # often used for missing values
```

Now change `xs`, predict what will happen, then re-run the code.

#### **Vectors**. {-}

Make your first vector
```{r}
x <- c(0,1,3,10,6) # Your First Vector
x # Print the vector
x[2] # Print the 2nd Element; 1
x+2 # Print simple calculation; 2,3,5,8,12
x*2
x^2
```

Apply mathematical calculations elementwise
```{r}
x+x
x*x
x^x
```

In R, scalars are treated as a vector with one element.
```{r}
c(1)
```

Sometimes, we will use vectors that are entirely ordered. 
```{r}
seq(1,7,by=1) #1:7
seq(1,7,by=0.5)

# Ordering data
sort(x)
x[order(x)]
```


#### **Matrices**. {-}

Matrices are also common objects
```{r}
x1 <- c(1,4,9)
x2 <- c(3,0,2)
x_mat <- rbind(x1, x2)

x_mat       # Print full matrix
x_mat[2,]   # Print Second Row
x_mat[,2]   # Print Second Column
x_mat[2,2]  # Print Element in Second Column and Second Row
```

There are elementwise calculations
```{r}
x_mat+2
x_mat*2
x_mat^2

x_mat + x_mat
x_mat*x_mat #NOT classical matrix multiplication
x_mat^x_mat
```


And you can also use matrix algebra
```{r}
x_mat1 <- matrix( seq(2,7), 2, 3)
x_mat1

x_mat2 <- matrix( seq(4,-1), 2, 3)
x_mat2

tcrossprod(x_mat1, x_mat2) #x_mat1 %*% t(x_mat2)

crossprod(x_mat1, x_mat2)
```


## Mathematical Functions

#### **Simple Functions**. {-}

Functions are applied to objects
```{r}
# Define a function that adds two to any vector
add_two <- function(input_vector) { #input_vector is a placeholder
    output_vector <- input_vector + 2 # new object defined locally 
    return(output_vector) # return new object 
}
# Apply that function to a vector
x <- c(0,1,3,10,6)
add_two(input_vector=x) #same as add_two(x)
```

Common mistakes:
```{r, eval=F}
print(output_vector)
# This is not available globally

# Double check your spelling
x < - add_two(input_vector=X) 

# Seeing "+" in the bottom console
# often means you forgot to close the function with "}" 
# press "Escape" and try again
add_two <- function(input_vector) { 
    output_vector <- input_vector + 2 
    return(output_vector)
x <- c(0,1,3,10,6)
add_two(x)
```

There are many different functions
```{r}
add_vec <- function(input_vector1, input_vector2) {
    output_vector <- input_vector1 + input_vector2
    return(output_vector)
}
add_vec(x,3)
add_vec(x,x)

sum_squared <- function(x1, x2) {
    y <- (x1 + x2)^2
    return(y)
}

sum_squared(1, 3)
sum_squared(x, 2)
sum_squared(x, NA) 
sum_squared(x, x)
sum_squared(x, 2*x)
```


Functions can take functions as arguments. Note that a *statistic* is defined as a function of data.
```{r}
statistic <- function(x, f){
    y <- f(x)
    return(y)
}
statistic(x, sum)
```


There are many possible functions you can make and use. More complicated functions often have defaults.
```{r}
fun_of_seq <- function(f, constant=2){
    x1 <- seq(1,3, length.out=12)
    x2 <- x1+constant
    x <- cbind(x1,x2)
    y <- f(x)
    return(y)
}
fun_of_seq(sum)
fun_of_seq(sum, 3)
fun_of_seq(prod)
fun_of_seq(prod, 3)
```


You can also apply functions to matrices
```{r}
sum_squared(x_mat, x_mat)

# Apply function to each matrix row
y <- apply(x_mat, 1, sum)^2 
# ?apply  #checks the function details
```


#### **Loops**. {-}

Applying the same function over and over again
```{r}
# Example 1: simple division
x <- vector(length=3)
#Fill empty vector
for(i in seq(1,3)){
    x[i] <- i/2
}
x
# Compare

# Example 2: exponential
#Create empty vector
x <- vector(length=3)
#Fill empty vector
for(i in seq(1,3)){
    x[i] <- exp(i)
}
# Compare
x
c( exp(1), exp(2), exp(3))

# Example 3: using existing data
x <- c(1,3,9,2)
y <- vector(length=length(x))
for(i in seq_along(x) ){
    y[i] <- x[i] + 1
}
y
```

A more complicated example
```{r}
complicated_fun <- function(i, j=0){
    x <- i^(i-1)
    y <- x + mean( seq(j,i) )
    z <- log(y)/i
    return(z)
}
complicated_vector <- vector(length=10)
for(i in seq(1,10) ){
    complicated_vector[i] <- complicated_fun(i)
}
```

A recursive example
```{r}
x <- vector(length=4)
x[1] <- 1
for(i in seq(2,4) ){
    x[i] <- (x[i-1]+1)^2
}
x
```

<!---
```{r, eval=F, echo=F}
# mapply takes multiple vectors
# x <- seq(1,3)
# mapply(sum, x, exp(x) )
```

```{r}
# for loop in a function
r_fun <- function(n){
    x <- rep(1,n)
    for(i in seq(2,length(x)) ){
        x[i] <- (x[i-1]+1)^2
    }
    return(x)
}
r_fun(5)
```
--->


#### **Basic Logic**. {-}

TRUE/FALSE
```{r}
x <- c(1,2,3,NA)
x > 2
x==2

any(x==2)
all(x==2)
2 %in% x

2==TRUE
2==FALSE
 
is.numeric(x)
is.na(x)
```

The "&" and "|" commands are logical calculations that compare vectors to the left and right.
```{r}
x <- seq(1,3)
(x >= 1) & (x < 2)
(x >= 1) | (x < 2)

if( all(x >= 1) ){
    print("ok")
} else {
    print("not ok")
}

logic_fun <- function(x){
    if( all(x >= 1) ){
        print("ok")
    } else {
        print("not ok")
    }
}
logic_fun( seq(1,3) )
logic_fun( seq(0,2) )
```


## Further Reading

There are many good and free programming materials online. For help setting up, see any of the following links

* <https://learnr-examples.shinyapps.io/ex-setup-r/>
* <https://rstudio-education.github.io/hopr/starting.html>
* <https://a-little-book-of-r-for-bioinformatics.readthedocs.io/en/latest/src/installr.html>
* <https://cran.r-project.org/doc/manuals/R-admin.html>
* <https://courses.edx.org/courses/UTAustinX/UT.7.01x/3T2014/56c5437b88fa43cf828bff5371c6a924/>
* <https://owi.usgs.gov/R/training-curriculum/installr/>
* <https://www.earthdatascience.org/courses/earth-analytics/document-your-science/setup-r-rstudio/>

The most common tasks can be found <https://github.com/rstudio/cheatsheets/blob/main/rstudio-ide.pdf>

Some of my programming examples originally come from <https://r4ds.had.co.nz/> and I recommend <https://intro2r.com>. 

I have also used online material from many places over the years, as there are many good yet free-online tutorials and courses specifically on R programming. See e.g., 

* <https://cran.r-project.org/doc/manuals/R-intro.html>
* R Graphics Cookbook, 2nd edition. Winston Chang. 2021. <https://r-graphics.org/>
* R for Data Science. H. Wickham and G. Grolemund. 2017. <https://r4ds.had.co.nz/index.html>
* An Introduction to R. W. N. Venables, D. M. Smith, R Core Team. 2017. <https://colinfay.me/intro-to-r/>
* Introduction to R for Econometrics. Kieran Marray. <https://bookdown.org/kieranmarray/intro_to_r_for_econometrics/>
* Wollschläger, D. (2020). Grundlagen der Datenanalyse mit R: eine anwendungsorientierte Einführung. <http://www.dwoll.de/rexrepos/>
* Spatial Data Science with R: Introduction to R. Robert J. Hijmans. 2021. <https://rspatial.org/intr/index.html>
* <https://www.econometrics-with-r.org/1.2-a-very-short-introduction-to-r-and-rstudio.html>
* <https://rafalab.github.io/dsbook/>
* <https://moderndive.com/foreword.html>
* <https://rstudio.cloud/learn/primers/1.2>
* <https://cran.r-project.org/manuals.html>
* <https://stats.idre.ucla.edu/stat/data/intro_r/intro_r_interactive_flat.html>
* <https://cswr.nrhstat.org/app-r>


For more on why to program in R, see

* <http://www.r-bloggers.com/the-reproducibility-crisis-in-science-and-prospects-for-r/>
* <http://fmwww.bc.edu/GStat/docs/pointclick.html>
* <https://github.com/qinwf/awesome-R\#reproducible-research>
* A Guide to Reproducible Code in Ecology and Evolution
* <https://biostat.app.vumc.org/wiki/pub/Main/TheresaScott/ReproducibleResearch.TAScott.handout.pdf>
