# Confidence Intervals
***

The sampling distribution describes how a statistic varies across samples; the *confidence interval* turns knowledge about that sampling distribution into a statement about the unknown parameter.
We will focus on data-driven approaches first, then theory-driven approaches, then theoretical properties of confidence intervals.

:::{.callout-important icon=false title="Key Definition" collapse="false"}
A *confidence interval* is a range of values, computed from the sample, intended to contain the unknown population parameter.

The *coverage* level $1-\alpha$ is the long-run proportion of intervals (constructed from repeated samples by the same procedure) that contain the true parameter.
:::

Confidence intervals are useful for *reporting uncertainty* alongside an estimate: instead of saying only "the mean is $\hat{M}$", we report a range and a coverage level, communicating both the best guess and how much sampling noise could plausibly move it.
The width depends on the sample size, the spread of the data, and the coverage level $1-\alpha$ that the user chooses.


## Resampling Intervals

If we have a sampling distribution for a statistic, the simplest way to read off a confidence interval is to take its middle quantiles.

:::{.callout-important icon=false title="Key Definition" collapse="false"}
The *critical values* of a sampling distribution are the lower and upper quantiles that bracket the middle $Z\%$ of the distribution.
For a $95\%$ confidence interval, the critical values are the $2.5$th and $97.5$th percentiles.
:::

Critical values are useful because they translate any coverage level into a pair of numbers that can be looked up or simulated, without committing to a parametric model for the sampling distribution.
Often, we have only one sample, in which case we can construct a bootstrap or jackknife sampling distribution (as in [Sampling & Resampling](https://jadamso.github.io/Rbooks/01_05_Sampling.html#resampling-distributions)) and compute the critical values directly from it:

* The $5^{th}$ and $95^{th}$ percentiles are the critical values for the $90\%$ confidence interval.
* The $2.5^{th}$ and $97.5^{th}$ percentiles are the critical values for the $95\%$ confidence interval.
* The $0.5^{th}$ and $99.5^{th}$ percentiles are the critical values for the $99\%$ confidence interval.


```{r}
sample_dat <- USArrests[, 'Murder']
sample_mean <- mean(sample_dat)
sample_mean

# Bootstrap Distribution
set.seed(1) # to be replicable
bootstrap_means <- rep(NA, 9999)
for(b in seq_along(bootstrap_means)){
    dat_id <- seq_along(sample_dat)
    boot_id <- sample(dat_id , replace=T)
    dat_b  <- sample_dat[boot_id] # c.f. jackknife
    mean_b <- mean(dat_b)
    bootstrap_means[b] <-mean_b
}

# 95% CI
boot_ci <- quantile(bootstrap_means, probs=c(.025, .975), na.rm=T)
boot_ci

# Visualize
hist(bootstrap_means, freq=F,
    main=NA,
    border=NA, breaks=50)
title('Bootstrap CI', font.main=1)
abline(v=boot_ci)
```

:::{.callout-note icon=false collapse="true" title="Must Know"}
Using the bootstrap distribution above, here are the $90\%$ and $99\%$ confidence intervals.
```{r}
# 90% CI: 5th and 95th percentiles
quantile(bootstrap_means, probs=c(.05, .95), na.rm=T)

# 99% CI: 0.5th and 99.5th percentiles
quantile(bootstrap_means, probs=c(.005, .995), na.rm=T)
```
The $99\%$ interval is wider than the $90\%$ interval: demanding higher coverage forces a longer interval.
:::

This approach can be easily extended to other statistics. For example, a $Z\%$ confidence interval for the median is an interval where $Z\%$ of those generated will contain the population median. 

:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
The following code creates a bootstrap CI for the median. 
Edit it to create a jackknife CI instead. 
Then edit it to create a bootstrap CI for the upper quartile. 

```{r}
sample_dat <- USArrests[, 'Murder']
sample_stat <- quantile(sample_dat, probs=0.5) #median
sample_stat

# Bootstrap Distribution
set.seed(1) # to be replicable
bootstrap_stats <- rep(NA, 9999)
for(b in seq_along(bootstrap_stats)){
    dat_id <- seq_along(sample_dat)
    boot_id <- sample(dat_id, replace=T)
    dat_b  <- sample_dat[boot_id] # c.f. jackknife
    stat_b <- quantile(dat_b, probs=0.5)
    bootstrap_stats[b] <-stat_b
}

# 95% CI
boot_ci <- quantile(bootstrap_stats, probs=c(.025, .975), na.rm=T)
boot_ci

# Visualize
hist(bootstrap_stats, freq=F,
    main=NA,
    border=NA, breaks=50)
title('Bootstrap CI', font.main=1)
abline(v=boot_ci)
```
:::

## Normal Approximation

When the sampling distribution is well approximated by a Normal curve, the confidence interval has a simple closed form built from the standard error.

:::{.callout-important icon=false title="Key Definition" collapse="false"}
The *margin of error* $E$ is the half-width of a Normal-approximation confidence interval, so the interval is $\hat{M} \pm E$.
It equals the critical value times the standard error, $E = q(\alpha/2) \cdot SE(M)$, and shrinks as $n$ grows.
:::

The Normal-approximation interval is useful because it requires only an estimate $\hat{M}$, a standard error $SE(M)$, and a single Normal quantile (no resampling loop, no full sampling distribution).
The $95\%$ interval is the most common and uses critical value $q=1.96$ (the $97.5$th percentile of the standard Normal); the $90\%$ interval uses $q=1.645$.
$SE(M)$ can come from the classical formula $\hat{S}/\sqrt{n}$, the bootstrap, or the jackknife.

A *coverage* level of $1-\alpha$ means that if the same sampling procedure were repeated $100$ times from the same population, approximately $1-\alpha$ percent of the intervals are expected to contain the true population mean.
^[Generally, a coverage level of $1-\alpha$ means $Prob( M - E < \mu < M + E)=1-\alpha$.
Notice that $Prob( M - E < \mu < M + E) = Prob( - E < \mu - M < + E) = Prob( \mu + E > M > \mu - E)$. So if the interval $[\mu - 10, \mu + 10]$ contains $95\%$ of all $M$, then the interval $[M-10, M+10]$ will also contain $\mu$ in $95\%$ of the samples because whenever $M$ is within $10$ of $\mu$, the value $\mu$ is also within $10$ of $M$. But for any particular sample, the interval $[\hat{M}-10, \hat{M}+10]$ either does or does not contain $\mu$. Similarly, if you compute $\hat{M}=9$ for your particular sample, a coverage level of $1-\alpha=95\%$ does not mean $Prob(9 - E < \mu < 9 + E)=95\%$.]

:::{.callout-note icon=false collapse="true" title="Must Know"}
For example, suppose we sample $n=36$ monthly electricity bills (in dollars) and find $\hat{M}=142$ and $\hat{S}=30$.
The classical standard error is $SE(M) \approx \hat{S}/\sqrt{n} = 30/\sqrt{36} = 5$.
From theory, we also know that $95\%$ confidence interval has upper and lower critical value of 
\begin{eqnarray}
F^{-1}_{Normal}\left(0.975, SE(M) \right) &=& +1.96 \cdot SE(M) \\
F^{-1}_{Normal}\left(0.025, SE(M) \right) &=& -1.96 \cdot SE(M)
\end{eqnarray}
Together, we can calculate
\begin{eqnarray}
E &=& q(\alpha/2) \cdot SE(M) = 1.96 \cdot 5 = 9.8\\
[\hat{M} - E, \hat{M} + E] &=& [142 - 9.8, 142 + 9.8] = [132.2,~ 151.8].
\end{eqnarray}
We are $95\%$ confident that the population mean electricity bill is between $\$132.20$ and $\$151.80$.
```{r}
M_hat <- 142
S_hat <- 30
n <- 36
SE <- S_hat / sqrt(n)
SE

# 95% CI Manual Calculation
M_hat + c(-1.96, 1.96) * SE

## 95% CI Intuition
qnorm( c(0.025, 0.975), M_hat, SE)
```
:::



The main advantages of the Normal approximation is that

* can be computed formulaically as above
* works well for estimating extreme probabilities, where resampling methods tend to be worse.

The main disadvantage is that 

* the sampling distribution might be far from normal.

In the example below, they are all quite similar, but that does not always need to be the case.
The Normal approximation is least reliable when the sample is small or the data are strongly skewed: the sampling distribution is then asymmetric, and a symmetric $\hat{M}\pm E$ interval cannot capture that.
The percentile bootstrap interval reads quantiles directly off the resampling distribution, so it can be lopsided when the data call for it.


```{r}
# Bootstrap Distribution with Percentile CI
hist(bootstrap_means, breaks=25,
    main=NA, border=NA,
    freq=F, ylim=c(0, 0.7),
    xlab=expression(hat(b)[b]))
title('Percentile vs Normal 95% CIs', font.main=1)
boot_ci_percentile <- quantile(bootstrap_means, probs=c(.025, .975))
abline(v=boot_ci_percentile, lty=1)

# Normal Approximation with IID Theory SEs
classic_se <- sd(sample_dat)/sqrt(length(sample_dat))
x <- seq(5, 10, by=0.01)
fx2 <- dnorm(x, sample_mean, classic_se)
lines(x, fx2, col=rgb(1, 0, 0, .8), lty=1)
ci_normal <- qnorm(c(.025, .975), sample_mean, classic_se)
#ci_normal <- sample_mean+c(-1.96, +1.96)*classic_se
abline(v=ci_normal, col=rgb(1, 0, 0, .8), lty=3)
```


:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
Note that you can also plug in Bootstrap or Jacknife SEs to the Normal Approximation.
```{r}
## Bootstrap Distribution
hist(bootstrap_means, breaks=25,
    main=NA, border=NA,
    freq=F, ylim=c(0, 0.7),
    xlab=expression(hat(b)[b]))
abline(v=boot_ci_percentile, lty=1)
title('Percentile vs Normal 95% CIs (with Jackknife SE)', font.main=1)

# Jackknife SE Estimate
n <- length(sample_dat)
jackknife_means <- rep(NA, n)
for(i in seq_along(jackknife_means)){
    dat_noti <- sample_dat[-i]
    mean_noti <- mean(dat_noti)
    jackknife_means[i] <- mean_noti
}
se_jack <- sd(jackknife_means)*sqrt(n)

# Normal Approx with Jack SE
x <- seq(5, 10, by=0.01)
fx <- dnorm(x, sample_mean, se_jack)
lines(x, fx, col=rgb(0, 0, 1, .8), lty=1)
ci_normal_alt <- qnorm(c(.025, .975), sample_mean, se_jack)
abline(v=ci_normal_alt, col=rgb(0, 0, 1, .8), lty=3)
```
Update the example to use `se_boot <- sd(bootstrap_means)` in place of `se_jack`
:::



## Theoretical Properties

A $Z\%$ confidence interval for the mean is an interval where $Z\%$ of the intervals generated on different samples will contain the population mean $\mu$.

```{r, echo=T}
n <- 30
B <- 100
sample_CIs <- matrix(nrow=B, ncol=2)
for(i in seq(1, B)){
    # Random Sample
    X <- runif(n)
    # Confidence Interval for each sample
    M <- mean(X)
    SE <- sd(X)/sqrt(n)
    CI <- qnorm( c(0.05, 0.95), M, SE)
    sample_CIs[i, ] <- CI
}


# Explicit calculation: whether the true mean is in each CI
mu_true <- 0.5 # theoretical result for uniform samples
covered <- mu_true >= sample_CIs[, 1] & mu_true <= sample_CIs[, 2]

# Visualize confidence intervals
plot.new()
plot.window(xlim = range(sample_CIs), ylim = c(0, B))
for(i in seq(1, B) ) {
  col_i <- if (covered[i]) rgb(0, 0, 0, 0.3) else rgb(1, 0, 0, 0.5)
  segments(
    sample_CIs[i, 1], i,
    sample_CIs[i, 2], i,
    col = col_i, lwd = 2)
}
abline(v = mu_true, col = rgb(0, 0, 1, .8), lwd = 2)
axis(1)
title('90% Coverage (Red = Missed)', font.main=1)
```

We typically have only one sample and compute a single confidence interval with the data we have. The interval you computed either contains the true mean or it does not.^[As such, a $Z\%$ confidence interval does not imply a $Z\%$ probability that the true parameter lies within a particular calculated interval. In practice, people often interpret confidence intervals informally as "showing the uncertainty around our estimate": wider intervals correspond to higher sampling variability and less precise information about $\mu$. ] So "$Z\%$ confident" translates to "$Z\%$ of intervals constructed on different samples will contain $\mu$".

:::{.callout-note icon=false collapse="true" title="Must Know"}
A confidence interval contains many values, and every one of them is a hypothesis the data cannot rule out.
Suppose a $95\%$ confidence interval for a population mean comes out as $[6.6, 9.0]$.
The data are consistent with $\mu=7$, and also with $\mu=8$, and also with $\mu=8.9$: none of these competing values can be ruled out at the $5\%$ level.
You would, however, rule out $\mu=5$ and $\mu=10$, since they fall outside the interval.
"Failing to reject" a value is therefore a weak statement: it says only that the value is one of many still compatible with the data, not that it is correct.
We make this link between intervals and hypotheses precise in [Hypothesis Testing](https://jadamso.github.io/Rbooks/01_08_HypothesisTests.html).
:::

#### **Sampling Distribution Quantiles**. {-}

To better understand the link between our theoretical interpretation and our calculations, we will compute the quantiles of the sampling distribution. E.g., We construct a $90\%$ confidence interval of the mean by taking the $5^{th}$ and $95^{th}$ percentiles of the sampling distribution of the mean. We conduct a simulation that shows the sampling distribution of the sample mean for a uniform random sample with a sample size of $n=1000$, and then construct a confidence interval using quantiles.
```{r}
# Create 300 samples, each with 1000 random uniform variables
X_samples <- matrix(nrow=300, ncol=1000)
for(i in seq(1, nrow(X_samples))){
    X_samples[i, ] <- runif(1000)
}
sample_means <- apply(X_samples, 1, mean) # mean for each sample (row)

# Middle 90%
mq <- quantile(sample_means, probs=c(.05, .95))
paste0('we are 90% confident that the mean is between ', 
    round(mq[1], 2), ' and ', round(mq[2], 2) )

hist(sample_means,
    breaks=seq(.4, .6, by=.001),
    border=NA, freq=F,
    col=rgb(0, 0, 0, .25),
    main=NA)
title('90% Confidence Interval for the Mean', font.main=1)
abline(v=mq)
```

:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
For another example, consider the median. We now repeat the above process to estimate the median for each sample, instead of the mean. 
```{r}
## Sample Quantiles (medians)
sample_quants <- apply(X_samples, 1, quantile, probs=0.5) #quantile for each sample (row)

# Middle 90% of estimates
mq <- quantile(sample_quants, probs=c(.05, .95))
paste0('we are 90% confident that the median is between ', 
    round(mq[1], 2), ' and ', round(mq[2], 2) )

hist(sample_quants,
    breaks=seq(.4, .6, by=.001),
    border=NA, freq=F,
    col=rgb(0, 0, 0, .25),
    main=NA)
title('90% Confidence Interval for the Median', font.main=1)
abline(v=mq)
```
:::


#### **Interval Size**.  {-}

Confidence intervals shrink with more data, as averaging washes out random fluctuations. 

```{r}
# Create 300 samples, each of size n
par(mfrow=c(1, 3))
for(n in c(25, 100, 250)){
    # Simulate 300 samples
    X_samples <- matrix(nrow=300, ncol=n)
    for(i in seq(1, nrow(X_samples))){
        X_samples[i, ] <- runif(n)
    }

    # Compute means for each sample (row)
    sample_means <- apply(X_samples, 1, mean)
    # 90% Confidence Interval
    mq <- quantile(sample_means, probs=c(.05, .95))
    hist(sample_means,
        breaks=seq(.1, .9, by=.005),
        border=NA, freq=F,
        col=rgb(0, 0, 0, .25),
        main=NA)
    title(paste0('n=', n), font.main=1)
    abline(v=mq)
}
```

Confidence intervals also shrink with lower probabilities of containing the theoretical value. For a fixed sample size $n$, there is a trade-off between *precision*: the width of a confidence interval, and *accuracy*: the probability that a confidence interval contains the theoretical value.


:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
Here is an example showing what a Normal CI estimates. 
Theoretically, $\pm 1 sd$ has 2/3 coverage.
Change to $\pm 2 sd$ to see Precision-Accuracy tradeoff.

```{r}
# Normal CI for each sample
xq <- apply(X_samples, 1, function(r){ #theoretical se's
    mean(r) + c(-1, 1)*sd(r)/sqrt(length(r))
})
# First 3 interval estimates
xq[, c(1, 2, 3)]

# Explicit calculation
mu_true <- 0.5 # theoretical result for uniform samples
# Logical vector: whether the true mean is in each CI
covered <- mu_true >= xq[1, ] & mu_true <= xq[2, ]
# Empirical coverage rate
coverage_rate <- mean(covered)
cat(sprintf('Estimated coverage probability: %.2f%%\n', 100 * coverage_rate))
```
:::

## Misc. Topics

#### **One-Sided Intervals**. {-}

Above, our confidence intervals were two-sided: they contained the middle $Z\%$ of the sampling distribution. We can also construct one-sided intervals that extend to infinity in one direction.

A one-sided interval is shifted to one side, containing one tail rather than the middle. For example, an upper-bounded interval uses $(-\infty, q_{0.95}]$, where $q_{0.95}$ is the $95^{\text{th}}$ percentile of the bootstrap distribution: we are $95\%$ confident the true value is at most $q_{0.95}$. A lower-bounded interval uses $[q_{0.05}, \infty)$, where $q_{0.05}$ is the $5^{\text{th}}$ percentile of the bootstrap distribution: we are $95\%$ confident the true value is at least $q_{0.05}$.

:::{.callout-note icon=false collapse="true" title="Must Know"}
For the electricity bill example above with $\hat{M}=142$ and $SE=5$, a one-sided $95\%$ upper-bounded confidence interval is
$$(-\infty,~ \hat{M} + 1.645 \times SE] = (-\infty,~ 142 + 1.645 \times 5] = (-\infty,~ 150.2].$$
We are $95\%$ confident the population mean bill is at most $\$150.20$. Note that $1.645$ is the $95^{\text{th}}$ percentile of the standard normal (compared to $1.96$ for the two-sided case).
```{r}
M_hat <- 142
SE <- 5
# One-sided 95% upper bound
M_hat + qnorm(0.95) * SE
```
:::

:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
Here is a bootstrap right-tail example
```{r}
# Bootstrap Distribution
set.seed(1) # to be replicable
bootstrap_means <- rep(NA, 9999)
for(b in seq_along(bootstrap_means)){
    dat_id <- seq(1, length(sample_dat))
    boot_id <- sample(dat_id, replace=T)
    dat_b  <- sample_dat[boot_id] # c.f. jackknife
    mean_b <- mean(dat_b)
    bootstrap_means[b] <-mean_b
}

# One-sided 95% lower bound
# We are 95% confident the mean is at least q_05
hist(bootstrap_means, border=NA, breaks=50,
    freq=F, main=NA, xlab='Bootstrap')
abline(v=sample_mean, col=rgb(0, 0, 1, .8))
ci_95 <- quantile(bootstrap_means, probs=c(0.05, 1) )
abline(v=ci_95, lwd=2)
```
:::


#### **Prediction Intervals**. {-}

Sometimes the question is not "what is the population mean?" but "what range should I expect a single new observation to fall in?".

:::{.callout-important icon=false title="Key Definition" collapse="false"}
A *prediction interval* is a range that aims to contain a future single observation, not a statistic.
We compute it from the lower and upper quantiles of the data themselves, so its width reflects the spread of individual observations rather than the spread of a sampling distribution.
:::

Prediction intervals are useful when the decision depends on individual outcomes rather than averages: forecasting whether a single warranty claim, a single weather day, or a single test score will fall in a tolerable range.
Note that $Z\%$ confidence intervals do not generally cover $Z\%$ of the data: in the examples above, the $90\%$ confidence interval for the mean is roughly $[0.48, 0.52]$, which theoretically covers only a $0.04$ proportion of uniform random data, far less than $0.9$.
```{r}
x <- runif(1000)
# Middle 90% of values
xq0 <- quantile(x, probs=c(.05, .95))
paste0('we are 90% confident that a future data point will be between ', 
    round(xq0[1], 2), ' and ', round(xq0[2], 2) )

hist(x,
    breaks=seq(0, 1, by=.01), border=NA,
    freq=F, main=NA)
title('Prediction Interval', font.main=1)
abline(v=xq0)
```

Confidence intervals also apply to regression coefficients. See [Simple Regression](https://jadamso.github.io/Rbooks/02_13_SimpleRegression.html) for how these ideas extend to slopes and intercepts.


## Exercises

1. A 95% confidence interval does not mean there is a 95% probability the true parameter lies inside that particular interval. Explain what "95% confident" actually means in terms of repeated sampling.

2. Suppose a sample of $n = 49$ light bulbs has a mean lifetime of $\hat{M} = 1200$ hours with a sample standard deviation of $\hat{S} = 140$ hours. Compute the 90% and 99% Normal-approximation confidence intervals for the population mean lifetime. Which interval is wider, and why?

3. Using the `USArrests` dataset in R, construct a bootstrap 95% confidence interval for the median of the `UrbanPop` variable. Use at least 9999 bootstrap resamples and plot the bootstrap distribution with the interval boundaries.

#### **Further Reading**. {-}

* <https://learningstatisticswithr.com/book/estimation.html#interval-estimates> -- Confidence intervals explained with intuition and R code.
* <https://www.khanacademy.org/math/statistics-probability/confidence-intervals-one-sample> -- Video walkthroughs of confidence interval construction and interpretation.
* <https://www.econometrics-with-r.org/3.1-hypothesis-tests-and-confidence-intervals-in-the-simple-linear-regression-model.html> -- Confidence intervals applied to regression, bridging descriptive and inferential statistics.

#### **Recall** {-}

This chapter built *confidence intervals* two ways (the *bootstrap percentile* method and the *Normal approximation*), explained what "$Z\%$ confident" means in repeated sampling, and showed the precision–accuracy trade-off.
The electricity-bill example tied the Normal formula to numbers: with $\hat{M}=142$, $\hat{S}=30$, and $n=36$, the classical standard error is $SE(M) = 30/\sqrt{36} = 5$, and the $95\%$ margin of error is $E = 1.96 \times 5 = 9.8$, giving a confidence interval of $[\$132.20, \$151.80]$.
In the next chapter we use the same machinery to test specific hypotheses about the population.
