# Data & Visualization
***

This chapter classifies the kinds of data you will work with and gives you the basic tools to visualize and summarize a single variable.
We start with data types (cardinal vs factor), then look at the structures that hold datasets in R (lists and data.frames), and finish with the standard ways to depict a distribution: proportions, histograms, ECDFs, quantiles, and boxplots.
Which depiction fits depends on the data type: proportions suit discrete data, histograms suit continuous data, and the ECDF, quantiles, and boxplot work for either.

## Data Types

Before any analysis, it helps to know whether the values in our dataset support arithmetic (their differences carry consistent meaning) or whether they are merely labels.

:::{.callout-important icon=false title="Key Definition" collapse="false"}
*Cardinal* (or *numeric*) data are values where the difference between elements always means the same thing.

*Factor* data are values where the difference between elements does not always mean the same thing.
:::

The type of a variable governs what summaries are meaningful: cardinal data can be averaged, scattered, and used in regressions, while factor data is best counted and tabulated.
Trying to compute the mean of factor data is a category error rather than a meaningful summary, and most statistical software will refuse to do so.

#### **Basic Types** {-}
Cardinal data come in two flavors (*discrete* or *continuous*) and factor data come in two flavors (*ordered* or *unordered*), giving four sub-types in all:

* *Cardinal (Numeric)*
    * *Discrete*: E.g. $\{ 1,2,3\}$ and notice that $2-1=3-2$.
    * *Continuous*: E.g., $\{1.4348, 2.4348, 2.9, 3.9 \}$ and notice that $2.9-1.4348=3.9-2.4348$
* *Factor*
    * *Ordered*: E.g., $\{1^{st}, 2^{nd}, 3^{rd}\}$ place in a race and notice that $1^{st}$ - $2^{nd}$ place does not equal $2^{nd}$ - $3^{rd}$ place for a very competitive person who cares only about winning.
    * *Unordered* (categorical): E.g., $\{Amanda, Bert, Charlie\}$ and notice that $Amanda - Bert$ never makes sense.

Note that for theoretical analysis, the types are sometimes grouped differently as

* Discrete (discrete cardinal, ordered factor, and unordered factor data).
  You can count the potential values.
  E.g., the set $\{A,B,C\}$ has three potential values.
* Continuous (continuous cardinal data).
  There are uncountably infinite potential values.
  E.g., Try counting the numbers between $0$ and $1$ including decimal points, and notice that any two numbers have another potential number between them.

:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
Here are some examples
```{r}
dat_card1 <- c(1, 2, 3) # Cardinal data (Discrete)
dat_card1

dat_card2 <- c(1.1, 2/3, 3) # Cardinal data (Continuous)
dat_card2

dat_fact1 <- factor( c('A', 'B', 'C'), ordered=TRUE) # Factor data (Ordinal)
dat_fact1

dat_fact2 <- factor( c('Leipzig', 'Los Angeles', 'Logan'), ordered=FALSE) # Factor data (Categorical)
dat_fact2

dat_fact3 <- factor( c(TRUE, FALSE), ordered=FALSE) # Factor data (Categorical)
dat_fact3

# Explicitly check the data types:
#class(dat_card1)
#class(dat_card2)
```
:::

#### **Strings** {-} 
Note that R allows for unstructured plain text, called *character strings*, which we can then format as factors
```{r}
c('A', 'B', 'C')  # character strings
c('Leipzig', 'Los Angeles', 'Logan')  # character strings
```
Also note that strings are encounter in a variety of settings, and you often have to format them after reading them into R (the [RStudio regex cheatsheet](https://raw.githubusercontent.com/rstudio/cheatsheets/main/regex.pdf) is a handy reference).^[We will not cover the statistical analysis of text in this book, but strings are amenable to statistical analysis.]
```{r}
# Strings
paste( 'hi', 'mom')
paste( c('hi', 'mom'), collapse='--')

kingText <- 'The king infringes the law on playing curling.'
gsub(pattern='ing', replacement='', kingText)
# advanced usage
#gsub('[aeiouy]', '_', kingText)
#gsub('([[:alpha:]]{3})ing\\b', '\\1', kingText) 
```

## Datasets

Datasets can be stored in a variety of formats on your computer.
But they can be analyzed in R in three basic ways.

#### **Lists** {-}

Lists are probably the most basic type
```{r}
x0_hat <- seq(1, 10)                       # vector of 1..10
y_hat <- 2*x0_hat                              # element-wise doubling
list(x0_hat, y_hat)                            # list holds two vectors of equal length

x_mat1 <- matrix( seq(2, 7), 2, 3)    # 2x3 matrix
x_mat2 <- matrix( seq(4, -1), 2, 3)
list(x_mat1, x_mat2)                  # list can also hold matrices
```

Lists are useful for storing unstructured data
```{r}
list(list(x_mat1), list(x_mat2))  # list of lists

list(x_mat1, list(x_mat1, x_mat2)) # list of different objects

# ...inception...
list(x_mat1,
    list(x_mat1, x_mat2), 
    list(x_mat1, list(x_mat2)
    )) 
```

#### **Data.frames** {-}

A *data.frame* looks like a matrix but each column can be a different type of vector (or even list).
This allows you to combine different data types into a single object for analysis, which is why it might be your most common object.

```{r}
# data.frames: your most common data type
    # matrix of different data-types
    # well-ordered lists
data.frame(x=x0_hat, y=y_hat)  # list of vectors
```

:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
Create a data.frame storing two different types of data.
Then show print only the second column
```{r}
dat <- data.frame(x=dat_fact2, y=dat_card2)
dat

dat[, 'y']
```
:::

Data.frames stack two dimensions, rows and columns.
The generalization to three or more dimensions is the *array*.

#### **Initial Data Inspection** {-}
Regardless of the data types you have, you typically begin by inspecting your data by examining the first few observations.
 
Consider, for example, historical data on crime in the US.

```{r}
head(USArrests) # Actual Data

# Check NA values
x0_na <- c(3, 3.1, NA, 0.02) # Small dataset with a missing value
sum( is.na(x0_na) )
```




## Proportions



In what follows, we will often work with data as a vector, where there are $n$ observations and $\hat{x}_{i}$ is the value of the $i$th one.
The hat marks a value computed from observed data.
We often analyze observations in comparison to some value $x$.

```{r}
x0_hat <- c(3, 3.1, 0.02) # Small dataset we will use in numerical examples
x <- 2 # particular value: 'little x'
sum(x0_hat <= x)
sum(x0_hat == x)
```

#### **Empirical Mass Function**  {-}

If you have factor data, or discrete cardinal data, you can directly plot the proportions.
For each unique outcome $x$ we compute
$$
\hat{p}(x)=\sum_{i=1}^{n}\mathbf{1}\left(\hat{x}_{i}=x\right)/n.
$$
where $n$ is the number of observations and $\sum_{i=1}^{n}\mathbf{1}\left(\hat{x}_{i}=x\right)$ counts the number of observations equal to $x$.

:::{.callout-note icon=false collapse="true" title="Must Know"}
For example, consider the dataset $\{1,2,1,3,1\}$ with $n=5$ observations.
The value $x=1$ appears three times, so $\hat{p}(1)=3/5$.
The value $x=2$ appears once, so $\hat{p}(2)=1/5$.
The value $x=3$ appears once, so $\hat{p}(3)=1/5$.
The proportions sum to one: $3/5 + 1/5 + 1/5 = 1$.
```{r}
x0_hat <- c(1, 2, 1, 3, 1)
table(x0_hat)
table(x0_hat)/length(x0_hat)
```
:::

```{r}
# Discretized data
x_hat <- USArrests[, 'Murder'] # lab data: murder arrests per 100,000
x_hat[1:3]
x_rounded <- floor(x_hat) #rounded down, to be discrete
x_rounded[1:3]

p_hat <- table(x_rounded)/length(x_rounded)
plot(p_hat, col=grey(0, .5),
    xlab='Murder Arrests (Discretized)',
    ylab='Proportion of States with each value')
```

There are several stylistic variations: a *bar plot* uses thick bars for each outcome, a *lollipop plot* above uses skinny lines with a circle on top.
E.g., add `points(names(p_hat), p_hat, pch=16)`
The most important point is that the height of each line equals the proportion of data with a specific value.

#### **Histogram Density** {-}

For continuous data, no single value is likely to repeat, so we summarize the distribution by counting how many observations land in each of several equal-width intervals.

:::{.callout-important icon=false title="Key Definition" collapse="false"}
A *histogram density* divides the range of the data into bins of equal half-width $h$.
For an exclusive bin $(x-h, x+h]$ with midpoint $x$, it reports the rescaled count
$$\hat{f}(x) = \frac{\sum_{i=1}^{n} \mathbf{1}\left( \hat{x}_{i}~\text{in}~(x-h, x+h] \right)}{n \cdot 2h},$$
where $\mathbf{1}(\cdot)$ is an *indicator function* that equals $1$ if the expression inside is TRUE and $0$ otherwise.
:::

The *[histogram](https://en.wikipedia.org/wiki/Histogram)* is the workhorse picture for the *shape* of a continuous distribution: where observations are concentrated, whether there are multiple peaks, whether one tail is longer than the other, and whether there are gaps with no observations.
These features are hard to spot in a list of numbers but jump out at a glance from a histogram.
This rescaling makes the total area of all bars equal to one, so bar heights have units of probability per unit of $x$.
It also makes histogram heights comparable across samples of different sizes and lets us interpret the bar height as a *density*: the proportion of data per unit of $x$.
The half-width $h$ acts as a *bandwidth*: a small $h$ produces a jagged picture sensitive to single observations, while a large $h$ smooths over real features.
Note that $h$ is measured from the bin midpoint out to one edge, so a bin is $2h$ wide.

:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
E.g., suppose $\hat{x}_{i}=3.8$ and $h=1/2$.

For $x=1$ we have $\mathbf{1}\left( \hat{x}_{i}~\text{in}~\left(1-h, 1+h \right] \right)=\mathbf{1}\left( 3.8~\text{in}~\left(0.5, 1.5\right] \right)=0$.

For $x=4$ $\mathbf{1}\left( \hat{x}_{i}~\text{in}~\left(4-h, 4+h \right] \right)=\mathbf{1}\left(  3.8~\text{in}~\left(3.5, 4.5\right] \right)=1$.
:::

Note that rectangle area equals the proportion of data in the bin.
Recalling that the area of the rectangle is "base x height":
$$\begin{aligned}
2h \times \hat{f}(x) &= \frac{\sum_{i=1}^{n} \mathbf{1}\left( \hat{x}_{i}~\text{in}~\left(x-h, x+h \right] \right)}{n}.
\end{aligned}$$
We compute $\hat{f}(x)$ for each bin midpoint $x$, and the area of all rectangles sums to one.
^[If $L$ distinct bins exactly cover the range, then $2h=[\text{max}(\hat{x}_{i}) - \text{min}(\hat{x}_{i})]/L$ and we can write the design points as $x~\text{in}~\left\{ \text{min}(\hat{x}_{i}) + (2\ell-1)h \right\}_{\ell=1}^{L}$, since consecutive midpoints sit $2h$ apart.]

:::{.callout-note icon=false collapse="true" title="Must Know"}
For example, consider the dataset $\{3,3.1,0.02\}$ and use bins $(0,1], (1,2], (2,3], (3,4]$.
In this case, the midpoints are $x=(0.5,1.5,2.5,3.5)$ and $h=1/2$.
Then the counts at each midpoints are $(1,0,1,1)$.
Since $\frac{1}{n 2h}=\frac{1}{3\times1}=\frac{1}{3}$, we can rescale the counts to compute the density as $\hat{f}(x)=(1,0,1,1) \frac{1}{3}=(1/3,0,1/3,1/3)$.

```{r}
# Intuitive Examples
x0_hat <- c(3, 3.1, 0.02)
f_hat <- hist(x0_hat, breaks=c(0, 1, 2, 3, 4), plot=FALSE, freq=FALSE)
f_hat

# base x height
base <- 1
height <- f_hat[['density']]
sum(base*height)
```

For another example, use the bins $(0,2]$ and $(2,4]$.
So the midpoints are $1$ and $3$, and the bin half-width is $1$.
Only one observation, $0.02$, falls in the bin $(0,2]$.
The other two observations, $3$ and $3.1$, fall into the bin $(2,4]$.
The scaling factor is $\frac{1}{n 2h}=\frac{1}{3\times 2 \times 1}=\frac{1}{6}$.
So the first bin has density $\hat{f}(1)=1\frac{1}{6}=1/6$ and the second bin has density $\hat{f}(3)=2\frac{1}{6}=2/6$.
The area of the first bin's rectangle is $2 \times \hat{f}(1)=2/6=1/3$ and the area of the second rectangle is $2 \times \hat{f}(3)=4/6=2/3$.

Now intuitively work through an example with three bins instead of four.
Compute the areas 
```{r}
f_hat <- hist(x0_hat, breaks=c(0, 4/3, 8/3, 4), plot=FALSE, freq=FALSE)
base <- 4/3
height <- f_hat[['density']]
sum(base*height)

# as a default, R uses bins (, ] instead of [, )
# but you can change that with 'right=F'
# hist(x0_hat, breaks=c(0, 4/3, 8/3, 4), plot=F, right=F)
```
:::

```{r}
# Practical Example
hist(x_hat,
    breaks=seq(0, 20, by=1), # bin width =1, so the half-width is h=1/2
    freq=FALSE,
    border=NA,
    main=NA,
    xlab='Murder Arrests')
title('Murder Arrests', font.main=1)
# Raw Observations
rug(x_hat, col=grey(0, .5))

# Since 2h=1, the density equals the proportion of states in each bin
# Redo this example with h=1 (i.e., breaks spaced 2h=2 apart)
```
 
Note that histograms qualitatively depict proportions over multiple bins (since each bar represents the proportion of data in its bins), but you can also directly compute proportions.

:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
E.g., the proportion of murder rates in $(2,4)$ is
```{r}
mean( x_hat > 2 & x_hat <4)
```
:::



## Distributions

#### **Empirical *Cumulative* Distribution Function** {-}

Another way to summarize a distribution asks, for each candidate value $x$, what proportion of the sample lies at or below it.
It is a *step function* (see this [ECDF walkthrough](https://library.virginia.edu/data/articles/understanding-empirical-cumulative-distribution-functions) for visual examples) and contains exactly the same information as the sorted dataset, just packaged as a function of $x$ rather than a list of values.

:::{.callout-important icon=false title="Key Definition" collapse="false"}
The *empirical cumulative distribution function* (ECDF) is the proportion of observations in the sample whose values are less than or equal to $x$,
$$\hat{F}(x) = \frac{1}{n}\sum_{i=1}^{n}\mathbf{1}(\hat{x}_{i} \leq x).$$
It jumps up by $1/n$ at each data point and equals $1$ once $x$ passes the largest observation.
:::



:::{.callout-note icon=false collapse="true" title="Must Know"}
For example, consider the dataset $\{3,3.1,0.02 \}$.
We reorder the observations as $\{0.02, 3, 3.1 \}$, so that there are discrete jumps of $1/n=1/3$ at each value.
Consider the points $x~\text{in}~\{0.5,2.5,3.5\}$.
At $x=0.5$, $\hat{F}(0.5)$ measures the proportion of the data $\leq 0.5$.
Since only one observation, $0.02$, of three is $\leq 0.5$, we can compute $\hat{F}(0.5)=1/3$.
Similarly, since only one observation, $0.02$, of three is $\leq 2.5$, we can compute $\hat{F}(2.5)=1/3$.
Since all observations are $\leq 3.5$, we can compute $\hat{F}(3.5)=1$.

```{r}
F_hat <- ecdf(x0_hat)

# Visualized
plot(F_hat)

# Evaluated at each data point
x <- x0_hat
F_hat(x0_hat)
#sum(x0_hat<=3)/length(x0_hat)
#sum(x0_hat<=3.1)/length(x0_hat)
#sum(x0_hat<=0.02)/length(x0_hat)

# Evaluated at other points
x <- c(0.5, 2.5, 3.5)
F_hat(x)
#sum(x0_hat<=0.5)/length(x0_hat)
#sum(x0_hat<=2.5)/length(x0_hat)
#sum(x0_hat<=3.5)/length(x0_hat)

# Evaluated at the histogram bin edges
x <- c(0, 1, 2, 3, 4)
F_hat(x)
diff(F_hat(x)) # proportion of data in each bin, as in the histogram
```
:::

```{r}
F_hat <- ecdf(x_hat)
# proportion of murders <= 10
F_hat(10)
# proportion of murders <= x, for all x
plot(F_hat, main=NA,
    xlab='Murder Arrests (x)',
    ylab='Proportion of States with Murder Arrests <= x',
    pch=16, col=grey(0, .5))
title('Murder Arrests ECDF', font.main=1)
rug(x_hat)
```

The ECDF is useful when we want precise numbers rather than a picture of shape: "what fraction of states have at most $10$ murders per $100$k?" reads off directly without choosing bin widths or smoothing parameters.
The ECDF also applies to more data types than the two previous depictions.
The mass function $\hat{p}$ needs values that repeat, so it suits discrete data, and the histogram $\hat{f}$ needs a bin width, so it suits continuous data.
The ECDF needs neither, since its only operation is checking whether $\hat{x}_{i} \leq x$, so it works for discrete and continuous data alike.
The ECDF does not work for unordered factors, where "$\leq$" has no meaning, but it does work for ordered factors such as letter grades.

Still, the ECDF and the histogram are two views of the same counts.
The area of a histogram bar is the proportion of data in the bin $(x-h, x+h]$, which is also the rise in the ECDF across that bin.
$$\begin{aligned}
2h \times \hat{f}(x) &= \hat{F}(x+h) - \hat{F}(x-h).
\end{aligned}$$
Conversely, adding up bar areas from left to right recovers $\hat{F}$ at each bin edge.^[
Summing the identity above over the bin midpoints $x'$ at or below $x$ telescopes, since the upper edge of each bin is the lower edge of the next:
$\sum_{x' \leq x} 2h \times \hat{f}(x') = \sum_{x' \leq x} \left[ \hat{F}(x'+h) - \hat{F}(x'-h) \right] = \hat{F}(x+h) - \hat{F}(x_{1}-h)$
where $x_{1}$ is the first midpoint.
The lowest bin edge $x_{1}-h$ lies below every observation, so $\hat{F}(x_{1}-h)=0$ and the bar areas up to $x$ add up to $\hat{F}(x+h)$.
For discrete data the same relationship holds without bins: $\hat{F}(x)=\sum_{x' \leq x} \hat{p}(x')$ accumulates the proportions, and the size of each jump in $\hat{F}$ is $\hat{p}(x)$.
]
For a continuous example, the shaded bars on the histogram cover the states with at most $5$ murder arrests, and their total area equals the height of the colored point on the ECDF, $\hat{F}(5)$.


```{r}
F_hat <- ecdf(x_hat)
breaks <- seq(0, 20, by=1) # bin width 2h=1
f_hat <- hist(x_hat, breaks=breaks, plot=FALSE)

# Same data, stacked: the shaded area on top equals the height of the point below
x <- 5 # particular value
below <- f_hat[['mids']] < x # bins at or below x
col_x <- rgb(1, 0, 0, .5)
par(mfrow=c(2, 1), mar=c(4, 4, 1, 1))

# Numerically verify shaded area equals point height
sum(f_hat[['density']][below]*diff(breaks)[below])
F_hat(x)

# Histogram: shade the bins at or below x
hist(x_hat, breaks=breaks, freq=FALSE, border=NA,
    col=ifelse(below, col_x, grey(.8)),
    main=NA, xlab=NA, ylab='Density')

# ECDF: mark the point (x, F_hat(x))
plot(F_hat, xlim=range(breaks), pch=16, col=grey(0, .5),
    main=NA, xlab='Murder Arrests', ylab='Proportion <= x')
segments(x, F_hat(x), -10, F_hat(x), col=col_x)
segments(x, F_hat(x), x, -1, col=col_x)
points(x, F_hat(x), pch=16, col=rgb(1, 0, 0, .8), cex=1.5)
```



:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
For a concrete example, consider the dataset $\{3,3.1,0.02 \}$.
At the bin edges $x~\text{in}~\{0,1,2,3,4\}$, the ECDF is $\hat{F}=(0,1/3,1/3,2/3,1)$, so it rises by $(1/3, 0, 1/3, 1/3)$ across the bins $(0,1], (1,2], (2,3], (3,4]$.
Those rises are exactly the bin proportions that the histogram rescaled by $1/(2h)$ to get $\hat{f}(x)=(1/3,0,1/3,1/3)$.
The relationship between $\hat{f}$ and $\hat{F}$ is easy to check with real data.
```{r}
# Histogram density from the ECDF, and back again
F_hat <- ecdf(x_hat)
breaks <- seq(0, 20, by=1) # bin width 2h=1
f_hat <- hist(x_hat, breaks=breaks, plot=FALSE)

# Rise in F_hat across each bin, divided by the bin width
f_from_F <- diff(F_hat(breaks))/diff(breaks)
all.equal(f_from_F, f_hat[['density']])

# Bar areas added up from the left, evaluated at the upper bin edges
F_from_f <- cumsum(f_hat[['density']]*diff(breaks))
all.equal(F_from_f, F_hat(breaks[-1]))
```

To see this generally, start from the definition of $\hat{f}$ and write the indicator for the bin as the difference of two "at or below" indicators:
$$\begin{aligned}
2h \times \hat{f}(x) &= 2h \times \frac{\sum_{i=1}^{n} \mathbf{1}\left( \hat{x}_{i}~\text{in}~(x-h, x+h] \right)}{n \cdot 2h} \\
&= \frac{1}{n}\sum_{i=1}^{n} \mathbf{1}\left( x-h < \hat{x}_{i} \leq x+h \right) \\
&= \frac{1}{n}\sum_{i=1}^{n} \left[ \mathbf{1}\left( \hat{x}_{i} \leq x+h \right) - \mathbf{1}\left( \hat{x}_{i} \leq x-h \right) \right] \\
&= \frac{1}{n}\sum_{i=1}^{n} \mathbf{1}\left( \hat{x}_{i} \leq x+h \right) - \frac{1}{n}\sum_{i=1}^{n} \mathbf{1}\left( \hat{x}_{i} \leq x-h \right) \\
&= \hat{F}(x+h) - \hat{F}(x-h).
\end{aligned}$$
The step that splits the indicator holds observation by observation: an $\hat{x}_{i}$ at or below $x-h$ contributes $1-1=0$, one inside the bin contributes $1-0=1$, and one above $x+h$ contributes $0-0=0$.
Dividing by the bin width, $\hat{f}(x)$ is the slope of $\hat{F}$ across the bin ("rise over run"): where $\hat{F}$ climbs steeply the histogram is tall, and where $\hat{F}$ is flat the histogram is near zero.
:::



#### **Quantiles** {-}

Reading the ECDF in reverse asks: at what value $x$ does a given proportion $p$ of the data lie below?

:::{.callout-important icon=false title="Key Definition" collapse="false"}
The *$p$th quantile* of a sample is the value $x$ where $p$ percent of the data fall below $x$ and $(1-p)$ percent fall above.
The *empirical quantile function* $\hat{q}(p) = \hat{F}^{-1}(p)$ inverts the ECDF, returning the quantile at any probability $p~\text{in}~[0,1]$.
:::

Quantiles are useful for precise numerical anchors: "the top $10\%$ of cities", "half of households earn less than...", or "the middle $50\%$ of test scores fall between $x_1$ and $x_2$".
They also describe spread (range, $IQR$) and shape (the gap between median and mean) without committing to a smooth distributional model.

Five quantiles get their own names because they summarize the distribution at a glance:

* The *min* (the $p=0$ quantile) is the smallest value, where $0\%$ of the data have lower values.
* The *lower quartile* (the $p=0.25$ quantile) is where $25\%$ of the data have lower values.
* The *median* (the $p=0.5$ quantile) is the middle value, where $50\%$ of the data have lower values.
* The *upper quartile* (the $p=0.75$ quantile) is where $75\%$ of the data have lower values.
* The *max* (the $p=1$ quantile) is the largest value, where $100\%$ of the data have lower values.

We also speak of *deciles* at $p=0.1, 0.2, \ldots, 0.9$ when we want a finer split.

:::{.callout-note icon=false collapse="true" title="Must Know"}
For example, that dataset $\{0,0,0.02,3,5\}$ has a median of $0.02$, with the lower quartile $0$ and the upper quartile $3$.
(The number $0$ is also special: the most frequent observation is called the *mode*.)

```{r}
quantile(x0_hat, probs=c(0, .5, 1))
```

Now work through an intuitive example with observations $\{1,2,...,13\}$.
Hint: split the ordered observations into four groups.
:::

```{r}
# common quantiles
quantile(x_hat)

# All deciles are quantiles
quantile(x_hat, probs=seq(0, 1, by=.1))

# Visualized: Inverting the Empirical Distribution
F_hat <- ecdf(x_hat)
plot(F_hat, lwd=2, xlim=c(0, 20),
    pch=16, col=grey(0, .5), main=NA)
title('Inverting the ECDF', font.main=1)
# Two Examples of Quantiles
p <- c(.25, .9) # Lower Quartile, Upper Decile
cols <- c(rgb(1, 0, 0, .8), rgb(0, 0, 1, .8))
q_hat <- quantile(x_hat, p, type=1)
q_hat
segments(q_hat, p, -10, p, col=cols)
segments(q_hat, p, q_hat, 0, col=cols)
mtext( round(q_hat, 2), 1, at=q_hat, col=cols)
```

There are some issues with quantiles with smaller datasets.
E.g., to compute the median of $\{3,3.1,0,1\}$, we need some ways to break ties (of which there are many options).
The most common way is perhaps a simple average of the tied values: $(3+1)/2=2$.

:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
To calculate quantiles, the computer defaults to sorting the observations from smallest to largest as $\hat{x}_{(1)}, \hat{x}_{(2)},... \hat{x}_{(n)}$, and then computes quantiles as $\hat{x}_{ (p \times n) }$, where $p~\text{in}~[0,1]$ refers to probability and $(p\times n)$ is rounded.
There are different ways to break ties and the min ($p=0$) is a special case.
```{r}
x_sorted_hat <- sort(x_hat)
x_sorted_hat

# median
x_sorted_hat[ceiling(length(x_sorted_hat)*.5)]
quantile(x_hat, probs=.5, type=1) #tie break rule #1

## arbitrary quantiles
p <- 0.11
x_sorted_hat[ceiling(length(x_sorted_hat)*p)]
quantile(x_hat, probs=p, type=1)

# min is a special case
min(x_sorted_hat)
x_sorted_hat[ceiling(length(x_sorted_hat)*0)]
quantile(x_hat, probs=0, type=1)
x_sorted_hat[1]
```
:::

#### **Boxplots** {-}
*[Boxplots](https://en.wikipedia.org/wiki/Box_plot)* also summarize the distribution.
The boxplot shows the median (solid black line) and *interquartile range* ($IQR=$ upper quartile $-$ lower quartile; filled box).^[
Technically, the upper and lower *hinges* use two different versions of the first and third quartile.
Statistical packages differ in how they compute both the quartiles and the fences that flag outliers [@frigge1989].
See also <https://stackoverflow.com/questions/40634693/lower-and-upper-quartiles-in-boxplot-in-r>.
]
As a default, whiskers are shown as $1.5\times IQR$ and values beyond that are highlighted as *outliers* (so whiskers do not typically show the data range).
No state sits more than $1.5\times IQR$ beyond a quartile, so the boxplot below draws no outlier circles.
Like the ECDF, the boxplot works for discrete and continuous data alike, since it is built from quantiles rather than from bins or repeated values.
It does need cardinal data: the whiskers extend $1.5\times IQR$ beyond the box, and that distance is only meaningful when differences are.

```{r}
# Default: median, box (IQR), whiskers at 1.5*IQR, outliers as circles
boxplot(x_hat,
    main=NA, ylab='Murder Arrests')
title('Default', font.main=1)
```

The default boxplot compresses $n$ observations into five numbers, which discards information about shape.
You can alternatively show all the raw data points instead of whisker+outliers.
This exposes the gaps, clusters, and repeated values that a five-number summary hides [@weissgerber2015].

```{r}
# Alternative: every observation, in place of whiskers and outliers
boxplot(x_hat,
    main=NA, ylab='Murder Arrests',
    whisklty=0, staplelty=0, outline=FALSE)
title('With Raw Data', font.main=1)
# Raw Observations
stripchart(x_hat,
    pch='-', col=grey(0, .5), cex=2,
    vert=TRUE, add=TRUE)
```

Each depiction in this chapter suits some data types and not others.
Boxplots are generally not well suited for factor data but better suited for comparing groups (discussed later in Bivariate Data).
For a single data vector, the mass function needs repeating values but works for unordered factors whereas the ECDF works for discrete and continuous data requires an ordering.^[
In R, `ecdf` and `quantile(..., type=1)` accept ordered factors, but `boxplot` silently treats the levels as the numbers $1, 2, 3, \ldots$.
]
Here is a table to help you recall

| Visual | Unordered factor | Ordered factor | Discrete cardinal | Continuous cardinal |
|-----------|------------------|----------------|-------------------|---------------------|
| Mass Function $\hat{p}(x)$ | Yes | Yes | Yes | No |
| Histogram $\hat{f}(x)$ | No | No | Yes | Yes |
| ECDF $\hat{F}(x)$ | No | Yes | Yes | Yes |
| Boxplot | No | No | Yes | Yes |


## Exercises

1. Comment the script you wrote for this chapter, then restart R and check that it runs from a clean session, then check the script with AI as explained in [Working with AI](04_A_WorkingWithAI.qmd).
Write three sentences from memory on the main statistical idea of this chapter, and ask the assistant what is wrong, vague, or missing.
Finish with your own questions about whatever you found hardest.

2. Classify each of the following as cardinal (discrete or continuous) or factor (ordered or unordered): (a) temperature in Celsius, (b) letter grade on a transcript, (c) number of siblings, (d) province of residence.
   Briefly explain each choice.

3. The built-in dataset `mtcars` has a column `mpg`.
   Compute the histogram density by hand for the bins $(10,15]$, $(15,20]$, $(20,25]$, $(25,30]$, $(30,35]$.
   That is, for each bin count the number of observations, then divide by $n \times 2h$ where $2h=5$.
   Verify your answer matches `hist(mtcars$mpg, breaks=seq(10,35,by=5), freq=FALSE)`.

4. Load `USArrests` in R.
   Compute the ECDF of the `Assault` variable and use it to find the proportion of states with 200 or fewer assault arrests.
   Then find the upper quartile of `Assault` using `quantile`.

#### **Further Reading** {-}

* <https://library.virginia.edu/data/articles/understanding-empirical-cumulative-distribution-functions> -- Clear walkthrough of the ECDF with visual examples.
* <https://raw.githubusercontent.com/rstudio/cheatsheets/main/regex.pdf> -- RStudio cheat sheet for regular expressions and string handling in R.
* <https://www.econometrics-with-r.org/2.1-random-variables-and-probability-distributions.html> -- Review of data types and distributions in an econometrics context.

#### **Recall** {-}

This chapter sorted observations into *cardinal* and *factor* types and then showed several complementary ways to picture the distribution of a single variable: bar plots and lollipops for proportions, the *histogram density*, the *empirical CDF*, *quantiles* (min, median, max, quartiles), and the boxplot.
Proportions suit discrete data and histograms suit continuous data, while the ECDF, quantiles, and boxplot work for either.
The running three-point dataset $\{3, 3.1, 0.02\}$ ties them together: the ECDF jumps by $1/3$ at each of $0.02$, $3$, and $3.1$ before flattening at $1$, its rises across the unit bins $(1/3, 0, 1/3, 1/3)$ are the histogram bar areas, and the median reads off as $3$.
Each picture answers a different question about the same data, and switching between them is a habit worth building.
