# Advanced Probability

## Discrete Factorial Distributions

#### **Binomial**. {-}

The sum of $n$ Bernoulli's (number of successes) in $n$ trials.

\begin{eqnarray}
X_{i} \in \{0,1,\ldots,n\} \\
Prob(X_{i}=x) &=& \binom{n}{x}p^{x}(1-p)^{n-x} \\
F(x) = Prob(X_{i} \leq x) &=& \sum_{k=0}^{\lfloor x \rfloor} Prob(X_{i}=x) 
\end{eqnarray}

The simplest case: how many heads will I get when I flip a coin twice? We can first see that there are four potential outcomes: $\{ (1,1), (1,0), (0,1), (0,0) \}$, each of which are equally likely (assuming the coin is fair). The corresponding sums are $\{2,1,1,0 \}$, which leads to probabilities $Prob(X_{i}=0)=1/4$, $Prob(X_{i}=1)=2/4$, and $Prob(X_{i}=2)=1/4$.

:::{.callout-tip icon=false collapse="true"}
When I flip a coin three times, the potential outcomes are 
\begin{eqnarray}
\{
(1,1,1), (1,1,0), (1,0,1), (1,0,0),\\
(0,1,1), (0,1,0), (0,0,1), (0,0,0) 
\}
\end{eqnarray}
Assuming the coin is fair, compute the probabilities for each possible sum.
:::

In general, we use the [Binomial_distribution](https://en.wikipedia.org/wiki/Binomial_distribution). For example, suppose $n=15$ people walk past a store each hour and each has a $p=30\%$ chance of entering. The number of customers who enter is $X_{i} \sim \text{Binom}(15, 0.3)$.

```{r}
## Store: 15 passers-by, 30% chance each enters
n <- 15
p <- 0.3

# PMF plot
x <- seq(0, n)
f_x <- dbinom(x, n, p)
plot(x, f_x, type = "h", col = "blue",
    main='',  xlab = "Customers entering", ylab = "Prob(X = x)")
title(bquote(paste('Binom(',.(n),', ',.(p), ')' ) ))
points(x, f_x, pch = 16, col = "blue")

## Theoretical CDF
F_x <- pbinom(x, n, p)
plot(x, F_x, type="s", col="blue",
    xlab = "Customers entering", ylab = "F(x)")
points(x, F_x, pch=16, col="blue")
title(bquote(paste('Binom(',.(n),', ',.(p), ') CDF')))
```

:::{.callout-note icon=false collapse="true"}
A quality inspector checks $n=20$ items from a production line with a $5\%$ defect rate ($p=0.05$). What is the probability of finding $2$ or fewer defectives?

We need $Prob(X_{i} \leq 2) = F(2)$ where $X_{i} \sim \text{Binom}(20, 0.05)$. By hand, the first few terms are
\begin{eqnarray}
Prob(X_{i}=0) &=& \binom{20}{0}(0.05)^{0}(0.95)^{20} \approx 0.36 \\
Prob(X_{i}=1) &=& \binom{20}{1}(0.05)^{1}(0.95)^{19} \approx 0.38 \\
Prob(X_{i}=2) &=& \binom{20}{2}(0.05)^{2}(0.95)^{18} \approx 0.19
\end{eqnarray}
So $Prob(X_{i} \leq 2) \approx 0.36 + 0.38 + 0.19 = 0.93$. There is roughly a $93\%$ chance the inspector finds $2$ or fewer defectives.
```{r}
# Individual terms
x <- c(0,1,2)
px <- dbinom(x, size=20, prob=0.05)
round(px,2)

# CDF: Prob(X <= 2)
pbinom(2, size=20, prob=0.05)
```
:::

:::{.callout-tip icon=false collapse="true"}
Suppose that employees at a company are $70%$ female and $30%$ male. If we select a random sample of eight employees, what is the probability that $2$ in the sample are female? That $\geq 2$ are female? 

Try conducting a simulation first, 
```{r}
n <- 8
p <- 0.3
X <- rbinom(1000, n, p)
```
:::


:::{.callout-tip icon=false collapse="true"}
For $n$ coin flips, show that $\mathbb{E}[X_{i}]=np$ and $\mathbb{V}[X_{i}]=np(1-p)$. Build on the case $n=2$ above, then consider case $n=3$, and then the general case of any $n$.

```{r}
# Compare empirical simulation vs theoretical
n <- 2
p <- 3/4
X <-  rbinom(1000, size=n, prob=p)

# Mean
c(empirical = mean(X), theoretical = n*p)

# Variance
c(empirical  = var(X),  theoretical  = n*p*(1-p))
```
:::


#### **Poisson**. {-}

The Poisson distribution models how many events occur in a fixed interval, when each individual event is rare and independent and has equal probabilities at any point on the interval.

In general, the [Poisson distribution](https://en.wikipedia.org/wiki/Poisson_distribution) gives
\begin{eqnarray}
X_{i} \in \{0,1,\ldots\} \\
Prob(X_{i}=x) &=& e^{-\lambda}\lambda^x/x! \\
F(x) = Prob(X_{i} \leq x) &=& \sum_{k=0}^{\lfloor x \rfloor} Prob(X_{i}=x)
\end{eqnarray}
The single parameter $\lambda$ is the average number of events per interval.
Note the infinite discrete support $\{0,1,2,\ldots\}$: there is no upper bound on the count, though large values become vanishingly unlikely.

For example, a small store averages $\lambda = 3.5$ customers per hour. In any given hour the actual count might be 2, or 5, or 7. 
For our store, $\lambda = 3.5$ means we expect 3.5 customers per hour on average and the following assumptions hold

- Arrivals are independent of each other (one customer entering does not make the next more or less likely)
- The average rate $\lambda$ is constant over the interval
- Two customers cannot arrive at exactly the same instant

We can then compute a few probabilities by hand.E.g, the chance that zero customers arrive in a given hour is
\begin{eqnarray}
Prob(X_{i}=0) &=& e^{-3.5} \cdot 3.5^{0} / 0! = e^{-3.5} \approx 0.030
\end{eqnarray}
The chance of exactly 2 customers is
\begin{eqnarray}
Prob(X_{i}=2) &=& e^{-3.5} \cdot 3.5^{2} / 2! = e^{-3.5} \cdot 6.125 \approx 0.185
\end{eqnarray}


```{r}
## Store: lambda = 3.5 customers per hour
lambda <- 3.5

# PMF plot
x <- seq(0, 15)
f_x <- dpois(x, lambda)
plot(x, f_x, type="h", col="blue",
     xlab = "Customers per hour", ylab = "Prob(X = x)")
points(x, f_x, pch = 16, col = "blue")
title(bquote(paste('Pois(',.(lambda), ')')))

# CDF plot
F_x <- ppois(x, lambda)
plot(x, F_x, type="s", col="blue",
    xlab = "Customers per hour", ylab = "F(x)")
points(x, F_x, pch=16, col="blue")
title(bquote(paste('Pois(',.(lambda), ') CDF')))
```

:::{.callout-tip icon=false collapse="true"}
Show that $\mathbb{E}[X_{i}] = \mathbb{V}[X_{i}]= \lambda$. That is, the Poisson mean and variance are both equal to the rate parameter.
```{r}
# Compare empirical simulation vs theoretical
lambda <- 3.5
X <- rpois(1e4, lambda)
c(emp_mean = mean(X), th_mean = lambda)
c(emp_var  = var(X),  th_var  = lambda)
```
:::

:::{.callout-note icon=false collapse="true"}
During the lunch rush, the store averages $\lambda=4.5$ customers per hour. What is the probability that more than $7$ customers arrive in a given hour?

We need $Prob(X_{i} > 7) = 1 - Prob(X_{i} \leq 7) = 1 - F(7)$.
\begin{eqnarray}
Prob(X_{i} > 7) &=& 1 - \sum_{k=0}^{7} \frac{e^{-4.5} \cdot 4.5^{k}}{k!}
\end{eqnarray}
```{r}
# Prob(X > 7)
1 - ppois(7, lambda=4.5)

# Verify: sum the PMF from 0 to 7
sum(dpois(0:7, lambda=4.5))
```
There is roughly a $9\%$ chance of more than $7$ arrivals in a given hour.
:::


:::{.callout-tip icon=false collapse="true"}
The `warpbreaks` dataset records the number of warp breaks per loom in a textile factory. We can check whether a Poisson distribution fits this count data by overlaying the fitted PMF on a histogram.
```{r}
# Count data: number of warp breaks per loom
breaks <- warpbreaks$breaks
hist(breaks, breaks=20, border=NA, freq=F,
    main='Warp Breaks per Loom', font.main=1,
    xlab='Number of breaks')

# Fit a Poisson and overlay
lambda_hat <- mean(breaks)
x <- seq(0, max(breaks))
points(x, dpois(x, lambda_hat), pch=16, col="blue")
legend("topright",
    legend=bquote(hat(lambda) == .(round(lambda_hat,1))),
    pch=16, col="blue")
```
:::



The shape of the Poisson also changes with $\lambda$. A quiet specialty shop ($\lambda=1$) has a strongly right-skewed arrival pattern, while a busy supermarket ($\lambda=15$) looks nearly Normal.

```{r}
# Shape varies with lambda
par(mfrow=c(1,3))
for(lam in c(1, 5, 15)) {
    x <- seq(0, 30)
    f_x <- dpois(x, lam)
    plot(x, f_x, type="h", col="blue",
        xlab = "Customers per hour", ylab = "Prob(X = x)")
    points(x, f_x, pch=16, col="blue")
    title(bquote(paste('Pois(',.(lam), ')')))
}
```


#### **Large Sample Approximations**. {-}

The de Moivre–Laplace Theorem says that as $n$ grows large, with $p \in (0,1)$ staying fixed, the Binomial distribution is approximately Normal with mean $np$ and variance $np(1-p)$

:::{.callout-tip icon=false collapse="true"}
The unemployment rate is $10%$. Suppose that $100$ employable people are selected randomly. What is the probability that this sample contains between $9$ and $12$ unemployed people. Use the normal approximation to binomial probabilities.

Parameters are 
\begin{eqnarray}
\mu &\approx& n p = 10 \\
\sigma^2 &=& n p (1−p) = 100 (0.1)(0.9)=9 \\
\sigma &=& \sqrt{9}=3.
\end{eqnarray}
:::

The Poisson Limit Theorem says that as $n$ grows large, with $p$ growing smaller at the same time, and $n p$ converging to a fixed number $\lambda$, the Binomial distribution is approximately Poisson with parameter $\lambda$.

:::{.callout-tip icon=false collapse="true"}
Imagine dividing one hour into $n$ tiny intervals. If the store averages $\lambda = 3.5$ customers per hour, each interval has a small probability $p = \lambda/n = 3.5/n$ of an arrival. With $n$ independent intervals, the number of arrivals is Binomial:
$$Prob(X=x) = \binom{n}{x} \left(\frac{\lambda}{n}\right)^{x} \left(1-\frac{\lambda}{n}\right)^{n-x}$$

As $n$ grows large, three things happen:

1. $\binom{n}{x} \cdot \frac{1}{n^{x}} \to \frac{1}{x!}$, because $\binom{n}{x} = \frac{n(n-1)\cdots(n-x+1)}{x!}$ and each factor in the numerator is approximately $n$
2. $\left(1 - \frac{\lambda}{n}\right)^{n} \to e^{-\lambda}$, by the definition of $e$
3. $\left(1 - \frac{\lambda}{n}\right)^{-x} \to 1$, since $x$ is fixed and $\lambda/n \to 0$

Multiplying the pieces together: $\frac{1}{x!} \cdot \lambda^{x} \cdot e^{-\lambda} = e^{-\lambda}\lambda^{x}/x!$, which is the Poisson PMF. In other words, the Poisson is what the Binomial becomes when there are many opportunities for an event but each one is individually unlikely (see also Large Sample Approximations below).

```{r}
# Binomial approaches Poisson as n grows
lambda <- 3.5
x <- seq(0, 12)
f_pois <- dpois(x, lambda)

par(mfrow=c(1,3))
for(n in c(10, 50, 1000)) {
    f_binom <- dbinom(x, size=n, prob=lambda/n)
    plot(x, f_binom, type="h", col="grey60", lwd=4,
        xlab="Customers", ylab="Prob(X = x)")
    points(x + 0.2, f_pois, type="h", col="blue", lwd=2)
    points(x + 0.2, f_pois, pch=16, col="blue", cex=0.7)
    title(bquote(paste('Binom(',.(n),', ',.(round(lambda/n,4)),')')))
}
legend("topright", c("Binomial","Poisson"),
    col=c("grey60","blue"), lwd=c(4,2), cex=0.8)
```
:::


:::{.callout-tip icon=false collapse="true"}
For another example, consider $n=1000$ independent trials each with probability $p=0.004$, so $\lambda = np = 4$. We can compare the exact Binomial to both approximations at once: the Poisson ($\lambda=4$) and the Normal ($\mu=np=4$, $\sigma^{2}=np(1-p)\approx 3.98$).
```{r}
# Binom(1000, 0.004) vs Pois(4) vs Normal(4, 3.98)
n <- 1000
p <- 0.004
x <- seq(0, 15)
f_binom <- dbinom(x, size=n, prob=p)
f_pois  <- dpois(x, lambda=n*p)
f_norm  <- dnorm(x, mean=n*p, sd=sqrt(n*p*(1-p)))

plot(x, f_binom, type="h", col="blue", lwd=4,
    xlab="x", ylab="Prob(X = x)",
    main="Approximating the Binomial", font.main=1)
points(x + 0.15, f_pois, type="h", col="red", lwd=2)
lines(x, f_norm, col="darkgreen", lwd=2)
legend("topright",
    c("Binom(1000, 0.004)", "Pois(4)", "Normal(4, 3.98)"),
    col=c("blue","red","darkgreen"), lwd=c(4,2,2))
```
The Poisson is nearly exact here because $p$ is very small. The Normal is a rougher fit because $\lambda=4$ is still fairly small and the distribution is noticeably skewed. As $\lambda$ grows, the Poisson itself becomes more symmetric and the Normal approximation improves.
:::



## Continuous Factorial Distributions

#### **Irwin–Hall**. {-}

The sum of $n$ i.i.d. $\text{Uniform}(0,1)$.  

* Continuous, support $[0,n]$
* Probability Density Function: $f(x) = \dfrac{1}{(n-1)!}
\displaystyle\sum_{k=0}^{\lfloor x \rfloor} (-1)^k
\binom{n}{k} (x - k)^{n-1}$ for $x \in [0,n]$, and $0$ otherwise
* See <https://en.wikipedia.org/wiki/Irwin%E2%80%93Hall_distribution>
* A common use case: representing the sum of many small independent financial shocks

```{r}
## Minimal example (base R)
# Irwin–Hall PDF function
dirwinhall <- function(x, n) {
    f_x <- vector(length=length(x))
    for(i in seq(x)) {
        xx <- x[i]
        if(xx < 0 | xx > n){
            f_x[i] <- 0 
        } else {
            k <- seq(0, floor(xx))
            f_k <- sum((-1)^k*choose(n, k)*(xx-k)^(n-1))/factorial(n-1)
            f_x[i] <- f_k
        }
    }
    return(f_x)
}

# Parameters
n <- 2
x <- seq(0, n, length.out = 500)

# Compute and plot PDF
f_x <- dirwinhall(x, n)
plot(x, f_x, type="l", col="blue",
    main='', xlab = "x", ylab = "f(x)")
title(bquote(paste('IrwinHall(',.(n), ')')))
```

See also the [Bates](https://en.wikipedia.org/wiki/Bates_distribution) distribution, which is for the mean  of $n$ i.i.d. $\text{Uniform}(0,1)$ random variables. It essentially rescales the Irwin-Hall distribution to range between $0$ and $1$.


#### **Beta**. {-}
The sample space is any number on the unit interval, $X_{i} \in [0,1]$, but with non-uniform probabilities. 


```{r}
X4 <- rbeta(2000,2,2) ## two shape parameters
hist(X4, breaks=20, border=NA, main=NA, freq=F)

#See the underlying probabilities
#f_25 <- dbeta(.25, 2, 2)

x <- seq(0,1,by=.01)
fx <- dbeta(x, 2, 2)
lines(x, fx, col='blue')
```

The [Beta](https://en.wikipedia.org/wiki/Beta_distribution) distribution is mathematically complicated to write, and so we omit it. However, we can find the probability graphically using either the probability density function or cumulative distribution function. 

:::{.callout-tip icon=false collapse="true"}
Suppose $X_{i}$ is a random variable with a beta distribution. Intuitively depict $Prob(X_{i} \in [0.2, 0.8])$ by drawing an area under the density function. Numerically estimate that same probability using the CDF.
```{r}
plot( ecdf(X4), main=NA) # Empirical

x <- seq(0,1,by=.01) # Theoretical
Fx <- pbeta(x, 2, 2)
lines(x, Fx, col='blue')

# Middle Interval Example 
F2 <- pbeta(0.2, 2, 2)
F8 <- pbeta(0.8, 2, 2)
F_2_8 <- F8 - F2
F_2_8

# Visualize
title('Middle between 0.2 and 0.8')
segments( 0.2, F2, -1, F2, col='red')
segments( 0.8, F8, -1, F8, col='red')
```
:::

This distribution is often used, as the probability density function has two parameters that allow it to take many different shapes.

:::{.callout-tip icon=false collapse="true"}
For each example below, intuitively depict $Prob(X_{i} \leq 0.5)$ using the PDF. Repeat the exercise using a CDF instead of a PDF to calculate a numerical value.
```{r, results='hide'}
#| fig-width: 8
#| fig-height: 6
op <- par(no.readonly = TRUE); on.exit(par(op), add = TRUE)
x <- seq(0,1,by=.01)
pars <- expand.grid( c(.5,1,2), c(.5,1,2) )
par(mfrow=c(3,3))
apply(pars, 1, function(p){
    fx <- dbeta( x,p[1], p[2])
    plot(x, fx, type='l', xlim=c(0,1), ylim=c(0,4), lwd=2, col='blue')
    #hist(rbeta(2000, p[1], p[2]), breaks=50, border=NA, main=NA, freq=F)
})
title('Beta densities', outer=T, line=-1)
```
:::



## Set Theory

Let the sample space contain events $A$ and $B$, with their probability denoted as $Prob(A)$ and $Prob(B)$. Then

- the *union*: $A\cup B$, refers to either event occurring
- the *intersection*: $A\cap B$, refers to both events occurring
- the events $A$ and $B$ are *mutually exclusive* if and only if $A\cap B$ is empty
- the *inclusion–exclusion* rule is: $Prob(A\cup B)=Prob(A)+Prob(B)-Prob(A\cap B)$
- the events $A$ and $B$ are *mutually independent* if and only if $Prob(A\cap B)=Prob(A) Prob(B)$
- the *conditional probability* is $Prob(A \mid B) = Prob(A \cap B)/ P(B)$

For example, consider a six sided die where events are the number $>2$ or the number is even.

- The sets are $A=\{3,4,5,6\}$ and $B=\{2,4,6\}$
- To check mutually exclusive: notice $A\cap B=\{4,6\}$ which is not empty. So event $A$ and event $B$ are not mutually exclusive

Further assuming the die are fair, each side has an equal $1/6$ probability. This lets us compute

- $Prob(A)=4/6=2/3$ and $Prob(B)=3/6=1/2$
- $A\cup B=\{2,3,4,5,6\}$, which implies $Prob(A\cup B)=5/6$
- $A\cap B=\{4,6\}$, which implies $Prob(A\cap B)=2/6=1/3$
- To check independence: notice $Prob(A\cap B)=1/3 = (2/3) (1/2) = Prob(A)Prob(B)$. So event $A$ and event $B$ are mutually independent
- To check inclusion–exclusion: notice $Prob(A) + Prob(B) - Prob(A\cap B) = 2/3 + 1/2 - 1/3 = 4/6 + 3/6 - 2/6 = 5/6 = Prob(A\cup B)$
- Finally, the conditional probability of $Prob(\text{die is} >2 \mid \text{die shows even number}) = Prob(A \mid B) = \frac{Prob(A \cap B)}{P(B)} = \frac{1/3}{1/2} = 2/3$

```{r}
# Simulation verification
set.seed(123)
x <- seq(1,6)
rolls <- sample(x, 1e6, replace = TRUE)
A <- rolls > 2
B <- rolls %% 2 == 0
P_AcondB <- mean(A & B) / mean(B)
round(P_AcondB, 3)
```

:::{.callout-tip icon=false collapse="true"}
Consider a six sided die with sample space $\{1,2,3,4,5,6\}$ where each outcome is each equally likely. What is $Prob(\text{die is odd or }<5)$?
Denote the odd events as $A=\{1,3,5\}$ and the less than five events as  $B=\{1,2,3,4\}$. Then

- $A\cap B=\{1,3\}$, which implies $Prob(A\cap B)=2/6=1/3$.
- $Prob(A)=3/6=1/2$ and $Prob(B)=4/6=2/3$.
- By inclusion–exclusion: $Prob(A\cup B)=Prob(A)+Prob(B)-Prob(A\cap B)=1/2+2/3-1/3=5/6$.

Now find $Prob(\text{die is even and }<5)$. Verify your answer with a computer simulation.
:::



#### **Law of Total Probability**. {-}

This law states that $Prob(A)=Prob(A \mid B) P(B) + Prob(A \mid \lnot B) P(\lnot B)$, where $\lnot B$ is the event "not $B$".

For example, consider a six sided die where events are the number $>2$ or the number is even.
- The sets are $A=\{3,4,5,6\}$ and $B=\{2,4,6\}$ and $\lnot B = \{1,3,5\}$ 
From before, we know that $Prob(A)=2/3$, $Prob(B)=1/2$, and $Prob(A \mid B) = 2/3$.
Then see that $Prob(A \mid B) Prob(B) = \frac{2}{3} \frac{1}{2} = 2/6$ 

We can also directly compute $Prob(\lnot B)=1/2$, or infer it from $Prob(\lnot B)=1-Prob(B)$.
The conditional probability of $Prob(\text{die is} >2 \mid \text{die shows odd number}) = Prob(A \mid \lnot B) = \frac{Prob(A \cap \lnot B)}{P(\lnot B)} = 2/3$.

Then we can see that
$Prob(A \mid B) Prob(B) + Prob(A \mid \lnot B) Prob(\lnot B) = 2/6 + 2/6 = 2/3 = Prob(A)$.


#### **Bayes' Theorem**. {-}

*Bayes' theorem* reverses the direction of a conditional probability. From the definition, $Prob(A \cap B) = Prob(B \mid A) \cdot Prob(A)$. Substituting into $Prob(A \mid B) = Prob(A \cap B) / Prob(B)$ gives
\begin{eqnarray}
Prob(A \mid B) &=& \frac{Prob(B \mid A) \cdot Prob(A)}{Prob(B)}
\end{eqnarray}
Substituting the Law of Total Probability into the denominator,
\begin{eqnarray}
Prob(A \mid B) &=& \frac{Prob(B \mid A) \cdot Prob(A)}{Prob(B \mid A) \cdot Prob(A) + Prob(B \mid \lnot A) \cdot Prob(\lnot A)}
\end{eqnarray}

In the die example above, we computed $Prob(A \mid B) = 2/3$: the probability the die is $>2$, given it is even. We can use Bayes' theorem to reverse this: what is the probability the die is even, given it is $>2$?

:::{.callout-note icon=false collapse="true"}
Using the same die with $A=\{3,4,5,6\}$ and $B=\{2,4,6\}$, we know $Prob(A \mid B)=2/3$, $Prob(B)=1/2$, and $Prob(A)=2/3$. By Bayes' theorem,
\begin{eqnarray}
Prob(B \mid A) &=& \frac{Prob(A \mid B) \cdot Prob(B)}{Prob(A)} = \frac{(2/3)(1/2)}{2/3} = \frac{1}{2}.
\end{eqnarray}
We can verify directly: given $A=\{3,4,5,6\}$, the even outcomes are $\{4,6\}$, so $Prob(B \mid A)=2/4=1/2$. Since $A$ and $B$ are independent, $Prob(B \mid A) = Prob(B)$ and flipping the conditioning changes nothing.
```{r}
# Simulation verification
set.seed(123)
rolls <- sample(seq(1,6), 1e6, replace=TRUE)
A <- rolls > 2
B <- rolls %% 2 == 0
# Prob(B | A)
mean(B[A])
```
:::

Bayes' theorem is most useful when events are not independent, so that knowing $B$ occurred changes what we believe about $A$.

:::{.callout-tip icon=false collapse="true"}
A company sources parts from two suppliers. Supplier $A$ provides $60\%$ of parts and supplier $B$ provides $40\%$. The defect rate from supplier $A$ is $3\%$ and from supplier $B$ is $5\%$. If a randomly selected part is defective ($D$), what is the probability it came from supplier $A$?

We know $Prob(A)=0.6$, $Prob(\lnot A)=0.4$, $Prob(D \mid A)=0.03$, and $Prob(D \mid \lnot A)=0.05$. By the Law of Total Probability,
\begin{eqnarray}
Prob(D) &=& Prob(D \mid A) \cdot Prob(A) + Prob(D \mid \lnot A) \cdot Prob(\lnot A) \\
&=& 0.03 \times 0.6 + 0.05 \times 0.4 = 0.038.
\end{eqnarray}
By Bayes' theorem,
\begin{eqnarray}
Prob(A \mid D) &=& \frac{Prob(D \mid A) \cdot Prob(A)}{Prob(D)} = \frac{0.03 \times 0.6}{0.038} \approx 0.474.
\end{eqnarray}
Even though supplier $A$ provides $60\%$ of all parts, they account for only about $47\%$ of defectives because their defect rate is lower.
```{r}
# Simulation verification
set.seed(42)
n <- 1e6
supplier <- sample(c("A","B"), n, replace=TRUE, prob=c(0.6, 0.4))
defect_rate <- ifelse(supplier == "A", 0.03, 0.05)
defective <- runif(n) < defect_rate

# Prob(A | D)
mean(supplier[defective] == "A")
```
:::

