# Data & Visualization
***

This chapter classifies the kinds of data you will work with and gives you the basic tools to visualize and summarize a single variable.
We start with data types (cardinal vs factor), then look at the structures that hold datasets in R (lists and data.frames), and finish with the standard ways to depict a distribution: proportions, histograms, ECDFs, quantiles, and boxplots.

## Data Types

Before any analysis, it helps to know whether the values in our dataset support arithmetic (their differences carry consistent meaning) or whether they are merely labels.

:::{.callout-important icon=false title="Key Definition" collapse="false"}
*Cardinal* (or *numeric*) data are values where the difference between elements always means the same thing.

*Factor* data are values where the difference between elements does not always mean the same thing.
:::

The type of a variable governs what summaries are meaningful: cardinal data can be averaged, scattered, and used in regressions, while factor data is best counted and tabulated.
Trying to compute the mean of factor data is a category error rather than a meaningful summary, and most statistical software will refuse to do so.

#### **Basic Types** {-}
Cardinal data come in two flavors (*discrete* or *continuous*) and factor data come in two flavors (*ordered* or *unordered*), giving four sub-types in all:

* *Cardinal (Numeric)*
    * *Discrete*: E.g. $\{ 1,2,3\}$ and notice that $2-1=3-2$.
    * *Continuous*: E.g., $\{1.4348, 2.4348, 2.9, 3.9 \}$ and notice that $2.9-1.4348=3.9-2.4348$
* *Factor*
    * *Ordered*: E.g., $\{1^{st}, 2^{nd}, 3^{rd}\}$ place in a race and notice that $1^{st}$ - $2^{nd}$ place does not equal $2^{nd}$ - $3^{rd}$ place for a very competitive person who cares only about winning.
    * *Unordered* (categorical): E.g., $\{Amanda, Bert, Charlie\}$ and notice that $Amanda - Bert$ never makes sense.

Note that for theoretical analysis, the types are sometimes grouped differently as

* Discrete (discrete cardinal, ordered factor, and unordered factor data).
  You can count the potential values.
  E.g., the set $\{A,B,C\}$ has three potential values.
* Continuous (continuous cardinal data).
  There are uncountably infinite potential values.
  E.g., Try counting the numbers between $0$ and $1$ including decimal points, and notice that any two numbers have another potential number between them.

:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
Here are some examples
```{r}
dat_card1 <- c(1, 2, 3) # Cardinal data (Discrete)
dat_card1

dat_card2 <- c(1.1, 2/3, 3) # Cardinal data (Continuous)
dat_card2

dat_fact1 <- factor( c('A', 'B', 'C'), ordered=TRUE) # Factor data (Ordinal)
dat_fact1

dat_fact2 <- factor( c('Leipzig', 'Los Angeles', 'Logan'), ordered=FALSE) # Factor data (Categorical)
dat_fact2

dat_fact3 <- factor( c(TRUE, FALSE), ordered=FALSE) # Factor data (Categorical)
dat_fact3

# Explicitly check the data types:
#class(dat_card1)
#class(dat_card2)
```
:::

#### **Strings** {-} 
Note that R allows for unstructured plain text, called *character strings*, which we can then format as factors
```{r}
c('A', 'B', 'C')  # character strings
c('Leipzig', 'Los Angeles', 'Logan')  # character strings
```
Also note that strings are encounter in a variety of settings, and you often have to format them after reading them into R (the [RStudio regex cheatsheet](https://raw.githubusercontent.com/rstudio/cheatsheets/main/regex.pdf) is a handy reference).^[We will not cover the statistical analysis of text in this book, but strings are amenable to statistical analysis.]
```{r}
# Strings
paste( 'hi', 'mom')
paste( c('hi', 'mom'), collapse='--')

kingText <- 'The king infringes the law on playing curling.'
gsub(pattern='ing', replacement='', kingText)
# advanced usage
#gsub('[aeiouy]', '_', kingText)
#gsub('([[:alpha:]]{3})ing\\b', '\\1', kingText) 
```

## Datasets

Datasets can be stored in a variety of formats on your computer.
But they can be analyzed in R in three basic ways.

#### **Lists** {-}

Lists are probably the most basic type
```{r}
x_hat <- seq(1, 10)                       # vector of 1..10
y_hat <- 2*x_hat                              # element-wise doubling
list(x_hat, y_hat)                            # list holds two vectors of equal length

x_mat1 <- matrix( seq(2, 7), 2, 3)    # 2x3 matrix
x_mat2 <- matrix( seq(4, -1), 2, 3)
list(x_mat1, x_mat2)                  # list can also hold matrices
```

Lists are useful for storing unstructured data
```{r}
list(list(x_mat1), list(x_mat2))  # list of lists

list(x_mat1, list(x_mat1, x_mat2)) # list of different objects

# ...inception...
list(x_mat1,
    list(x_mat1, x_mat2), 
    list(x_mat1, list(x_mat2)
    )) 
```

#### **Data.frames** {-}

A *data.frame* looks like a matrix but each column can be a different type of vector (or even list).
This allows you to combine different data types into a single object for analysis, which is why it might be your most common object.

```{r}
# data.frames: your most common data type
    # matrix of different data-types
    # well-ordered lists
data.frame(x=x_hat, y=y_hat)  # list of vectors
```

:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
Create a data.frame storing two different types of data.
Then show print only the second column
```{r}
dat <- data.frame(x=dat_fact2, y=dat_card2)
dat

dat[, 'y']
```
:::

Data.frames stack two dimensions, rows and columns.
The generalization to three or more dimensions is the *array*, covered in [Misc. Multivariate Topics](03_30_MiscTopics.qmd#arrays).

#### **Initial Data Inspection** {-}
Regardless of the data types you have, you typically begin by inspecting your data by examining the first few observations.
 
Consider, for example, historical data on crime in the US.

```{r}
head(USArrests) # Actual Data

# Check NA values
x_hat <- c(3, 3.1, NA, 0.02) #Small dataset we will use in numerical examples
sum( is.na(x_hat) )
```




## Proportions



In what follows, we will often work with data as a vector, where there are $n$ observations and $\hat{x}_{i}$ is the value of the $i$th one.
The hat marks a value computed from observed data; we will contrast it with theoretical quantities in [Population Statistics](01_06_PopStatistics.qmd#statistical-theory).
We often analyze observations in comparison to some value $x$.

```{r}
x_hat <- c(3, 3.1, 0.02) # Observed values
x <- 2 # particular value: 'little x'
sum(x_hat <= x)
sum(x_hat == x)
```

#### **Empirical Mass Function**  {-}

If you have factor data, or discrete cardinal data, you can directly plot the proportions.
For each unique outcome $x$ we compute
$$
\hat{p}(x)=\sum_{i=1}^{n}\mathbf{1}\left(\hat{x}_{i}=x\right)/n.
$$
where $n$ is the number of observations and $\sum_{i=1}^{n}\mathbf{1}\left(\hat{x}_{i}=x\right)$ counts the number of observations equal to $x$.

:::{.callout-note icon=false collapse="true" title="Must Know"}
For example, consider the dataset $\{1,2,1,3,1\}$ with $n=5$ observations.
The value $x=1$ appears three times, so $\hat{p}(1)=3/5$.
The value $x=2$ appears once, so $\hat{p}(2)=1/5$.
The value $x=3$ appears once, so $\hat{p}(3)=1/5$.
The proportions sum to one: $3/5 + 1/5 + 1/5 = 1$.
```{r}
x_hat <- c(1, 2, 1, 3, 1)
table(x_hat)
table(x_hat)/length(x_hat)
```
:::

```{r}
# Discretized data
x_hat <- USArrests[, 'Murder']
x_hat[1:3]
x_rounded <- floor(x_hat) #rounded down, to be discrete
x_rounded[1:3]

p_hat <- table(x_rounded)/length(x_rounded)
plot(p_hat, col=grey(0, .5),
    xlab='Murder Arrests (Discretized)',
    ylab='Proportion of States with each value')
```

There are several stylistic variations: a *bar plot* uses thick bars for each outcome, a *lollipop plot* above uses skinny lines with a circle on top.
E.g., add `points(names(p_hat), p_hat, pch=16)`
The most important point is that the height of each line equals the proportion of data with a specific value.

#### **Histogram Density** {-}

For continuous data, no single value is likely to repeat, so we summarize the distribution by counting how many observations land in each of several equal-width intervals.

:::{.callout-important icon=false title="Key Definition" collapse="false"}
A *histogram density* divides the range of the data into bins of equal half-width $h$.
For an exclusive bin $(x-h, x+h]$ with midpoint $x$, it reports the rescaled count
$$\hat{f}(x) = \frac{\sum_{i=1}^{n} \mathbf{1}\left( \hat{x}_{i} \in (x-h, x+h] \right)}{n \cdot 2h},$$
where $\mathbf{1}(\cdot)$ is an *indicator function* that equals $1$ if the expression inside is TRUE and $0$ otherwise.
:::

The *[histogram](https://en.wikipedia.org/wiki/Histogram)* is the workhorse picture for the *shape* of a continuous distribution: where observations are concentrated, whether there are multiple peaks, whether one tail is longer than the other, and whether there are gaps with no observations.
These features are hard to spot in a list of numbers but jump out at a glance from a histogram.
This rescaling makes the total area of all bars equal to one, so bar heights have units of probability per unit of $x$.
It also makes histogram heights comparable across samples of different sizes and lets us interpret the bar height as a *density*: the proportion of data per unit of $x$.
The half-width $h$ acts as a *bandwidth*: a small $h$ produces a jagged picture sensitive to single observations, while a large $h$ smooths over real features.
Note that $h$ is measured from the bin midpoint out to one edge, so a bin is $2h$ wide.
The same $h$ reappears in the [kernel density](01_10_AdvancedEmpirics.qmd#kernel-density), and again in [Local Regression](02_16_LocalRegression.qmd#regressograms), where it sets how much data around a point a local model uses.

:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
E.g., suppose $\hat{x}_{i}=3.8$ and $h=1/2$.

For $x=1$ we have $\mathbf{1}\left( \hat{x}_{i} \in \left(1-h, 1+h \right] \right)=\mathbf{1}\left( 3.8 \in \left(0.5, 1.5\right] \right)=0$.

For $x=4$ $\mathbf{1}\left( \hat{x}_{i} \in \left(4-h, 4+h \right] \right)=\mathbf{1}\left(  3.8 \in \left(3.5, 4.5\right] \right)=1$.
:::

Note that rectangle area equals the proportion of data in the bin.
Recalling that the area of the rectangle is "base x height":
$$\begin{aligned}
2h \times \hat{f}(x) &= \frac{\sum_{i=1}^{n} \mathbf{1}\left( \hat{x}_{i} \in \left(x-h, x+h \right] \right)}{n}.
\end{aligned}$$
We compute $\hat{f}(x)$ for each bin midpoint $x$, and the area of all rectangles sums to one.
^[If $L$ distinct bins exactly cover the range, then $2h=[\text{max}(\hat{x}_{i}) - \text{min}(\hat{x}_{i})]/L$ and we can write the design points as $x \in \left\{ \text{min}(\hat{x}_{i}) + (2\ell-1)h \right\}_{\ell=1}^{L}$, since consecutive midpoints sit $2h$ apart.]

:::{.callout-note icon=false collapse="true" title="Must Know"}
For example, consider the dataset $\{3,3.1,0.02\}$ and use bins $(0,1], (1,2], (2,3], (3,4]$.
In this case, the midpoints are $x=(0.5,1.5,2.5,3.5)$ and $h=1/2$.
Then the counts at each midpoints are $(1,0,1,1)$.
Since $\frac{1}{n 2h}=\frac{1}{3\times1}=\frac{1}{3}$, we can rescale the counts to compute the density as $\hat{f}(x)=(1,0,1,1) \frac{1}{3}=(1/3,0,1/3,1/3)$.

```{r}
# Intuitive Examples
x_hat <- c(3, 3.1, 0.02)
f_hat <- hist(x_hat, breaks=c(0, 1, 2, 3, 4), plot=FALSE, freq=FALSE)
f_hat

# base x height
base <- 1
height <- f_hat[['density']]
sum(base*height)
```

For another example, use the bins $(0,2]$ and $(2,4]$.
So the midpoints are $1$ and $3$, and the bin half-width is $1$.
Only one observation, $0.02$, falls in the bin $(0,2]$.
The other two observations, $3$ and $3.1$, fall into the bin $(2,4]$.
The scaling factor is $\frac{1}{n 2h}=\frac{1}{3\times 2 \times 1}=\frac{1}{6}$.
So the first bin has density $\hat{f}(1)=1\frac{1}{6}=1/6$ and the second bin has density $\hat{f}(3)=2\frac{1}{6}=2/6$.
The area of the first bin's rectangle is $2 \times \hat{f}(1)=2/6=1/3$ and the area of the second rectangle is $2 \times \hat{f}(3)=4/6=2/3$.

Now intuitively work through an example with three bins instead of four.
Compute the areas 
```{r}
f_hat <- hist(x_hat, breaks=c(0, 4/3, 8/3, 4), plot=FALSE, freq=FALSE)
base <- 4/3
height <- f_hat[['density']]
sum(base*height)

# as a default, R uses bins (, ] instead of [, )
# but you can change that with 'right=F'
# hist(x_hat, breaks=c(0, 4/3, 8/3, 4), plot=F, right=F)
```
:::

```{r}
# Practical Example
x_hat <- USArrests[, 'Murder']
hist(x_hat,
    breaks=seq(0, 20, by=1), # bin width =1, so the half-width is h=1/2
    freq=FALSE,
    border=NA,
    main=NA,
    xlab='Murder Arrests')
title('Murder Arrests', font.main=1)
# Raw Observations
rug(x_hat, col=grey(0, .5))

# Since 2h=1, the density equals the proportion of states in each bin
# Redo this example with h=1 (i.e., breaks spaced 2h=2 apart)
```
 
Note that histograms qualitatively depict proportions over multiple bins (since each bar represents the proportion of data in its bins), but you can also directly compute proportions.

:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
E.g., the proportion of murder rates in $(2,4)$ is
```{r}
mean( x_hat > 2 & x_hat <4)
```
:::



## Distributions

#### **Empirical *Cumulative* Distribution Function** {-}

Another way to summarize a distribution asks, for each candidate value $x$, what proportion of the sample lies at or below it.

:::{.callout-important icon=false title="Key Definition" collapse="false"}
The *empirical cumulative distribution function* (ECDF) is the proportion of observations in the sample whose values are less than or equal to $x$,
$$\hat{F}(x) = \frac{1}{n}\sum_{i=1}^{n}\mathbf{1}(\hat{x}_{i} \leq x).$$
It jumps up by $1/n$ at each data point and equals $1$ once $x$ passes the largest observation.
:::

The ECDF is useful when we want precise numbers rather than a picture of shape: "what fraction of states have at most $10$ murders per $100$k?" reads off directly without choosing bin widths or smoothing parameters.
It is a *step function* (see this [ECDF walkthrough](https://library.virginia.edu/data/articles/understanding-empirical-cumulative-distribution-functions) for visual examples) and contains exactly the same information as the sorted dataset, just packaged as a function of $x$ rather than a list of values.
Inverting the picture horizontally (asking "at what value $x$ does the proportion reach $p$?") gives the quantiles introduced next.

:::{.callout-note icon=false collapse="true" title="Must Know"}
For example, consider the dataset $\{3,3.1,0.02 \}$.
We reorder the observations as $\{0.02, 3, 3.1 \}$, so that there are discrete jumps of $1/n=1/3$ at each value.
Consider the points $x \in \{0.5,2.5,3.5\}$.
At $x=0.5$, $F(0.5)$ measures the proportion of the data $\leq 0.5$.
Since only one observations, $0.02$, of three is $\leq 0.5$, we can compute $F(0.5)=1/3$.
Similarly, since only one observations, $0.02$, of three is $\leq 2.5$, we can compute $F(2.5)=1/3$.
Since all observations are $\leq 3.5$, we can compute $F(3.5)=1$.
```{r}
x_hat <- c(3, 3.1, 0.02)
F_hat <- ecdf(x_hat)

# Visualized
plot(F_hat)

# Evaluated at each data point
x <- x_hat
F_hat(x_hat)
#sum(x_hat<=3)/length(x_hat)
#sum(x_hat<=3.1)/length(x_hat)
#sum(x_hat<=0.02)/length(x_hat)

# Evaluated at other points
x <- c(0.5, 2.5, 3.5)
F_hat(x)
#sum(x_hat<=0.5)/length(x_hat)
#sum(x_hat<=2.5)/length(x_hat)
#sum(x_hat<=3.5)/length(x_hat)
```
:::

```{r}
F_hat <- ecdf(USArrests[, 'Murder'])
# proportion of murders <= 10
F_hat(10)
# proportion of murders <= x, for all x
plot(F_hat, main=NA,
    xlab='Murder Arrests (x)',
    ylab='Proportion of States with Murder Arrests <= x',
    pch=16, col=grey(0, .5))
title('Murder Arrests ECDF', font.main=1)
rug(USArrests[, 'Murder'])
```

#### **Quantiles** {-}

Reading the ECDF in reverse asks: at what value $x$ does a given proportion $p$ of the data lie below?

:::{.callout-important icon=false title="Key Definition" collapse="false"}
The *$p$th quantile* of a sample is the value $x$ where $p$ percent of the data fall below $x$ and $(1-p)$ percent fall above.
The *empirical quantile function* $\hat{q}(p) = \hat{F}^{-1}(p)$ inverts the ECDF, returning the quantile at any probability $p \in [0,1]$.
:::

Quantiles are useful for precise numerical anchors: "the top $10\%$ of cities", "half of households earn less than...", or "the middle $50\%$ of test scores fall between $x_1$ and $x_2$".
They also describe spread (range, $IQR$) and shape (the gap between median and mean) without committing to a smooth distributional model.

Five quantiles get their own names because they summarize the distribution at a glance:

* The *min* (the $p=0$ quantile) is the smallest value, where $0\%$ of the data have lower values.
* The *lower quartile* (the $p=0.25$ quantile) is where $25\%$ of the data have lower values.
* The *median* (the $p=0.5$ quantile) is the middle value, where $50\%$ of the data have lower values.
* The *upper quartile* (the $p=0.75$ quantile) is where $75\%$ of the data have lower values.
* The *max* (the $p=1$ quantile) is the largest value, where $100\%$ of the data have lower values.

We also speak of *deciles* at $p=0.1, 0.2, \ldots, 0.9$ when we want a finer split.

:::{.callout-note icon=false collapse="true" title="Must Know"}
For example, that dataset $\{0,0,0.02,3,5\}$ has a median of $0.02$, with the lower quartile $0$ and the upper quartile $3$.
(The number $0$ is also special: the most frequent observation is called the *mode*, which we revisit in [Numerical Statistics](01_03_DescriptiveStatistics.qmd#mode-and-share-concentration).)

```{r}
x_hat <-  c(3.1, 3, 0.02)
quantile(x_hat, probs=c(0, .5, 1))
```

Now work through an intuitive example with observations $\{1,2,...,13\}$.
Hint: split the ordered observations into four groups.
:::

```{r}
# common quantiles
x_hat <- USArrests[, 'Murder']
quantile(x_hat)

# All deciles are quantiles
quantile(x_hat, probs=seq(0, 1, by=.1))

# Visualized: Inverting the Empirical Distribution
F_hat <- ecdf(x_hat)
plot(F_hat, lwd=2, xlim=c(0, 20),
    pch=16, col=grey(0, .5), main=NA)
title('Inverting the ECDF', font.main=1)
# Two Examples of Quantiles
p <- c(.25, .9) # Lower Quartile, Upper Decile
cols <- c(rgb(1, 0, 0, .8), rgb(0, 0, 1, .8))
q_hat <- quantile(x_hat, p, type=1)
q_hat
segments(q_hat, p, -10, p, col=cols)
segments(q_hat, p, q_hat, 0, col=cols)
mtext( round(q_hat, 2), 1, at=q_hat, col=cols)
```

There are some issues with quantiles with smaller datasets.
E.g., to compute the median of $\{3,3.1,0,1\}$, we need some ways to break ties (of which there are many options).
The most common way is perhaps a simple average of the tied values: $(3+1)/2=2$.




:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
To calculate quantiles, the computer defaults to sorting the observations from smallest to largest as $\hat{x}_{(1)}, \hat{x}_{(2)},... \hat{x}_{(n)}$, and then computes quantiles as $\hat{x}_{ (p \times n) }$, where $p \in [0,1]$ refers to probability and $(p\times n)$ is rounded.
There are different ways to break ties and the min ($p=0$) is a special case.
```{r}
x_hat <- USArrests[, 'Murder']
x_sorted_hat <- sort(x_hat)
x_sorted_hat

# median
x_sorted_hat[ceiling(length(x_sorted_hat)*.5)]
quantile(x_hat, probs=.5, type=1) #tie break rule #1

## arbitrary quantiles
p <- 0.11
x_sorted_hat[ceiling(length(x_sorted_hat)*p)]
quantile(x_hat, probs=p, type=1)

# min is a special case
min(x_sorted_hat)
x_sorted_hat[ceiling(length(x_sorted_hat)*0)]
quantile(x_hat, probs=0, type=1)
x_sorted_hat[1]
```
:::

#### **Boxplots** {-}
*[Boxplots](https://en.wikipedia.org/wiki/Box_plot)* also summarize the distribution.
The boxplot shows the median (solid black line) and *interquartile range* ($IQR=$ upper quartile $-$ lower quartile; filled box).^[
Technically, the upper and lower *hinges* use two different versions of the first and third quartile.
Statistical packages differ in how they compute both the quartiles and the fences that flag outliers [@frigge1989].
See also <https://stackoverflow.com/questions/40634693/lower-and-upper-quartiles-in-boxplot-in-r>.
]
As a default, whiskers are shown as $1.5\times IQR$ and values beyond that are highlighted as *outliers* (so whiskers do not typically show the data range).
No state sits more than $1.5\times IQR$ beyond a quartile, so the boxplot below draws no outlier circles.
They show up in [Numerical Statistics](01_03_DescriptiveStatistics.qmd#kurtosis), where heavier-tailed data push observations past the whiskers.

```{r}
# Default: median, box (IQR), whiskers at 1.5*IQR, outliers as circles
boxplot(USArrests[, 'Murder'],
    main=NA, ylab='Murder Arrests')
title('Default', font.main=1)
```

The default boxplot compresses $n$ observations into five numbers, which discards information about shape.
You can alternatively show all the raw data points instead of whisker+outliers.
This exposes the gaps, clusters, and repeated values that a five-number summary hides [@weissgerber2015].

```{r}
# Alternative: every observation, in place of whiskers and outliers
boxplot(USArrests[, 'Murder'],
    main=NA, ylab='Murder Arrests',
    whisklty=0, staplelty=0, outline=FALSE)
title('With Raw Data', font.main=1)
# Raw Observations
stripchart(USArrests[, 'Murder'],
    pch='-', col=grey(0, .5), cex=2,
    vert=TRUE, add=TRUE)
```

These univariate visualization tools extend to examining relationships between two variables.
See [Bivariate Distributions](02_11_BivariateDistributions.qmd) for how histograms, ECDFs, and boxplots adapt to bivariate data.


## Exercises

1. Comment the script you wrote for this chapter, then restart R and check that it runs from a clean session, then check the script with AI as explained in [Working with AI](04_A_WorkingWithAI.qmd).
Write three sentences from memory on the main statistical idea of this chapter, and ask the assistant what is wrong, vague, or missing.
Finish with your own questions about whatever you found hardest.

2. Classify each of the following as cardinal (discrete or continuous) or factor (ordered or unordered): (a) temperature in Celsius, (b) letter grade on a transcript, (c) number of siblings, (d) province of residence.
   Briefly explain each choice.

3. The built-in dataset `mtcars` has a column `mpg`.
   Compute the histogram density by hand for the bins $(10,15]$, $(15,20]$, $(20,25]$, $(25,30]$, $(30,35]$.
   That is, for each bin count the number of observations, then divide by $n \times 2h$ where $2h=5$.
   Verify your answer matches `hist(mtcars$mpg, breaks=seq(10,35,by=5), freq=FALSE)`.

4. Load `USArrests` in R.
   Compute the ECDF of the `Assault` variable and use it to find the proportion of states with 200 or fewer assault arrests.
   Then find the upper quartile of `Assault` using `quantile`.

#### **Further Reading** {-}

* <https://library.virginia.edu/data/articles/understanding-empirical-cumulative-distribution-functions> -- Clear walkthrough of the ECDF with visual examples.
* <https://raw.githubusercontent.com/rstudio/cheatsheets/main/regex.pdf> -- RStudio cheat sheet for regular expressions and string handling in R.
* <https://www.econometrics-with-r.org/2.1-random-variables-and-probability-distributions.html> -- Review of data types and distributions in an econometrics context.

#### **Recall** {-}

This chapter sorted observations into *cardinal* and *factor* types and then showed several complementary ways to picture the distribution of a single variable: bar plots and lollipops for proportions, the *histogram density*, the *empirical CDF*, *quantiles* (min, median, max, quartiles), and the boxplot.
The running three-point dataset $\{3, 3.1, 0.02\}$ ties two of them together: the ECDF jumps by $1/3$ at each of $0.02$, $3$, and $3.1$ before flattening at $1$, and the median reads off as $3$.
Each picture answers a different question about the same data, and switching between them is a habit worth building.
In the next chapter we summarize the same distributions with single numbers: mean, variance, median, MAD, skewness, and kurtosis.
