# Advanced Probability
***

This chapter rounds out the probability toolbox with distributions that use factorials (!): the Binomial and Poisson (discrete), and the Irwin–Hall and Beta (continuous).
It then formalizes basic set theory, Bayes' theorem, and the Law of Total Probability so that conditional reasoning has a clear language.


## Discrete Distributions

#### **Binomial**. {-}

A single Bernoulli trial gives a $0$ or a $1$; counting how many successes occur across $n$ such trials is the next-most-common discrete distribution.

:::{.callout-important icon=false title="Key Definition" collapse="false"}
A *Binomial* random variable counts the number of successes in $n$ independent Bernoulli trials, each with success probability $p$.
Its PMF is $Prob(X_{i} = x) = \binom{n}{x} p^{x} (1-p)^{n-x}$ for $x \in \{0, 1, \ldots, n\}$, mean $np$, and variance $np(1-p)$.
:::

The *[Binomial distribution](https://en.wikipedia.org/wiki/Binomial_distribution)* is useful for any situation that aggregates independent yes/no events: defective items in a batch, customers who enter a store, voters who pick a candidate, free throws that go in.
The full specification gathers sample space, PMF, and CDF in one place:
$$\begin{aligned}
X_{i} &\in \{0,1,\ldots,n\} \\
Prob(X_{i}=x) &= \binom{n}{x}p^{x}(1-p)^{n-x} \\
F(x) = Prob(X_{i} \leq x) &= \sum_{k=0}^{\lfloor x \rfloor} Prob(X_{i}=x)
\end{aligned}$$

The simplest case: how many heads will I get when I flip a fair coin twice?
We can first see that there are four potential outcomes: $\{ (1,1), (1,0), (0,1), (0,0) \}$, each of which are equally likely.
The corresponding sums of $\{2,1,1,0\}$ are equally likely, which leads to probabilities $Prob(X_{i}=0)=1/4$, $Prob(X_{i}=1)=2/4$, and $Prob(X_{i}=2)=1/4$.

If the coin is unfair, with probability of heads equal to $0.25$, then the four outcomes $\{(1,1), (1,0), (0,1), (0,0)\}$ are no longer equally likely: they have probabilities $\{0.0625, 0.1875, 0.1875, 0.5625\}$.
Collecting the matching sums, $Prob(X_{i}=2)=0.0625$, $Prob(X_{i}=1)=0.1875+0.1875=0.375$, and $Prob(X_{i}=0)=0.5625$.

:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
When I flip a coin three times, the potential outcomes are 
$$\begin{gathered}
\{
(1,1,1), (1,1,0), (1,0,1), (1,0,0),\\
(0,1,1), (0,1,0), (0,0,1), (0,0,0) 
\}
\end{gathered}$$
Assuming the coin is fair, compute the probabilities for each possible sum.
Assuming the coin is unfair, with probability of heads equal to $0.25$, compute the probabilities for each possible sum.
:::

In general, we use the *Binomial distribution*.
The $\binom{n}{x}$ term is called the [binomial coefficient](01_05_Sampling.qmd#sampling-distributions) and it counts all the different groups that lead to the same sum.
The other terms intuitively reflect the probabilities of getting $x$ successes and $n-x$ failures.

For example, suppose $n=15$ people walk past a store each hour and each has a $p=30\%$ chance of entering.
The number of customers who enter is $X_{i} \sim \text{Binom}(15, 0.3)$.

```{r}
## Store: 15 passers-by, 30% chance each enters
n <- 15
p <- 0.3

# PMF plot
x <- seq(0, n)
f_x <- dbinom(x, n, p)
plot(x, f_x, type = 'h', col = rgb(0, 0, 1, .8),
    main=NA, xlab = 'Customers entering', ylab = 'Prob(X = x)')
title(bquote(paste('Binom(', .(n), ', ', .(p), ')')), font.main=1)
points(x, f_x, pch = 16, col = rgb(0, 0, 1, .8))

## Theoretical CDF
F_x <- pbinom(x, n, p)
plot(x, F_x, type='s', col=rgb(0, 0, 1, .8),
    xlab = 'Customers entering', ylab = 'F(x)', main=NA)
points(x, F_x, pch=16, col=rgb(0, 0, 1, .8))
title(bquote(paste('Binom(', .(n), ', ', .(p), ') CDF')), font.main=1)
```

:::{.callout-note icon=false collapse="true" title="Must Know"}
A quality inspector checks $n=20$ items from a production line with a $5\%$ defect rate ($p=0.05$).
What is the probability of finding $2$ or fewer defectives?

We need $Prob(X_{i} \leq 2) = F(2)$ where $X_{i} \sim \text{Binom}(20, 0.05)$.
By hand, the first few terms are
$$\begin{aligned}
Prob(X_{i}=0) &= \binom{20}{0}(0.05)^{0}(0.95)^{20} \approx 0.36 \\
Prob(X_{i}=1) &= \binom{20}{1}(0.05)^{1}(0.95)^{19} \approx 0.38 \\
Prob(X_{i}=2) &= \binom{20}{2}(0.05)^{2}(0.95)^{18} \approx 0.19
\end{aligned}$$
So $Prob(X_{i} \leq 2) \approx 0.36 + 0.38 + 0.19 = 0.93$.
There is roughly a $93\%$ chance the inspector finds $2$ or fewer defectives.
```{r}
# Individual terms
x <- c(0, 1, 2)
px <- dbinom(x, size=20, prob=0.05)
round(px, 2)

# CDF: Prob(X <= 2)
pbinom(2, size=20, prob=0.05)
```
:::

:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
Suppose that employees at a company are $70%$ female and $30%$ male.
If we select a random sample of eight employees, what is the probability that $2$ in the sample are female?
That $\geq 2$ are female? 

Try conducting a simulation first, 
```{r}
n <- 8
p <- 0.3
X <- rbinom(1000, n, p)
```
:::


:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
For $n$ coin flips, show that $\mathbb{E}[X_{i}]=np$ and $\mathbb{V}[X_{i}]=np(1-p)$.
Build on the case $n=2$ above, then consider case $n=3$, and then the general case of any $n$.

```{r}
# Compare empirical simulation vs theoretical
n <- 2
p <- 3/4
X <-  rbinom(1000, size=n, prob=p)

# Mean
c(empirical = mean(X), theoretical = n*p)

# Variance
c(empirical  = var(X),  theoretical  = n*p*(1-p))
```
:::


#### **Poisson**. {-}

When the trials are too many and individually too rare to count, the Binomial limits to a simpler model parameterized by a single rate.

:::{.callout-important icon=false title="Key Definition" collapse="false"}
A *Poisson* random variable counts events in a fixed interval when events are rare, independent, and arrive at a constant rate $\lambda$.
Its PMF is $Prob(X_{i} = x) = e^{-\lambda} \lambda^{x} / x!$ for $x \in \{0, 1, 2, \ldots\}$, and the mean and variance are both $\lambda$.
:::

The *[Poisson distribution](https://en.wikipedia.org/wiki/Poisson_distribution)* is useful for count outcomes where events are sparse but the opportunities are many: arrivals at a store, decay events from a radioactive source, typos per page.
The single parameter $\lambda$ is the average number of events per interval, and the full specification adds the CDF:
$$\begin{aligned}
X_{i} &\in \{0,1,\ldots\} \\
Prob(X_{i}=x) &= e^{-\lambda}\lambda^x/x! \\
F(x) = Prob(X_{i} \leq x) &= \sum_{k=0}^{\lfloor x \rfloor} Prob(X_{i}=x)
\end{aligned}$$
Note the infinite discrete support $\{0,1,2,\ldots\}$: there is no upper bound on the count, though large values become vanishingly unlikely.

For example, a small store averages $\lambda = 3.5$ customers per hour.
In any given hour the actual count might be 2, or 5, or 7. 
For our store, $\lambda = 3.5$ means we expect 3.5 customers per hour on average and the following assumptions hold

- Arrivals are independent of each other (one customer entering does not make the next more or less likely)
- The average rate $\lambda$ is constant over the interval
- Two customers cannot arrive at exactly the same instant

We can then compute a few probabilities by hand.
For example, the chance that zero customers arrive in a given hour is
$$\begin{aligned}
Prob(X_{i}=0) &= e^{-3.5} \cdot 3.5^{0} / 0! = e^{-3.5} \approx 0.030
\end{aligned}$$
The chance of exactly 2 customers is
$$\begin{aligned}
Prob(X_{i}=2) &= e^{-3.5} \cdot 3.5^{2} / 2! = e^{-3.5} \cdot 6.125 \approx 0.185
\end{aligned}$$


```{r}
## Store: lambda = 3.5 customers per hour
lambda <- 3.5

# PMF plot
x <- seq(0, 15)
f_x <- dpois(x, lambda)
plot(x, f_x, type='h', col=rgb(0, 0, 1, .8),
     xlab = 'Customers per hour', ylab = 'Prob(X = x)', main=NA)
points(x, f_x, pch = 16, col = rgb(0, 0, 1, .8))
title(bquote(paste('Pois(', .(lambda), ')')), font.main=1)

# CDF plot
F_x <- ppois(x, lambda)
plot(x, F_x, type='s', col=rgb(0, 0, 1, .8),
    xlab = 'Customers per hour', ylab = 'F(x)', main=NA)
points(x, F_x, pch=16, col=rgb(0, 0, 1, .8))
title(bquote(paste('Pois(', .(lambda), ') CDF')), font.main=1)
```

:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
Show that $\mathbb{E}[X_{i}] = \mathbb{V}[X_{i}]= \lambda$.
That is, the Poisson mean and variance are both equal to the rate parameter.
```{r}
# Compare empirical simulation vs theoretical
lambda <- 3.5
X <- rpois(1e4, lambda)
c(emp_mean = mean(X), th_mean = lambda)
c(emp_var  = var(X),  th_var  = lambda)
```
:::

:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
During the lunch rush, the store averages $\lambda=4.5$ customers per hour.
What is the probability that more than $7$ customers arrive in a given hour?

We need $Prob(X_{i} > 7) = 1 - Prob(X_{i} \leq 7) = 1 - F(7)$.
$$\begin{aligned}
Prob(X_{i} > 7) &= 1 - \sum_{k=0}^{7} \frac{e^{-4.5} \cdot 4.5^{k}}{k!}
\end{aligned}$$
```{r}
# Prob(X > 7)
1 - ppois(7, lambda=4.5)

# Verify: sum the PMF from 0 to 7
sum(dpois(0:7, lambda=4.5))
```
There is roughly a $9\%$ chance of more than $7$ arrivals in a given hour.
:::

The Poisson assumes the mean equals the variance.
In real count data the variance is often larger than the mean (called *overdispersion*), because real-world arrivals are bunchier than the Poisson assumes.
Before using a Poisson, compute the ratio $\text{var}(x)/\text{mean}(x)$ on the data; a ratio well above $1$ is a warning sign.



The shape of the Poisson also changes with $\lambda$.
A quiet specialty shop ($\lambda=1$) has a strongly right-skewed arrival pattern, while a busy supermarket ($\lambda=15$) looks nearly Normal.

```{r}
# Shape varies with lambda
par(mfrow=c(1, 3))
for(lam in c(1, 5, 15)) {
    x <- seq(0, 30)
    f_x <- dpois(x, lam)
    plot(x, f_x, type='h', col=rgb(0, 0, 1, .8),
        xlab = 'Customers per hour', ylab = 'Prob(X = x)', main=NA)
    points(x, f_x, pch=16, col=rgb(0, 0, 1, .8))
    title(bquote(paste('Pois(', .(lam), ')')), font.main=1)
}
```


#### **Large Sample Approximations**. {-}

The *de Moivre–Laplace Theorem* says that as $n$ grows large, with $p \in (0,1)$ staying fixed, the Binomial distribution is approximately Normal with mean $np$ and variance $np(1-p)$

:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
The unemployment rate is $10%$.
Suppose that $100$ employable people are selected randomly.
What is the probability that this sample contains between $9$ and $12$ unemployed people.
Use the normal approximation to binomial probabilities.

Parameters are 
$$\begin{aligned}
\mu &\approx n p = 10 \\
\sigma^2 &= n p (1−p) = 100 (0.1)(0.9)=9 \\
\sigma &= \sqrt{9}=3.
\end{aligned}$$
:::

The *Poisson Limit Theorem* says that as $n$ grows large, with $p$ growing smaller at the same time, and $n p$ converging to a fixed number $\lambda$, the Binomial distribution is approximately Poisson with parameter $\lambda$.

:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
Imagine dividing one hour into $n$ tiny intervals.
If the store averages $\lambda = 3.5$ customers per hour, each interval has a small probability $p = \lambda/n = 3.5/n$ of an arrival.
With $n$ independent intervals, the number of arrivals is Binomial:
$$Prob(X=x) = \binom{n}{x} \left(\frac{\lambda}{n}\right)^{x} \left(1-\frac{\lambda}{n}\right)^{n-x}$$

As $n$ grows large, three things happen:

1. $\binom{n}{x} \cdot \frac{1}{n^{x}} \to \frac{1}{x!}$, because $\binom{n}{x} = \frac{n(n-1)\cdots(n-x+1)}{x!}$ and each factor in the numerator is approximately $n$
2. $\left(1 - \frac{\lambda}{n}\right)^{n} \to e^{-\lambda}$, by the definition of $e$
3. $\left(1 - \frac{\lambda}{n}\right)^{-x} \to 1$, since $x$ is fixed and $\lambda/n \to 0$

Multiplying the pieces together: $\frac{1}{x!} \cdot \lambda^{x} \cdot e^{-\lambda} = e^{-\lambda}\lambda^{x}/x!$, which is the Poisson PMF.
In other words, the Poisson is what the Binomial becomes when there are many opportunities for an event but each one is individually unlikely (see also Large Sample Approximations below).

```{r}
# Binomial approaches Poisson as n grows
lambda <- 3.5
x <- seq(0, 12)
f_pois <- dpois(x, lambda)

par(mfrow=c(1, 3))
for(n in c(10, 50, 1000)) {
    f_binom <- dbinom(x, size=n, prob=lambda/n)
    comp_cols <- c(rgb(.8, 0, 0, .5), rgb(0, 0, .8, .5))
    plot(x, f_binom, type='h', col=comp_cols[1], lwd=4,
        xlab='Customers', ylab='Prob(X = x)', main=NA)
    points(x + 0.2, f_pois, type='h', col=comp_cols[2], lwd=2)
    points(x + 0.2, f_pois, pch=16, col=comp_cols[2], cex=0.7)
    title(bquote(paste('Binom(', .(n), ', ', .(round(lambda/n, 4)), ')')), font.main=1)
}
legend('topright', c('Binomial', 'Poisson'),
    col=comp_cols, lwd=c(4, 2), cex=0.8)
```
:::


:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
For another example, consider $n=1000$ independent trials each with probability $p=0.004$, so $\lambda = np = 4$.
We can compare the exact Binomial to both approximations at once: the Poisson ($\lambda=4$) and the Normal ($\mu=np=4$, $\sigma^{2}=np(1-p)\approx 3.98$).
```{r}
# Binom(1000, 0.004) vs Pois(4) vs Normal(4, 3.98)
n <- 1000
p <- 0.004
x <- seq(0, 15)
f_binom <- dbinom(x, size=n, prob=p)
f_pois  <- dpois(x, lambda=n*p)
f_norm  <- dnorm(x, mean=n*p, sd=sqrt(n*p*(1-p)))

approx_cols <- hcl.colors(3, alpha=.45)
plot(x, f_binom, type='h', col=approx_cols[1], lwd=4,
    xlab='x', ylab='Prob(X = x)',
    main=NA)
title('Approximating the Binomial', font.main=1)
points(x + 0.15, f_pois, type='h', col=approx_cols[2], lwd=2)
lines(x, f_norm, col=approx_cols[3], lwd=2)
legend('topright',
    c('Binom(1000, 0.004)', 'Pois(4)', 'Normal(4, 3.98)'),
    col=approx_cols, lwd=c(4, 2, 2))
```
The Poisson is nearly exact here because $p$ is very small.
The Normal is a rougher fit because $\lambda=4$ is still fairly small and the distribution is noticeably skewed.
As $\lambda$ grows, the Poisson itself becomes more symmetric and the Normal approximation improves.
:::



## Continuous Distributions

#### **Irwin–Hall**. {-}

Suppose the store has a tip jar and each customer drops in a random amount between $\$0$ and $\$1$.
If $n=2$ customers visit, the total tips are the sum of two independent $\text{Uniform}(0,1)$ random variables.
The [Irwin-Hall distribution](https://en.wikipedia.org/wiki/Irwin%E2%80%93Hall_distribution) describes this sum for any $n$.

The support is continuous on $[0,n]$ and the probability density function is
$$f(x) = \dfrac{1}{(n-1)!} \displaystyle\sum_{k=0}^{\lfloor x \rfloor} (-1)^k \binom{n}{k} (x - k)^{n-1}$$
for $x \in [0,n]$, and $0$ otherwise.
The parameter $n$ is the number of independent uniform contributions.
When $n=1$ this is just the Uniform density; when $n=2$ it is a triangle peaking at $1$.

```{r}
## Tip jar: total tips from n customers, each Uniform(0, 1)
# Irwin-Hall PDF function
dirwinhall <- function(x, n) {
    f_x <- rep(NA, length(x))
    for(i in seq(x)) {
        xx <- x[i]
        if(xx < 0 | xx > n){
            f_x[i] <- 0
        } else {
            k <- seq(0, floor(xx))
            f_k <- sum((-1)^k*choose(n, k)*(xx-k)^(n-1))/factorial(n-1)
            f_x[i] <- f_k
        }
    }
    return(f_x)
}

# Two customers
n <- 2
x <- seq(0, n, length.out = 500)

# Compute and plot PDF
f_x <- dirwinhall(x, n)
plot(x, f_x, type='l', col=rgb(0, 0, 1, .8),
    main=NA, xlab = 'Total tips ($)', ylab = 'f(x)')
title(bquote(paste('IrwinHall(', .(n), ')')), font.main=1)
```

See also the [Bates](https://en.wikipedia.org/wiki/Bates_distribution) distribution, which is for the mean of $n$ i.i.d. $\text{Uniform}(0,1)$ random variables.
It essentially rescales the Irwin-Hall distribution to range between $0$ and $1$.


#### **Beta**. {-}

When the variable of interest is itself a proportion that can vary from one occasion to the next, we need a distribution supported on the unit interval $[0,1]$ rather than the whole real line.

:::{.callout-important icon=false title="Key Definition" collapse="false"}
A *Beta* random variable lives on the unit interval $[0,1]$ and has two shape parameters $a, b > 0$ that control the location and skewness of the density.
It is the natural distribution for a proportion (e.g., a daily purchase rate).
:::

The *[Beta distribution](https://en.wikipedia.org/wiki/Beta_distribution)* is useful whenever the quantity of interest is bounded between $0$ and $1$: a fraction of customers who buy, a click-through rate, a survival probability.
When both shape parameters equal $2$, the distribution is symmetric and bell-shaped on $[0,1]$, centering around $50\%$; varying the shape parameters allows everything from sharply right-skewed to U-shaped.
The density formula is mathematically involved, so we omit it and work graphically instead.

```{r}
## Daily purchase rate: Beta(2, 2)
X4 <- rbeta(2000, 2, 2)
hist(X4, breaks=20, border=NA, main=NA, freq=FALSE,
    xlab='Purchase rate')

x <- seq(0, 1, by=.01)
fx <- dbeta(x, 2, 2)
lines(x, fx, col=rgb(0, 0, 1, .8))
```

We can find probabilities using either the density (area under the curve) or the CDF.

:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
What is the probability that the daily purchase rate falls between $20\%$ and $80\%$?
Intuitively depict this as the area under the density.
Then compute it numerically using the CDF.
```{r}
plot(ecdf(X4), main=NA,
    xlab='Purchase rate') # Empirical

x <- seq(0, 1, by=.01) # Theoretical
Fx <- pbeta(x, 2, 2)
lines(x, Fx, col=rgb(0, 0, 1, .8))

# Middle Interval Example
F2 <- pbeta(0.2, 2, 2)
F8 <- pbeta(0.8, 2, 2)
F_2_8 <- F8 - F2
F_2_8

# Visualize
title('Purchase rate between 0.2 and 0.8', font.main=1)
segments(0.2, F2, -1, F2, col=rgb(1, 0, 0, .8))
segments(0.8, F8, -1, F8, col=rgb(1, 0, 0, .8))
```
:::

The two shape parameters allow the Beta to take many different shapes, making it useful whenever the quantity of interest is a proportion.

:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
Different shape parameters represent different beliefs about the store's purchase rate.
For each density below, intuitively depict $Prob(X_{i} \leq 0.5)$ using the PDF.
Repeat the exercise using a CDF instead of a PDF to calculate a numerical value.
```{r, results='hide'}
#| fig-width: 8
#| fig-height: 6
op <- par(no.readonly = TRUE); on.exit(par(op), add = TRUE)
x <- seq(0, 1, by=.01)
pars <- expand.grid(c(.5, 1, 2), c(.5, 1, 2))
par(mfrow=c(3, 3))
apply(pars, 1, function(p){
    fx <- dbeta(x, p[1], p[2])
    plot(x, fx, type='l', xlim=c(0, 1), ylim=c(0, 4), lwd=2, col=rgb(0, 0, 1, .8),
        xlab='Purchase rate', ylab='f(x)', main=NA)
})
title('Beta densities', outer=TRUE, line=-1, font.main=1)
```
:::



## Set Theory

#### **Independence and Conditional Probability**. {-}

Two ideas turn raw events into a calculus: when one event tells us something about another, and when it does not.

:::{.callout-important icon=false title="Key Definition" collapse="false"}
Suppose there are two events $A,B$ with $Prob(A)>0, Prob(B) > 0$).

The *conditional probability* of $A$ given $B$ is $Prob(A \mid B) = Prob(A \cap B) / Prob(B)$. 

The events are *independent* if $Prob(A \cap B) = Prob(A) \cdot Prob(B)$. 
Equivalently stated, independent events have: $Prob(A \mid B)=Prob(A)$.
:::

These two concepts are useful as the foundation for the rest of probability: independence makes joint probabilities multiplicative, while conditional probability lets us update beliefs once new information arrives (Bayes' theorem below builds on this).
Together with the basic set operations on a sample space, they give us the full vocabulary for events:

- the *union*: $A\cup B$, refers to either event occurring
- the *intersection*: $A\cap B$, refers to both events occurring
- the events $A$ and $B$ are *mutually exclusive* if and only if $A\cap B$ is empty
- the *inclusion–exclusion* rule is: $Prob(A\cup B)=Prob(A)+Prob(B)-Prob(A\cap B)$

For example, consider a six sided die where events are the number $>2$ or the number is even.

- The sets are $A=\{3,4,5,6\}$ and $B=\{2,4,6\}$
- To check mutually exclusive: notice $A\cap B=\{4,6\}$ which is not empty.
  So event $A$ and event $B$ are not mutually exclusive

Further assuming the die are fair, each side has an equal $1/6$ probability.
This lets us compute

- $Prob(A)=4/6=2/3$ and $Prob(B)=3/6=1/2$
- $A\cup B=\{2,3,4,5,6\}$, which implies $Prob(A\cup B)=5/6$
- $A\cap B=\{4,6\}$, which implies $Prob(A\cap B)=2/6=1/3$
- To check independence: notice $Prob(A\cap B)=1/3 = (2/3) (1/2) = Prob(A)Prob(B)$.
  So event $A$ and event $B$ are mutually independent
- To check inclusion–exclusion: notice $Prob(A) + Prob(B) - Prob(A\cap B) = 2/3 + 1/2 - 1/3 = 4/6 + 3/6 - 2/6 = 5/6 = Prob(A\cup B)$
- Finally, the conditional probability of $Prob(\text{die is} >2 \mid \text{die shows even number}) = Prob(A \mid B) = \frac{Prob(A \cap B)}{P(B)} = \frac{1/3}{1/2} = 2/3$

```{r}
# Simulation verification
set.seed(123)
x <- seq(1, 6)
rolls <- sample(x, 1e6, replace = TRUE)
A <- rolls > 2
B <- rolls %% 2 == 0
P_AcondB <- mean(A & B) / mean(B)
round(P_AcondB, 3)
```

:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
Consider a six sided die with sample space $\{1,2,3,4,5,6\}$ where each outcome is each equally likely.
What is $Prob(\text{die is odd or }<5)$?
Denote the odd events as $A=\{1,3,5\}$ and the less than five events as  $B=\{1,2,3,4\}$.
Then

- $A\cap B=\{1,3\}$, which implies $Prob(A\cap B)=2/6=1/3$.
- $Prob(A)=3/6=1/2$ and $Prob(B)=4/6=2/3$.
- By inclusion–exclusion: $Prob(A\cup B)=Prob(A)+Prob(B)-Prob(A\cap B)=1/2+2/3-1/3=5/6$.

Now find $Prob(\text{die is even and }<5)$.
Verify your answer with a computer simulation.
:::




#### **Bayes' Theorem**. {-}

Often, we know $Prob(B \mid A)$ but want $Prob(A \mid B)$: a diagnostic test reports its accuracy in one direction but we need the answer in the other.

:::{.callout-important icon=false title="Key Definition" collapse="false"}
*Bayes' theorem* reverses the direction of a conditional probability:
$$Prob(A \mid B) = \frac{Prob(B \mid A) \cdot Prob(A)}{Prob(B)}.$$
It lets us update a prior belief about $A$ into a posterior belief once we observe $B$.
:::

Bayes' theorem is useful as the workhorse of *evidence updating*: any time data $B$ are observed and we want to revise the probability of some hypothesis $A$, this is the formula that does it.
The derivation is short: from the definition of conditional probability, $Prob(A \cap B) = Prob(B \mid A) \cdot Prob(A)$, and substituting into $Prob(A \mid B) = Prob(A \cap B) / Prob(B)$ gives the formula above.

In the die example, we computed $Prob(A \mid B) = 2/3$: the probability the die is $>2$, given it is even.
We can use Bayes' theorem to reverse this: what is the probability the die is even, given it is $>2$?

:::{.callout-note icon=false collapse="true" title="Must Know"}
Using the same die with $A=\{3,4,5,6\}$ and $B=\{2,4,6\}$, we know $Prob(A \mid B)=2/3$, $Prob(B)=1/2$, and $Prob(A)=2/3$.
By Bayes' theorem,
$$\begin{aligned}
Prob(B \mid A) &= \frac{Prob(A \mid B) \cdot Prob(B)}{Prob(A)} = \frac{(2/3)(1/2)}{2/3} = \frac{1}{2}.
\end{aligned}$$
We can verify directly: given $A=\{3,4,5,6\}$, the even outcomes are $\{4,6\}$, so $Prob(B \mid A)=2/4=1/2$.
Since $A$ and $B$ are independent, $Prob(B \mid A) = Prob(B)$ and flipping the conditioning changes nothing.
```{r}
# Simulation verification
set.seed(123)
rolls <- sample(seq(1, 6), 1e6, replace=TRUE)
A <- rolls > 2
B <- rolls %% 2 == 0
# Prob(B | A)
mean(B[A])
```
:::

Bayes' theorem is most useful when events are not independent, so that knowing $B$ occurred changes what we believe about $A$.

:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
A company sources parts from two suppliers.
Supplier $A$ provides $60\%$ of parts and supplier $B$ provides $40\%$.
The defect rate from supplier $A$ is $3\%$ and from supplier $B$ is $5\%$.
If a randomly selected part is defective ($D$), what is the probability it came from supplier $A$?

We know $Prob(A)=0.6$, $Prob(\lnot A)=0.4$, $Prob(D \mid A)=0.03$, and $Prob(D \mid \lnot A)=0.05$.
By the Law of Total Probability,
$$\begin{aligned}
Prob(D) &= Prob(D \mid A) \cdot Prob(A) + Prob(D \mid \lnot A) \cdot Prob(\lnot A) \\
&= 0.03 \times 0.6 + 0.05 \times 0.4 = 0.038.
\end{aligned}$$
By Bayes' theorem,
$$\begin{aligned}
Prob(A \mid D) &= \frac{Prob(D \mid A) \cdot Prob(A)}{Prob(D)} = \frac{0.03 \times 0.6}{0.038} \approx 0.474.
\end{aligned}$$
Even though supplier $A$ provides $60\%$ of all parts, they account for only about $47\%$ of defectives because their defect rate is lower.
```{r}
# Simulation verification
set.seed(42)
n <- 1e6
supplier <- sample(c('A', 'B'), n, replace=TRUE, prob=c(0.6, 0.4))
defect_rate <- ifelse(supplier == 'A', 0.03, 0.05)
defective <- runif(n) < defect_rate

# Prob(A | D)
mean(supplier[defective] == 'A')
```
:::


#### **Law of Total Probability**. {-}

This law states that $Prob(A)=Prob(A \mid B) P(B) + Prob(A \mid \lnot B) P(\lnot B)$, where $\lnot B$ is the event "not $B$".

For example, consider a six sided die where events are the number $>2$ or the number is even.
The sets are $A=\{3,4,5,6\}$ and $B=\{2,4,6\}$ and $\lnot B = \{1,3,5\}$.
From before, we know that $Prob(A)=2/3$, $Prob(B)=1/2$, and $Prob(A \mid B) = 2/3$.
Then see that $Prob(A \mid B) Prob(B) = \frac{2}{3} \frac{1}{2} = 2/6$ 

We can also directly compute $Prob(\lnot B)=1/2$, or infer it from $Prob(\lnot B)=1-Prob(B)$.
The conditional probability of $Prob(\text{die is} >2 \mid \text{die shows odd number}) = Prob(A \mid \lnot B) = \frac{Prob(A \cap \lnot B)}{P(\lnot B)} = 2/3$.

Then we can see that
$Prob(A \mid B) Prob(B) + Prob(A \mid \lnot B) Prob(\lnot B) = 2/6 + 2/6 = 2/3 = Prob(A)$.


Note that substituting the Law of Total Probability into the denominator of Bayes Theorem yields
$$\begin{aligned}
Prob(A \mid B) &= \frac{Prob(B \mid A) \cdot Prob(A)}{Prob(B \mid A) \cdot Prob(A) + Prob(B \mid \lnot A) \cdot Prob(\lnot A)}
\end{aligned}$$


## Exercises

1. Comment the script you wrote for this chapter, then restart R and check that it runs from a clean session, then check the script with AI as explained in [Working with AI](04_A_WorkingWithAI.qmd).
   Write three sentences from memory on the main statistical idea of this chapter, and ask the assistant what is wrong, vague, or missing.
   Finish with your own questions about whatever you found hardest.

2. The Poisson distribution assumes events are independent and occur at a constant rate.
   Give an example of a real-world count variable where these assumptions are likely violated, and explain which assumption fails.

3. A call center receives an average of $\lambda = 6$ calls per hour.
   Using the Poisson distribution, compute $Prob(X_{i} \leq 3)$ and $Prob(X_{i} > 10)$ by hand (set up the sums) and then verify with `ppois` in R.

4. Using the `warpbreaks` dataset in R, compute the sample mean and variance of `breaks`.
   Compare their ratio to what the Poisson distribution predicts (i.e., $\mathbb{E}[X_{i}] = \mathbb{V}[X_{i}] = \lambda$).
   Does the Poisson seem like a good fit?
   Simulate 10000 draws from a Poisson with $\lambda = \bar{X}$ and compare the simulated histogram to the observed data.

#### **Further Reading**. {-}

* <https://learningstatisticswithr.com/book/probability.html#the-binomial-distribution> -- Binomial distribution mechanics and connection to Bernoulli trials, with R code.
* <https://www.khanacademy.org/math/statistics-probability/random-variables-stats-library/poisson-distribution/a/poisson-distribution> -- Intuition and examples for the Poisson distribution.
* <https://www.econometrics-with-r.org/2.1-random-variables-and-probability-distributions.html> -- Probability distributions including discrete families used in econometrics.

#### **Recall** {-}

This chapter rounded out univariate probability with the *Binomial* and *Poisson* (discrete counts), the *Irwin–Hall* and *Beta* (continuous random variables built from Uniforms), and the set-theoretic machinery of *independence*, *conditional probability*, *Bayes' theorem*, and the *Law of Total Probability*.
The defective-parts example wired these ideas together: with $Prob(A)=0.6$, $Prob(D \mid A)=0.03$, and $Prob(D \mid \lnot A)=0.05$, the Law of Total Probability gives $Prob(D)= 0.03 \times 0.6 + 0.05 \times 0.4 = 0.038$, and Bayes' theorem then reverses the conditioning: $Prob(A \mid D) = (0.03 \times 0.6)/0.038 \approx 0.474$.
In the next chapter we collect three advanced empirical tools: kernel densities, data transformations, and drawing samples by inverting a distribution function.

