# Statistics
***

We often summarize distributions with *statistics*: functions of data. The most basic way to do this is with `summary`, whose values can all be calculated individually. 

```{r}
# A random sample (real data)
X <- USArrests[,'Murder']
summary(X)

# A random sample (computer simulation)
X <-  runif(1000)
summary(X)

# Another random sample (computer simulation)
X <- rnorm(1000) 
summary(X)
```

Together, the mean and variance statistics summarize the central tendency and dispersion of a distribution. In some special cases, such as with the normal distribution, they completely describe the distribution. Other distributions are better described with other statistics, either as an alternative or in addition to the mean and variance. After discussing those other statistics, we will return to the two most basic statistics in theoretical detail.

## Mean and Variance

The mean and variance are the two most basic statistics that summarize the center and how spread apart the values are. As before, we represent data as vector $\hat{X}=(\hat{X}_{1}, \hat{X}_{2}, ....\hat{X}_{n})$, where there are $n$ observations and $\hat{X}_{i}$ is the value of the $i$th one.



#### **Mean**. {-}
Perhaps the most common statistic is the mean, which is the [sum of all values] divided by [number of values];
\begin{eqnarray}
\bar{X} &=& \frac{\sum_{i=1}^{n}\hat{X}_{i}}{n},
\end{eqnarray}
where $\hat{X}_{i}$ denotes the value of the $i$th observation.

:::{.callout-note icon=false collapse="true"}
For example, a dataset of $\{1,4,10\}$ has a mean of $[1+4+10]/3=5$.
```{r}
X <- c(1,4,10)
sum(X)/length(X)
mean(X)
```
:::

```{r}
# compute the mean of a random sample
X1 <- USArrests[,'Murder']
X1_bar <- mean(X1)
X1_bar

# visualize on a histogram
hist(X1, border=NA, main=NA)
abline(v=X1_bar, col=2, lwd=2)
title(paste0('mean= ', round(X1_bar,2)), font.main=1)
```

#### **Variance**.{-}
Perhaps the second most common statistic is the variance: the average squared deviation from the mean
\begin{eqnarray}
V_{X} &=&\frac{\sum_{i=1}^{n} [\hat{X}_{i} - \bar{X}]^2}{n}.
\end{eqnarray}
The standard deviation is simply $s_{X} = \sqrt{V_{X}}$, which can be interpreted as the average distance from the mean.

:::{.callout-note icon=false collapse="true"}
For example, a dataset of $\{1,4,10\}$ has a mean of $[1+4+10]/3=5$. The variance is $[(1-5)^2+(4-5)^2+(10-5)^2]/3=14$ and the standard deviation is $\sqrt{14}=3.742$.
```{r}
X <- c(1,4,10)
X_bar <- mean(X)
X_var <- mean( (X - X_bar)^2 )
sqrt(X_var)
```
:::

```{r}
X1_s <- sd(X1) # sqrt(var(X))
hist(X1, border=NA, main=NA, freq=F)
X1_s_lh <- c(X1_bar - X1_s,  X1_bar + X1_s)
abline(v=X1_s_lh, col=4)
text(X1_s_lh, -.02,
    c( expression(bar(X)-s[X]), expression(bar(X)+s[X])),
    col=4, adj=0)
title(paste0('sd= ', round(X1_s,2)), font.main=1)
```

:::{.callout-tip icon=false collapse="true"}
Note that a "corrected version" is used by R and many statisticians: $V_{X}' =\frac{\sum_{i=1}^{n} [\hat{X}_{i} - \bar{X}]^2}{n-1}$ and $s_{X}' = \sqrt{V_{X}'}$. In this class, we use the "unocorrect version" as defined previously.
```{r}
X <- c(1,4,10)

X_bar <- mean(X)
X_var <- mean( (X - X_bar)^2 )
X_var

# Corrected Version
n <- length(X)  
X_var2 <- sum( (X - X_bar)^2 )/(n-1)
X_var2

var(X) # R-Version
```
:::

## Other Center/Spread Statistics

A general rule of applied statistics is that there are multiple ways to measure something. Mean and Variance are measurements of Center and Spread, but there are others that have different theoretical properties and may be better suited for your dataset.

#### **Medians and Absolute Deviations**. {-}
We can use the *Median* as a "robust alternative" to means that is especially useful for data with asymmetric distributions and extreme values. Recall that the $q$th quantile is the value where $q$ percent of the data are below and ($1-q$) percent are above. The median ($q=.5$) is the point where half of the data is lower values and the other half is higher. This means that median is not sensitive to extreme values (whereas the mean is).



```{r}
X <- rexp(50, 0.4)
X
hist(X, freq=F, breaks=seq(0,16),
    border=NA, main='')

# For discrete numbers
# X <- rgeom(50, .4)
#proportions <- table(X)/length(X)
#plot(proportions, ylab='proportion')
```

:::{.callout-note icon=false collapse="true"}
Examine robustness to an extreme value
```{r}

X_extreme <- c(X, 1000) # add one extreme value
#par(mfrow=c(1,2)) # visualize side-by-side
#hist(X)
#hist(X_extreme)


# Which measures of central tendency are robust
# to a single extreme value?
mean(X)
mean( X_extreme )

quantile(X, prob=0.5)
quantile(X_extreme, prob=0.5)
```
:::


:::{.callout-tip icon=false collapse="true"}
Just like weighted means, we can also compute weighted quantiles
```{r}
weighted.quantile <- function(x, w, probs){
    #See spatstat.univar::weighted.quantile
    oo <- order(x)
    x <- x[oo]
    w <- w[oo]
    Fx <- cumsum(w)/sum(w)
    quantile_id <- max(which(Fx <= probs))+1
    xq <- x[quantile_id] 
    return(xq)
    
}

## Unweighted
quantile(X, probs=.5)
weights <- rep(1, length(X))
weighted.quantile(x=X, w=weights, probs=.5)

## Weighted
weights <- seq(X)
weights <- weights/sum(weights) # normalize
weighted.quantile(x=X, w=weights, probs=.5)
```
:::

We can also use the *Interquartile Range* or *Median Absolute Deviation* as an alternative to variance. The difference between the first and third quartiles ($q=.25$ and $q=.75$) measure is range of the middle $50%$ of the data, which is how the boxplot measures "spread". The median absolute deviation is another statistic that also measures "spread".
\begin{eqnarray}
\tilde{X} &=& Med( \hat{X}_{i}) \\
MAD_{X} &=& Med\left( | \hat{X}_{i} - \tilde{X} | \right).
\end{eqnarray}

:::{.callout-note icon=false collapse="true"}
```{r}
X <- c(1,4,10)
# An alternative to sd(X), used in the boxplot
quants <- quantile(X, probs=c(.25,.75))
quants[2]-quants[1]
IQR(X)

#Another alternative to sd(X)
mad(X, constant=1)
# Intuitively:
# = median( abs(1-4), abs(4-4), abs(10-4) ) 
# = median( 3,0,6 ) = 3

# Computationally equivalent
# x_med <- quantile(X, probs=.5)
# x_absdev <- abs(X -x_med)
# quantile(x_absdev, probs=.5)
```
:::

:::{.callout-note icon=false collapse="true"}
Examine robustness to an extreme value
```{r}
X <- rgeom(50, .4)
X_extreme <- c(X, 1000) # add one extreme value

sd(X)
sd(X_extreme)

mad(X, constant=1)
mad(X_extreme, constant=1)

IQR(X)
IQR(X_extreme)
```
:::


:::{.callout-tip icon=false collapse="true"}
Note that there other "absolute deviation" statistics
```{r, eval=F}
# sometimes seen elsewhere
mean( abs(X - mean(X)) )
mean( abs(X - median(X)) )
median( abs(X - mean(X)) )
```
:::

#### **Mode and Share Concentration**. {-}
Sometimes, none of the above work well. With categorical data, for example, distributions are easier to describe with other statistics. The *mode* is the most common observation: the value with the highest observed frequency. We can also measure the spread of the frequencies or concentration at the mode vs elsewhere.

```{r}
# Draw 3 Random Letters
x <- LETTERS
x_probs <- 1:length(x)
x_probs <- x_probs/sum(x_probs) # probs must sum to 1
X <- sample(x, 1, prob=x_probs, replace=T)
X

# Draw Random Letters 100 Times
X <- sample(x, 1000, prob=x_probs, replace=T)
X <- factor(unlist(X), levels=LETTERS, ordered=F)

proportions <- table(X)/length(X)
plot(proportions, col=grey(0,0.5))

# mode(s)
mode_id <- which(proportions==max(proportions))
names(proportions)[ mode_id ]

# freq. spreads
sd(proportions)
sum(proportions^2)
# freq. concentration at mode
max(proportions)/mean(proportions)
```

## Shape Statistics

Central tendency and dispersion are often insufficient to describe a distribution. To further describe shape, we can compute skew and kurtosis to measure asymmetry and extreme values. There are many other statistics we could compute on an ad-hoc basis.

#### **Skewness**. {-}
The skew statistic captures how asymmetric the distribution is by measuring the average cubed deviation from the mean, normalized the standard deviation cubed
\begin{eqnarray}
W_{X} &=& \frac{\sum_{i=1}^{n} [\hat{X}_{i} - \bar{X}]^3 / n}{ [s_{X}]^3 }
\end{eqnarray}

```{r}
X <- rweibull(1000, shape=1)
hist(X, border=NA, main=NA, freq=F, breaks=20)

skewness <-  function(X){
 X_bar <- mean(X)
 m3 <- mean((X - X_bar)^3)
 s3 <- sd(X)^3
 skew <- m3/s3
 return(skew)
}

skewness( rweibull(1000, shape=1))
skewness( rweibull(1000, shape=10) )
```

We can automatically compare against the normal distribution, which has a skew of 0.

#### **Kurtosis**. {-}
This statistic captures how many "outliers" there are like skew but looking at quartic deviations instead of cubed ones.
\begin{eqnarray}
K_{X} &=& \frac{\sum_{i=1}^{n} [\hat{X}_{i} - \bar{X}]^4 / n}{ [s_{X}]^4 }.
\end{eqnarray}

Some authors further subtract 3 to explicitly compare against the normal distribution (the normal distribution has a kurtosis of 3).

```{r}
X <- rweibull(1000, shape=1)
boxplot(X, main=NA)

kurtosis <- function(X){  
 X_bar <- mean(X)
 m4 <- mean((X - X_bar)^4) 
 s4 <- sd(X)^4
 kurt <- m4/s4
 excess_kurt <- kurt - 3 # compare against normal
 return(kurt)
}

kurtosis( rweibull(1000, shape=1) )
kurtosis( rweibull(1000, shape=10) )
```


#### **Clusters/Gaps**. {-}
You can also describe distributions in terms of how clustered the values are, including the number of modes, bunching, and many other statistics. But remember that "a picture is worth a thousand words".

```{r}
# Complicated Random Variable
r_ugly0 <- function(n, populationMean=c(-2, 3), populationVar=c(1, 4), prob=c(.5,.5)){
  # Which distribution is each observation coming from
  di <- sample(1:2, size=n, replace=TRUE, prob=prob)
  rnorm(n, mean=populationMean[di], sd=sqrt(populationVar)[di])
}

X <- r_ugly0(6000)
hist(X, breaks=60,
    freq=F, border=NA,
    xlab="x", main='')

#hist( rbeta(1000, .6, .6), border=NA, main=NA, freq=F, breaks=20)
```

```{r}
# Another Complicated Random Variable
r_ugly1 <- function(n, theta1=c(-8,-1), theta2=c(-2,2), rho=.25){
    omega   <- rbinom(n, size=1, rho)
    epsilon <- omega * runif(n, theta1[1], theta2[1]) +
        (1-omega) * rnorm(n, theta1[2], theta2[2])
    return(epsilon)
}
# Very Large Sample
X <- r_ugly1(1000000)
hist(X, breaks=1000,  freq=F, border=NA,
    xlab="x", main='')
```

```{r, eval=F}
# Show True Density
d_ugly1 <- function(x, theta1=c(-8,-1), theta2=c(-2,2), rho=.25){
    rho     * dunif(x, theta1[1], theta2[1]) +
    (1-rho) * dnorm(x, theta1[2], theta2[2]) }
x <- seq(-12,6,by=.001)
dx <- d_ugly1(x)
lines(x, dx, col=1)
```



## Probability Theory

You were already introduced to this with [random variables](https://jadamso.github.io/Rbooks/random-variables.html). In this section, we will dig a little deeper theoretically into the statistics we compute. When we know how the data are generated theoretically, we can often compute the theoretical value of the two most basic and often-used statistics: the mean and variance. To see this, we separately analyze how they are computed for discrete and continuous random variables.

Here, we denote $X_{i}$ as a random variable that can take on specific values $x$ from the sample space with known probabilities.

#### **Discrete Random Variables**. {-} 
If the sample space is discrete, we can compute the theoretical mean (or expected value) as
$$
\mathbb{E}[X_{i}] = \sum_{x} x Prob(X_{i}=x),
$$
where $Prob(X_{i}=x)$ is the probability the random variable $X_{i}$ takes the particular value $x$. Similarly, we can compute the theoretical variance as
$$
\mathbb{V}[X_{i}] = \sum_{x} \left(x - \mathbb{E}[X_{i}] \right)^2 Prob(X_{i}=x).
$$
The theoretical standard deviation is $\mathbb{s}[X_{i}] = \sqrt{\mathbb{V}[X_{i}]}$.

:::{.callout-note icon=false collapse="true"}
For example, consider an unfair coin with a $3/4$ probability of heads ($x=1$) and a $1/4$ probability of tails ($x=0$) has a theoretical mean of 
$$
\mathbb{E}[X_{i}] = 1\frac{3}{4} + 0\frac{1}{4} = .75
$$
and a theoretical variance of 
$$
\mathbb{V}[X_{i}] = [1 - .75]^2 \frac{3}{4} + [0 - .75]^2 \frac{1}{4} = 0.1875
$$

```{r}
# A simulation of many coin flips
x <- c(0,1)
x_probs <- c(1/4, 3/4)
X <- sample(x, 1000, prob=x_probs, replace=T)

round( mean(X), 4)

round( var(X), 4)
```

Now consider a four-sided die with outcomes $(1,2,3,4)$ and probabilities $(1/8, 2/8, 1/8, 4/8)$ . What is the mean?
$$
\mathbb{E}[X_{i}] = 1\times 1/8 + 2 \times 2/8 + 3 \times 1/8 + 4 \times 4/8  = 3
$$
What is the variance? Verify both the mean and variance by simulation.
:::


:::{.callout-tip icon=false collapse="true"}
Consider a six-sided die with equal probability of landing on each side. What is the mean?

```{r}
# Manual Way
1*1/6 + 2*1/6 + 3*1/6 + 4*1/6 + 5*1/6 + 6*1/6

# Another Manual Way (only works with equal probabilities)
(1+2+3+4+5+6)/6

# Computerized Way
x <- c(1,2,3,4,5,6)
x_probs <- c(1/6, 1/6, 1/6, 1/6, 1/6, 1/6)
X_bar <- sum(x*x_probs)
X_bar

# Verified by simulation
Xsim <- sample(x, prob=x_probs,
          size=10000, replace=T)
mean(Xsim)
```

What is the variance and standard deviation?
```{r}
# Manual Way
Xvar <- (1-X_bar)^2*1/6 +
  (2 - X_bar)^2*1/6 +
  (3 - X_bar)^2*1/6 +
  (4 - X_bar)^2*1/6 +
  (5 - X_bar)^2*1/6 + 
  (6 - X_bar)^2*1/6
Xvar
Xsd <- sqrt(Xvar)
Xsd

# Verified by simulation
sd(Xsim)
```
:::

#### **Weighted Data**. {-} 
Sometimes, you may have a dataset of values and probability weights. Othertimes, you can calculate them yourself. In either case, you can explicitly do the computations for discrete data. Given data on unique outcome $x$ and their frequency $\widehat{p}_{x}=\sum_{i=1}^{n}\mathbf{1}\left(X_{i}=x\right)/n$, we compute 
\begin{eqnarray}
\bar{X} &=& \sum_{x} x \widehat{p}_{x}.
\end{eqnarray}

:::{.callout-note icon=false collapse="true"}
Suppose we flipped a coin $100$ times and found that $76$ were Heads and $23$ were Tails. The estimated probabilities are $76/100$ for the outcome $X_{1}=1$ and $24/100$ for the outcome $X_{i}=0$. We compute the mean as $\bar{X}=1\times 0.76 + 0 \times 0.24 = 0.76$.

```{r}
# Compute probability weights for unique values
P  <- table(X) #table of counts
p <- c(P)/length(X) #frequencies (must sum to 1)
x <- as.numeric(names(p)) #unique values
cbind(x,p)

# Compute Mean
X_mean <- sum(x*p)
X_mean
```
:::

:::{.callout-tip icon=false collapse="true"}
Try computing the two different ways for another random sample
```{r}
x <- c(0,1,2)
x_probs <- c(1/3,1/3,1/3)
X  <-  sample(x, prob=x_probs, 1000, replace=T)

# First Way (Computerized)
mean(X)

# Second Way (Explicit Calculations)
# start with a table of counts like the previous example
```
:::

This idea generalizes to a *weighted mean*: an average where different values contribute to the final result with varying levels of importance. For each outcome $x$ we have a weight $W_{x}$ and compute
\begin{eqnarray}
\bar{X} &=& \frac{\sum_{x} x W_{x}}{\sum_{x} W_{x}} = \sum_{x} x w_{x},
\end{eqnarray}
where $w_{x}=\frac{W_{x}}{\sum_{x'} W_{x'}}$ is normalized version of $W_{x}$ that implies $\sum_{x}w_{x}=1$.

:::{.callout-note icon=false collapse="true"}
For example, suppose a student has these scores
```{r}
Homework1 <- c(score=88, weight=0.25)
Homework2 <- c(score=92, weight=0.25)
Exam1 <- c(score=67, weight=0.2)
Exam2 <- c(score=90, weight=0.3)

Grades <- rbind(Homework1, Homework2, Exam1, Exam2)
Grades
```

We can compute the final grade as a Weighted Mean
```{r}
# Manual Way
88*0.25 + 92*0.25 + 67*0.2 + 90*0.3

# Computerized Way
Values <- Grades[,'score'] * Grades[,'weight']
FinalGrade <- sum(Values)
FinalGrade
```
:::

Note that if there are $K$ unique outcomes and $W_{x}=1$ then $\sum_{x}W_{x}=K$ and $w_{x}=1/K$. This means $\bar{X} = \sum_{x} x w_{x} = \sum_{x} x /K$, which is just a simple mean. 

The weighting idea generalizes to other statistics. E.g., we can also computed a weighted variance. 

:::{.callout-tip icon=false collapse="false"}
Provide an example of computing a weighted variance building on this code below
```{r, eval=F}
x_diff <- (x - x_mean)^2
p <- P/sum(P)
x_var <- sum(p * x_diff)
```
:::


#### **Continuous Random Variables**. {-}

Many continuous random variables are parameterized by their means and variances. For example, the exponential distribution has parameter $\lambda$, which corresponds to the theoretical mean $1/\lambda$ and variance $1/\lambda^2$. For another example, the two parameters of the normal distribution are $\mu$ and $\sigma$, which corresponds to the theoretical mean and variance. 
```{r}
# Exponential Random Variable
X <- rexp(5000, 2)
m <- mean(X)
round(m, 2)

# Normal Random Variable
X <- rnorm(5000, 1, 2)
round(mean(X), 2)
round(var(X), 2)
```

<details>
<summary> Advanced and Optional </summary>
  <p>
If the sample space is continuous, we can compute the theoretical mean (or expected value) as
$$
\mathbb{E}[X] = \int x f(x) d x,
$$
where $f(x)$ is the probability the random variable takes the particular value $x$. Similarly, we can compute the theoretical variance as
$$
\mathbb{V}[X_{i}]= \int \left(x - \mathbb{E}[X_{i}] \right)^2 f(x) d x,
$$

For example, consider a random variable with a continuous uniform distribution over $[-1, 1]$. In this case, $f(x)=1/[1 - (-1)]=1/2$ for each $x \in [-1, 1]$ and 
$$
\mathbb{E}[X_{i}] = \int_{-1}^{1} \frac{x}{2} d x = \int_{-1}^{0} \frac{x}{2} d x + \int_{0}^{1} \frac{x}{2} d x = 0
$$
and 
$$
\mathbb{V}[X_{i}]= \int_{-1}^{1} x^2 \frac{1}{2} d x = \frac{1}{2} \frac{x^3}{3}|_{-1}^{1} = \frac{1}{6}[1 - (-1)] = 2/6 =1/3
$$
```{r}
X <- USArrests[,'Murder']
round( mean(X), 4)
round( var(X), 4)
```
:::{.callout-tip icon=false collapse="false"}
You can again explicitly do the computations with weighted data, but here we have an additional approximation error
```{r}
# values and probabilities
h  <- hist(X, plot=F)
wt <- h[['counts']]/length(X) 
xt <- h[['mids']]
# Weighted mean
X_mean <- sum(wt*xt)
X_mean

# Compare to "mean(x)"
```

Try it yourself with 
```{r}
X <- runif(2000, -1, 1)
```
:::

  </p>
</details>

## Further Reading

Probability Theory

* [Refresher] <https://www.khanacademy.org/math/statistics-probability/probability-library/basic-theoretical-probability/a/probability-the-basics>
* <https://book.stat420.org/probability-and-statistics-in-r.html>
* <https://bookdown.org/speegled/foundations-of-statistics/>
* <https://math.dartmouth.edu/~prob/prob/prob.pdf>
* <https://bookdown.org/probability/beta/discrete-random-variables.html>
* <https://www.econometrics-with-r.org/2.1-random-variables-and-probability-distributions.html>
* <https://probability4datascience.com/ch02.html>
* <https://statsthinking21.github.io/statsthinking21-R-site/probability-in-r-with-lucy-king.html>
* <https://bookdown.org/probability/statistics/>
* <https://www.atmos.albany.edu/facstaff/timm/ATM315spring14/R/IPSUR.pdf>
* <https://rc2e.com/probability>
* <https://bookdown.org/probability/beta/>
* <https://bookdown.org/a_shaker/STM1001_Topic_3/>
* <https://bookdown.org/fsancier/bookdown-demo/>
* <https://bookdown.org/kevin_davisross/probsim-book/>
* <https://bookdown.org/machar1991/ITER/2-pt.html>
* <https://www.atmos.albany.edu/facstaff/timm/ATM315spring14/R/IPSUR.pdf>
* <https://math.dartmouth.edu/~prob/prob/prob.pdf>

For weighted statistics, see

* <https://seismo.berkeley.edu/~kirchner/Toolkits/Toolkit_12.pdf>
* <https://www.bookdown.org/rwnahhas/RMPH/survey-desc.html>

