# Numerical Statistics
***

We often summarize distributions with *statistics*: functions of data.
We will go through some specific examples mathematically below, but can intuitively understand that they are generally computed as `statistic <- function(x){ .... }`.
This chapter covers measures of center (mean, median, mode), spread (variance, IQR, MAD), and shape (skewness, kurtosis).

:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
Note that functions can take functions as arguments, meaning we can also program statistics generally as 
```{r}
statistic <- function(x_hat, f){
    stat_hat <- f(x_hat)
    return(stat_hat)
}

x_hat <- c(0, 1, 3, 10, 6) # Data
statistic(x_hat, sum)
```
:::


The most basic way to compute statistics is with `summary`, which reports multiple values that can all be calculated individually. 

```{r}
# A random sample (real data)
x_hat <- USArrests[, 'Murder']
summary(x_hat)

# A random sample (computer simulation)
x_hat <-  runif(1000)
summary(x_hat)

# Another random sample (computer simulation)
x_hat <- rnorm(1000)
summary(x_hat)
```

Together, the sample mean and variance statistics summarize the central tendency and dispersion of a distribution.
In some special cases, such as with the normal distribution, they completely describe the distribution.
Other distributions are better described with other statistics, either as an alternative or in addition to the mean and variance.
After discussing those other statistics, we will return to the two most basic statistics in theoretical detail.

## Mean and Variance

The mean and variance are the two most basic statistics that summarize the center and how spread apart the values are for data in your sample.
As before, we represent data as vector $\hat{x}=(\hat{x}_{1}, \hat{x}_{2}, ....\hat{x}_{n})$, where there are $n$ observations and $\hat{x}_{i}$ is the value of the $i$th one.


#### **Mean** {-}

The first step in describing a sample is to reduce its center to a single number.

:::{.callout-important icon=false title="Key Definition" collapse="false"}
The sample *mean* is the sum of the observations divided by the number of observations,
$$\hat{m} = \frac{1}{n} \sum_{i=1}^{n} \hat{x}_{i}.$$
:::

The mean (also called the *sample mean* or *empirical mean*) is the simplest and most commonly reported single-number summary of a sample, and the natural measure of central tendency when the data are roughly symmetric.
The mean is useful when totals matter: total income across a population equals the mean income times the number of people, a relation that does not hold for the median or other rank-based summaries.
Because every observation contributes equally to the sum, a single extreme value can pull the mean substantially in its direction.

:::{.callout-note icon=false collapse="true" title="Must Know"}
For example, a dataset of $\{1,4,10\}$ has a mean of $[1+4+10]/3=5$.
```{r}
x_hat <- c(1, 4, 10)
sum(x_hat)/length(x_hat)
mean(x_hat)
```
:::

```{r}
# compute the mean of a random sample
x1_hat <- USArrests[, 'Murder']
m1_hat <- mean(x1_hat)
m1_hat

# visualize on a histogram
hist(x1_hat, border=NA, main=NA, freq=FALSE)
abline(v=m1_hat, col=rgb(1, 0, 0, .8), lwd=2)
title(paste0('mean= ', round(m1_hat, 2)), font.main=1)
```

#### **Variance**{-}

Once we know where the data are centered, the next question is how widely they vary around that center.

:::{.callout-important icon=false title="Key Definition" collapse="false"}
The sample *variance* is the average squared deviation from the mean,
$$\hat{v} = \frac{1}{n} \sum_{i=1}^{n} (\hat{x}_{i} - \hat{m})^2.$$
The sample *standard deviation* is its square root, $\hat{s} = \sqrt{\hat{v}}$, and is in the same units as the data.
:::

The variance and standard deviation are the second-most-common single-number summaries of a sample.
They tell us how concentrated or spread out the data are around the mean: a small variance means the observations cluster tightly, a large variance means they fan out.
The standard deviation in particular is useful because it sits in the original units of the data, making sentences like "monthly returns averaged $0.6\%$ with a standard deviation of $4\%$" interpretable at a glance.
Squaring the deviations also means a few extreme values can dominate the result, which is one reason the median absolute deviation (below) is often preferred for heavy-tailed data.

:::{.callout-note icon=false collapse="true" title="Must Know"}
For example, a dataset of $\{1,4,10\}$ has a mean of $[1+4+10]/3=5$.
The variance is $[(1-5)^2+(4-5)^2+(10-5)^2]/3=[16+1+25]/3=42/3=14$ and the standard deviation is $\sqrt{14}$.
```{r}
x_hat <- c(1, 4, 10)
m_hat <- mean(x_hat)
v_hat <- mean( (x_hat - m_hat)^2 )
sqrt(v_hat)
```
:::

```{r}
s1_hat <- sd(x1_hat) # sqrt(var(x_hat))
hist(x1_hat, border=NA, main=NA, freq=FALSE)
m1_sd_bounds <- c(m1_hat - s1_hat,  m1_hat + s1_hat)
abline(v=m1_hat, col=rgb(1, 0, 0, .8), lwd=2) #mean
abline(v=m1_sd_bounds, col=rgb(0, 0, 1, .8)) # +- sd
text(m1_sd_bounds, -.02,
    c( expression(bar(X)-s[X]), expression(bar(X)+s[X])),
    col=rgb(0, 0, 1, .8), adj=0)
title(paste0('sd= ', round(s1_hat, 2)), font.main=1)
```

:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
Note that a "unbiased version" of the empirical variance is used by R and many statisticians: $\hat{v}' =\frac{\sum_{i=1}^{n} [\hat{x}_{i} - \hat{m}]^2}{n-1}$ and $\hat{s}' = \sqrt{\hat{v}'}$.
In this class, we use the version defined previously because we do not yet know about "bias/unbiased" statistics.
Do not be concerned, as there is hardly any difference when $n$ is large: e.g., $\frac{1}{n}\approx \frac{1}{n-1}$ for $n=100,000$.
```{r}
x_hat <- c(1, 4, 10)

m_hat <- mean(x_hat)
v_hat <- mean( (x_hat - m_hat)^2 )
v_hat

# Corrected Version
n <- length(x_hat)
v_corrected_hat <- sum( (x_hat - m_hat)^2 )/(n-1)
v_corrected_hat

var(x_hat) # R-Version
```
:::

## Other Center/Spread Statistics

A general rule of applied statistics is that there are multiple ways to measure something.
Mean and Variance are measurements of Center and Spread, but there are others that have different theoretical properties and may be better suited for your dataset.

#### **Medians and Absolute Deviations** {-}

When the data are skewed or contain outliers, statistics built from quantiles often give a more representative summary of center and spread than the mean and variance.

:::{.callout-important icon=false title="Key Definition" collapse="false"}
The sample *median* $\tilde{m}$ is the value where one half of the data fall below and one half above; it is the $0.5$ quantile.

The *interquartile range* ($IQR$) is the upper quartile minus the lower quartile, the spread of the middle half of the data.

The *median absolute deviation* ($\text{MAD}$) is the median of the absolute deviations from the median, $\text{Med}(|\hat{x}_{i} - \tilde{m}|)$.
:::

These three statistics share a common rationale: by using ranks rather than arithmetic averages, they resist the influence of extreme values.
A single outlier can pull the *mean* by an arbitrary amount, but moves the median by at most one position; the $IQR$ and $\text{MAD}$ inherit the same protection because they are built from medians and quantiles (recall the [$p$-th quantile](01_02_Data.qmd#quantiles) from the previous chapter).
Reach for these robust summaries when the distribution is skewed or contains values you cannot trust to be typical: income, house prices, response times, and similar heavy-tailed quantities where a small fraction of large values would swamp the mean.

```{r}
x1_hat <- USArrests[, 'Murder']
median(x1_hat)
quantile(x1_hat, prob=0.5)
```

:::{.callout-note icon=false collapse="true" title="Must Know"}
Examine robustness to an extreme value
```{r}
x1_extreme_hat <- c(x1_hat, 1000) # add one extreme value
#par(mfrow=c(1, 2)) # visualize side-by-side
#hist(x1_hat)
#hist(x1_extreme_hat)

# Which measures of central tendency are robust
# to a single extreme value?
mean( x1_hat)
mean( x1_extreme_hat )

quantile(x1_hat, prob=0.5)
quantile(x1_extreme_hat, prob=0.5)
```
:::



The $IQR$ corresponds to the box in the boxplot, and the $\text{MAD}$ uses the same robust idea but measures distances from the median directly rather than reading off quartile cut-offs.

```{r}
IQR(x1_hat)
mad(x1_hat)
```

:::{.callout-note icon=false collapse="true" title="Must Know"}
Compute the $IQR$ statistic for the dataset $\{-100,1,4,10,10\}$. 

$IQR =$ Upper quartile $-$ Lower quartile $= 10 - 1 = 9$. 

```{r}
x_hat <- c(-100, 1, 4, 10, 10)
# An intuitive alternative to sd(x_hat), used in the boxplot
quants <- quantile(x_hat, probs=c(.25, .75))
quants[2]-quants[1]
IQR(x_hat)
```
Compute the $\text{MAD}$ statistic for the dataset $\{1,4,10\}$.
First compute the median, $\text{Med}(1,4,10)=4$.
Then compute $\text{Med}( |1-4|,~ |4-4|,~ |10-4| )= \text{Med}( 3, 0, 6 ) = 3$.
```{r}
#Another alternative to sd(x_hat)
mad(x_hat, constant=1)

# Computationally equivalent
# med_hat <- quantile(x_hat, probs=.5)
# absdev_hat <- abs(x_hat -med_hat)
# quantile(absdev_hat, probs=.5)
```
:::

:::{.callout-note icon=false collapse="true" title="Must Know"}
Compare the robustness of various "spread" metrics to an extreme value
```{r}
sd(x1_hat)
sd(x1_extreme_hat)

IQR(x1_hat)
IQR(x1_extreme_hat)

mad(x1_hat, constant=1)
mad(x1_extreme_hat, constant=1)
```
:::


:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
Note that there other "absolute deviation" statistics
```{r, eval=FALSE}
# sometimes seen elsewhere
mean( abs(x1_hat - mean(x1_hat)) )
mean( abs(x1_hat - median(x1_hat)) )
```
:::


#### **Weighted Statistics** {-}

The mean generalizes to a *weighted mean*: an average where different values contribute to the final result with varying levels of importance.
For each outcome $x$ we have a weight $W_{x}$ and compute
$$\begin{aligned}
\hat{m} &= \frac{\sum_{x} x W_{x}}{\sum_{x} W_{x}} = \sum_{x} x w_{x},
\end{aligned}$$
where $w_{x}=\frac{W_{x}}{\sum_{x'} W_{x'}}$ is normalized version of $W_{x}$ that implies $\sum_{x}w_{x}=1$.


:::{.callout-note icon=false collapse="true" title="Must Know"}
For another example, suppose a student has these scores
```{r}
Homework1 <- c(score=88, weight=0.25)
Homework2 <- c(score=92, weight=0.25)
Exam1 <- c(score=67, weight=0.2)
Exam2 <- c(score=90, weight=0.3)

Grades <- rbind(Homework1, Homework2, Exam1, Exam2)
Grades
```

We can compute the final grade as a weighted mean
```{r}
# Manual Way
88*0.25 + 92*0.25 + 67*0.2 + 90*0.3

# Computerized Way
Values <- Grades[, 'score'] * Grades[, 'weight']
FinalGrade <- sum(Values)
FinalGrade
```
:::


When data are discrete, we can also compute the mean using "probability weights" $w_{x} = \hat{p}(x)=\sum_{i=1}^{n}\mathbf{1}\left(\hat{x}_{i}=x\right)/n$.

:::{.callout-note icon=false collapse="true" title="Must Know"}
E.g., the dataset $\{1,2,1,3,1\}$ has $\hat{p}(1)=\frac{3}{5}$, $\hat{p}(2)=\frac{1}{5}$, $\hat{p}(3)=\frac{1}{5}$.
Then, sorting the dataset as \{1,1,1,2,3\}, we can see that $\hat{m} = [1+1+1+2+3]/5 = [1+1+1]/5 + 2/5 + 3/5 = 1 \frac{3}{5} + 2 \frac{1}{5} + 3 \frac{1}{5} = 1 \hat{p}(1) + 2 \hat{p}(2) + 3 \hat{p}(3)$.
In either case, we end up with $\frac{3+2+3}{5}=8/5=1.6$

```{r}
x_hat <- c(1, 2, 1, 3, 1)
mean(x_hat)

proportions <- table(x_hat)/length(x_hat)
proportions
vals <- sort(unique(x_hat))
sum(vals*proportions)
```
:::


In principle, we can also compute other weighted statistics such as a weighted median.

:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
See that we can also compute weighted quantiles
```{r}
weighted.quantile <- function(x_hat, w, probs){
    #See spatstat.univar::weighted.quantile
    oo <- order(x_hat)
    x_hat <- x_hat[oo]
    w <- w[oo]
    F_hat <- cumsum(w)/sum(w)
    quantile_id <- max(which(F_hat <= probs))+1
    q_hat <- x_hat[quantile_id]
    return(q_hat)
}

## Unweighted
quantile(x_hat, probs=.5)
weights <- rep(1, length(x_hat))
weighted.quantile(x_hat=x_hat, w=weights, probs=.5)

## Weighted
weights <- seq(x_hat)
weights <- weights/sum(weights) # normalize
weighted.quantile(x_hat=x_hat, w=weights, probs=.5)
```
:::

#### **Mode and Share Concentration** {-}
Sometimes, none of the above work well.
With categorical data, for example, distributions are easier to describe with other statistics.
The sample mode is the most common observation: the value with the highest observed frequency.
We can also measure the spread of the frequencies or concentration at the mode vs elsewhere.

:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
Here is an example of the mode.
In the data below, $A$ appears three times, $C$ twice, and $B$ once, so the mode is $A$.
Notice the mode is not the alphabetically middle letter, $B$.
```{r}
x_hat <- c('A', 'B', 'A', 'C', 'C', 'A')
proportions <- table(x_hat)/length(x_hat)
plot(proportions, col=grey(0, 0.5))

# mode(s)
mode_id <- which(proportions==max(proportions))
names(proportions)[ mode_id ]
```

Here is an example of spread for categorical data.
```{r}
# freq. spreads
sd(proportions)
sum(proportions^2)

# freq. concentration at mode
max(proportions)/mean(proportions)
```
:::

## Shape Statistics

Central tendency and dispersion are often insufficient to describe a distribution.
To further describe shape, we can compute sample skew and kurtosis to measure asymmetry and extreme values. 

There are many other statistics we could compute on an ad-hoc basis.
However, shape is often best understood with graphical descriptions: histogram, ECDF, Boxplot.
These should be made before numerical descriptions: skewness and kurtosis statistics.

#### **Skewness** {-}

Center and spread alone leave open whether the distribution is symmetric; the first shape question asks whether one tail extends further than the other.

:::{.callout-important icon=false title="Key Definition" collapse="false"}
The sample *skewness* measures asymmetry: the average cubed deviation from the mean, scaled by the cube of the standard deviation,
$$\text{skew} = \frac{\sum_{i=1}^{n}(\hat{x}_{i}-\hat{m})^3 / n}{\hat{s}^3}.$$
Positive skew means a long right tail, negative skew a long left tail, and zero skew (e.g., the Normal) is symmetric.
:::

Cubing the deviations preserves their sign: positive deviations stay positive and negative deviations stay negative.
The numerator is therefore close to zero when the data are roughly balanced around the mean, and large in absolute value when one tail is much longer than the other.
Dividing by the cubed standard deviation makes skewness dimensionless, so the values are comparable across datasets measured in different units.
Skewness is useful for flagging whether the mean and median will agree closely (low skew) or diverge sharply (high skew), and for deciding whether a log or square-root transformation might tame the distribution before further analysis.

:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
Consider the right-skewed dataset $\{0,1,2,3,9\}$ with mean $\hat{m}=15/5=3$.
The cubed deviations from the mean are $(-3)^3, (-2)^3, (-1)^3, 0^3, 6^3$, which equal $-27, -8, -1, 0, 216$.
A cube keeps the sign of its deviation, so the single large positive deviation outweighs the small negative ones.
Their average is the numerator: $\sum_{i=1}^{n}[\hat{x}_{i}-\hat{m}]^3/n = 180/5 = 36$.
Dividing by $\hat{s}^3 \approx 44.2$ gives a skewness of $36/44.2 \approx 0.81$.
The positive value reflects the long right tail created by the value $9$.
```{r}
x_demo_hat <- c(0, 1, 2, 3, 9)
m3_hat <- mean( (x_demo_hat - mean(x_demo_hat))^3 ) # numerator
m3_hat
m3_hat / sd(x_demo_hat)^3 # skewness
```
:::

```{r}
x_hat <- USArrests[, 'Murder']
hist( x_hat^2, border=NA,
	xlab='[Murder Rate]^2',
	main=NA, freq=FALSE, breaks=20)

skewness <-  function(x_hat){
 m_hat <- mean(x_hat)
 m3_hat <- mean((x_hat - m_hat)^3)
 s3_hat <- sd(x_hat)^3
 skew <- m3_hat/s3_hat
 return(skew)
}

skewness( x_hat^2 )
skewness( x_hat^3 )
```

We can automatically compare against the normal distribution, which has a skew of 0.

#### **Kurtosis** {-}

A second shape question, independent of asymmetry, asks how much of the variability comes from a few far-from-center observations rather than from the typical bulk.

:::{.callout-important icon=false title="Key Definition" collapse="false"}
The sample *kurtosis* measures how heavy the tails are: the average fourth power of deviations from the mean, scaled by $\hat{s}^4$,
$$\text{kurt} = \frac{\sum_{i=1}^{n}(\hat{x}_{i}-\hat{m})^4 / n}{\hat{s}^4}.$$
The Normal distribution has kurtosis $3$, so subtracting $3$ gives a comparison-to-Normal version (excess kurtosis).
:::

Where skewness uses the cube (and so can be negative), kurtosis uses the fourth power, which forces every term to be positive and disproportionately amplifies the largest deviations.
A handful of far-from-the-mean values can drive kurtosis up substantially even when the bulk of the distribution looks unremarkable, which is why it is often described as a "tail-heaviness" measure.
Kurtosis is useful for distinguishing distributions that look similar in their bulk but differ in how often extreme values occur: daily stock returns, for example, have moderate variance but kurtosis well above $3$, reflecting occasional crashes that a Normal model would treat as vanishingly rare.

:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
Using the same dataset $\{0,1,2,3,9\}$ with mean $\hat{m}=3$, the *quartic* deviations are $(-3)^4, (-2)^4, (-1)^4, 0^4, 6^4$, which equal $81, 16, 1, 0, 1296$.
Because the power is even, every term is positive, and the large deviation from the value $9$ dominates the sum.
Their average is the numerator: $\sum_{i=1}^{n}[\hat{x}_{i}-\hat{m}]^4/n = 1394/5 = 278.8$.
Dividing by $\hat{s}^4 = 156.25$ gives a kurtosis of $278.8/156.25 \approx 1.78$.
```{r}
x_demo_hat <- c(0, 1, 2, 3, 9)
m4_hat <- mean( (x_demo_hat - mean(x_demo_hat))^4 ) # numerator
m4_hat
m4_hat / sd(x_demo_hat)^4 # kurtosis
```
:::

Some authors further subtract $3$ to explicitly compare against the normal distribution (the normal distribution has a *kurtosis* of $3$).

Boxplot whiskers are a great way to examine kurtosis, with more circle ``outliers'' indicating more kurtosis.
You can also see skew in the boxplot when one quartile is much further from the median than the other. 

```{r}
x1_hat <- x_hat^1/mean(x_hat^1)
x2_hat <- x_hat^2/mean(x_hat^2)
x3_hat <- x_hat^3/mean(x_hat^3)
x4_hat <- x_hat^4/mean(x_hat^4)
boxplot(x1_hat, x2_hat, x3_hat, x4_hat,
	names=c(1, 2, 3, 4), xlab='Data Transformations',
	main=NA)


kurtosis <- function(x_hat){
 m_hat <- mean(x_hat)
 m4_hat <- mean((x_hat - m_hat)^4)
 s4_hat <- sd(x_hat)^4
 kurt <- m4_hat/s4_hat
 return(kurt)
 # use instead to compare against normal
 # excess_kurt <- kurt - 3 
}

kurtosis( x1_hat )
kurtosis( x2_hat )
kurtosis( x3_hat )
kurtosis( x4_hat )
```


#### **Clusters/Gaps** {-}
You can also describe distributions in terms of how clustered the values are, including the number of modes (peaks), bunching, and many other statistics.
A *gap* is a range with few or no observations, and a *cluster* is a group of values bunched together.
A distribution with one peak is *unimodal*, and one with two peaks is *bimodal*.
The two histograms below are each bimodal: two clusters of values separated by a gap.
The mean and variance alone would miss this structure, since two very different distributions can share the same mean and variance.
So remember that "a picture is worth a thousand words": plot your data before summarizing it numerically.

```{r, echo=FALSE}
# Complicated Random Variable
r_ugly0 <- function(n, population_means=c(-2, 3), population_vars=c(1, 4), prob=c(.5, .5)){
  # Which distribution is each observation coming from
  di <- sample(c(1, 2), size=n, replace=TRUE, prob=prob)
  rnorm(n, mean=population_means[di], sd=sqrt(population_vars)[di])
}

x_hat <- r_ugly0(6000)
hist(x_hat, breaks=60,
    freq=FALSE, border=NA,
    xlab='x', main=NA)

#hist( rbeta(1000, .6, .6), border=NA, main=NA, freq=F, breaks=20)
```

```{r, echo=FALSE}
# Another Complicated Random Variable
r_ugly1 <- function(n, theta1=c(-8, -1), theta2=c(-2, 2), rho=.25){
    omega   <- rbinom(n, size=1, rho)
    x_hat <- omega * runif(n, theta1[1], theta2[1]) +
        (1-omega) * rnorm(n, theta1[2], theta2[2])
    return(x_hat)
}
# Very Large Sample
x_hat <- r_ugly1(1000000)
hist(x_hat, breaks=1000, freq=FALSE, border=NA,
    xlab='x', main=NA)
```



## Exercises

1. Comment the script you wrote for this chapter, then restart R and check that it runs from a clean session, then check the script with AI as explained in [Working with AI](04_A_WorkingWithAI.qmd).
   Write three sentences from memory on the main statistical idea of this chapter, and ask the assistant what is wrong, vague, or missing.
   Finish with your own questions about whatever you found hardest.

2. Explain in your own words why the median is more robust to extreme values than the mean.
   Give a concrete example with a small dataset where adding one outlier changes the mean substantially but barely moves the median.

3. Using the dataset $\{2, 5, 5, 8, 12\}$, compute by hand: (a) the sample mean $\hat{m}$, (b) the sample variance $\hat{v}$, (c) the sample standard deviation $\hat{s}$, and (d) the $IQR$.
   Show your work for each step.

4. Load `USArrests` in R.
   Compute the mean, standard deviation, and MAD (with `constant=1`) of the `Assault` variable.
   Then apply a log transformation with `log(USArrests[,'Assault'])` and compute the skewness of both the original and transformed data using the `skewness` function defined in the chapter.
   Which version is less skewed?

#### **Further Reading** {-}

* <https://learningstatisticswithr.com/book/descriptives.html> -- Thorough treatment of measures of center, spread, and skewness with R examples.
* <https://www.khanacademy.org/math/statistics-probability/summarizing-quantitative-data> -- Short video lessons on mean, median, variance, and standard deviation.
* <https://www.econometrics-with-r.org/2.1-random-variables-and-probability-distributions.html> -- Descriptive statistics in the context of econometric data analysis.

#### **Recall** {-}

This chapter compressed a distribution into single numbers: the *mean* and *variance* (and *standard deviation*) for center and spread, the *median*, *IQR*, and *MAD* as robust alternatives, and the *skewness* and *kurtosis* for shape.
The running five-point dataset $\{0, 1, 2, 3, 9\}$ tied the shape statistics together: its mean is $3$, its cubed deviations average to $36$ giving a skewness of about $0.81$ (long right tail), and its quartic deviations give a kurtosis of about $1.78$ (both driven by the single outlier at $9$).
On real data such as `USArrests[, 'Murder']`, a histogram plus the mean and standard deviation already reveal most of the shape; the median, MAD, skew, and kurtosis fill in what symmetry assumptions hide.
In the next chapter we flip the script: instead of computing statistics from observed data, we generate data from a known random-variable model.
