# Advanced Probability

## Discrete Factorial Distributions

#### **Binomial**. {-}

The sum of $n$ Bernoulli's (number of successes)

\begin{eqnarray}
X_{i} \in \{0,1,\ldots,n\} \\
Prob(X_{i}=x) &=& \binom{n}{x}p^{x}(1-p)^{n-x}
\end{eqnarray}

The simplest case: how many heads will I get when I flip a coin twice? We can first see that there are four potential outcomes: $\{ (1,1), (1,0), (0,1), (0,0) \}$, each of which are equally likely (assuming the coin is fair). The corresponding sums are $\{2,1,1,0 \}$, which leads to probabilities $Prob(X_{i}=0)=1/4$, $Prob(X_{i}=1)=2/4$, and $Prob(X_{i}=2)=1/4$

:::{.callout-tip icon=false collapse="true"}
When I flip a coin three times, the potential outcomes are 
\begin{eqnarray}
\{
(1,1,1), (1,1,0), (1,0,1), (1,0,0),\\
(0,1,1), (0,1,0), (0,0,1), (0,0,0) 
\}
\end{eqnarray}
Assuming the coin is fair, compute the probabilities for each possible sum.
:::

In general, we use the [Binomial_distribution](https://en.wikipedia.org/wiki/Binomial_distribution).


```{r}
## Minimal example
n <- 15
p <- 0.3

# PMF plot
x <- seq(0, n)
f_x <- dbinom(x, n, p)
plot(x, f_x, type = "h", col = "blue",
    main='',  xlab = "x", ylab = "Prob(X = x)")
title(bquote(paste('Binom(',.(n),', ',.(p), ')' ) ))
points(x, f_x, pch = 16, col = "blue")

## Simulation
X <- rbinom(1000, n, p)
plot(ecdf(X))
```

:::{.callout-tip icon=false collapse="true"}
Suppose that employees at a company are $70%$ female and $30%$ male. If we select a random sample of eight employees, what is the probability that $2$ in the sample are female? That $\geq 2$ are female? 

Try conducting a simulation first, 
```
n <- 8
p <- 0.3
X <- rbinom(1000, n, p)
```
:::


:::{.callout-tip icon=false collapse="true"}
For $n$ coin flips, show that $\mathbb{E}[X_{i}]=np$ and $\mathbb{V}[X_{i}]=np(1-p)$. Build on the case $n=2$ above, then consider case $n=3$, and then the general case of any $n$.

```{r}
# Simulate:  Compare empirical vs theoretical
x <- c(0,1)
x_probs <- c(1/4, 3/4)
X <- sample(x, 1000, prob=x_probs, replace=T)
#c(emp_mean = mean(X), th_mean = n*p)
#c(emp_var  = var(X),  th_var  = n*p*(1-p))
```
:::

The Binomial Limit Theorem (de Moivre–Laplace theorem) says that as $n$ grows large, with $p \in (0,1)$ staying fixed, the Binomial distribution is approximately normal with mean $np$ and variance $np(1-p)$

:::{.callout-tip icon=false collapse="true"}
The unemployment rate is $10%$. Suppose that $100$ employable people are selected randomly. What is the probability that this sample contains between $9$ and $12$ unemployed people. Use the normal approximation to binomial probabilities.

Parameters are 
$\mu \approx n p = 10$ 
$\sigma^2= n p (1−p) = 100 (0.1)(0.9)=9$ 
$\sigma=\sqrt{9}=3$.
:::



#### **Poisson**. {-}

The number of events in a fixed interval

* Discrete, support $\{0,1,2,\ldots\}$
* Probability Mass Function: $Prob(X_{i}=x)=e^{-\lambda}\lambda^x/x!$
* See <https://en.wikipedia.org/wiki/Poisson_distribution>
* A common use case: How many cars will show up in the lot tomorrow?

```{r}
## Minimal example
lambda <- 3.5

# PMF plot
x <- seq(0, 15)
f_x <- dpois(x, lambda)
plot(x, f_x, type="h", col="blue",
     xlab = "x", ylab = "Prob(X = x)")
points(x, f_x, pch = 16, col = "blue")
title(bquote(paste('Pois(',.(lambda), ')')))


# Simulate: Compare empirical vs theoretical
#X <- rpois(1e4, lambda)
#c(emp_mean = mean(X), th_mean = lambda)
#c(emp_var  = var(X),  th_var  = lambda)
```

:::{.callout-tip icon=false collapse="true"}
Show that $\mathbb{E}[X_{i}] = \mathbb{V}[X_{i}]= \lambda$.
:::


## Continuous Factorial Distributions

#### **Irwin–Hall**. {-}

The sum of $n$ i.i.d. $\text{Uniform}(0,1)$.  

* Continuous, support $[0,n]$
* Probability Density Function: $f(x) = \dfrac{1}{(n-1)!}
\displaystyle\sum_{k=0}^{\lfloor x \rfloor} (-1)^k
\binom{n}{k} (x - k)^{n-1}$ for $x \in [0,n]$, and $0$ otherwise
* See <https://en.wikipedia.org/wiki/Irwin%E2%80%93Hall_distribution>
* A common use case: representing the sum of many small independent financial shocks

```{r}
## Minimal example (base R)
# Irwin–Hall PDF function
dirwinhall <- function(x, n) {
    f_x <- vector(length=length(x))
    for(i in seq(x)) {
        xx <- x[i]
        if(xx < 0 | xx > n){
            f_x[i] <- 0 
        } else {
            k <- seq(0, floor(xx))
            f_k <- sum((-1)^k*choose(n, k)*(xx-k)^(n-1))/factorial(n-1)
            f_x[i] <- f_k
        }
    }
    return(f_x)
}

# Parameters
n <- 2
x <- seq(0, n, length.out = 500)

# Compute and plot PDF
f_x <- dirwinhall(x, n)
plot(x, f_x, type="l", col="blue",
    main='', xlab = "x", ylab = "f(x)")
title(bquote(paste('IrwinHall(',.(n), ')')))
```

See also the [Bates](https://en.wikipedia.org/wiki/Bates_distribution) distribution, which is for the mean  of $n$ i.i.d. $\text{Uniform}(0,1)$ random variables. It essentially rescales the Irwin-Hall distribution to range between $0$ and $1$.


#### **Beta**. {-}
The sample space is any number on the unit interval, $X_{i} \in [0,1]$, but with non-uniform probabilities. 


```{r}
X4 <- rbeta(2000,2,2) ## two shape parameters
hist(X4, breaks=20, border=NA, main=NA, freq=F)

#See the underlying probabilities
#f_25 <- dbeta(.25, 2, 2)

x <- seq(0,1,by=.01)
fx <- dbeta(x, 2, 2)
lines(x, fx, col='blue')
```

The [Beta](https://en.wikipedia.org/wiki/Beta_distribution) distribution is mathematically complicated to write, and so we omit it. However, we can find the probability graphically using either the probability density function or cumulative distribution function. 

:::{.callout-tip icon=false collapse="true"}
Suppose $X_{i}$ is a random variable with a beta distribution. Intuitively depict $Prob(X_{i} \in [0.2, 0.8])$ by drawing an area under the density function. Numerically estimate that same probability using the CDF.
```{r}
plot( ecdf(X4), main=NA) # Empirical

x <- seq(0,1,by=.01) # Theoretical
Fx <- pbeta(x, 2, 2)
lines(x, Fx, col='blue')

# Middle Interval Example 
F2 <- pbeta(0.2, 2, 2)
F8 <- pbeta(0.8, 2, 2)
F_2_8 <- F8 - F2
F_2_8

# Visualize
title('Middle between 0.2 and 0.8')
segments( 0.2, F2, -1, F2, col='red')
segments( 0.8, F8, -1, F8, col='red')
```
:::

This distribution is often used, as the probability density function has two parameters that allow it to take many different shapes.

:::{.callout-tip icon=false collapse="true"}
For each example below, intuitively depict $Prob(X_{i} \leq 0.5)$ using the PDF. Repeat the exercise using a CDF instead of a PDF to calculate a numerical value.
```{r, results='hide'}
#| fig-width: 8
#| fig-height: 6
op <- par(no.readonly = TRUE); on.exit(par(op), add = TRUE)
x <- seq(0,1,by=.01)
pars <- expand.grid( c(.5,1,2), c(.5,1,2) )
par(mfrow=c(3,3))
apply(pars, 1, function(p){
    fx <- dbeta( x,p[1], p[2])
    plot(x, fx, type='l', xlim=c(0,1), ylim=c(0,4), lwd=2, col='blue')
    #hist(rbeta(2000, p[1], p[2]), breaks=50, border=NA, main=NA, freq=F)
})
title('Beta densities', outer=T, line=-1)
```
:::



## Set Theory

Let the sample space contain events $A$ and $B$, with their probability denoted as $Prob(A)$ and $Prob(B)$. Then

- the *union*: $A\cup B$, refers to either event occurring
- the *intersection*: $A\cap B$, refers to both events occurring
- the events $A$ and $B$ are *mutually exclusive* if and only if $A\cap B$ is empty
- the *inclusion–exclusion* rule is: $Prob(A\cup B)=Prob(A)+Prob(B)-Prob(A\cap B)$
- the events $A$ and $B$ are *mutually independent* if and only if $Prob(A\cap B)=Prob(A) Prob(B)$
- the *conditional probability* is $Prob(A \mid B) = Prob(A \cap B)/ P(B)$

For example, consider a six sided die where events are the number $>2$ or the number is even.

- The sets are $A=\{3,4,5,6\}$ and $B=\{2,4,6\}$
- To check mutually exclusive: notice $A\cap B=\{4,6\}$ which is not empty. So event $A$ and event $B$ are not mutually exclusive

Further assuming the die are fair, each side has an equal $1/6$ probability. This lets us compute

- $Prob(A)=4/6=2/3$ and $Prob(B)=3/6=1/2$
- $A\cup B=\{2,3,4,5,6\}$, which implies $Prob(A\cup B)=5/6$
- $A\cap B=\{4,6\}$, which implies $Prob(A\cap B)=2/6=1/3$
- To check independence: notice $Prob(A\cap B)=1/3 = (2/3) (1/2) = Prob(A)Prob(B)$. So event $A$ and event $B$ are mutually independent
- To check inclusion–exclusion: notice $Prob(A) + Prob(B) - Prob(A\cap B) = 2/3 + 1/2 - 1/3 = 4/6 + 3/6 - 2/6 = 5/6 = Prob(A\cup B)$
- Finally, the conditional probability of $Prob(\text{die is} >2 \mid \text{die shows even number}) = Prob(A \mid B) = \frac{Prob(A \cap B)}{P(B)} = \frac{1/3}{1/2} = 2/3$

```{r}
# Simulation verification
set.seed(123)
x <- seq(1,6)
rolls <- sample(x, 1e6, replace = TRUE)
A <- rolls > 2
B <- rolls %% 2 == 0
P_AcondB <- mean(A & B) / mean(B)
round(P_AcondB, 3)
```

:::{.callout-tip icon=false collapse="true"}
Consider a six sided die with sample space $\{1,2,3,4,5,6\}$ where each outcome is each equally likely. What is $Prob(\text{die is odd or }<5)$?
Denote the odd events as $A=\{1,3,5\}$ and the less than five events as  $B=\{1,2,3,4\}$. Then

- $A\cap B=\{1,3\}$, which implies $Prob(A\cap B)=2/6=1/3$.
- $Prob(A)=3/6=1/2$ and $Prob(B)=4/6=2/3$.
- By inclusion–exclusion: $Prob(A\cup B)=Prob(A)+Prob(B)-Prob(A\cap B)=1/2+2/3-1/3=5/6$.

Now find $Prob(\text{die is even and }<5)$. Verify your answer with a computer simulation.
:::



#### **Law of Total Probability**. {-}

This law states that $Prob(A)=Prob(A \mid B) P(B) + Prob(A \mid \lnot B) P(\lnot B)$, where $\lnot B$ is the event "not $B$".

For example, consider a six sided die where events are the number $>2$ or the number is even.
- The sets are $A=\{3,4,5,6\}$ and $B=\{2,4,6\}$ and $\lnot B = \{1,3,5\}$ 
From before, we know that $Prob(A)=2/3$, $Prob(B)=1/2$, and $Prob(A \mid B) = 2/3$.
Then see that $Prob(A \mid B) Prob(B) = \frac{2}{3} \frac{1}{2} = 2/6$ 

We can also directly compute $Prob(\lnot B)=1/2$, or infer it from $Prob(\lnot B)=1-Prob(B)$.
The conditional probability of $Prob(\text{die is} >2 \mid \text{die shows odd number}) = Prob(A \mid \lnot B) = \frac{Prob(A \cap \lnot B)}{P(\lnot B)} = 2/3$.

Then we can see that
$Prob(A \mid B) Prob(B) + Prob(A \mid \lnot B) Prob(\lnot B) = 2/6 + 2/6 = 2/3 = Prob(A)$.

