# Misc. Univariate Topics



## Data Transformations

Transformations can stabilize variance, reduce skewness, and make model errors closer to Gaussian.

Perhaps the most common examples are *power transformations*: $y= x^\lambda$, which includes $\sqrt{x}$ and $x^2$.

Other examples include the *exponential transformation*: $y=\exp(x)$ for any $x\in (-\infty, \infty)$ and *logarithmic transformation*: $y=\log(x)$ for any $x>0$.

The *Box–Cox Transform* nests many cases. For $x>0$ and parameter $\lambda$,
\begin{eqnarray}
y=\begin{cases}
\dfrac{x^\lambda-1}{\lambda}, & \lambda\neq 0,\\
\log\left(x\right) & \lambda=0.
\end{cases}
\end{eqnarray}
This function is continuous over $\lambda$.

```{r}
# Box–Cox transform and inverse
bc_transform <- function(x, lambda) {
  if (any(x <= 0)) stop("Box-Cox requires x > 0")
  if (abs(lambda) < 1e-8) log(x) else (x^lambda - 1)/lambda
}
bc_inverse <- function(t, lambda) {
  if (abs(lambda) < 1e-8) exp(t) else (lambda*t + 1)^(1/lambda)
}

X <- USArrests[,'Murder']
hist(X, main='', border=NA, freq=F)

par(mfrow=c(1,3))
for(lambda in c(-1,0,1)){
    Y <- bc_transform(X, lambda)
    hist(Y, 
        main=bquote(paste(lambda,'=',.(lambda))),
        border=NA, freq=F)
}
```


#### **Law of the Unconscious Statistician (LOTUS)**. {-}

As before, we will represent the random variable as $X_{i}$, which can take on values $x$ from the sample space. If $X_{i}$ is a discrete random variable (a random variable with a discrete sample space) and $g$ is a function, then
$\mathbb E[g(X_{i})] = \sum_x g(x)Prob(X_{i}=x)$.

:::{.callout-tip icon=false collapse="true"}
Let $X_{i}$ take values $\{1,2,3\}$ with
\begin{eqnarray}
Pr(X_{i}=1)=0.2,\quad Prob(X_{i}=2)=0.5,\quad Prob(X_{i}=3)=0.3.
\end{eqnarray}
Let $g(x)=x^2+1$. Then $g(1)=1^2+1=2$, $g(2)=2^2+1=5$, $g(3)=3^2+1=10$. 

Then, by LOTUS,
\begin{eqnarray}
\mathbb E[g(X_{i})]=\sum_x g(x)Prob(X_{i}=x)
&=& g(1)\cdot 0.2 + g(2)\cdot 0.5 + g(3)\cdot 0.3 \\
&=& 2 \cdot 0.2 + 5 \cdot 0.5 + 10 \cdot 0.3 \\
&=& 0.4 + 2.5 + 3 = 5.9.
\end{eqnarray}
```{r}
x  <- c(1,2,3)
x_probs <- c(0.2,0.5,0.3)
g  <- function(x) x^2 + 1
sum(g(x) * x_probs) 
```
:::

```{r}
g  <- function(x) x^2 + 1

# A theoretical example
x <- c(1,2,3,4)
x_probs <- c(1/4, 1/4, 1/4, 1/4)
sum(g(x) * x_probs) 

# A simulation example
X <- sample(x, prob=x_probs, size=1000, replace=T)
mean(g(X))
```

:::{.callout-tip icon=false collapse="true"}
If $X_{i}$ is a continuous random variable (a random variable with a continuous sample space) and $g$ is a function, then
$\mathbb E[g(X_{i})] = \int_{-\infty}^{\infty} g(x)f(x) dx$.
```{r}
x <- rexp(5e5, rate = 1)           # X ~ Exp(1)
mean(sqrt(x))                      # LOTUS Simulation
sqrt(pi) / 2                       # Exact via LOTUS integral
```
:::

Note that you have already seen the special case where $g(X_{i})=\left(X_{i}-\mathbb{E}[X_{i}]\right)^2$.

#### **Jensen’s inequality**. {-}

Concave functions curve inwards, like the inside of a cave.
Convex functions curve outward, the opposite of concave.

If $g$ is a *concave* function, then $g(\mathbb E[X_{i}]) \geq \mathbb E[g(X_{i})]$.
```{r}
# Continuous Example 1
mean( sqrt(x) )
sqrt( mean(x) ) 

# Continuous Example 2
mean( log(x) )
log( mean(x) ) 
```

```{r}
## Discrete Example
x  <- c(1,2,3)
px <- c(0.2,0.5,0.3)
EX <- sum(x * px)
EX

g  <- sqrt
gEX <- g(EX)
EgX <- sum(g(x) * px)
c(gEX, EgX)
```


If $g$ is a *convex* function, then the inequality reverses: $g(\mathbb E[X_{i}]) \leq \mathbb E[g(X_{i})]$.
```{r}
mean( exp(x) )
exp( mean(x) )  
```

## Drawing Samples

To generate a random variable from known distributions, you can use some type of physical machine. E.g., you can roll a fair die to generate Discrete Uniform data or you can roll weighted die to generate Categorical data.

There are also several ways to computationally generate random variables from a probability distribution. Perhaps the most common one is "inverse sampling". To generate a random variable using inverse sampling, first sample $p$ from a uniform distribution and then find the associated quantile  quantile function $\hat{F}^{-1}(p)$.^[Drawing random uniform samples with computers is actually quite complex and beyond the scope of this course.]


#### **Using Data**. {-}

You can generate a random variable from a known empirical distribution. Inverse sampling randomly selects observations from the dataset with equal probabilities. To implement this, we 

* order the data and associate each observation with an ECDF value
* draw $p \in [0,1]$ as a uniform random variable
* find the associated quantile via the ECDF

:::{.callout-tip icon=false collapse="true"}
Here is an example of generating random murder rates for US states.
```{r}
# Empirical Distribution
X <- USArrests[,'Murder']
FX_hat <- ecdf(X)
plot(FX_hat, lwd=2, xlim=c(0,20),
    pch=16, col=grey(0,.5), main='')

# Generating random variables via inverse ECDF
p <- runif(3000) ## Multiple Draws
QX_hat <- quantile(FX_hat, p, type=1)
QX_hat[c(1,2,3)]

## Can also do directly from the data
QX_hat <- quantile(X, p, type=1)
QX_hat[c(1,2,3)]
```
:::

#### **Using Math**. {-}

If you know the distribution function that generates the data, then you can derive the quantile function and do inverse sampling. 
That is how computers generate random data from a distribution.
```{r}
# 4 random data points from 3 different distributions
qunif(4)
qexp(4)
qnorm(4)
```

Here is an in-depth example of the [Dagum distribution](https://en.wikipedia.org/wiki/Dagum_distribution). The distribution function is $F(x)=(1+(x/b)^{-a})^{-c}$. For a given probability $p$, we can then solve for the quantile as $F^{-1}(p)=\frac{ b p^{\frac{1}{ac}} }{(1-p^{1/c})^{1/a}}$. Afterwhich, we sample $p$ from a uniform distribution and then find the associated quantile.

```{r}
# Theoretical Quantile Function (from VGAM::qdagum)
qdagum <- function(p, scale.b=1, shape1.a, shape2.c) {
  # Quantile function (theoretically derived from the CDF)
  ans <- scale.b * (expm1(-log(p) / shape2.c))^(-1 / shape1.a)
  # Special known cases
  ans[p == 0] <- 0
  ans[p == 1] <- Inf
  # Safety Checks
  ans[p < 0] <- NaN
  ans[p > 1] <- NaN
  if(scale.b <= 0 | shape1.a <= 0 | shape2.c <= 0){ ans <- ans*NaN }
  # Return
  return(ans)
}

# Generate Random Variables (VGAM::rdagum)
rdagum <-function(n, scale.b=1, shape1.a, shape2.c){
    p <- runif(n) # generate random probabilities
    x <- qdagum(p, scale.b=scale.b, shape1.a=shape1.a, shape2.c=shape2.c) #find the inverses
    return(x)
}

# Example
set.seed(123)
X <- rdagum(3000,1,3,1)
X[c(1,2,3)]
```


## Set Theory

Let the sample space contain events $A$ and $B$, with their probability denoted as $Prob(A)$ and $Prob(B)$. Then

- the *union*: $A\cup B$, refers to either event occurring
- the *intersection*: $A\cap B$, refers to both events occurring
- the events $A$ and $B$ are *mutually exclusive* if and only if $A\cap B$ is empty
- the *inclusion–exclusion* rule is: $Prob(A\cup B)=Prob(A)+Prob(B)-Prob(A\cap B)$
- the events $A$ and $B$ are *mutually independent* if and only if $Prob(A\cap B)=Prob(A) Prob(B)$
- the *conditional probability* is $Prob(A \mid B) = Prob(A \cap B)/ P(B)$

For example, consider a six sided die where events are the number $>2$ or the number is even.

- The sets are $A=\{3,4,5,6\}$ and $B=\{2,4,6\}$
- To check mutually exclusive: notice $A\cap B=\{4,6\}$ which is not empty. So event $A$ and event $B$ are not mutually exclusive

Further assuming the die are fair, each side has an equal $1/6$ probability. This lets us compute

- $Prob(A)=4/6=2/3$ and $Prob(B)=3/6=1/2$
- $A\cup B=\{2,3,4,5,6\}$, which implies $Prob(A\cup B)=5/6$
- $A\cap B=\{4,6\}$, which implies $Prob(A\cap B)=2/6=1/3$
- To check independence: notice $Prob(A\cap B)=1/3 = (2/3) (1/2) = Prob(A)Prob(B)$. So event $A$ and event $B$ are mutually independent
- To check inclusion–exclusion: notice $Prob(A) + Prob(B) - Prob(A\cap B) = 2/3 + 1/2 - 1/3 = 4/6 + 3/6 - 2/6 = 5/6 = Prob(A\cup B)$
- Finally, the conditional probability of $Prob(\text{die is} >2 \mid \text{die shows even number}) = Prob(A \mid B) = \frac{Prob(A \cap B)}{P(B)} = \frac{1/3}{1/2} = 2/3$

```{r}
# Simulation verification
set.seed(123)
x <- seq(1,6)
rolls <- sample(x, 1e6, replace = TRUE)
A <- rolls > 2
B <- rolls %% 2 == 0
P_AcondB <- mean(A & B) / mean(B)
round(P_AcondB, 3)
```

:::{.callout-tip icon=false collapse="true"}
Consider a six sided die with sample space $\{1,2,3,4,5,6\}$ where each outcome is each equally likely. What is $Prob(\text{die is odd or }<5)$?
Denote the odd events as $A=\{1,3,5\}$ and the less than five events as  $B=\{1,2,3,4\}$. Then

- $A\cup B=\{1,3\}$, which implies $Prob(A\cap B)=2/6=1/3$.
- $Prob(A)=3/6=1/2$ and $Prob(B)=4/6=2/3$.
- By inclusion–exclusion: $Prob(A\cup B)=Prob(A)+Prob(B)-Prob(A\cap B)=1/2+2/3-1/3=5/6$.

Now find $Prob(\text{die is even and }<5)$. Verify your answer with a computer simulation.
:::

#### ***Law of Total Probability*. {-}

This law states that $Prob(A)=Prob(A \mid B) P(B) + Prob(A \mid \not B) P(\not B)$, where $\not B$ is the event "not $B$".

For example, consider a six sided die where events are the number $>2$ or the number is even.
- The sets are $A=\{3,4,5,6\}$ and $B=\{2,4,6\}$ and $\not B = \{1,3,5\}$ 
From before, we know that $Prob(A)=2/3$, $Prob(B)=1/2$, and $Prob(A \mid B) = 2/3$.
Then see that $Prob(A \mid B) Prob(B) = \frac{2}{3} \frac{1}{2} = 2/6$ 

We can also directly compute $Prob(\not B)=1/2$, or infer it from $Prob(\not B)=1-Prob(B)$.
The conditional probability of $Prob(\text{die is} >2 \mid \text{die shows odd number}) = Prob(A \mid \not B) = \frac{Prob(A \cap \not B)}{P(\not B)} = 2/3$.

Then we can see that
$Prob(A \mid B) Prob(B) + Prob(A \mid \not B) Prob(\not B) = 2/6 + 2/6 = 2/3 = Prob(A)$.


## Factorial Distributions

#### **Binomial**. {-}

The sum of $n$ Bernoulli trials (number of successes)

* Discrete, support $\{0,1,\ldots,n\}$
* Probability Mass Function: $Prob(X_{i}=x)=\binom{n}{x}p^{x}(1-p)^{n-x}$
* See <https://en.wikipedia.org/wiki/Binomial_distribution>
* A common use case: How many heads will I get when I flip a coin twice?

```{r}
## Minimal example
n <- 15
p <- 0.3

# PMF plot
x <- seq(0, n)
f_x <- dbinom(x, n, p)
plot(x, f_x, type = "h", col = "blue",
    main='',  xlab = "x", ylab = "Prob(X = x)")
title(bquote(paste('Binom(',.(n),', ',.(p), ')' ) ))
points(x, f_x, pch = 16, col = "blue")

# Simulate:  Compare empirical vs theoretical
#X <- rbinom(1e4, size = n, prob = p)
#c(emp_mean = mean(X), th_mean = n*p)
#c(emp_var  = var(X),  th_var  = n*p*(1-p))
```

:::{.callout-tip icon=false collapse="true"}
Suppose that employees at a company are $70%$ female and $30%$ male. If we select a random sample of eight employees, what is the probability that more than $2$ in the sample are female?
:::

:::{.callout-tip icon=false collapse="true"}
Show that 
$\mathbb{E}[X_{i}]=np$ and $\mathbb{V}[X_{i}]=np(1-p)$. Start with the case $n=1$, then the case $n=2$, then case $n=3$, then the general case of any $n$.
:::

The Binomial Limit Theorem (de Moivre–Laplace theorem) says that as $n$ grows large, with $p \in (0,1)$ staying fixed, the Binomial distribution is approximately normal with mean $np$ and variance $np(1-p)$

:::{.callout-tip icon=false collapse="true"}
The unemployment rate is $10%$. Suppose that $100$ employable people are selected randomly. What is the probability that this sample contains between $9$ and $12$ unemployed people. Use the normal approximation to binomial probabilities.

Parameters are 
$\mu \approx n p = 10$
$\sigma^2= n p (1−p) = 100 (0.1)(0.9)=9$
$\sigma=\sqrt{9}=3$).
:::

#### **Poisson**. {-}

The number of events in a fixed interval

* Discrete, support $\{0,1,2,\ldots\}$
* Probability Mass Function: $Prob(X_{i}=x)=e^{-\lambda}\lambda^x/x!$
* See <https://en.wikipedia.org/wiki/Poisson_distribution>
* A common use case: How many cars will show up in the lot tomorrow?

```{r}
## Minimal example
lambda <- 3.5

# PMF plot
x <- seq(0, 15)
f_x <- dpois(x, lambda)
plot(x, f_x, type="h", col="blue",
     xlab = "x", ylab = "Prob(X = x)")
points(x, f_x, pch = 16, col = "blue")
title(bquote(paste('Pois(',.(lambda), ')')))


# Simulate: Compare empirical vs theoretical
#X <- rpois(1e4, lambda)
#c(emp_mean = mean(X), th_mean = lambda)
#c(emp_var  = var(X),  th_var  = lambda)
```

:::{.callout-tip icon=false collapse="true"}
Show that $\mathbb{E}[X_{i}] = \mathbb{V}[X_{i}]= \lambda$.
:::


#### **Irwin–Hall**. {-}

The sum of $n$ i.i.d. $\text{Uniform}(0,1)$.  

* Continuous, support $[0,n]$
* Probability Density Function: $f(x) = \dfrac{1}{(n-1)!}
\displaystyle\sum_{k=0}^{\lfloor x \rfloor} (-1)^k
\binom{n}{k} (x - k)^{n-1}$ for $x \in [0,n]$, and $0$ otherwise
* See <https://en.wikipedia.org/wiki/Irwin%E2%80%93Hall_distribution>
* A common use case: representing the sum of many small independent financial shocks

```{r}
## Minimal example (base R)
# Irwin–Hall PDF function
dirwinhall <- function(x, n) {
    f_x <- vector(length=length(x))
    for(i in seq(x)) {
        xx <- x[i]
        if(xx < 0 | xx > n){
            f_x[i] <- 0 
        } else {
            k <- seq(0, floor(xx))
            f_k <- sum((-1)^k*choose(n, k)*(xx-k)^(n-1))/factorial(n-1)
            f_x[i] <- f_k
        }
    }
    return(f_x)
}

# Parameters
n <- 2
x <- seq(0, n, length.out = 500)

# Compute and plot PDF
f_x <- dirwinhall(x, n)
plot(x, f_x, type="l", col="blue",
    main='', xlab = "x", ylab = "f(x)")
title(bquote(paste('IrwinHall(',.(n), ')')))
```

See also the [Bates](https://en.wikipedia.org/wiki/Bates_distribution) distribution, which is for the mean  of $n$ i.i.d. $\text{Uniform}(0,1)$ random variables. It essentially rescales the Irwin-Hall distribution to range between $0$ and $1$.


#### **Beta**. {-}
The sample space is any number on the unit interval, $X_{i} \in [0,1]$, but with non-uniform probabilities. 


```{r}
X4 <- rbeta(2000,2,2) ## two shape parameters
hist(X4, breaks=20, border=NA, main=NA, freq=F)

#See the underlying probabilities
#f_25 <- dbeta(.25, 2, 2)

x <- seq(0,1,by=.01)
fx <- dbeta(x, 2, 2)
lines(x, fx, col='blue')
```

The [Beta](https://en.wikipedia.org/wiki/Beta_distribution) distribution is mathematically complicated to write, and so we omit it. However, we can find the probability graphically using either the probability density function or cumulative distribution function. 

:::{.callout-tip icon=false collapse="true"}
Suppose $X_{i}$ is a random variable with a beta distribution. Intuitively depict $Prob(X_{i} \in [0.2, 0.8])$ by drawing an area under the density function. Numerically estimate that same probability using the CDF.
```{r}
plot( ecdf(X4), main=NA) # Empirical

x <- seq(0,1,by=.01) # Theoretical
Fx <- pbeta(x, 2, 2)
lines(x, Fx, col='blue')

# Middle Interval Example 
F2 <- pbeta(0.2, 2, 2)
F8 <- pbeta(0.8, 2, 2)
F_2_8 <- F8 - F2
F_2_8

# Visualize
title('Middle between 0.2 and 0.8')
segments( 0.2, F2, -1, F2, col='red')
segments( 0.8, F8, -1, F8, col='red')
```
:::

This distribution is often used, as the probability density function has two parameters that allow it to take many different shapes.

:::{.callout-tip icon=false collapse="true"}
For each example below, intuitively depict $Prob(X_{i} \leq 0.5)$ using the PDF. Repeat the exercise using a CDF instead of a PDF to calculate a numerical value.
```{r, results='hide'}
#| fig-width: 8
#| fig-height: 6
op <- par(no.readonly = TRUE); on.exit(par(op), add = TRUE)
x <- seq(0,1,by=.01)
pars <- expand.grid( c(.5,1,2), c(.5,1,2) )
par(mfrow=c(3,3))
apply(pars, 1, function(p){
    fx <- dbeta( x,p[1], p[2])
    plot(x, fx, type='l', xlim=c(0,1), ylim=c(0,4), lwd=2, col='blue')
    #hist(rbeta(2000, p[1], p[2]), breaks=50, border=NA, main=NA, freq=F)
})
title('Beta densities', outer=T, line=-1)
```
:::

## Beyond Basic Programming

#### **Packages**. {-}
Expansion "packages" allow for more functionality.

Most packages can be easily installed from the internet via [CRAN](https://cran.r-project.org/web/packages/available_packages_by_name.html).
```{r, eval=FALSE}
# common packages for tables/figures
install.packages("plotly")
install.packages("reactable")
install.packages("stargazer")

# common packages for data/teaching
install.packages("AER")
install.packages("Ecdat")
install.packages('wooldridge')

# other packages for statistics and data handling
install.packages("extraDistr")
install.packages("twosamples")
install.packages("data.table")
```

You only need to install packages once. But you need to load the package via `library` every time you want the extended functionality. For example, to generate exotic probability distributions
```{r}
# install.packages("extraDistr"), once ever
library(extraDistr) # every time you need it

par(mfrow=c(1,2))
for(p in c(-.5,0)){
    x <- rgev(2000, mu=0, sigma=1, xi=p)
    hist(x, breaks=50, border=NA, main=NA, freq=F)
}
title('GEV densities', outer=T, line=-1)
```

```{r}
library(extraDistr)

par(mfrow=c(1,3))
for(p in c(-1, 0,2)){
    x <- rtlambda(2000, p)
    hist(x, breaks=100, border=NA, main=NA, freq=F)
}
title('Tukey-Lambda densities', outer=T, line=-1)
```

The most common packages also have [cheatsheets](https://www.rstudio.com/resources/cheatsheets/) you can use. 

#### **Updating**. {-}
Make sure R and your packages are up to date. The current version of R and any packages used can be found (and recorded) with 

```{r, eval=F}
sessionInfo()
```

To update your R packages, use 
```{r, eval=F}
update.packages()
```

After updating R, you can update *all* packages stored on your computer (in all `.libPaths()`)
```{r, eval=FALSE}
update.packages(checkBuilt=T, ask=F)
# install.packages(old.packages(checkBuilt=T)[,"Package"])
```


#### **Base**. {-}
While additional packages can make your code faster, they also create dependencies that can lead to problems. So learn base R well before becoming dependent on other packages

* <https://bitsofanalytics.org/posts/base-vs-tidy/>
* <https://jtr13.github.io/cc21fall2/comparison-among-base-r-tidyverse-and-datatable.html>


#### **Github Packages**. {-}

<details>
<summary> Advanced and Optional </summary>
  <p>

Sometimes you will want to install a package from GitHub. For this, you can use [devtools](https://devtools.r-lib.org/) or its light-weight version [remotes](https://remotes.r-lib.org/)

To color terminal output on Linux systems, for example, you can use the colorout package
```{r, eval=FALSE}
# install.packages("remotes")
library(remotes)
# Install <https://github.com/jalvesaq/colorout
# to .libPaths()[1]
install_github('jalvesaq/colorout')
library(colorout)
```

To use `devtools` instead, you also need to have software developer tools installed on your computer.

* Windows: [Rtools](https://cran.r-project.org/bin/windows/Rtools/)
* Mac: [Xcode](https://apps.apple.com/us/app/xcode/id497799835)


  </p>
</details>

