
# Random Variables
***

Previously, we computed a distribution given the data, whereas now we generate individual data points given the distribution.
This chapter introduces *random variables* and the two basic kinds of sample space (discrete and continuous), along with the most common distributions in each family.

To talk about data we have not yet observed, we need an object that names the set of possible outcomes and the long-run frequency of each.

:::{.callout-important icon=false title="Key Definition" collapse="false"}
A *random variable* $X_{i}$ is a variable whose value is generated according to a probability distribution.
Its *sample space* is the set of all values it can take, and its *probability* assigns a long-run frequency to each subset of the sample space.
A *realized value* $\hat{X}_{i}$ is a specific observed value drawn from the random variable's distribution.
:::

The random-variable framing is useful for reasoning about a *data generating process* before any data are observed.
For example, consider flipping a coin before knowing whether it lands on heads or tails: we denote the unflipped coin as the random variable $X_{i}$, with sample space $\{0, 1\}$ ($x=1$ is heads, $x=0$ is tails) and probability $1/2$ on each outcome.
A specific flipped coin that has landed heads or tails is denoted $\hat{X}_{i}$, mirroring the hat notation from the previous chapter.

```{r, echo=FALSE}
x <- c(0, 1)
x_probs <- c(0.5, 0.5)
X0 <- sample(x, 400, prob=x_probs, replace=T)
# Plot Cumulative Proportion
X0_t <- seq(1, length(X0))
X0_mt <- cumsum(X0)/X0_t
par(mar=c(4, 4, 1, 4))
plot(X0_t, X0_mt, type='l',
    ylab='Cumulative Proportion Heads',
    xlab='Flip #', 
    ylim=c(0, 1), 
    lwd=2)
# Show individual flip outcomes
points(X0_t, X0, col=grey(0, .5),
    pch='|', cex=.3)
# Show theoretical proportions of heads
abline(h=x_probs[2], col=rgb(0, 0, 1, .8))
```

There are two basic types of sample spaces: discrete (encompassing cardinal-discrete, factor-ordered, and factor-unordered data) and continuous. This leads to two types of random variables: discrete and continuous. However, each type has many different 
[probability distributions](https://en.wikipedia.org/wiki/List_of_probability_distributions).

#### **Cumulative Probability**. {-}

The first job of any random-variable description is to assign a probability to events of the form "the value falls at or below $x$".

:::{.callout-important icon=false title="Key Definition" collapse="false"}
The *cumulative distribution function* (CDF) of a random variable $X_{i}$ is
$$F(x) = Prob(X_{i} \leq x),$$
the theoretical probability that $X_{i}$ takes a value at or below $x$.
The CDF is the theoretical analogue of the empirical CDF $\hat{F}$ from the previous chapter.
:::

The CDF is useful because it packages every probability of interest into a single function: $Prob(X_{i} > x) = 1 - F(x)$, $Prob(a < X_{i} \leq b) = F(b) - F(a)$, and so on.
You can think of $F(x)$ as the ECDF for a dataset with an infinite number of observations; equivalently, the ECDF is an empirical version of the CDF applied to observed data.
The most common random variables in R are [easily accessible](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/Distributions.html) through their CDFs.

#### **Probability Rules**. {-}

After introducing different random variables, we will also cover some basic implications of their CDF. Intuitively, probabilities must sum up to one. So we can compute $Prob(X_{i} > x) = 1- F(x)$. We also have two "in" and "out" probabilities.

The probability of $X_{i}\leq b$ and $X_{i}> a$ can be written in terms of falling into a range $Prob(X_{i} \in (a,b])=Prob(a < X_{i} \leq b) = F(b) - F(a)$. 

The opposite probability of $X_{i} > b$ or $X_{i} \leq a$ is $Prob(X_{i} \leq a \text{ or } X_{i} > b) = F(a) + [1- F(b)]$. Notice that this opposite probability $F(a) + [1- F(b)] =1 - [F(b) - F(a)]$, so that $Prob(X_{i} \text{ out of } (a,b]) = 1 - Prob( X_{i} \in (a,b])$. Each of these rules is applied to concrete numbers in the four-sided-die example below.


## Discrete Random Variables

A discrete random variable can take one of several values in a set. E.g., any number in $\{1,2,3,...\}$ or any letter in $\{A,B,C,...\}$. Theoretical proportions are referred to as a *probability mass function*, which can be thought of as a proportions bar plot for an infinitely large dataset. Equivalently, the bar plot is an empirical version of the probability mass function that is applied to observed data. The CDF accumulates the mass function, $F(x)=\sum_{x' \leq x} Prob(X_{i}=x')$, and conversely the probability of any value is the size of the jump in $F$ at that value.

In R, you can simulate random discrete data using the `sample` function. To do so, you must specify each outcome $x$ and its probability $Prob(X_{i}=x)$.

#### **Bernoulli**. {-}

The simplest random variable models a single yes/no event, like one flip of a coin.

:::{.callout-important icon=false title="Key Definition" collapse="false"}
A *Bernoulli* random variable takes two values, $x=1$ (success/heads) with probability $p$ and $x=0$ (failure/tails) with probability $1-p$.
The single parameter $p \in [0,1]$ controls the probability of success.
:::

Bernoulli variables are useful whenever an outcome is binary (heads or tails, success or failure, present or absent), and are also the building block from which richer discrete distributions are assembled out of independent yes/no events.
Writing the distribution out in full helps fix the notation we will reuse for richer cases:
\begin{eqnarray}
X_{i} &\in& \{0,1\} \\
Prob(X_{i} =0) &=& 1-p \\
Prob(X_{i} =1) &=& p \\
F(x) &=& \begin{cases}
    0   & x<0 \\
    1-p & x \in [0,1) \\
    1   & x\geq 1
\end{cases}
\end{eqnarray}

The theoretical mean of a Bernoulli variable works out to $p$, as derived in [Population Statistics](https://jadamso.github.io/Rbooks/01_06_PopStatistics.html#population-parameters).

Here is an example of the [Bernoulli](https://en.wikipedia.org/wiki/Bernoulli_distribution) distribution. While you might get all heads (or all tails) in the first few coin flips, the ratios level out to their theoretical values after many flips.
```{r}
x <- c(0, 1)
x_probs <- c(3/4, 1/4)

sample(x, 1, prob=x_probs, replace=T) # 1 Flip
sample(x, 4, prob=x_probs, replace=T) # 4 Flips 
X0 <- sample(x, 400, prob=x_probs, replace=T)

# Plot proportions
proportions <- table(X0)/length(X0)
plot(proportions, 
    col=grey(0, 0.5),
    xlab='Flip Outcome', 
    ylab='Proportion',
    main=NA)
points(c(0, 1), c(.75, .25), pch=16, col=rgb(0, 0, 1, .8)) # Theoretical values

# Plot CDF
plot( ecdf(X0), col=grey(0, .5),
    pch=16, main=NA) #Empirical
#points(c(0, 1), c(.75, 1), pch=16, col='blue') # Theoretical
```


#### **Discrete Uniform**. {-}

Discrete numbers with equal probability, such as a die with $K$ sides. 
\begin{eqnarray}
X_{i} &\in& \{1,...K\} \\
Prob(X_{i} =1) &=& Prob(X_{i} =2) = ... = 1/K\\
F(x) &=& \begin{cases}
    0   & x<1 \\
    1/K & x \in [1,2) \\
    2/K & x \in [2,3) \\
    \vdots & \\
    1   & x\geq K
\end{cases}
\end{eqnarray}

:::{.callout-note icon=false collapse="true" title="Must Know"}
Here is an example with $K=4$. E.g., rolling a four-sided die.

The probability of a value smaller than or equal to $3$ is $Prob(X_{i} \leq 3)=1/4 + 1/4 + 1/4 = 3/4$.

The probability of a value larger than $3$ is $Prob(X_{i} > 3) = 1-Prob(X_{i} \leq 3)=1/4$.

The probability of a value $>$ 1 and $\leq 3$ is $Prob(1 < X_{i} \leq 3) = Prob(X_{i} \leq 3) - Prob(X_{i} \leq 1) = 3/4 - 1/4 = 2/4$.^[This is the general formula using CDFs, and you can verify it works in this instance by directly adding the probability of each 2 or 3 event: $Prob(X_{i} = 2) +  Prob(X_{i} = 3) = 1/4 + 1/4 = 2/4$.]

The probability of a value $\leq$ 1 or $> 3$ is $Prob(X_{i} \leq 1 \text{ or } X_{i} > 3) =  Prob(X_{i} \leq 1) +  \left[ 1- Prob(X_{i} \leq 3) \right] = 1/4 + [1 - 3/4]=2/4$.
:::

```{r}
x <- c(1, 2, 3, 4)
x_probs <- c(1/4, 1/4, 1/4, 1/4)
# sample(x, 1, prob=x_probs, replace=T) # 1 roll
X1 <- sample(x, 2000, prob=x_probs, replace=T) # 2000 rolls

# Plot Long run proportions
proportions <- table(X1)/length(X1)
plot(proportions, col=grey(0, .5),
    xlab='Outcome', ylab='Proportion', main=NA)
points(x, x_probs, pch=16, col=rgb(0, 0, 1, .8)) # Theoretical values

# Hist w/ Theoretical Counts
# hist(X1, breaks=50, border=NA, main=NA, ylab='Count')
# points(x, x_probs*length(X1), pch='-') 

# Alternative Plot
plot( ecdf(X1), pch=16, col=grey(0, .5), main=NA)

# Alternative Plot 2
#props <- table(X1)
#barplot(props, ylim = c(0, 0.35), ylab = 'Proportion', xlab = 'Value')
#abline(h = 1/4, lty = 2)
```

Note that the *[Discrete Uniform](https://en.wikipedia.org/wiki/Discrete_uniform_distribution)* distribution generalizes to arbitrary sets, such as $\{0.0, 0.25, 0.5, 0.75, 1.0\}$ instead of $\{1,2,3,4,5\}$. We will not exploit the generalization in this class.




#### **Multinoulli**. {-}
Numbers $1,...K$ with unequal probabilities.
\begin{eqnarray}
X_{i} &\in& \{1,...K\} \\
Prob(X_{i} =1) &=& p_{1} \\
Prob(X_{i} =2) &=& p_{2} \\
        &\vdots& \\
p_{1} + p_{2} + ... &=& 1\\
F(x) &=& \begin{cases}
    0   & x<1 \\
    p_{1} & x \in [1,2) \\
    p_{1} + p_{2} & x \in [2,3) \\
    \vdots & \\
    1   & x\geq K
\end{cases}
\end{eqnarray}

Here is an empirical example with three outcomes

```{r}
x <- c(1, 2, 3)
x_probs <- c(3/10, 1/10, 6/10)
sum(x_probs)
X2 <- sample(x, 2000, prob=x_probs, replace=T) # sample of 2000

# Plot Long run proportions
proportions <- table(X2)/length(X2)
plot(proportions, col=grey(0, .5), ylim=c(0, 0.7),
    xlab='Outcome', ylab='Proportion', main=NA)
points(seq(x), x_probs, pch=16, col=rgb(0, 0, 1, .8)) # Theoretical values

# Histogram version
# X2_alt <- X1
# X2_alt[X2_alt=='A'] <- 1
#X2_alt[X2_alt=='B'] <- 2
#X2_alt[X2_alt=='C'] <- 3
#X2_alt <- as.numeric(X1_alt)
#hist(X2_alt, breaks=50, border=NA, 
#    main=NA, ylab='Count')
#points(x, x_probs*length(X2_alt), pch=16) ## Theoretical Counts

# Alternative Plot
# plot( ecdf(X2), pch=16, col=grey(0, .5), main=NA)
```

We can also replace numbers with letters $(A,...Z)$ or names $(John, Jamie, ...)$ although we must be careful with the CDF when there is no longer a natural ordering. When the outcomes are not cardinal data, the Multinoulli random variable is typically referred to as a *Categorical* random variable.
The CDF $F(x)=Prob(X_{i} \leq x)$ requires an ordering on the outcomes, and with labels like "my car dies" and "it rains next Tuesday", you do not have any inherent order and the CDF is an arbitrary artifact.

:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
Suppose there are three possible outcomes of an experiment, $\{\text{my car dies}, \text{it rains next Tuesday},\text{a cat is born}\}$, which have corresponding probabilities $\{3/10, 1/10, 6/10 \}$ that are known theoretically. Compute the probability that my car dies or a cat is born.
:::

:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
Draw Random Letters with increasing probability
```{r}
x <- LETTERS
x_probs <- seq(1, length(x))
x_probs <- x_probs/sum(x_probs) # probs must sum to 1
X <- sample(x, 1, prob=x_probs, replace=T)
X

# Draw Random Letters 100 Times
X <- sample(x, 1000, prob=x_probs, replace=T)
X <- factor(unlist(X), levels=LETTERS, ordered=F)
```
:::




## Continuous Random Variables

A continuous random variable can take one value out of an uncountably infinite number. E.g., any number between $0$ and $1$ with any number of decimal points.
With a continuous random variable, the probability of any individual point is zero, so we cannot describe the distribution by listing $Prob(X_{i} = x)$ for each $x$; instead, we describe it with the CDF or with a density.

:::{.callout-important icon=false title="Key Definition" collapse="false"}
The *probability density function* (PDF) $f(x)$ of a continuous random variable describes how concentrated the probability is near each value of $x$.
The probability that $X_{i}$ falls in an interval $[a,b]$ is the area under $f$ from $a$ to $b$, equivalently $F(b) - F(a)$.
$f(x)$ itself is not a probability; only its area over an interval is.
:::

The PDF is useful for intuition: probabilities are pictured as areas, peaks of $f$ mark where outcomes are likely, and the shape of $f$ tells you whether the distribution is symmetric, skewed, or heavy-tailed.
The CDF is the workhorse for numerical calculation, since $F(x)$ already encodes the cumulative area under $f$ from $-\infty$ to $x$: where $F$ rises steeply $f$ is large, and where $F$ is flat $f$ is near zero.
Just as $F$ can be thought of as the ECDF $\hat{F}$ with an infinite amount of data, $f$ can be thought of as a histogram $\hat{f}$ with an infinite amount of data; equivalently, the histogram is an empirical version of the PDF applied to observed data.


In R, you work with continuous distributions using these functions

* `dXXX` density, $f(x)$ for particular values $x$
* `pXXX` distribution, $F(x)$ for particular values $x$
* `rXXX` generate random variables according to the distribution

where `XXX` is the distribution name (`unif`, `exp`, `norm`)

#### **Continuous Uniform**. {-}

The simplest continuous distribution is the one with no preferred region: every value in an interval is equally likely.

:::{.callout-important icon=false title="Key Definition" collapse="false"}
A *Continuous Uniform* random variable on the interval $[a,b]$ is equally likely to land anywhere in $[a,b]$, with density $f(x) = 1/(b-a)$ for $x \in [a,b]$ and $0$ otherwise.
The unit interval $[0,1]$ is the special case $a=0, b=1$.
:::

The continuous uniform is useful as a *reference distribution*: a flat density makes it a natural "no information" baseline, and R's `runif` is the foundation underlying many other random-number routines.
The full specification gathers sample space, density, and cumulative function in one place:
\begin{eqnarray}
X_{i} &\in& [a,b] \\
f(x) &=& \begin{cases}
    1/(b-a) & x \in [a,b] \\
    0 & \text{Otherwise}
\end{cases}\\
F(x) &=& \begin{cases}
    0 & x < a \\
    (x-a)/(b-a) & x \in [a,b] \\
    1 & x > b.
\end{cases}
\end{eqnarray}

:::{.callout-note icon=false collapse="true" title="Must Know"}
The density $f$ and the CDF $F$ hold the same information, and you can move between them.
Consider a Continuous Uniform on $[-1, 2]$, so its density is flat: $f(x)=1/3$ for $x \in [-1, 2]$.

To build $F$ from $f$, accumulate area. The area under $f$ from $-1$ to $x$ is a rectangle of width $(x-(-1))=x+1$ and height $1/3$, so $F(x)=(x+1)/3$.
Check: $F(-1)=0$, $F(2)=1$, and $F(0.5)=1.5/3=0.5$.

To recover $f$ from $F$, take the slope. The CDF $F(x)=(x+1)/3$ is a straight line of slope $1/3$, so $f(x)=1/3$.
```{r}
x <- seq(-1, 2, by=0.01)
par(mfrow=c(1, 2))
# Density f(x): flat at 1/3
plot(x, dunif(x, -1, 2), type='l', col=rgb(0, 0, 1, .8),
    ylim=c(0, 0.5), xlab='x', ylab='f(x)', main=NA)
title('Density f(x)', font.main=1)
# CDF F(x): accumulated area under f
plot(x, punif(x, -1, 2), type='l', col=rgb(0, 0, 1, .8),
    xlab='x', ylab='F(x)', main=NA)
title('CDF F(x)', font.main=1)
```
The unit-interval case $[0,1]$ has density $f(x)=1$ and CDF $F(x)=x$, a special case of the above with $a=0$ and $b=1$.
:::

:::{.callout-note icon=false collapse="true" title="Must Know"}
For the unit-interval special case ($a=0, b=1$), the probability of a value being exactly $0.25$ is $Prob(X_{i} =0.25)=0$.

The probability of a value smaller than $0.25$ is $F(0.25)=0.25$.

The probability of a value larger than $0.25$ is $1-F(0.25)=0.75$.

The probability of a value in $(0.25,0.75]$ is $Prob(0.25 < X_{i} \leq 0.75) = Prob(X_{i} \leq 0.75) - Prob(X_{i} \leq 0.25) = 0.75 - 0.25 = 0.5$. 

The probability of a value in $(0.2,0.7]$ is $Prob(0.2 < X_{i} \leq 0.7) = Prob(X_{i} \leq 0.7) - Prob(X_{i} \leq 0.2) = 0.7 - 0.2 = 0.5$.

The probability of a value outside of $(0.2,0.7]$ is $Prob(X_{i} \leq 0.2 \text{ or } x > 0.7) = 0.2 + [1-0.7]=0.5$. Alternatively, you can compute $1- Prob(0.2 < X_{i} \leq 0.7)=1-0.5=0.5$.

```{r}
# Prob. < 0.25
punif(0.25)

# Prob. > 0.25
1-punif(0.25)

# Prob. in (0.2, 0.7]
F_two <- punif( c(0.2, 0.7) )
F_in <- F_two[2] - F_two[1]
F_in

# Prob. out of (0.2, 0.7]
F_out <- 1- F_in
F_out
```
:::


```{r}
runif(3) # 3 draws

# Empirical Density 
X3 <- runif(2000)
hist(X3, breaks=20, border=NA, main=NA, freq=F)
# Theoretical Density
x <- seq(-0.1, 1.1, by=.001)
fx <- dunif(x)
lines(x, fx, col=rgb(0, 0, 1, .8))

# CDF example 1
P_low <- punif(0.25)
P_low
# Uncomment to show via PDF
# x_low <- seq(0, 0.25, by=.001)
# fx_low <- dunif(x_low)
# polygon( c(x_low, rev(x_low)), c(fx_low, fx_low*0),
#    col=rgb(0, 0, 1, .25), border=NA)
    
# CDF example 2
P_high <- 1-punif(0.25)
P_high
# Uncomment to show via PDF
# x_high <- seq(0.25, 1, by=.001)
# fx_high <- dunif(x_high)
# polygon( c(x_high, rev(x_high)), c(fx_high, fx_high*0),
#    col=rgb(0, 0, 1, .25), border=NA)
    
# CDF example 3
P_mid <- punif(0.75) - punif(0.25)
P_mid
# Uncomment to show via PDF
# x_mid <-  seq(0.25, 0.75, by=.001)
# fx_mid <- dunif(x_mid)
# polygon( c(x_mid, rev(x_mid)), c(fx_mid, fx_mid*0),
#    col=rgb(0, 0, 1, .25), border=NA)
```


Note that the *[Continuous Uniform](https://en.wikipedia.org/wiki/Continuous_uniform_distribution)* distribution generalizes to an arbitrary interval, $X_{i} \in [a,b]$. In this case, $f(x)=1/[b-a]$ if $x \in [a,b]$ and $F(x)=[x-a]/[b-a]$ if $x \in [a,b]$.

:::{.callout-note icon=false collapse="true" title="Must Know"}
Suppose $X_{i}$ is a random variable continuously distributed over $a=-2$ and $b=2$. What is the probability of a value smaller than $0.25$? First use the computer to suggest an answer: simulate $1000$ draws and then make a histogram and an ECDF. Then find the answer mathematically using the CDF. Finally, verify the answer is intuitively correct in a figure of the PDF. You should draw by hand both the CDF and the PDF with correct axes labels and marking clearly the probability of a value smaller than $0.25$.
```{r}
# Simulation
X <- runif(1000, -2, 2)
# ... 

# Draw by computer the CDF, F(x)
x <- seq(-3, 3, by=0.005)
Fx <- punif(x, -2, 2)
plot(x, Fx, type='l', col=rgb(0, 0, 1, .8))
# Answer
Fstar <- punif(0.25, -2, 2)
# Visualize Answer
segments(0.25, 0, 0.25, Fstar,
    col=rgb(0, 0, 1, 0.25))

# Draw by computer the PDF, f(x)
fx <- dunif(x, -2, 2)
plot(x, fx, type='l', col=rgb(0, 0, 1, .8))
# Visualize Answer
fstar <- dunif(0.25, -2, 2)
polygon(
    c(-2, 0.25, 0.25, -2), 
    c(0, 0, fstar, fstar),
    col=rgb(0, 0, 1, 0.25), border=NA)
```
:::

:::{.callout-note icon=false collapse="true" title="Must Know"}
Suppose the flight time between Calgary and Kamloops is Uniformly distributed between $68$ and $78$ minutes. According to Air Canada the flight takes $70$ minutes. What is the probability that the flight will be late?
:::


#### **Exponential**. {-}
The sample space is any positive number.^[In other classes, you may further distinguish types of random variables based on whether their maximum value is theoretically finite or infinite.] An *[Exponential](https://en.wikipedia.org/wiki/Exponential_distribution)* random variable has a single *rate* parameter, $\lambda>0$.
\begin{eqnarray}
X_{i} &\in& [0,\infty) \\
f(x) &=& \lambda exp\left\{ -\lambda x \right\} \\
F(x) &=& \begin{cases} 
    0 & x < 0 \\
    1-  exp\left\{ -\lambda x \right\} & x \geq 0.
\end{cases}
\end{eqnarray}
```{r}
rexp(3) # 3 draws

X5 <- rexp(2000)
hist(X5, breaks=20,
    border=NA, main=NA,
    freq=F, ylim=c(0, 1), xlim=c(0, 10))

x <- seq(0, 10, by=.1)
fx <- dexp(x)
lines(x, fx, col=rgb(0, 0, 1, .8))
```

:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
Suppose the lifetime of a battery is an exponential random variable with $\lambda=1/50$. Using the computer, find the probability that the lifetime is $< 10$ hours.
Find the probability that the lifetime is $\geq 100$ hours. Use the computer to find the probability that the lifetime is between $10$ and $100$ hours.

```{r}
pexp(10, 1/50)
```
:::


:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
Note that there is a discrete version of the exponential, called the *geometric distribution*.
```{r}
Y <- rexp(5000, rate=.2)

X <- rgeom(5000, 1-exp(-.2))

par(mfrow=c(1, 2))
bks <- seq(0, 50, by=0.25)
hist(X, breaks=bks, freq=F, border=NA, main=NA, xlab='x')
title('Geometric', font.main=1)
hist(floor(Y), breaks=bks, freq=F, border=NA, main=NA, xlab='x')
title('Exponential (floored)', font.main=1)
```
:::


#### **Normal (Gaussian)**. {-}

For continuous data spread symmetrically around a center, the default model is the bell curve.

:::{.callout-important icon=false title="Key Definition" collapse="false"}
A *Normal* (or *Gaussian*) random variable has a bell-shaped density on the real line, with two parameters: the mean $\mu$ (the center) and the standard deviation $\sigma$ (the spread).
The density is $f(x) = \frac{1}{\sqrt{2\pi\sigma^{2}}}\exp\{-(x-\mu)^{2}/(2\sigma^{2})\}$.
:::

The *[Normal](https://en.wikipedia.org/wiki/Normal_distribution)* is useful as the default model for measurements that cluster around a typical value with symmetric scatter (heights, test scores, measurement errors), and is sometimes called the *Gaussian* distribution.
We call it "Normal" because it appears again and again, both as a description of real data and as the limiting shape for sample averages.
The full PDF spells out the parameters $\mu \in (-\infty,\infty)$ and $\sigma > 0$:
\begin{eqnarray}
X_{i} &\in& (-\infty,\infty) \\
f(x) &=& \frac{1}{\sqrt{2\pi \sigma^2}} exp\left\{ \frac{-(x-\mu)^2}{2\sigma^2} \right\}
\end{eqnarray}

Despite the complicated formula, the shape is simple: a bell centered at $\mu$, with most of the probability within a few multiples of $\sigma$ on either side.

```{r}
rnorm(3) # 3 draws

X6 <- rnorm(2000)
hist(X6, breaks=20,
    border=NA, main=NA,
    freq=F, ylim=c(0, .4), xlim=c(-4, 4))

x <- seq(-10, 10, by=.025)
fx <- dnorm(x)
lines(x, fx, col=rgb(0, 0, 1, .8))
```

Even thought the distribution function is complex, we can compute CDF values using the computer.
```{r}
pnorm( c(-1.645, 1.645) ) # 10%
pnorm( c(-2.576, 2.576) ) #  1%
```


:::{.callout-note icon=false collapse="true" title="Must Know"}
Suppose $X_{i}$ is a random variable with a normal distribution with $\mu=0$ and $\sigma=1$. Intuitively depict $Prob(X_{i} \in [0.2, 0.8])$ by drawing an area under the density function. Numerically estimate that same probability using the CDF.
```{r}
plot( ecdf(X6), main=NA) # Empirical

x <- seq(-4, 4, by=.01) # Theoretical
Fx <- pnorm(x, 0, 1)
lines(x, Fx, col=rgb(0, 0, 1, .8))

# Middle Interval Example
F2 <- pnorm(0.2, 0, 1)
F8 <- pnorm(0.8, 0, 1)
F_2_8 <- F8 - F2
F_2_8

# Visualize
title('Middle between 0.2 and 0.8', font.main=1)
segments( 0.2, F2, -5, F2, col=rgb(1, 0, 0, .8))
segments( 0.8, F8, -5, F8, col=rgb(1, 0, 0, .8))
```
:::


:::{.callout-note icon=false collapse="true" title="Must Know"}
Suppose that your health status is a normally distributed random variable with $\mu=2$ and $\sigma=3$. If we randomly sample one person, what is the probability there health status is higher than $4$?
```{r}
# Start with a simulation of 1000 people to build intuition
X <- rnorm(1000, 2, 3)
hist(X, freq=F, border=NA, main=NA)
sum(X>4)/1000

# Do an exact calculation
1-pnorm(4, 2, 3)
```
:::


:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
Draw the Middle $90\%$ of a normal distribution

```{r}
# PDF
x <- seq(-10, 10, by=.025)
fx <- dnorm(x)
plot(x, fx,
    col=rgb(0, 0, 1, .8), type='l',
    ylab='Theoretical Density: f(x)',
    main=NA)
title('Middle 90%', font.main=1)
# Show Middle 90%
x_90 <- seq(-1.645, 1.645, by=.025)
fx_90 <- dnorm(x_90)
polygon( c(x_90, rev(x_90)), c(fx_90, fx_90*0),
    col=rgb(0, 0, 1, .25), border=NA)
pnorm(1.645)-pnorm(-1.645)
```
:::


:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
Suppose scores in a math class are approximately normally distributed with $\mu=70$ and $\sigma=10$.
If you select one student at random, what is the probability their score is higher than $90$?
Is $Prob(X_{i}\geq 90)$ higher if $\mu=80, \sigma=10$? What about $\mu=70, \sigma=20$?
```{r}
# Case 1: baseline, mu=70 and sigma=10
1 - pnorm(90, mean=70, sd=10)

# Case 2: shift the mean up to 80, same sigma
1 - pnorm(90, mean=80, sd=10)

# Case 3: keep mean at 70, double the spread to sigma=20
1 - pnorm(90, mean=70, sd=20)
```
Raising $\mu$ from $70$ to $80$ moves the bell closer to $90$, so the right-tail probability rises sharply.
Doubling $\sigma$ at fixed $\mu$ also raises the right-tail probability because the bell flattens and more mass spreads into the tails.
:::


## Exercises

1. A Multinoulli random variable has outcomes $\{1,2,3,4\}$ with probabilities $\{0.1, 0.3, 0.4, 0.2\}$. Using the CDF, compute $Prob(X_{i} \leq 2)$, $Prob(X_{i} > 3)$, and $Prob(1 < X_{i} \leq 3)$. Show your work.

2. Suppose $X_{i}$ is normally distributed with $\mu=10$ and $\sigma=4$. Using `pnorm`, compute the probability that $X_{i}$ falls between $6$ and $14$. Then compute the probability that $X_{i}$ is more than two standard deviations away from the mean, i.e., $Prob(X_{i} < 2 \text{ or } X_{i} > 18)$.

3. In R, generate $2000$ draws from an exponential distribution with $\lambda=0.5$ using `rexp(2000, rate=0.5)`. Plot a histogram with `freq=F` and overlay the theoretical density using `lines` and `dexp`. Then use `pexp` to compute the exact probability that a draw exceeds $4$.

#### **Further Reading**. {-}

Many random variables are related to each other. Also note that numbers randomly generated on your computer cannot be truly random, they are "pseudorandom."

* <https://en.wikipedia.org/wiki/Relationships_among_probability_distributions> -- Overview of how common distributions (Normal, Chi-squared, Exponential, etc.) connect to each other.
* <https://www.math.wm.edu/~leemis/chart/UDR/UDR.html> -- Interactive chart showing transformations and limiting relationships among univariate distributions.
* <https://learningstatisticswithr.com/book/probability.html> -- Introduction to probability distributions with worked R examples.

#### **Recall** {-}

This chapter introduced *random variables* and distinguished *discrete* families (Bernoulli, Uniform, Multinoulli) from *continuous* families (Uniform, Exponential, Normal), describing each through the *CDF* $F(x)=Prob(X_{i} \leq x)$ and (for continuous variables) the *PDF* $f(x)$.
The four-sided die example tied the CDF rules together: with $Prob(X_{i} \leq 3) = 3/4$, the right-tail probability is $Prob(X_{i} > 3) = 1/4$, the in-interval probability is $Prob(1 < X_{i} \leq 3) = 3/4 - 1/4 = 2/4$, and the out-of-interval probability is $Prob(X_{i} \leq 1 \text{ or } X_{i} > 3) = 1/4 + 1/4 = 2/4$.
In the next chapter we draw repeated samples from these distributions and ask how the resulting statistics behave from sample to sample.

