
# Bivariate Distributions
***

We will now study two variables. The data for each observation data can be grouped together as a vector $(\hat{X}_{i}, \hat{Y}_{i})$.
```{r}
# Bivariate Data from USArrests
xy <- USArrests[,c('Murder','UrbanPop')]
xy[1,]
```
The vector $(\hat{X}_{i}, \hat{Y}_{i})$ has a *joint distribution* that describes the relationship between $\hat{X}_{i}$ and $\hat{Y}_{i}$.

## Types of Distributions

#### **Joint Distributions**.{-}

Scatterplots are used frequently to summarize the joint relationship between two variables, multiple observations of $(\hat{X}_{i}, \hat{Y}_{i})$. They can be enhanced in several ways. As a default, use semi-transparent points so as not to hide any points (and perhaps see if your observations are concentrated anywhere). You can also add other features that help summarize the relationship, although I will defer this until later.

```{r}
plot(Murder~UrbanPop, USArrests, pch=16, col=grey(0.,.5))
```



#### **Marginal Distributions**.{-}
You can also show the distributions of each variable along each axis.

```{r}
# Setup Plot
layout( matrix(c(2,0,1,3), ncol=2, byrow=TRUE),
    widths=c(9/10,1/10), heights=c(1/10,9/10))

# Scatterplot
par(mar=c(4,4,1,1))
plot(Murder~UrbanPop, USArrests, pch=16, col=rgb(0,0,0,.5))

# Add Marginals
par(mar=c(0,4,1,1))
xhist <- hist(USArrests[,'UrbanPop'], plot=FALSE)
barplot(xhist[['counts']], axes=FALSE, space=0, border=NA)

par(mar=c(4,0,1,1))
yhist <- hist(USArrests[,'Murder'], plot=FALSE)
barplot(yhist[['counts']], axes=FALSE, space=0, horiz=TRUE, border=NA)
```

#### **Conditional Distributions**.{-}

We can show how distributions and densities change according to a second (or even third) variable using data splits. E.g., 

```{r}
# Tailored Histogram 
ylim <- c(0,8)
xbks <-  seq(min(USArrests[,'Murder'])-1, max(USArrests[,'Murder'])+1, by=1)

# Also show more information
# Split Data by Urban Population above/below mean
pop_mean <- mean(USArrests[,'UrbanPop'])
pop_cut <- USArrests[,'UrbanPop']< pop_mean
murder_lowpop <- USArrests[pop_cut,'Murder']
murder_highpop <- USArrests[!pop_cut,'Murder']
cols <- c(low=rgb(0,0,1,.75), high=rgb(1,0,0,.75))

par(mfrow=c(1,2))
hist(murder_lowpop,
    breaks=xbks, col=cols[1],
    main='Urban Pop >= Mean', font.main=1,
    xlab='Murder Arrests',
    border=NA, ylim=ylim)

hist(murder_highpop,
    breaks=xbks, col=cols[2],
    main='Urban Pop < Mean', font.main=1,
    xlab='Murder Arrests',
    border=NA, ylim=ylim)
```

It is sometimes it is preferable to show the ECDF instead. And you can glue various combinations together to convey more information all at once

```{r}
par(mfrow=c(1,2))
# Full Sample Density
hist(USArrests[,'Murder'], 
    main='Density Function Estimate', font.main=1,
    xlab='Murder Arrests',
    breaks=xbks, freq=F, border=NA)

# Split Sample Distribution Comparison
F_lowpop <- ecdf(murder_lowpop)
plot(F_lowpop, col=cols[1],
    pch=16, xlab='Murder Arrests',
    main='Distribution Function Estimates',
    font.main=1, bty='n')
F_highpop <- ecdf(murder_highpop)
plot(F_highpop, add=T, col=cols[2], pch=16)

legend('bottomright', col=cols,
    pch=16, bty='n', inset=c(0,.1),
    title='% Urban Pop.',
    legend=c('Low (<= Mean)','High (>= Mean)'))
```

```{r, message=F}
# Simple Interactive Scatter Plot
# plot(Assault~UrbanPop, USArrests, col=grey(0,.5), pch=16,
#    cex=USArrests[,'Murder']/diff(range(USArrests[,'Murder']))*2,
#    main='US Murder arrests (per 100,000)')
```


You can also split data into grouped boxplots in the same way
```{r}
layout( t(c(1,2,2)))
boxplot(USArrests[,'Murder'], main='',
    xlab='All Data', ylab='Murder Arrests')

# K Groups with even spacing
K <- 3
USArrests[,'UrbanPop_Kcut'] <- cut(USArrests[,'UrbanPop'],K)
Kcols <- hcl.colors(K,alpha=.5)
boxplot(Murder~UrbanPop_Kcut, USArrests,
    main='', col=Kcols,
    xlab='Urban Population', ylab='')

# 4 Groups with equal numbers of observations
#Qcuts <- c(
#    '0%'=min(USArrests[,'UrbanPop'])-10*.Machine[['double.eps']],
#    quantile(USArrests[,'UrbanPop'], probs=c(.25,.5,.75,1)))
#USArrests[,'UrbanPop']_cut <- cut(USArrests[,'UrbanPop'], Qcuts)
#boxplot(Murder~UrbanPop_cut, USArrests, col=hcl.colors(4,alpha=.5))
```


## Probability Theory

We now consider a bivariate *random vector* $(X_{i}, Y_{i})$, which is a theoretical version of the bivariate observations $(\hat{X}_{i}, \hat{Y}_{i})$. E.g., If we are going to flip two coins, then $(X_{i}, Y_{i})$ corresponds to the unflipped coins and $(\hat{X}_{i}, \hat{Y}_{i})$ corresponds to concrete values after they are flipped.


#### **Definitions for Discrete Data**. {-}
The *joint distribution* is defined as
\begin{eqnarray}
Prob(X_{i} = x, Y_{i} = y)
\end{eqnarray}
Variables are *statistically independent* if $Prob(X_{i} = x, Y_{i} = y)= Prob(X_{i} = x) Prob(Y_{i} = y)$ for all $x, y$. Independence is sometimes assumed for mathematical simplicity, not because it generally fits data well.^[The same can be said about assuming normally distributed errors, although at least that can be motivated by the Central Limit Theorems.]


The *conditional distributions* are defined as
\begin{eqnarray}
Prob(X_{i} = x | Y_{i} = y) = \frac{ Prob(X_{i} = x, Y_{i} = y)}{ Prob( Y_{i} = y )}\\
Prob(Y_{i} = y | X_{i} = x) = \frac{ Prob(X_{i} = x, Y_{i} = y)}{ Prob( X_{i} = x )}
\end{eqnarray}
The *marginal distributions* are then defined as
\begin{eqnarray}
Prob(X_{i} = x) = \sum_{y} Prob(X_{i} = x | Y_{i} = y) Prob( Y_{i} = y ) \\
Prob(Y_{i} = y) = \sum_{x} Prob(Y_{i} = y | X_{i} = x) Prob( X_{i} = x ),
\end{eqnarray}
which is also known as the *law of total probability*.



#### **Coin Flips Example**. {-}

For one example, Consider flipping two coins, where we mark whether "heads" is face up with a $1$ and "tail" with a $0$. E.g., the first coin has a value of $x=1$ if it shows heads and $x=0$ if it shows tails.
This table shows both the joint distribution and also each marginal distribution.

|    | $x=0$  | $x=1$   | Marginal    |
|:------:|:---:|:---:|:---:|
| $y=0$    |$Prob(X_{i}=0,Y_{i}=0)$|$Prob(X_{i}=1,Y_{i}=0)$|$Prob(Y_{i}=0)$|
| $y=1$    |$Prob(X_{i}=0,Y_{i}=1)$|$Prob(X_{i}=1,Y_{i}=1)$|$Prob(Y_{i}=1)$|
| **Marginal** |$Prob(X_{i}=0)$       |$Prob(X_{i}=1)$        | $1$           |

Note that different joint distributions can have the same marginal distributions.

:::{.callout-note icon=false collapse="true"}
Suppose both coins are "fair": $Prob(X_{i}=1)= 1/2$ and $Prob(Y_{i}=1|X_{i}=x)=1/2$ for either $x=1$ or $x=0$, then the four potential outcomes have equal probabilities. 
\begin{eqnarray}
Prob(X_{i} = 0, Y_{i} = 0) &=& 1/2 \times 1/2 = 1/4 \\
Prob(X_{i} = 0, Y_{i} = 1) &=& 1/4 \\
Prob(X_{i} = 1, Y_{i} = 0) &=& 1/4 \\
Prob(X_{i} = 1, Y_{i} = 1) &=& 1/4 .
\end{eqnarray}
The joint distribution is written generally as
\begin{eqnarray}
Prob(X_{i} = x, Y_{i} = y) &=& Prob(X_{i} = x) Prob(Y_{i} = y).
\end{eqnarray}

The marginal distribution of the second coin is 
\begin{eqnarray}
Prob(Y_{i} = 0) &=& Prob(Y_{i} = 0 | X_{i} = 0) Prob(X_{i}=0) + Prob(Y_{i} = 0 | X_{i} = 1) Prob(X_{i}=1)\\
&=& 1/2 \times 1/2 + 1/2 \times 1/2 = 1/2\\
Prob(Y_{i} = 1) &=& Prob(Y_{i} = 1 | X_{i} = 0) Prob(X_{i}=0) + Prob(Y_{i} = 1 | X_{i} = 1) Prob(X_{i}=1)\\
&=& 1/2 \times 1/2 + 1/2 \times 1/2 = 1/2
\end{eqnarray}

```{r}
# Create a 2x2 matrix for the joint distribution.
# Rows correspond to X1 (coin 1), and columns correspond to X2 (coin 2).
P_fair <- matrix(1/4, nrow = 2, ncol = 2)
rownames(P_fair) <- c("X1=0", "X1=1")
colnames(P_fair) <- c("X2=0", "X2=1")
P_fair

# Compute the marginal distributions.
# Marginal for X1: sum across columns.
P_X1 <- rowSums(P_fair)
P_X1
# Marginal for X2: sum across rows.
P_X2 <- colSums(P_fair)
P_X2

# Compute the conditional probabilities Prob(X2 | X1).
cond_X2_given_X1 <- matrix(0, nrow = 2, ncol = 2)
for (j in c(1,2)) {
  cond_X2_given_X1[, j] <- P_fair[, j] / P_X1[j]
}
rownames(cond_X2_given_X1) <- c("X2=0", "X2=1")
colnames(cond_X2_given_X1) <- c("given X1=0", "given X1=1")
cond_X2_given_X1
```
:::

:::{.callout-note icon=false collapse="true"}
Now consider a second example, where the second coin is "Completely Unfair", so that it is always the same as the first. The outcomes generated with a Completely Unfair coin are the same as if we only flipped one coin.
\begin{eqnarray}
Prob(X_{i} = 0, Y_{i} = 0) &=& 1/2 \\
Prob(X_{i} = 0, Y_{i} = 1) &=& 0 \\
Prob(X_{i} = 1, Y_{i} = 0) &=& 0 \\
Prob(X_{i} = 1, Y_{i} = 1) &=& 1/2 .
\end{eqnarray}
The joint distribution is written generally as
\begin{eqnarray}
Prob(X_{i} = x, Y_{i} = y) &=& Prob(X_{i} = x) \mathbf{1}( x=y ),
\end{eqnarray}
where $\mathbf{1}(X_{i}=1)$ means $X_{i}= 1$ and $0$ if $X_{i}\neq0$.
The marginal distribution of the second coin is 
\begin{eqnarray}
Prob(Y_{i} = 0) 
&=& Prob(Y_{i} = 0 | X_{i} = 0) Prob(X_{i}=0) + Prob(Y_{i} = 0 | X_{i} = 1) Prob(X_{i} = 1)\\
&=& 1/2 \times 1 + 0 \times 1/2 = 1/2 .\\
Prob(Y_{i} = 1)
&=& Prob(Y_{i} = 1 | X_{i} =0) Prob( X_{i} = 0) + Prob(Y_{i} = 1 | X_{i} = 1) Prob( X_{i} = 1)\\
&=& 0\times 1/2 + 1 \times 1/2 = 1/2 .
\end{eqnarray}
which is the same as in the first example!
:::

```{r}
# Create the joint distribution matrix for the unfair coin case.
P_unfair <- matrix(c(0.5, 0, 0, 0.5), nrow = 2, ncol = 2, byrow = TRUE)
rownames(P_unfair) <- c("X1=0", "X1=1")
colnames(P_unfair) <- c("X2=0", "X2=1")
P_unfair

# Compute the marginal distribution for X2 in the unfair case.
P_X2_unfair <- colSums(P_unfair)
P_X1_unfair <- rowSums(P_unfair)

# Compute the conditional probabilities Prob(X1 | X2) for the unfair coin.
cond_X2_given_X1_unfair <- matrix(NA, nrow = 2, ncol = 2)
for (j in c(1,2)) {
  if (P_X1_unfair[j] > 0) {
    cond_X2_given_X1_unfair[, j] <- P_unfair[, j] / P_X1_unfair[j]
  }
}
rownames(cond_X2_given_X1_unfair) <- c("X2=0", "X2=1")
colnames(cond_X2_given_X1_unfair) <- c("given X1=0", "given X1=1")
cond_X2_given_X1_unfair
```



#### **Definitions for Continuous Data**. {-}
The *joint distribution* is defined as
\begin{eqnarray}
F(x, y) &=& Prob(X_{i} \leq x, Y_{i} \leq y)
\end{eqnarray}
The *marginal distributions* are then defined as
\begin{eqnarray}
 F_{X}(x) &=& F(x, \infty)\\
 F_{Y}(y) &=& F(\infty, y).
\end{eqnarray}
which is also known as the *law of total probability*.
Variables are statistically independent if $F(x, y) = F_{X}(x)F_{Y}(y)$ for all $x, y$.


For example, suppose $(X_{i},Y_{i})$ is bivariate normal with  means $(\mu_{X}, \mu_{Y})$, variances $(\sigma_{X}, \sigma_{Y})$ and covariance $\rho$.

```{r}
# Simulate Bivariate Data
N <- 10000
Mu <- c(2,2) ## Means

Sigma1 <- matrix(c(2,-.8,-.8,1),2,2) ## CoVariance Matrix
MVdat1 <- mvtnorm::rmvnorm(N, Mu, Sigma1)

Sigma2 <- matrix(c(2,.4,.4,1),2,2) ## CoVariance Matrix
MVdat2 <- mvtnorm::rmvnorm(N, Mu, Sigma2)

par(mfrow=c(1,2))
## Different diagonals
plot(MVdat2, col=rgb(1,0,0,0.02), pch=16,
    main='Joint', font.main=1,
    ylim=c(-4,8), xlim=c(-4,8), xlab='X1', ylab='X2')
points(MVdat1,col=rgb(0,0,1,0.02),pch=16)
## Same marginal distributions
xbks <- seq(-4,8,by=.2)
hist(MVdat2[,2], col=rgb(1,0,0,0.5),
    breaks=xbks, border=NA, xlab='X2',
    main='Marginal', font.main=1)
hist(MVdat1[,2], col=rgb(0,0,1,0.5),
    add=T, breaks=xbks, border=NA)
```


```{r, eval=F}
# See that independent data are a special case
n <- 2e4
## 2 Indepenant RV
XYiid <- cbind( rnorm(n),  rnorm(n))
## As a single Joint Draw
XYjoint <- mvtnorm::rmvnorm(n, c(0,0))
## Plot
par(mfrow=c(1,2))
plot(XYiid, xlab=
    col=grey(0,.05), pch=16, xlim=c(-5,5), ylim=c(-5,5))
plot(XYjoint,
    col=grey(0,.05), pch=16, xlim=c(-5,5), ylim=c(-5,5))

# Compare densities
#d1 <- dnorm(XYiid[,1],0)*dnorm(XYiid[,2],0)
#d2 <- mvtnorm::dmvnorm(XYiid, c(0,0))
#head(cbind(d1,d2))
```


The multivariate normal is a workhorse for analytical work on multivariate random variables, but there are many more. See e.g., <https://cran.r-project.org/web/packages/NonNorMvtDist/NonNorMvtDist.pdf>


#### **Important Applications**. {-}

Note *Simpson's Paradox*:

Also note *Bayes' Theorem*:
\begin{eqnarray}
Prob(X_{i} = x | Y_{i} = y)  Prob( Y_{i} = y) 
    &=& Prob(X_{i} = x, Y_{i} = y) = Prob(Y_{i} = y | X_{i} = x) Prob(X_{i} = x).\\
Prob(X_{i} = x | Y_{i} = y)
    &=& \frac{ Prob(Y_{i} = y | X_{i} = x) Prob(X_{i}=x) }{ Prob( Y_{i} = y) }.
\end{eqnarray}
```{r}
# Verify Bayes' theorem for the unfair coin case:
# Compute Prob(X1=1 | X2=1) using the formula:
#   Prob(X1=1 | X2=1) = [Prob(X2=1 | X1=1) * Prob(X1=1)] / Prob(X2=1)

P_X1_1 <- 0.5
P_X2_1_given_X1_1 <- 1  # Since coin 2 copies coin 1.
P_X2_1 <- P_X2_unfair["X2=1"]

bayes_result <- (P_X2_1_given_X1_1 * P_X1_1) / P_X2_1
bayes_result
```


## Further Reading 

For plotting histograms and marginal distributions, see 

* <https://www.r-bloggers.com/2011/06/example-8-41-scatterplot-with-marginal-histograms/>
* <https://r-graph-gallery.com/histogram.html>
* <https://r-graph-gallery.com/74-margin-and-oma-cheatsheet.html>
* <https://jtr13.github.io/cc21fall2/tutorial-for-scatter-plot-with-marginal-distribution.html>

Many introductory econometrics textbooks have a good appendix on probability and statistics. There are many useful statistical texts online too

See the Further reading about Probability Theory in the Statistics chapter.

* <https://www.r-bloggers.com/2024/03/calculating-conditional-probability-in-r/>

