
# Bivariate Statistics
***

All of the univariate statistics we have covered apply to marginal distributions. For joint distributions, there are several ways to statistically describe the relationship between two variables.  The major differences surround whether the data are cardinal or an ordered/unordered factor.

## Statistics of Association

#### **Cardinal**. {-}
*Pearson (Linear) Correlation*.
Suppose $X$ and $Y$ are both cardinal data. As such, you can compute the most famous measure of association, the covariance:
$$
C_{XY} =  \sum_{i} [X_i - \overline{X}] [Y_i - \overline{Y}] / N
$$
```{r}
xy <- USArrests[,c('Murder','UrbanPop')]
#plot(xy, pch=16, col=grey(0,.25))
cov(xy)
```
Note that $C_{XX}=V_{X}$.
For ease of interpretation and comparison, we rescale this statistic to always lay between $-1$ and $1$ 
$$
r_{XY} = \frac{ C_{XY} }{ \sqrt{V_X} \sqrt{V_Y}}
$$
```{r}
cor(xy)[2]
```

*Falk Codeviance*.
The Codeviance is a robust alternative to Covariance. Instead of relying on means (which can be sensitive to outliers), it uses medians ($\tilde{X}$) to capture the central tendency.^[See also *Theil-Sen Estimator*, which may be seen as a precursor.] We can also scale the Codeviance by the median absolute deviation to compute the median correlation.
\begin{eqnarray}
\text{CoDev}(X,Y) = \text{Med}\left\{ |X_i - \tilde{X}| |Y_i - \tilde{Y}| \right\} \\
\tilde{r}_{XY} = \frac{ \text{CoDev}(X,Y) }{ \text{MAD}(X) \text{MAD}(Y) }.
\end{eqnarray}
```{r}
cor_m <- function(xy) {
  # Compute medians for each column
  med <- apply(xy, 2, median)
  # Subtract the medians from each column
  xm <- sweep(xy, 2, med, "-")
  # Compute CoDev
  CoDev <- median(xm[, 1] * xm[, 2])
  # Compute the medians of absolute deviation
  MadProd <- prod( apply(abs(xm), 2, median) )
  # Return the robust correlation measure
  return( CoDev / MadProd)
}
cor_m(xy)
```

#### **Ordered Factor**. {-}
Suppose $X$ and $Y$ are both *ordered* variables. *Kendall's Tau* measures the strength and direction of association by counting the number of concordant pairs (where the ranks agree) versus discordant pairs (where the ranks disagree). A value of $\tau = 1$ implies perfect agreement in rankings, $\tau = -1$ indicates perfect disagreement, and $\tau = 0$ suggests no association in the ordering.
$$
\tau = \frac{2}{n(n-1)} \sum_{i} \sum_{j > i} \text{sgn} \Bigl( (X_i - X_j)(Y_i - Y_j) \Bigr),
$$
where the sign function is:
$$
\text{sgn}(z) = 
\begin{cases}
+1 & \text{if } z > 0\\
0  & \text{if } z = 0 \\
-1 & \text{if} z < 0 
\end{cases}.
$$
```{r}
xy <- USArrests[,c('Murder','UrbanPop')]
xy[,1] <- rank(xy[,1] )
xy[,2] <- rank(xy[,2] )
# plot(xy, pch=16, col=grey(0,.25))
tau <- cor(xy[, 1], xy[, 2], method = "kendall")
round(tau, 3)
```

#### **Unordered Factor**. {-}
Suppose $X$ and $Y$ are both *categorical* variables; the value of $X$ is one of $1...r$ categories and the value of $Y$ is one of $1...k$ categories. *Cramer's V* quantifies the strength of association by adjusting a "chi-squared" statistic to provide a measure that ranges from $0$ to $1$; $0$ indicates no association while a value closer to $1$ signifies a strong association. 

First, consider a contingency table for $X$ and $Y$ with $r$ rows and $k$ columns. The chi-square statistic is then defined as:

$$
\chi^2 = \sum_{i=1}^{r} \sum_{j=1}^{k} \frac{(O_{ij} - E_{ij})^2}{E_{ij}}.
$$

where

- $O_{ij}$ denote the observed frequency in cell $(i, j)$,
- $E_{ij} = \frac{R_i \cdot C_j}{n}$ is the expected frequency for each cell if $X$ and $Y$ are independent
- $R_i$ denote the total frequency for row $i$ (i.e., $R_i = \sum_{j=1}^{k} O_{ij}$),
- $C_j$ denote the total frequency for column $j$ (i.e., $C_j = \sum_{i=1}^{r} O_{ij}$),
- $n$ be the grand total of observations, so that $n = \sum_{i=1}^{r} \sum_{j=1}^{k} O_{ij}$.


Second, normalize the chi-square statistic with the sample size and the degrees of freedom to compute Cramer's V. 

$$
V = \sqrt{\frac{\chi^2 / n}{\min(k - 1, \, r - 1)}},
$$

where:

- $n$ is the total sample size,
- $k$ is the number of categories for one variable,
- $r$ is the number of categories for the other variable.


```{r}
xy <- USArrests[,c('Murder','UrbanPop')]
xy[,1] <- cut(xy[,1],3)
xy[,2] <- cut(xy[,2],4)
table(xy)

cor_v <- function(xy){
    # Create a contingency table from the categorical variables
    tbl <- table(xy)
    # Compute the chi-square statistic (without Yates' continuity correction)
    chi2 <- chisq.test(tbl, correct=FALSE)[['statistic']]
    # Total sample size
    n <- sum(tbl)
    # Compute the minimum degrees of freedom (min(rows-1, columns-1))
    df_min <- min(nrow(tbl) - 1, ncol(tbl) - 1)
    # Calculate Cramer's V
    V <- sqrt((chi2 / n) / df_min)
    return(V)
}
cor_v(xy)

# DescTools::CramerV( table(xy) )
```


#### **Mixed**. {-}

For mixed data, where $Y_{i}$ is a cardinal variable and $X_{i}$ is an unordered factor variable, we analyze associations via group comparisons. The basic idea is seen in a comparison of two samples, which corresponds to an $X_{i}$ with two categories.


## Two-Sample Differences

Suppose we have 2 samples of data. 

Each $X_{is}$ is an individual observation $i$ from the sample $s=1,2$. (For example, the wages for men and women in Canada. For another example, homicide rates in two different American states.)

```{r, eval=F}
library(wooldridge)
x1 <- wage1[wage1$educ == 15, 'wage']
x2 <- wage1[wage1$educ == 16, 'wage']
```

For simplicity, we will assume that each observation is an independent observation. We will further assume the data from each group are normally distributed, but the mean and variance can be different across groups.

```{r}
# Sample 1 (e.g., males)
n1 <- 100
x1 <- rnorm(n1, 0, 2)
# Sample 2 (e.g., females)
n2 <- 80
x2 <- rnorm(n2, 1, 1)

par(mfrow=c(1,2))
bks <- seq(-7,7, by=.5)
hist(x1, border=NA, breaks=bks,
    main='Sample 1', font.main=1)

hist(x2, border=NA, breaks=bks, 
    main='Sample 2', font.main=1)
```
There may be several differences between these samples. Often, the first summary statistic we investigate is the difference in means. 

#### **Equal Means**. {-}
We often want to know if the means of different sample are different in . To test this hypothesis, we compute the sample mean $\overline{X}_{s}$ over all observations in each sample and then examine the differences term
\begin{eqnarray} 
D = \overline{X}_{1} - \overline{X}_{2},
\end{eqnarray}
with a null hypothesis of $D=0$.

```{r}
# Differences between means
m1 <- mean(x1)
m2 <- mean(x2)
d <- m1-m2
    
# Bootstrap Distribution
boot_d <- sapply(1:10^4, function(b){
    x1_b <- sample(x1, replace=T)
    x2_b <- sample(x2, replace=T)
    m1_b <- mean(x1_b)
    m2_b <- mean(x2_b)
    d_b <- m1_b - m2_b
    return(d_b)
})
hist(boot_d, border=NA, font.main=1,
    main='Difference in Means')

# 2-Sided Test
boot_ci <- quantile(boot_d, probs=c(.025, .975))
abline(v=boot_ci, lwd=2)
abline(v=0, lwd=2, col=2)
```{r}

```{r}
# p-value
1 - ecdf(boot_d)(0)
```


Just as with one sample tests, we can compute a *standardized differences*, where $D$ is converted into a $t$ statistic.  Note, however, that we have to compute the standard error for the difference statistic, which is a bit more complicated. However, this allows us to easily conduct one or two sided hypothesis tests using a standard normal approximation.
```{r, eval=F}
se_hat <- sqrt(var(x1)/n1 + var(x2)/n2);
t_obs <- d/se_hat
```

```{r, eval=F, echo=F}
t_2sample <- function(x1, x2){
    # Differences between means
    m1 <- mean(x1)
    m2 <- mean(x2)
    d <- (m1-m2)

    # SE estimate
    n1  <- length(x1)
    n2  <- length(x2)
    s1  <- var(x1)
    s2  <- var(x2)
    s   <- ((n1-1)*s1 + (n2-1)*s2)/(n1+n2-2)
    d_se <- sqrt(s*(1/n1+1/n2))

    # t stat
    t_stat <- d/d_se
    return(t_stat)
}
 
tstat <- twosam(data[,'male'], data[,'female'])
tstat
```


#### **Other Differences**. {-}
The above procedure generalized from differences in "means" to other statistics like "quantiles".

```{r}
# Bootstrap Distribution Function
boot_fun <- function( fun, B=10^4, ...){
    boot_d <- sapply(1:B, function(b){
        x1_b <- sample(x1, replace=T)
        x2_b <- sample(x2, replace=T)
        f1_b <- fun(x1_b, ...)
        f2_b <- fun(x2_b, ...)
        d_b <- f1_b - f2_b
        return(d_b)
    })
    return(boot_d)
}

# 2-Sided Test for Median Differences
# d <- median(x2) - median(x1)
boot_d <- boot_fun(median)
hist(boot_d, border=NA, font.main=1,
    main='Difference in Medians')
abline(v=quantile(boot_d, probs=c(.025, .975)), lwd=2)
abline(v=0, lwd=2, col=2)
1 - ecdf(boot_d)(0)
```

Note that these estimates suffer from a finite-sample bias, which we can correct for. Also note that bootstrap tests can perform poorly with highly unequal variances or skewed data.

```{r, eval=F}
# 2-Sided Test for SD Differences
#d <- sd(x2) - sd(x1)
boot_d <- boot_fun(sd)
hist(boot_d, border=NA, font.main=1,
    main='Difference in Standard Deviations')
abline(v=quantile(boot_d, probs=c(.025, .975)), lwd=2)
abline(v=0, lwd=2, col=2)
1 - ecdf(boot_d)(0)


# Try any function!
# boot_fun( function(xs) { IQR(xs)/median(xs) } )
```

#### **Distributional Comparisons**. {-}
We can also examine whether there are any differences between the entire *distributions*
```{r}
# Sample Wage Data
library(wooldridge)
x1 <- sort( wage1[wage1$educ == 15,  'wage'])  
x2 <- sort( wage1[wage1$educ == 16,  'wage'] )
x <- sort(c(x1, x2))

# Compute Quantiles
quants <- seq(0,1,length.out=101)
Q1 <- quantile(x1, probs=quants)
Q2 <- quantile(x2, probs=quants)

# Compare Distributions via Quantiles
rx <- range(c(x1, x2))
par(mfrow=c(1,2))
plot(rx, c(0,1), type='n', font.main=1,
    main='Distributional Comparison',
    xlab=expression(Q[s]),
    ylab=expression(F[s]))
lines(Q1, quants, col=2)
lines(Q2, quants, col=4)
legend('bottomright', col=c(2,4), lty=1,
legend=c('F1', 'F2'))

# Compare Quantiles
plot(Q1, Q2, xlim=rx, ylim=rx,
    main='Quantile-Quantile Plot', font.main=1,
pch=16, col=grey(0,.25))
abline(a=0,b=1,lty=2)
```

The starting point for hypothesis testing is the Kolmogorov-Smirnov Statistic: the maximum absolute difference between two CDF's over all sample data $x \in \{X_1\} \cup \{X_2\}$.
\begin{eqnarray}
KS &=& \max_{x} |F_{1}(x)- F_{2}(x)|^{p},
\end{eqnarray}
where $p$ is an integer (typically 1). An intuitive alternative is the Cramer-von Mises Statistic: the sum of absolute differences (raised to an integer, typically 2) between two CDF's. 
\begin{eqnarray}
CVM=\sum_{x} |F_{1}(x)- F_{2}(x)|^{p}.
\end{eqnarray}

```{r}
# Distributions
F1 <- ecdf(x1)(x)
F2 <- ecdf(x2)(x)

library(twosamples)

# Kolmogorov-Smirnov
KSq <- which.max(abs(F2 - F1))
KSqv <- round(twosamples::ks_stat(x1, x2),2)

# Cramer-von Mises Statistic (p=2)
CVMqv <- round(twosamples::cvm_stat(x1, x2, power=2), 2) 

# Visualize Differences
plot(range(x), c(0,1), type="n", xlab='x', ylab='ECDF')
lines(x, F1, col=2, lwd=2)
lines(x, F2, col=4, lwd=2)
# CVM
segments(x, F1, x, F2, lwd=.5, col=grey(0,.2))
# KS
segments(x[KSq], F1[KSq], x[KSq], F2[KSq], lwd=1.5, col=grey(0,.75), lty=2)
```

Just as before, you use bootstrapping for hypothesis testing.
```{r}
twosamples::cvm_test(x1, x2)
```

#### **Comparing Multiple Groups**. {-}
For multiple groups, we can tests the equality of all distributions (whether at least one group is different). The *Kruskal-Wallis* test examines $H_0:\; F_1 = F_2 = \dots = F_G$ versus $H_A:\; \text{at least one } F_g \text{ differs}$, where $F_g$ is the continuous distribution of group $g=1,...G$. This test does not tell us which group is different.

To conduct the test, first denote individuals $i=1,...n$ with overall ranks $r_1,....r_{n}$. Each individual belongs to group $g=1,...G$, and each group $g$ has $n_{g}$ individuals with average rank $\overline{r}_{g} = \sum_{i} r_{i} /n_{g}$. The Kruskal Wallis statistic is 
\begin{eqnarray}
KW &=& (N-1) \frac{\sum_{g=1}^{G} n_{g}( \overline{r}_{g} - \overline{r}  )^2  }{\sum_{i=1}^{N} ( r_{i} - \overline{r}  )^2}, 
\end{eqnarray}
where  $\overline{r} = \frac{N+1}{2}$ is the grand mean rank.

In the special case with only two groups, $G=2$, the Kruskal Wallis test reduces to the *Mann–Whitney U-test* (also known as the \textit{Wilcoxon rank-sum test}). In this case, we can write the hypotheses in terms of individual outcomes in each group, $Y_i$ in one group $Y_j$ in the other; $H_0: P(Y_i > Y_j)=P(Y_i > Y_i)$  versus $H_A: P(Y_i > Y_j) \neq P(Y_i > Y_j)$. The corresponding test statistic is
\begin{eqnarray}
U   &=& \min(U_1,U_2) \\
U_g &=& \sum_{i\in g}\sum_{j\in -g}
           \Bigl[\mathbf 1(Y_i > Y_j) + \tfrac12\mathbf 1(Y_i = Y_j)\Bigr].
\end{eqnarray}


```{r}
library(AER)
data(CASchools)
CASchools$stratio <- CASchools$students/CASchools$teachers

# Do student/teacher ratio differ for at least 1 county?
# Single test of multiple distributions
kruskal.test(CASchools$stratio, CASchools$county)

# Multiple pairwise tests
# pairwise.wilcox.test(CASchools$stratio, CASchools$county)
```




## Further Reading 

Other Statistics 

* <https://cran.r-project.org/web/packages/qualvar/vignettes/wilcox1973.html>


