
# Misc. Topics
***


## Locally Linear Multiple Regression


## Nonparametric Tests

#### **Distributional Comparisons**. {-}
We can also examine whether there are any differences between the entire *distributions*
```{r}
# Sample Wage Data
library(wooldridge)
x1 <- sort( wage1[wage1$educ == 15,  'wage'])  
x2 <- sort( wage1[wage1$educ == 16,  'wage'] )
x <- sort(c(x1, x2))

# Compute Quantiles
quants <- seq(0,1,length.out=101)
Q1 <- quantile(x1, probs=quants)
Q2 <- quantile(x2, probs=quants)

# Compare Distributions via Quantiles
rx <- range(c(x1, x2))
par(mfrow=c(1,2))
plot(rx, c(0,1), type='n', font.main=1,
    main='Distributional Comparison',
    xlab=expression(Q[s]),
    ylab=expression(F[s]))
lines(Q1, quants, col=2)
lines(Q2, quants, col=4)
legend('bottomright', col=c(2,4), lty=1,
legend=c('F1', 'F2'))

# Compare Quantiles
plot(Q1, Q2, xlim=rx, ylim=rx,
    main='Quantile-Quantile Plot', font.main=1,
pch=16, col=grey(0,.25))
abline(a=0,b=1,lty=2)
```

The starting point for hypothesis testing is the Kolmogorov-Smirnov Statistic: the maximum absolute difference between two CDF's over all sample data $x \in \{X_1\} \cup \{X_2\}$.
\begin{eqnarray}
KS &=& \max_{x} |F_{1}(x)- F_{2}(x)|^{p},
\end{eqnarray}
where $p$ is an integer (typically 1).

An intuitive alternative is the Cramer-von Mises Statistic: the sum of absolute differences (raised to a power, typically 2) between two CDF's. 
\begin{eqnarray}
CVM=\sum_{x} |F_{1}(x)- F_{2}(x)|^{p}.
\end{eqnarray}

```{r}
# Distributions
F1 <- ecdf(x1)(x)
F2 <- ecdf(x2)(x)

library(twosamples)

# Kolmogorov-Smirnov
KSq <- which.max(abs(F2 - F1))
KSqv <- round(twosamples::ks_stat(x1, x2),2)

# Cramer-von Mises Statistic (p=2)
CVMqv <- round(twosamples::cvm_stat(x1, x2, power=2), 2) 

# Visualize Differences
plot(range(x), c(0,1), type="n", xlab='x', ylab='ECDF')
lines(x, F1, col=2, lwd=2)
lines(x, F2, col=4, lwd=2)
# CVM
segments(x, F1, x, F2, lwd=.5, col=grey(0,.2))
# KS
segments(x[KSq], F1[KSq], x[KSq], F2[KSq], lwd=1.5, col=grey(0,.75), lty=2)
```

Just as before, you use bootstrapping for hypothesis testing.
```{r}
twosamples::cvm_test(x1, x2)
```

#### **Comparing Multiple Groups**. {-}
For multiple groups, we can tests the equality of all distributions (whether at least one group is different). The *Kruskal-Wallis* test examines
\[
H_0:\; F_1 = F_2 = \dots = F_G
\quad\text{versus}\quad
H_A:\; \text{at least one } F_g \text{ differs},
\]
where $F_g$ is the continuous distribution of group $g$. This test does not tell us which group is different.

To conduct the test, first denote individuals $i=1,...n$ with overall ranks $r_1,....r_{n}$. Each individual belongs to group $g=1,...G$, and each group $g$ has $n_{g}$ individuals with average rank $\overline{r}_{g} = \sum_{i} r_{i} /n_{g}$. The Kruskal Wallis statistic is 
\begin{eqnarray}
KW &=& (N-1) \frac{\sum_{g=1}^{G} n_{g}( \overline{r}_{g} - \overline{r}  )^2  }{\sum_{i=1}^{N} ( r_{i} - \overline{r}  )^2}, 
\end{eqnarray}
where  $\overline{r} = \frac{N+1}{2}$ is the grand mean rank.

In the special case with only two groups, $G=2$, the Kruskal Wallis test reduces to the *Mann–Whitney U-test* (also known as the \textit{Wilcoxon rank-sum test}). In this case, we can write the hypotheses in terms of individual outcomes in each group, $Y_i$ in one group $Y_j$ in the other;
\[
H_0: P(Y_i > Y_j)=P(Y_i > Y_i)
\quad\text{versus}\quad
H_A: P(Y_i > Y_j) \neq P(Y_i > Y_j) 
\]
The corresponding test statistic is
\begin{eqnarray}
U   &=& \min(U_1,U_2) \\
U_g &=& \sum_{i\in g}\sum_{j\in -g}
           \Bigl[\mathbf 1(Y_i > Y_j) + \tfrac12\mathbf 1(Y_i = Y_j)\Bigr].
\end{eqnarray}


```{r}
library(AER)
data(CASchools)
CASchools$stratio <- CASchools$students/CASchools$teachers

# Do student/teacher ratio differ for at least 1 county?
# Single test of multiple distributions
kruskal.test(CASchools$stratio, CASchools$county)

# Multiple pairwise tests
# pairwise.wilcox.test(CASchools$stratio, CASchools$county)
```





## Prediction

#### **Prediction Intervals**. {-}

In addition to confidence intervals, we can also compute a *prediction interval* which estimate the variability of new data rather than a statistic

In this example, we consider a single variable and compute the frequency each value was covered.
```{r}
x <- runif(1000)
# Middle 90% of values
xq0 <- quantile(x, probs=c(.05,.95))

bks <- seq(0,1,by=.01)
hist(x, breaks=bks, border=NA,
    main='Prediction Interval', font.main=1)
abline(v=xq0)

paste0('we are 90% confident that the a future data point will be between ', 
    round(xq0[1],2), ' and ', round(xq0[2],2) )
```
In this example, we consider a range for $y_{i}(x)$ rather than for $m(x)$. These intervals also take into account the residuals --- the variability of individuals around the mean. 
```{r}
# Bivariate Data from USArrests
xy <- USArrests[,c('Murder','UrbanPop')]
colnames(xy) <- c('y','x')
xy0 <- xy[order(xy$x),]
```

For a nice overview of different types of intervals, see https://www.jstor.org/stable/2685212. For an in-depth view, see "Statistical Intervals: A Guide for Practitioners and Researchers" or "Statistical Tolerance Regions: Theory, Applications, and Computation". See https://robjhyndman.com/hyndsight/intervals/ for constructing intervals for future observations in a time-series context. See Davison and Hinkley, chapters 5 and 6 (also Efron and Tibshirani, or Wehrens et al.)


```{r}
# From "Basic Regression"
xy0 <- xy[order(xy$x),]
X0 <- unique(xy0$x)
reg_lo <- loess(y~x, data=xy0, span=.8)
preds_lo <- predict(reg_lo, newdata=data.frame(x=X0))


# Jackknife CI
jack_lo <- sapply(1:nrow(xy), function(i){
    xy_i <- xy[-i,]
    reg_i <- loess(y~x, dat=xy_i, span=.8)
    predict(reg_i, newdata=data.frame(x=X0))
})
```

```{r}
boot_regs <- lapply(1:399, function(b){
    b_id <- sample( nrow(xy), replace=T)
    xy_b <- xy[b_id,]
    reg_b <- lm(y~x, dat=xy_b)
})

plot(y~x, pch=16, col=grey(0,.5),
    dat=xy0, ylim=c(0, 20))
lines(X0, preds_lo,
    col=hcl.colors(3,alpha=.75)[2],
    type='o', pch=2)

# Estimate Residuals CI at design points
res_lo <- sapply(1:nrow(xy), function(i){
    y_i <- xy[i,'y']
    preds_i <- jack_lo[,i]
    resids_i <- y_i - preds_i
})
res_cb <- apply(res_lo, 1, quantile,
    probs=c(.025,.975), na.rm=T)

# Plot
lines( X0, preds_lo +res_cb[1,],
    col=hcl.colors(3,alpha=.75)[2], lt=2)
lines( X0, preds_lo +res_cb[2,],
    col=hcl.colors(3,alpha=.75)[2], lty=2)



# Smooth estimates 
res_lo <- lapply(1:nrow(xy), function(i){
    y_i <- xy[i,'y']
    x_i <- xy[i,'x']
    preds_i <- jack_lo[,i]
    resids_i <- y_i - preds_i
    cbind(e=resids_i, x=x_i)
})
res_lo <- as.data.frame(do.call(rbind, res_lo))

res_fun <- function(x0, h, res_lo){
    # Assign equal weight to observations within h distance to x0
    # 0 weight for all other observations
    ki <- dunif(res_lo$x, x0-h, x0+h) 
    ei <- res_lo[ki!=0,'e']
    res_i <- quantile(ei, probs=c(.025,.975), na.rm=T)
}
res_lo2 <- sapply(X0, res_fun, h=15, res_lo=res_lo)

lines( X0, preds_lo + res_lo2[1,],
    col=hcl.colors(3,alpha=.75)[2], lty=1, lwd=2)
lines( X0, preds_lo + res_lo2[2,],
    col=hcl.colors(3,alpha=.75)[2], lty=1, lwd=2)
```



```{r, eval=F}
# Bootstrap Prediction Interval
boot_resids <- lapply(boot_regs, function(reg_b){
    e_b <- resid(reg_b)
    x_b <- reg_b$model$x
    res_b <- cbind(e_b, x_b)
})
boot_resids <- as.data.frame(do.call(rbind, boot_resids))
# Homoskedastic
ehat <- quantile(boot_resids$e_b, probs=c(.025, .975))
x <- quantile(xy$x,probs=seq(0,1,by=.1))
boot_pi <- coef(reg)[1] + x*coef(reg)['x']
boot_pi <- cbind(boot_pi + ehat[1], boot_pi + ehat[2])

# Plot Bootstrap PI
plot(y~x, dat=xy, pch=16, main='Prediction Intervals',
    ylim=c(-5,20), font.main=1)
polygon( c(x, rev(x)), c(boot_pi[,1], rev(boot_pi[,2])),
    col=grey(0,.2), border=NA)

# Parametric PI (For Comparison)
#pi <- predict(reg, interval='prediction', newdata=data.frame(x))
#lines( x, pi[,'lwr'], lty=2)
#lines( x, pi[,'upr'], lty=2)
```

#### **Forecasting and TSCV** {-}

## Decision Theory

#### **Statistical Power** {-}

#### **Quality Control** {-}

#### **Optimal Experiment Designs** {-}

