
# Comparing Groups
***


For mixed data, $\hat{Y}_{i}$ is a cardinal variable and $\hat{X}_{i}$ is a factor variable (typically unordered). For such data, we analyze associations via group comparisons. The basic idea is best seen in a comparison of two samples, which corresponds to an $\hat{X}_{i}$ with two categories. For example, the heights of men and women in Canada or the homicide rates in two different American states. For another example, the wages for people with and without completing a degree.

```{r}
library(wooldridge)
## Group 1
X1_id <- wage1[,'educ'] == 15
Y1 <- wage1[X1_id, 'wage']
## Group 2
X2_id <- wage1[,'educ'] == 16
Y2 <- wage1[X2_id, 'wage']

# Initial Summary Figure
bks <- seq(0, 24, by=1.5)
dlim <- c(0,.2)
cols <- c(rgb(1,0,0,.5), rgb(0,0,1,.5))

hist(Y1, breaks=bks, ylim=dlim,
     col=cols[1], xlab='Wages',
     freq=F, border=NA, main='')
hist(Y2, breaks=bks, ylim=dlim,
     col=cols[2],
     freq=F, border=NA, add=T)
legend('topright',
       col=cols, pch=15,
       legend=c('Grade 15', 'Grade 16'),
       title='School Completed')
```



## Differences

There may be several differences between these samples. Often, the first summary statistic we investigate is the difference in means. 

#### **Mean Differences**. {-}

We often want to know if the means of different sample are different. To test this hypothesis, we compute the means separately for each sample and then examine the differences term
\begin{eqnarray} 
\hat{D} = \hat{M}_{Y1} - \hat{M}_{Y2},
\end{eqnarray}
with a null hypothesis that there is no difference in the population means.

:::{.callout-note icon=false collapse="true"}
```{r}
# Differences between means
m1 <- mean(Y1)
m2 <- mean(Y2)
d <- m1-m2
    
# Bootstrap Distribution
bootstrap_diff <- vector(length=9999)
for(b in seq(bootstrap_diff) ){
    Y1_b <- sample(Y1, replace=T)
    Y2_b <- sample(Y2, replace=T)
    m1_b <- mean(Y1_b)
    m2_b <- mean(Y2_b)
    d_b <- m1_b - m2_b
    bootstrap_diff[b] <- d_b
}
hist(bootstrap_diff,
    border=NA, font.main=1,
    main='Difference in Means')

# 2-Sided Test via Confidence Interval
boot_ci <- quantile(bootstrap_diff, probs=c(.025, .975))
abline(v=boot_ci, lwd=2)
abline(v=0, lwd=2, col=2)
```
:::


Just as with one sample tests, we can compute a *standardized differences*, where $D$ is converted into a $t$ statistic. Note, however, that we have to compute the standard error for the difference statistic, which is a bit more complicated. However, this allows us to easily conduct one or two sided hypothesis tests using a standard normal approximation.

:::{.callout-tip icon=false collapse="true"}
```{r, eval=F}
se_hat <- sqrt(var(Y1)/n1 + var(Y2)/n2);
t_obs <- d/se_hat

t_2sample <- function(Y1, Y2){
    # Differences between means
    m1 <- mean(Y1)
    m2 <- mean(Y2)
    d <- (m1-m2)

    # SE estimate
    n1  <- length(Y1)
    n2  <- length(Y2)
    s1  <- var(Y1)
    s2  <- var(Y2)
    s   <- ((n1-1)*s1 + (n2-1)*s2)/(n1+n2-2)
    d_se <- sqrt(s*(1/n1+1/n2))

    # t stat
    t_stat <- d/d_se
    return(t_stat)
}
 
tstat <- twosam(data[,'male'], data[,'female'])
tstat
```
:::


#### **Quantile Differences**. {-}
The above procedure generalized from differences in means to other quantiles statistics like medians.

```{r}
# Quantile Comparison

## Distribution 1
F1 <- ecdf(Y1)
plot(F1, col=cols[1],
     pch=16, xlab='Wages',
     main='Comparing Medians',
     font.main=1, bty='n')
## Median 1
med1 <- quantile(F1, probs=0.5)
segments(med1, 0, med1, 0.5, col=cols[1], lty=2)
abline(h=0.5, lty=2)

## Distribution 2
F2 <- ecdf(Y2)
plot(F2, add=TRUE, col=cols[2], pch=16)
## Median 2
med2 <- quantile(F2, probs=0.5)
segments(med2, 0, med2, 0.5, col=cols[2], lty=2)

## Legend
legend('bottomright',
       col=cols, pch=15,
       legend=c('Grade 15', 'Grade 16'),
       title='School Completed')

```

:::{.callout-note icon=false collapse="true"}
```{r}
# Bootstrap Distribution Function
boot_quant <- function(Y1, Y2, B=9999, prob=0.5, ...){
    bootstrap_diff <- vector(length=B)
    for(b in seq(bootstrap_diff)){
        Y1_b <- sample(Y1, replace=T)
        Y2_b <- sample(Y2, replace=T)
        q1_b <- quantile(Y1_b, probs=0.5, ...)
        q2_b <- quantile(Y2_b, probs=0.5, ...)
        d_b <- q1_b - q2_b
        bootstrap_diff[b] <- d_b
    }
    return(bootstrap_diff)
}

# 2-Sided Test for Median Differences
# d <- median(Y2) - median(Y1)
boot_d <- boot_quant(Y1, Y1, B=999, prob=0.5)
hist(boot_d, border=NA, font.main=1,
    main='Difference in Medians')
abline(v=quantile(boot_d, probs=c(.025, .975)), lwd=2)
abline(v=0, lwd=2, col=2)
1 - ecdf(boot_d)(0)
```
:::


If we want to test for the differences in medians across groups with independent observations, we can also use notches in the boxplot. If the notches of two boxes do not overlap, then there is rough evidence that the difference in medians is statistically significant. The square root of the sample size is also shown as the bin width in each boxplot.^[Let each group $g$ have median $\tilde{M}_{g}$, interquartile range $\hat{IQR}_{g}$, observations $n_{g}$. We can compute standard deviation of the median as $\tilde{S}_{g}= \frac{1.25 \hat{IQR}_{g}}{1.35 \sqrt{n_{g}}}$. As a rough guess, the interval $\tilde{M}_{g} \pm 1.7 \tilde{S}_{g}$ is the historical default and displayed as a *notch* in the boxplot. See also <https://www.tandfonline.com/doi/abs/10.1080/00031305.1978.10479236>.]

```{r}
boxplot(Y1, Y2,
    col=cols,
    notch=T,
    varwidth=T)
```


Note that bootstrap tests can perform poorly with highly unequal variances or skewed data. To see this yourself, make a simulation with skewed data and unequal variances.


In principle, we can also examine whether there are differences in spread or shape statistics such as `sd` and `IQR`, or `skew` and `kurtosis`. More often, however, we examine whether there are any differences in the distributions.

:::{.callout-tip icon=false collapse="true"}
Here is an example to look at differences in "spread"
```{r, eval=F}
boot_fun <- function( fun, B=9999, ...){
    bootstrap_diff <- vector(length=B)
    for(b in seq(bootstrap_diff)){
        Y1_b <- sample(Y1, replace=T)
        Y2_b <- sample(Y2, replace=T)
        f1_b <- fun(Y1_b, ...)
        f2_b <- fun(Y2_b, ...)
        d_b <- f1_b - f2_b
        bootstrap_diff[b] <- d_b
    }
    return(bootstrap_diff)
}

# 2-Sided Test for SD Differences
#d <- sd(Y2) - sd(Y1)
boot_d <- boot_fun(sd)
hist(boot_d, border=NA, font.main=1,
    main='Difference in Standard Deviations')
abline(v=quantile(boot_d, probs=c(.025, .975)), lwd=2)
abline(v=0, lwd=2, col=2)
1 - ecdf(boot_d)(0)


# Try any function!
# boot_fun( function(xs) { IQR(xs)/median(xs) } )
```
:::



## Distributional Comparisons

We can also examine whether there are any differences between the entire *distributions*. Often, we simply plot the two CDF's. Another useful visualization is to plot the quantiles against one another: a quantile-quantile plot. I.e., the first data point on the bottom left shows the first quantile for both distributions. 

```{r}
# Wage Data (same as from before)
#library(wooldridge)
#Y1 <- sort( wage1[wage1[,'educ'] == 15,  'wage'])  
#Y2 <- sort( wage1[wage1[,'educ'] == 16,  'wage'] )

# Compute Quantiles
quants <- seq(0,1,length.out=101)
Q1 <- quantile(Y1, probs=quants)
Q2 <- quantile(Y2, probs=quants)

# Compare Distributions via Quantiles
#ry <- range(c(Y1, Y2))
#plot(ry, c(0,1), type='n', font.main=1,
#    main='Distributional Comparison',
#    xlab="Quantile",
#    ylab="Probability")
#lines(Q1, quants, col=2)
#lines(Q2, quants, col=4)
#legend('bottomright', col=c(2,4), lty=1,
#	legend=c(
#	    expression(hat(F)[1]),
#	    expression(hat(F)[2]) 
#))

# Compare Quantiles
ry <- range(c(Y1, Y2))
plot(Q1, Q2, xlim=ry, ylim=ry,
    xlab=expression(Q[1]),
    ylab=expression(Q[2]),
    main='Quantile-Quantile Plot',
    font.main=1,
    pch=16, col=grey(0,.25))
abline(a=0,b=1,lty=2)
```

The starting point for hypothesis testing is the Kolmogorov-Smirnov Statistic: the maximum absolute difference between two CDF's over all sample data $y \in \{Y_1\} \cup \{Y_2\}$.
\begin{eqnarray}
\hat{KS} &=& \max_{y} |\hat{F}_{1}(y)- \hat{F}_{2}(y)|^{p},
\end{eqnarray}
where $p$ is an integer (typically 1). An intuitive alternative is the Cramer-von Mises Statistic: the sum of absolute differences (raised to an integer, typically 2) between two CDF's. 
\begin{eqnarray}
\hat{CVM} &=& \sum_{y} | \hat{F}_{1}(y)- \hat{F}_{2}(y)|^{p}.
\end{eqnarray}

```{r}
# Distributions
y <- sort(c(Y1, Y2))
F1 <- ecdf(Y1)(y)
F2 <- ecdf(Y2)(y)

library(twosamples)

# Kolmogorov-Smirnov
KSq <- which.max(abs(F2 - F1))
KSqv <- round(twosamples::ks_stat(Y1, Y2),2)

# Cramer-von Mises Statistic (p=2)
CVMqv <- round(twosamples::cvm_stat(Y1, Y2, power=2), 2) 

# Visualize Differences
plot(range(y), c(0,1), type="n", xlab='x', ylab='ECDF')
lines(y, F1, col=cols[1], lwd=2)
lines(y, F2, col=cols[2], lwd=2)

# KS
title( paste0('KS: ', KSqv), adj=0, font.main=1)
segments(y[KSq], F1[KSq], y[KSq], F2[KSq], lwd=1.5, col=grey(0,.75), lty=2)

# CVM
title( paste0('CVM: ', CVMqv), adj=1, font.main=1)
segments(y, F1, y, F2, lwd=.5, col=grey(0,.2))
```

Just as before, you use bootstrapping for hypothesis testing.
```{r}
twosamples::ks_test(Y1, Y2)

twosamples::cvm_test(Y1, Y2)
```

:::{.callout-tip icon=false collapse="true"}
Compare the distribution of arrests for two different counties, each with data over time.

```{r, eval=F}
library(wooldridge)
countymurders
```

:::

#### **Comparing Multiple Groups**. {-}
For multiple groups, we can tests the equality of all distributions (whether at least one group is different). The *Kruskal-Wallis* test examines $H_0:\; F_1 = F_2 = \dots = F_G$ versus $H_A:\; \text{at least one } F_g \text{ differs}$, where $F_g$ is the continuous distribution of group $g=1,...G$. This test does not tell us which group is different.

To conduct the test, first denote individuals $i=1,...n$ with overall ranks $\hat{r}_1,....\hat{r}_{n}$. Each individual belongs to group $g=1,...G$, and each group $g$ has $n_{g}$ individuals with average rank $\bar{r}_{g} = \sum_{i} \hat{r}_{i} /n_{g}$. The Kruskal Wallis statistic is 
\begin{eqnarray}
\hat{KW} &=& (N-1) \frac{\sum_{g=1}^{G} n_{g}( \bar{r}_{g} - \bar{r}  )^2  }{\sum_{i=1}^{n} ( \hat{r}_{i} - \bar{r}  )^2}, 
\end{eqnarray}
where  $\bar{r} = \frac{n+1}{2}$ is the grand mean rank.

In the special case with only two groups, $G=2$, the Kruskal Wallis test reduces to the *Mann–Whitney U* test (also known as the *Wilcoxon rank-sum* test). In this case, we can write the hypotheses in terms of individual outcomes in each group, $Y_i$ in one group $Y_j$ in the other; $H_0: Prob(Y_i > Y_j)=Prob(Y_i > Y_i)$  versus $H_A: Prob(Y_i > Y_j) \neq Prob(Y_i > Y_j)$. The corresponding test statistic is
\begin{eqnarray}
\hat{U}   &=& \min(\hat{U}_1, \hat{U}_2) \\
\hat{U}_g &=& \sum_{i\in g}\sum_{j\in -g}
           \Bigl[\mathbf 1( \hat{Y}_{i} > \hat{Y}_{j}) + \tfrac12\mathbf 1(\hat{Y}_{i} = \hat{Y}_{j})\Bigr].
\end{eqnarray}


```{r}
library(AER)
data(CASchools)
CASchools[,'stratio'] <- CASchools[,'students']/CASchools[,'teachers']

# Do student/teacher ratio differ for at least 1 county?
# Single test of multiple distributions
kruskal.test(CASchools[,'stratio'], CASchools[,'county'])

# Multiple pairwise tests
# pairwise.wilcox.test(CASchools[,'stratio'], CASchools[,'county'])
```

## Theory

Under the assumption that both populations are normally distributed, we can analytically derive the sampling distribution for the difference-in-means. In particular, the $t$-statistic
\begin{eqnarray}
\hat{t} = \frac{
    \hat{M}_{Y1} - \hat{M}_{Y2}
}{
    \sqrt{\hat{S}_{Y1}+\hat{S}_{Y2}}/\sqrt{n}
}
\end{eqnarray}
follows Student's [t-distribution](https://en.wikipedia.org/wiki/Student%27s_t-distribution).
Welch's [$t$-statistic](https://en.wikipedia.org/wiki/Welch%27s_t-test) is an adjustment for two normally distributed populations with potentially unequal variances or sample sizes.

```{r}
# Sample 1 (e.g., males)
n1 <- 100
Y1 <- rnorm(n1, 0, 2)
# Sample 2 (e.g., females)
n2 <- 80
Y2 <- rnorm(n2, 1, 1)

par(mfrow=c(1,2))
bks <- seq(-8,8, by=.5)
hist(Y1, border=NA, breaks=bks,
    main='Sample 1', font.main=1)
hist(Y2, border=NA, breaks=bks, 
    main='Sample 2', font.main=1)

t.test(Y1, Y2, var.equal=F)
```

## Further Reading 

Other Statistics 

* <https://cran.r-project.org/web/packages/qualvar/vignettes/wilcox1973.html>


