
# Statistics of Association
***

There are several ways to statistically assess the relationship between two variables. The major differences surround whether the data are cardinal or an ordered/unordered factor. In all cases, we measure association but not causation.

## Cardinal Data

#### **Pearson (Linear) Correlation**. {-}

When two variables are cardinal, the foundational measure of how they move together is the covariance.

:::{.callout-important icon=false title="Key Definition" collapse="false"}
The *covariance* $\hat{C}_{XY}$ is the average product of paired deviations from each variable's mean (where $\hat{M}_{X}$ and $\hat{M}_{Y}$ are the sample means):
$$\hat{C}_{XY} =  \sum_{i=1}^{n} [\hat{X}_{i} - \hat{M}_{X}] [\hat{Y}_i - \hat{M}_{Y}] / n.$$
:::

The *covariance* is useful as the building block for every other linear two-variable statistic: it is positive when $\hat{X}_{i}$ and $\hat{Y}_{i}$ tend to be above or below their means together, negative when they tend to move apart, and zero when there is no linear co-movement. Its units are the units of $\hat{X}_{i}$ times the units of $\hat{Y}_{i}$, which makes raw values hard to compare across datasets. Note that the covariance of $\hat{X}$ with itself is just its variance, $\hat{C}_{XX}=\hat{V}_{X}$, and the standard deviation is $\hat{S}_{X}=\sqrt{\hat{V}_{X}}$.

For an interpretable, unit-free measure that is comparable across datasets, we rescale the covariance.

:::{.callout-important icon=false title="Key Definition" collapse="false"}
The *Pearson correlation* $\hat{R}_{XY}$ rescales the covariance by the product of the two standard deviations:
$$\hat{R}_{XY} = \frac{ \hat{C}_{XY} }{ \hat{S}_{X} \hat{S}_{Y}}.$$
:::

The *Pearson correlation* is useful as the workhorse unit-free summary of *linear* association: it is constrained to $[-1, 1]$, with $+1$ for a perfect rising line, $-1$ for a perfect falling line, and $0$ for no linear association. Because the bounds are the same in every dataset, it lets us compare the strength of two-variable associations across studies. A value close to $-1$ suggests negative association, a value close to $0$ suggests no linear association, and a value close to $+1$ suggests positive association.

:::{.callout-note icon=false collapse="true" title="Must Know"}
What is the correlation for the dataset $\{ (0,0.1) , (1, 0.3), (2, 0.2) \}$? Find the answer both mathematically and computationally.

Mathematically, there are five steps.

Step 1: Compute the means
\begin{eqnarray}
\hat{M}_{X} &=& \frac{0+1+2}{3} = 1 \\
\hat{M}_{Y} &=& \frac{0.1+0.3+0.2}{3} = 0.2
\end{eqnarray}

Step 2: Compute the deviances
\begin{eqnarray}
\begin{array}{c|rrrr}
\hat{X}_i   & 0   & 1 & 2 \\
\hat{X}_i-\hat{M}_{X} & -1  & 0 & 1 \\
\hat{Y}_i   & 0.1 & 0.3 & 0.2 \\
\hat{Y}_i-\hat{M}_{Y}   & -0.1 & 0.1 & 0 
\end{array}
\end{eqnarray}

Step 3: Compute the Covariance
\begin{eqnarray}
\hat{C}_{XY} &=&
\sum  (\hat{X}_i-\hat{M}_{X})(\hat{Y}_i-\hat{M}_{Y})/n 
= \left[ (-1)(-0.1) + 0(0.1) + 1(0) \right] \frac{1}{3} 
=  (-0.1) \frac{1}{3}
= 1/30
\end{eqnarray}

Step 4: Compute Standard Deviations
\begin{eqnarray}
\hat{V}_{X} &=& \sum_{i=1}^n \left(\hat{X}_i-\hat{M}_{X}\right)^2 / n
= \left[(-1)^2+0^2+1^2 \right]/3
= 2/3 \\
\hat{S}_{X} &=& \sqrt{2/3} \\
\hat{V}_{Y} &=& \sum_{i=1}^n \left( \hat{Y}_i-\hat{M}_{Y} \right)^2 / n
= \left[ (-0.1)^2+(0.1)^2+0^2 \right]/3
= \left[0.01+0.01\right]/3
= \frac{2}{100} \frac{1}{3} = 2/300 \\
\hat{S}_{Y} &=& \sqrt{2/300}
\end{eqnarray}

Step 5: Compute the Correlation
\begin{eqnarray}
\frac{\hat{C}_{XY}}{\hat{S}_X \hat{S}_Y}
&=& \frac{1/30}{ \sqrt{2/3}  \sqrt{2/300}}
= \frac{1/30}{ 2 /\sqrt{900}}
= \frac{1/30}{2/30} = 1/2
\end{eqnarray}

Note that this value suggests a positive relationship between the variables.

Computationally, we do the same steps
```{r}
# Create the Data
X <- c(0, 1, 2)
X
Y <- c(0.1, 0.3, 0.2)
Y

# Compute the Means
mX <- mean(X)
mY <- mean(Y)

# Compute the Deviances
dev_X <- X - mX
dev_Y <- Y - mY

# Compute the Covariance
cov_manual <-  sum(dev_X * dev_Y) / length(X)

# Compute the Standard Deviations
var_X <- sum(dev_X^2) / length(X)
sd_X <- sqrt(var_X)
var_Y <- sum(dev_Y^2) / length(Y)
sd_Y <- sqrt(var_Y)

# Compute the Correlation
cor_manual <- cov_manual / (sd_X * sd_Y)
cor_manual

# Verify with the built-in function
cor(X, Y)
```
:::

#### **Hypothesis Testing**. {-}

You can conduct hypothesis tests for these statistics using the same procedures we learned for univariate data, [ch.5](https://jadamso.github.io/Rbooks/01_07_Intervals.html). For example, by inverting a confidence interval.

:::{.callout-note icon=false collapse="true" title="Must Know"}
```{r}
# Load the Data
xy <- USArrests[, c('Murder', 'UrbanPop')]
xy_cor <- cor(xy[, 1], xy[, 2])
#plot(xy, pch=16, col=grey(0, .25))
    
# Bootstrap Distribution of Correlation
n <- nrow(xy)
bootstrap_cor <- rep(NA, 9999)
for(b in seq(bootstrap_cor) ){
    xy_b <- xy[sample(n, replace=T), ]
    xy_cor_b <- cor(xy_b[, 1], xy_b[, 2])
    bootstrap_cor[b] <- xy_cor_b
}
hist(bootstrap_cor, breaks=100,
    border=NA, freq=F,
    xlab='Correlation',
    main=NA)
title('Bootstrap Distribution', font.main=1)

## Test whether correlation is statistically different from 0
boot_ci <- quantile(bootstrap_cor, probs=c(0.025, 0.975))
abline(v=boot_ci)
abline(v=0, col=rgb(1, 0, 0, .8))
```
:::


Importantly, we can also impose the null of hypothesis of no association by reshuffling the data. If we resample without replacement, this is known as a *permutation test*.

:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
```{r}
xy <- USArrests[, c('Murder', 'UrbanPop')]
xy_cor <- cor(xy[, 1], xy[, 2])
#plot(xy, pch=16, col=grey(0, .25))
    
# Null Bootstrap Distribution of Correlation
n <- nrow(xy)
null_bootstrap_cor <- rep(NA, 9999)
for(b in seq(null_bootstrap_cor) ){
    xy_b <- xy
    xy_b[, 'UrbanPop'] <- xy[sample(n, replace=T), 'UrbanPop'] ## Reshuffle X
    xy_cor_b <- cor(xy_b[, 1], xy_b[, 2])
    null_bootstrap_cor[b] <- xy_cor_b
}
hist(null_bootstrap_cor, breaks=100,
    border=NA, freq=F,
    xlab='Correlation',
    main=NA)
title('Null Bootstrap Distribution', font.main=1)

## Test whether correlation is statistically different from 0
boot_ci <- quantile(null_bootstrap_cor, probs=c(0.025, 0.975))
abline(v=boot_ci)
abline(v=xy_cor, col=rgb(0, 0, 1, .8))
```
Because all dependence resides in the pairing, and permuting one margin destroys the pairing completely.

To construct a permutation test, we need to use `replace=F`. Rework the above code to make a Permutation Null Distribution and conduct a permutation test.
:::

So the bootstrap evaluates sampling variability in the world that generated your data. A permutation test constructs the distribution of the statistic in a world where the null is true. Altogether, we have

| Distribution| Sample Size per Iteration | Number of Iterations | Mechanism | Typical Purpose |
| -------- | ------- | ------- | ------- |----------------|
Jackknife       | $n-1$ | $n$  | $X,Y$: Deterministically leave-one-out observation | Variance estimate (after rescaling) and Normal CI estimate |
Bootstrap       | $n$   | $B$  | $X,Y$: Random resample with replacement | Percentile CI estimate |
Null Bootstrap  | $n$   | $B$  | $X,Y$: Random resample with replacement and shifted | Percentile CI under imposed null, $p$-values |
Permutation     | $n$   | $B$  | $X$: Random resample without replacement | Percentile CI under imposed null of no association, $p$-values |
: Types of resampling


#### **Falk Codeviance**. {-}

When the data contain outliers, the mean-based covariance can be misleading and a robust alternative is needed.

:::{.callout-important icon=false title="Key Definition" collapse="false"}
The *Falk codeviance* $\tilde{C}_{XY}$ replaces the mean-based average in the covariance with the median of paired deviations from each variable's median (where $\tilde{M}_{X}$ and $\tilde{M}_{Y}$ are the sample medians):
$$\tilde{C}_{XY} = \text{Med}\left\{ (\hat{X}_{i} - \tilde{M}_{X})(\hat{Y}_i - \tilde{M}_{Y}) \right\}.$$

The *median correlation* $\tilde{R}_{XY}$ rescales the codeviance by the product of the two median absolute deviations:
$$\tilde{R}_{XY} = \frac{ \tilde{C}_{XY} }{ \hat{\text{MAD}}_{X} \hat{\text{MAD}}_{Y}}.$$
:::

The Falk codeviance is useful for cardinal data with outliers or heavy tails (income, prices, response times), where one extreme observation can move $\hat{C}_{XY}$ by a lot but $\tilde{C}_{XY}$ by very little (the same robustness rationale as for the median, $IQR$, and $\text{MAD}$ from Part 1).^[See <https://jadamso.github.io/Rbooks/01_03_DescriptiveStatistics.html#medians-and-absolute-deviations>. See also the *Theil-Sen Estimator*, which may be seen as a precursor.] Unlike the Pearson correlation, the median correlation $\tilde{R}_{XY}$ typically lies in $[-1,1]$ but not always.
```{r}
codev <- function(xy) {
  # Compute medians for each column
  med <- apply(xy, 2, median)
  # Subtract the medians from each column
  xm <- sweep(xy, 2, med, '-')
  # Compute CoDev
  CoDev <- median(xm[, 1] * xm[, 2])
  # Compute the medians of absolute deviation
  MadProd <- prod( apply(abs(xm), 2, median) )
  # Return the robust correlation measure
  return( CoDev / MadProd)
}
xy_codev <- codev(xy)
xy_codev
```

:::{.callout-note icon=false collapse="true" title="Must Know"}
Compute the Codeviance for the dataset $\{(1,2),(2,1),(3,4),(4,3),(5,6)\}$.

The medians are $\tilde{M}_{X}=3$ and $\tilde{M}_{Y}=3$, giving signed deviations and products

\begin{array}{c|rrrrr}
\hat{X}_{i}-\tilde{M}_{X} & -2 & -1 & 0 & 1 & 2 \\
\hat{Y}_{i}-\tilde{M}_{Y} & -1 & -2 & 1 & 0 & 3 \\
(\hat{X}_{i}-\tilde{M}_{X})(\hat{Y}_{i}-\tilde{M}_{Y}) & 2 & 2 & 0 & 0 & 6
\end{array}

The Codeviance is the median of those products: $\tilde{C}_{XY}=\text{Med}\{0,0,2,2,6\}=2$.
The median absolute deviations are $\hat{\text{MAD}}_{X}=\text{Med}\{0,1,1,2,2\}=1$ and $\hat{\text{MAD}}_{Y}=\text{Med}\{0,1,1,2,3\}=1$, so the median correlation is $\tilde{R}_{XY}=2/(1\cdot1)=2$.
Here $\tilde{R}_{XY}$ exceeds $1$: unlike Pearson's correlation, the median correlation is not confined to $[-1,1]$.
```{r}
xy_small <- cbind(X=c(1,2,3,4,5), Y=c(2,1,4,3,6))
codev(xy_small)
```
:::

You construct sampling distributions and conduct hypothesis tests for Falk's Codeviance statistic in the same way you do for Pearson's Correlation statistic.

:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
```{r}
xy <- USArrests[, c('Murder', 'UrbanPop')]
xy_cor <- cor(xy[, 1], xy[, 2])
#plot(xy, pch=16, col=grey(0, .25))
    
# Null Permutation Distribution of Codeviance
n <- nrow(xy)
null_permutation_codev <- rep(NA, 9999)
for(b in seq(null_permutation_codev) ){
    xy_b <- xy
    xy_b[, 'UrbanPop'] <- xy[sample(n, replace=F), 'UrbanPop'] ## Reshuffle X
    xy_codev_b <- codev(xy_b)
    null_permutation_codev[b] <- xy_codev_b
}
hist(null_permutation_codev, breaks=100,
    border=NA, freq=F,
    xlab='Codeviance',
    main=NA)
title('Null Permutation Distribution', font.main=1)

## Test whether correlation is statistically different from 0
abline(v=quantile(null_permutation_codev, probs=c(0.025, 0.975)))
abline(v=xy_codev, col=rgb(0, 0, 1, .8))
```
:::


## Factor Data

#### **Two Ordered Factors**. {-}

When the data are ordered (rankings or ordered categories), we can summarize association by counting which pairs of observations agree in direction.

:::{.callout-important icon=false title="Key Definition" collapse="false"}
*Kendall's rank correlation* $\hat{KT}$ is the share of variable-pairs that agree in direction (concordant) minus the share that disagree (discordant):
$$\hat{KT} = \frac{2}{n(n-1)} \sum_{i} \sum_{j > i} \text{sgn} \Bigl( (\hat{X}_{i} - \hat{X}_{j})(\hat{Y}_i - \hat{Y}_j) \Bigr),$$
where the sign function is
$$\text{sgn}(z) = \begin{cases} +1 & \text{if } z > 0\\ 0  & \text{if } z = 0 \\ -1 & \text{if } z < 0 \end{cases}.$$
:::

Kendall's $\hat{KT}$ is useful for any *monotone* relationship: the values do not have to fall on a straight line, only have consistent ordering. Because it only uses signs, it works for any ordinal data (rankings, ordered categories) where arithmetic averages are not meaningful. The bounds $[-1, 1]$ make it directly comparable to the Pearson correlation: a value closer to $+1$ suggests positive association in rankings, $-1$ negative, and $0$ no association in the ordering.
```{r}
xy <- USArrests[, c('Murder', 'UrbanPop')]
xy[, 1] <- rank(xy[, 1] )
xy[, 2] <- rank(xy[, 2] )
# plot(xy, pch=16, col=grey(0, .25))
KT <- cor(xy[, 1], xy[, 2], method = 'kendall')
round(KT, 3)
```

:::{.callout-note icon=false collapse="true" title="Must Know"}
Compute Kendall's $\hat{KT}$ for the dataset $\{(1,1),(2,3),(3,2),(4,4)\}$.

With $n=4$ there are $n(n-1)/2=6$ pairs. For each pair we take the sign of $(\hat{X}_{i}-\hat{X}_{j})(\hat{Y}_{i}-\hat{Y}_{j})$: a pair is *concordant* ($+1$) if both variables move the same way, *discordant* ($-1$) if they move oppositely.

| Pair | $X$ moves | $Y$ moves | sign |
|:---:|:---:|:---:|:---:|
| (1,1),(2,3) | up | up   | $+1$ |
| (1,1),(3,2) | up | up   | $+1$ |
| (1,1),(4,4) | up | up   | $+1$ |
| (2,3),(3,2) | up | down | $-1$ |
| (2,3),(4,4) | up | up   | $+1$ |
| (3,2),(4,4) | up | up   | $+1$ |

There are $5$ concordant and $1$ discordant pair, so
\begin{eqnarray}
\hat{KT} = \frac{2}{n(n-1)}\sum_{i}\sum_{j>i}\text{sgn}(\cdots) = \frac{2}{4\cdot3}(5-1) = \frac{8}{12} = \frac{2}{3} \approx 0.67.
\end{eqnarray}
```{r}
cor(c(1,2,3,4), c(1,3,2,4), method='kendall')
```
:::

You construct sampling distributions and conduct hypothesis tests for Kendall's rank correlation statistic in the same way you do as for Pearson's Correlation statistic and Falk's Codeviance statistic.

:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
Test whether Kendal's correlation statistic is statistically different from $0$. Expand on the example below to use bootstrapping.
```{r}
xy <- USArrests[, c('Murder', 'UrbanPop')]
KT <- cor(xy[, 1], xy[, 2], method='kendall')
```
:::

Kendall's rank correlation coefficient can also be used for non-linear relationships, where Pearson's correlation coefficient often falls short. It almost always helps to visual your data first before summarizing it with a statistic.

```{r, results='hide', echo=F}
#| fig-width: 8
#| fig-height: 8
## https://commons.wikimedia.org/wiki/File:Correlation_examples2.svg

MyPlot <- function(xy, xlim = c(-4, 4), ylim = c(-3, 3)) {
   plot(xy, main = NA, xlab = '', ylab = '',
        col = grey(0, .25), pch=16, cex=.5,
        xaxt = 'n', yaxt = 'n', bty = 'n',
        xlim = xlim, ylim = ylim)
    box(lwd=.1)
    
    cor1 <- cor(xy[, 1], xy[, 2])
    cor2 <- codev(xy)
    cor3 <- cor(xy[, 1], xy[, 2], method='kendall')
    cors <- c(cor1, cor2, cor3)
    
    #cor3 <- abs(generalCorr::gmcmtx0( xy[, c(1, 2)])[2])
    #cor4 <- Rfast::dcor( xy[, 1], xy[, 2])[[4]]
    #cor5 <- XICOR::xicor( xy[, 1], xy[, 2]) ## cor( xy[, 1], xy[, 2], method='kendall') 

    fm <- cors*0 + 1
    fm[which.max(abs(cors))] <- 2
    title(paste0('Pearson:  ',  formatC(cor1, digits=2, format='f')),
        line=3, adj=0, font.main=fm[1])
    title(paste0('Falk:     ',  formatC(cor2, digits=2, format='f')),
        line=2, adj=0, font.main=fm[2])    
    title(paste0('Kendall:  ',  formatC(cor3, digits=2, format='f')),
        line=1, adj=0, font.main=fm[3])
}

MvNormal <- function(n = 1000, cor, f) {
   for (i in cor) {
      sd = matrix(c(1, i, i, 1), ncol = 2)
      x = mvtnorm::rmvnorm(n, c(0, 0), sd)
      x[, 2] <- f(x[, 2])
      MyPlot(x)
   }
}

output <- function() {
   par(mfrow = c(3, 2), oma = c(0, 0, 0, 0), mar=c(3, 2, 6, 1))
   cor <- c(0.99, 0.9)
   MvNormal(800, cor, function(y){y});
   MvNormal(800, cor, function(y){(y/2)^3});
   MvNormal(800, cor, function(y){log( (y-min(y)+.Machine[['double.eps']])/(max(y)-min(y)) ) });
}
output()
```



#### **Two Unordered Factors**. {-}

When neither variable has a meaningful order, we organize the data as a contingency table and summarize the strength of association with a single bounded score.

:::{.callout-important icon=false title="Key Definition" collapse="false"}
*Cramer's V* $\hat{CV}$ rescales the *chi-squared statistic* $\hat{\chi}^{2}$ into a $[0, 1]$ score for the strength of association between two categorical variables in a $K\times J$ contingency table:
$$\hat{\chi}^2 = \sum_{k=1}^{K} \sum_{j=1}^{J} \frac{(\hat{O}_{kj} - \hat{E}_{kj})^2}{\hat{E}_{kj}}, \qquad \hat{CV} = \sqrt{\frac{\hat{\chi}^2 / n}{\min(J - 1, \, K - 1)}},$$
where $\hat{O}_{kj}$ is the observed frequency in cell $(k, j)$, $\hat{E}_{kj} = \hat{RF}_{k} \cdot \hat{CF}_{j} / n$ is the expected frequency under independence, and $\hat{RF}_{k}=\sum_{j} \hat{O}_{kj}$, $\hat{CF}_{j}=\sum_{k} \hat{O}_{kj}$ are the row and column totals.
:::

Cramer's V is useful for two unordered categorical variables (occupation, region, brand), because it treats every reordering of rows or columns the same and makes no use of any ordering. The chi-squared piece measures how far observed counts are from what we would expect if $\hat{X}$ and $\hat{Y}$ were independent, and dividing by $n \cdot \min(J - 1, K - 1)$ rescales it into a bounded score so different tables are comparable: $0$ suggests no association, and a value closer to $1$ suggests a strong association.

```{r}
xy <- USArrests[, c('Murder', 'UrbanPop')]
xy[, 1] <- cut(xy[, 1], 3)
xy[, 2] <- cut(xy[, 2], 4)
table(xy)

CV <- function(xy){
    # Create a contingency table from the categorical variables
    tbl <- table(xy)
    # Compute the chi-square statistic (without Yates' continuity correction)
    chi2 <- chisq.test(tbl, correct=FALSE)[['statistic']]
    # Total sample size
    n <- sum(tbl)
    # Compute the minimum degrees of freedom (min(rows-1, columns-1))
    df_min <- min(nrow(tbl) - 1, ncol(tbl) - 1)
    # Calculate Cramer's V
    V <- sqrt((chi2 / n) / df_min)
    return(V)
}
CV(xy)

# DescTools::CramerV( table(xy) )
```

:::{.callout-note icon=false collapse="true" title="Must Know"}
Compute Cramer's V for this $2\times2$ table of $100$ workers, cross-classifying a college degree against employment.

\begin{array}{c|cc|c}
 & \text{employed} & \text{unemployed} & \text{Row total}\\
\hline
\text{degree}    & 30 & 10 & 40\\
\text{no degree} & 20 & 40 & 60\\
\hline
\text{Column total} & 50 & 50 & 100
\end{array}

If degree and employment were independent, the expected count in each cell is $\hat{E}_{kj}=\hat{RF}_{k}\hat{CF}_{j}/n$. For the top-left cell, $\hat{E}_{11}=40\cdot50/100=20$; all four expected counts are $(20,20,30,30)$.
\begin{eqnarray}
\hat{\chi}^2 &=& \frac{(30-20)^2}{20}+\frac{(10-20)^2}{20}+\frac{(20-30)^2}{30}+\frac{(40-30)^2}{30} = 5+5+3.33+3.33 = 16.67.
\end{eqnarray}
With $K=2$ rows and $J=2$ columns, $\min(J-1,K-1)=1$, so
\begin{eqnarray}
\hat{CV} = \sqrt{\frac{\hat{\chi}^2/n}{\min(J-1,K-1)}} = \sqrt{\frac{16.67/100}{1}} \approx 0.41.
\end{eqnarray}
```{r}
# Observed 2x2 table
O <- rbind(degree    = c(employed=30, unemployed=10),
           no_degree = c(20, 40))
n <- sum(O)

# Expected counts under independence
E <- outer(rowSums(O), colSums(O)) / n

# Chi-square statistic and Cramer's V
chi2 <- sum((O - E)^2 / E)
df_min <- min(nrow(O)-1, ncol(O)-1)
c(chi2=chi2, CramerV=sqrt((chi2/n)/df_min))
```
:::

You construct sampling distributions and conduct hypothesis tests for  Cramer's V statistic in the same way you do as the other statistics.

## Association is not Causation


One major issue is statistical significance: sometimes there are relationships in the population that do not show up in samples, or non-relationships that appear in samples. This should be familiar at this point.

Other major issues pertain to a) real relationships averaging out and b) mechanically inducing relationships with random data. To be concrete about these issue, I will focus on the most-used correlation statistic and examine

* Causation without correlation
* Correlation without causation


#### **Causation without correlation**. {-}

Examples of this first issue includes *nonlinear effects* and  *heterogeneous effects* averaging out.

```{r}
set.seed(123)
n <- 10000

# X causes Y via Y = X^2 + noise
X <- runif(n, min = -1, max = 1)
epsilon <- rnorm(n, mean = 0, sd = 0.1)
Y <- X^2 + epsilon  # clear causal effect of X on Y
plot(X, Y, pch=16, col=grey(0, .05),
    main=NA, xlab='X', ylab='Y')

# Correlation over the entire range
title( paste0('Cor: ', round( cor(X, Y), 1) ) ,
    font.main=1)
```

```{r}
# Heterogeneous Effects
X <- rnorm(n) # Randomized 'treatment'

# Heterogeneous effects based on group
group <- rbinom(n, size = 1, prob = 0.5)
epsilon <- rnorm(n, mean = 0, sd = 1)
Y <- ifelse(group == 1,
            X + epsilon,   # positive effect
            -X + epsilon)  # negative effect
plot(X, Y, pch=16, col=grey(0, .05),
    main=NA, xlab='X', ylab='Y')

# Correlation in the pooled sample
title( paste0('Cor: ', round( cor(X, Y), 1) ),
    font.main=1 )
```


#### **Correlation without causation**. {-}
Examples of this second issue includes *shared denominators* and  *selection bias* inducing correlations.

Consider three completely random variables. We can induce a mechanical relationship between the first two variables by dividing them both by the third variable.

```{r}
set.seed(123)
n <- 20000

# Independent components
A <- runif(n)
B <- runif(n)
C <- runif(n)
par(mfrow=c(1, 2))
plot(A, B, pch=16, col=grey(0, .05),
    main=NA, xlab='A', ylab='B')
title('Independent Variables', font.main=1)

# Ratios with a shared denominator
X <- A / C
Y <- B / C
plot(X, Y, pch=16, col=grey(0, .05),
    main=NA, xlab='X', ylab='Y')
title('With Common Divisor', font.main=1)

# Correlation
cor(X, Y)
```

Consider an admissions rule into university: applicants are accepted if they have either high test scores or strong extracurriculars. Even if there is no general relationship between test scores and extracurriculars, you will see one amongst university students.

```{r}

# Independent traits in the population
test_score        <- rnorm(n, mean = 0, sd = 1)
extracurriculars  <- rnorm(n, mean = 0, sd = 1)

# Selection above thresholds
threshold <- 1.0
admitted <- (test_score > threshold) | (extracurriculars > threshold)
mean(admitted)  # admission rate

par(mfrow = c(1, 2))
# Full population
plot(test_score, extracurriculars,
     pch=16, col=grey(0, .05),
     main=NA, xlab='Test Score', ylab='Extracurriculars')
title('General Sample', font.main=1)
# Admitted only
plot(test_score[admitted], extracurriculars[admitted],
     pch=16, col=grey(0, .05),
     main=NA, xlab='Test Score', ylab='Extracurriculars')
title('University Sample', font.main=1)
# Correlation among admitted applicants only
cor(test_score[admitted], extracurriculars[admitted])
```


:::{.callout-note icon=false collapse="true" title="Must Know"}
In the dataset below, compute the correlation statistic and comment on whether any relationship is causal. 
```{r}
plot(Murder ~ Assault, USArrests, pch=16, col=grey(0, .5),
    main=NA, xlab='Assault Arrests', ylab='Murder Arrests')
```
:::

Note that the examples above are not the only examples of "correlation does not mean causation". Many real datasets have temporal and spatial interdependence that create additional issues. Many real datasets also have economic interdependence, which also creates additional issues. We will delay covering these issues until much later.



## Exercises

1. Give an example of a dataset where Pearson's correlation $\hat{R}_{XY}$ is close to zero but there is a clear causal relationship between $X$ and $Y$. What feature of the relationship causes the correlation to miss it, and which alternative statistic from this chapter might detect it?

2. For the dataset $\{(1, 5),\ (2, 3),\ (3, 4),\ (4, 2),\ (5, 1)\}$, compute the Pearson correlation $\hat{R}_{XY}$ by hand. Show all five steps: means, deviances, covariance $\hat{C}_{XY}$, standard deviations $\hat{S}_{X}$ and $\hat{S}_{Y}$, and the final correlation.

3. Using the `USArrests` dataset, compute both the Pearson correlation and Kendall's rank correlation between `Murder` and `Assault`. Then write a bootstrap with $B = 9999$ iterations to construct a $95\%$ confidence interval for the Pearson correlation. Does the interval contain zero?

#### **Further Reading**. {-}

* <https://www.econometrics-with-r.org/2-2-an-application-to-the-gender-gap-of-earnings.html> -- worked example of computing and interpreting correlation in an economics setting.
* <https://learningstatisticswithr.com/book/regression.html#correlations> -- clear introduction to Pearson, Spearman, and other correlation measures with R code.
* <https://www.khanacademy.org/math/statistics-probability/describing-relationships-quantitative-data/scatterplots-and-correlation/v/correlation-and-causality> -- short video on why correlation does not imply causation.

#### **Recall** {-}

This chapter summarized two-variable association with a small toolbox matched to data type: Pearson correlation $\hat{R}_{XY}$ for cardinal data (worked out by hand on $\{(0, 0.1), (1, 0.3), (2, 0.2)\}$, yielding $\hat{R}_{XY}=1/2$), Falk codeviance $\tilde{C}_{XY}$ as the robust cousin for outlier-prone data, Kendall's $\hat{KT}$ for ordered factors (walked through the six pairs of $\{(1, 1), (2, 3), (3, 2), (4, 4)\}$ to get $\hat{KT}=2/3$), and Cramer's V $\hat{CV}$ for unordered factors (the $2\times 2$ degree-vs-employment table giving $\hat{CV}\approx 0.41$). We closed with simulations showing that association is not causation. The next chapter takes the cardinal case and fits a line, which moves from "how strongly do $X$ and $Y$ move together" to "by how much does $Y$ change per unit of $X$".
