
# Bivariate Distributions
***

We will now study two variables. The data for each observation data can be grouped together as a vector $(\hat{X}_{i}, \hat{Y}_{i})$.
```{r}
# USArrests has 50 rows (one per US state) and four crime columns;
# pick two of those columns to form a bivariate dataset.
xy <- USArrests[, c('Murder', 'UrbanPop')]
xy[1, ]                                # first state's two values
```

## Joint Distributions

#### **Discrete Data**.{-}

We have multiple observations of $(\hat{X}_{i}, \hat{Y}_{i})$, each of which corresponds to a unique value $(x,y)$. A *frequency table* shows how often each combination of values appear. The *joint distribution* counts up the number of pairs with the same values and divides by the number of observations, $n$;
\begin{eqnarray}
\hat{p}_{xy} = \sum_{i=1}^{n}\mathbf{1}\left( \hat{X}_{i}=x, \hat{Y}_{i}=y \right)/n,
\end{eqnarray}

:::{.callout-important title="Definition" collapse="false"}
The *joint distribution* of two discrete variables $X_{i}$ and $Y_{i}$ is the table $\hat{p}_{xy}$ giving the share of observations with $X_{i}=x$ and $Y_{i}=y$, for every combination of values $(x,y)$.
:::

The *marginal distributions* are just the univariate information, which can be computed independently (see [here](https://jadamso.github.io/Rbooks/01_06_PopStatistics.html#estimates) ) and also from the joint distribution
\begin{eqnarray}
\hat{p}_{x} = \sum_{y} \hat{p}_{xy}
\end{eqnarray}

:::{.callout-important title="Definition" collapse="false"}
The *marginal distribution* of $X_{i}$ is the univariate distribution $\hat{p}_{x}$ obtained by summing the joint distribution over all values of $Y_{i}$. It tells you the share of observations at each value of $X_{i}$, ignoring $Y_{i}$.
:::

For example, suppose we observe a sample of $n=13$ students with two discrete variables:

* $X_{i}$ depicts years of education, taking values in $\{12, 14\}$
* $Y_{i}$ depicts sex, where $y=1$ means female and $y=0$ means male

Assume the count data are summarized by the following table:

\begin{array}{c|cc|c}
 & y=0\ (\text{male}) & y=1\ (\text{female})\\
\hline
x=12 & 4 & 3 \\
x=14 & 1 & 5 \\
\hline
\end{array}

The joint distributions divides each cell count by the total number of observations to obtain
\begin{array}{c|cc}
 & y=0 & y=1\\
\hline
x=12 & 4/13 & 3/13 \\
x=14 & 1/13 & 5/13
\end{array}
The marginal distribution of $X_{i}$ is
\begin{eqnarray}
\hat{p}_{x=12} &=& \frac{4}{13} + \frac{3}{13} = \frac{7}{13},\\
\hat{p}_{x=14} &=& \frac{1}{13} + \frac{5}{13} = \frac{6}{13}.
\end{eqnarray}
The marginal distribution of $Y_{i}$ is
\begin{eqnarray}
\hat{p}_{y=0} &=& \frac{4}{13}  + \frac{1}{13} = \frac{5}{13} \approx 0.38,\\
\hat{p}_{y=1} &=& \frac{3}{13}  + \frac{5}{13} = \frac{8}{13} \approx 0.62.
\end{eqnarray}

Together, this yields a frequency table with marginals
\begin{array}{c|cc|c}
 & y=0\ (\text{male}) & y=1\ (\text{female}) & \text{Row total}\\
\hline
x=12 & 4/13 & 3/13 & 7/13\\
x=14 & 1/13 & 5/13 & 6/13\\
\hline
\text{Column total} & 5/13 & 8/13 & 1
\end{array}

```{r}
# Simple hand-built dataset: 13 students, two variables each
dat <- rbind(
    c(12, 0), c(12, 0), c(12, 0), c(12, 0), # 4 male with 12 years
    c(14, 0),                                # 1 male with 14 years
    c(12, 1), c(12, 1), c(12, 1),            # 3 female with 12 years
    c(14, 1), c(14, 1), c(14, 1), c(14, 1), c(14, 1)) # 5 female with 14 years
colnames(dat) <- c('educ', 'female')
dat <- as.data.frame(dat)

# Frequency table and joint distribution
tab  <- table(dat)          # raw counts: rows=educ, cols=female
prop <- tab / sum(tab)      # joint distribution p_xy (sums to 1)

# addmargins() appends row and column totals (i.e., the marginals)
prop_m <- addmargins(prop)
round(prop_m, 2)
```

Note that you can compute various univariate statistics using the marginal distributions. E.g., the mean of $\hat{X}_{i}$ is $\hat{M}_{X}=\sum_{x} x \hat{p}_{x}$. In the the above example, that equals $12 \frac{7}{13} + 14\frac{6}{13} \approx 12.9$. Try computing the standard deviation $\hat{S}_{X}$ yourself.
```{r}
# marginals and univariate statistics
p_x <- margin.table(prop, 1)
sum(p_x* as.numeric(names(p_x)) )
mean(dat[, 1])

# p_y <- margin.table(prop, 2)
```



:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
Summarize the distribution of sex and education attainment for highschool students
```{r}
library(Ecdat)

# Highschool Students, Sex and Education Attainment
dat <- Wages1[, c('school', 'sex')]
dat <- dat[ dat[, 'school'] >=10, ]
```
Try it yourself before looking at this answer
```{r}
tab <- table(dat) ## counts
tab

prop <- tab/sum(tab) ## frequencies
round(prop, 3)

prop_full <- addmargins(prop) ## column and row sums
round(prop_full, 3)
```
:::


#### **Continuous Data**.{-}

*Scatterplots* are used frequently to summarizes the joint distribution of continuous data. They can be enhanced in several ways. As a default, use semi-transparent points so as not to hide any points (and perhaps see if your observations are concentrated anywhere). You can also add other features that help summarize the relationship, although I will defer this until later.

```{r}
plot(Murder ~ UrbanPop, USArrests, pch=16, col=grey(0, .5),
    main=NA, xlab='Urban Population', ylab='Murder Arrests')
```

You can also show the marginal distributions of each variable along each axis (see this [margin and layout cheatsheet](https://r-graph-gallery.com/74-margin-and-oma-cheatsheet.html) for how `layout()`, `mar`, and `oma` work together).

```{r}
# Setup Plot
layout( matrix(c(2, 0, 1, 3), ncol=2, byrow=TRUE),
    widths=c(9/10, 1/10), heights=c(1/10, 9/10))

# Scatterplot
par(mar=c(4, 4, 1, 1))
plot(Murder ~ UrbanPop, USArrests, pch=16, col=grey(0, .5),
    main=NA, xlab='Urban Population', ylab='Murder Arrests')

# Add Marginals
par(mar=c(0, 4, 1, 1))
xhist <- hist(USArrests[, 'UrbanPop'], plot=FALSE)
barplot(xhist[['counts']], axes=FALSE, space=0, border=NA)

par(mar=c(4, 0, 1, 1))
yhist <- hist(USArrests[, 'Murder'], plot=FALSE)
barplot(yhist[['counts']], axes=FALSE, space=0, horiz=TRUE, border=NA)
```

## Conditional Distributions

#### **Discrete Data**.{-}

In contrast to marginal and joint distributions, *conditional* distributions describe how the distribution of one variable changes when we restrict attention to a subgroup defined by another variable. Formally, if $\hat{p}_{xy}$ denotes the empirical joint distribution, then the empirical conditional distribution of $Y_{i}$ given $X_{i}=x$ is
\begin{eqnarray}
\hat{p}_{y \mid x} = \frac{\hat{p}_{xy}}{\hat{p}_{x} },
\end{eqnarray}
where $\hat{p}_{x} = \sum_y \hat{p}_{xy}$.

:::{.callout-important title="Definition" collapse="false"}
The *conditional distribution* of $Y_{i}$ given $X_{i}=x$ is $\hat{p}_{y\mid x}=\hat{p}_{xy}/\hat{p}_{x}$: the joint share renormalized so the row at $X_{i}=x$ sums to one. It tells you the share of values of $Y_{i}$ *within the subgroup* where $X_{i}=x$.
:::

For example, suppose we observe a sample of $n=13$ students with two discrete variables:

* $X_{i}$ depicts years of education, taking values in $\{12, 14\}$
* $Y_{i}$ depicts sex, where $y=1$ means female and $y=0$ means male

Assume the data are summarized by the following joint distribution table:

\begin{array}{c|cc|c}
 & y=0\ (\text{male}) & y=1\ (\text{female}) & \text{Row total}\\
\hline
x=12 & 4/13 & 3/13 & 7/13\\
x=14 & 1/13 & 5/13 & 6/13\\
\hline
\text{Column total} & 5/13 & 8/13 & 1
\end{array}

To compute the conditional distribution of $Y_{i}$ given $X_{i}=x$, we rescale the numbers in each row so that each row sums to 1.
Specifically, for $x=12$, we compute
\begin{eqnarray}
\hat{p}_{y=0 \mid x=12}
&=& \frac{\hat{p}_{x=12, y=0}}{\hat{p}_{x=12}}
= \frac{4/13}{7/13}
= \frac{4}{7} \\
\hat{p}_{y=1\mid x=12}
&=& \frac{\hat p_{x=12, y=1}}{\hat{p}_{x=12}}
= \frac{3/13}{7/13}
= \frac{3}{7}
\end{eqnarray}
Similarly, for $x=14$, we compute
\begin{eqnarray}
\hat{p}_{y=0 \mid x=14}
&=& \frac{\hat{p}_{x=14, y=0}}{\hat{p}_{x=14}}
= \frac{1/13}{6/13}
= \frac{1}{6}. \\
\hat{p}_{y=1\mid x=14}
&=& \frac{\hat p_{x=14, y=1}}{\hat{p}_{x=14}}
= \frac{5/13}{6/13}
= \frac{5}{6}.
\end{eqnarray}
Together, we find
\begin{array}{c|cc|c}
 & y=0\ (\text{male}) & y=1\ (\text{female}) & \text{Row total}\\
\hline
x=12 & 4/7 & 3/7 & 7/7\\
x=14 & 1/6 & 5/6 & 6/6\\
\hline
\end{array}

In this example, we say that

* conditional on students having $12$ years of education, $\frac{4}{7}\approx 57\%$ are male and $\frac{3}{7}\approx 43\%$ are female.
* conditional on students having $14$ years of education, $\frac{1}{6} \approx 17\%$ are male and $\frac{5}{6} \approx 83\%$ are female.


We can also compute the conditional distribution of $X_{i}$ given $Y_{i}=y$, $\hat{p}_{x \mid y} = \hat{p}_{xy} / \hat{p}_{y}$,just as we did above. In this case, we rescale the numbers in each column of the joint distribution table so that each column sums to 1.

:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
Continuing the example above, show that

* among male students, $\approx 80\%$ have 12 years of education.
* among female students, $\approx 38\%$ have 12 years of education.

Try programming the results, especially if you are stuck or uncertain

```{r}
dat <- rbind(
    c(12, 0), c(12, 0), c(12, 0), c(12, 0), #Male
    c(14, 0), 
    c(12, 1), c(12, 1), c(12, 1),
    c(14, 1), c(14, 1), c(14, 1), c(14, 1), c(14, 1))
colnames(dat) <- c('educ', 'female')
dat <- as.data.frame(dat)

tab <- table(dat)
tab

# Joint distribution
round(prop.table(tab), 3)

# Conditional distribution of Y given X
round(prop.table(tab, 1), 3)

# Conditional distribution of X given Y
round(prop.table(tab, 2), 3)
```
:::

Conditional distributions change the unit of analysis:

* $\hat{p}_{y=1 \mid x=14}$ answers "what fraction of students are female within the subgroup with 14 years of education?"
* $\hat{p}_{x=14 \mid y=1}$ answers "what fraction of students have 14 years of education within the subgroup that is female?"

Mixing up these two directions is a common mistake known as the *base-rate fallacy*.

:::{.callout-important title="Definition" collapse="false"}
The *base-rate fallacy* is the mistake of confusing $\hat{p}_{y\mid x}$ with $\hat{p}_{x\mid y}$ — reading the wrong conditional — and forgetting how common $X_{i}=x$ is in the first place ($\hat{p}_{x}$, the *base rate*).
:::

:::{.callout-note icon=false collapse="true" title="Must Know"}
**The most-stolen car.** Suppose you read that the Honda Civic is the most-stolen car in a city. Does that mean a Civic is more likely to be stolen than other cars? Not necessarily.

Imagine a city with $100{,}000$ cars: $70{,}000$ Civics and $30{,}000$ of other models. Let $X_{i}$ be the model and $Y_{i}$ whether car $i$ was stolen. The counts are

\begin{array}{c|cc|c}
 & \text{not stolen} & \text{stolen} & \text{Row total}\\
\hline
\text{Civic} & 69300 & 700 & 70000\\
\text{other} & 29460 & 540 & 30000\\
\hline
\text{Column total} & 98760 & 1240 & 100000
\end{array}

More Civics are stolen than other cars ($700 > 540$), so the Civic is the "most-stolen" model. But the conditional distributions tell two different stories:

* $\hat{p}_{\text{stolen}\mid\text{Civic}} = \frac{700}{70000} = 0.010$ is the theft rate for a Civic.
* $\hat{p}_{\text{stolen}\mid\text{other}} = \frac{540}{30000} = 0.018$ is the theft rate for other cars.
* $\hat{p}_{\text{Civic}\mid\text{stolen}} = \frac{700}{1240} \approx 0.565$ is the share of stolen cars that are Civics.

A Civic is *less* likely to be stolen ($1.0\%$ vs $1.8\%$), yet most stolen cars are Civics. The reason is the *base rate*: $\hat{p}_{\text{Civic}} = 0.70$ of all cars are Civics. "Most-stolen" is a statement about $\hat{p}_{x\mid y}$ (the model given theft), while car-theft risk is a statement about $\hat{p}_{y\mid x}$ (theft given the model).

```{r}
# Count table: rows = model, columns = stolen
counts <- rbind(
    Civic = c(not=69300, stolen=700),
    other = c(not=29460, stolen=540))

# Theft rate by model: p(stolen | model)
counts[, 'stolen'] / rowSums(counts)

# Share of stolen cars by model: p(model | stolen)
counts[, 'stolen'] / sum(counts[, 'stolen'])
```
:::

#### **Continuous Data**.{-}

These describe the relationship between $\hat{Y}_{i}$ and $\hat{X}_{i}$. We show how $Y$ changes according to $X$ using a 
[histogram](https://jadamso.github.io/Rbooks/01_02_Data.html#histogram-density) or bar plot. When $X$ is continuous, as it often is, we split it into distinct bins and convert it to a factor variable. E.g.,  

```{r}
# Split Data by Urban Population above/below mean
pop_mean <- mean(USArrests[, 'UrbanPop'])
pop_cut <- USArrests[, 'UrbanPop'] <= pop_mean
murder_lowpop <- USArrests[pop_cut, 'Murder']
murder_highpop <- USArrests[!pop_cut, 'Murder']
cols <- c(low=rgb(0, 0, 1, .75), high=rgb(1, 0, 0, .75))

# Common Histogram 
ylim <- c(0, .25)
xbks <-  seq(min(USArrests[, 'Murder'])-1, max(USArrests[, 'Murder'])+1, by=1)

par(mfrow=c(1, 2))
hist(murder_lowpop,
    breaks=xbks, col=cols[1],
    main=NA,
    xlab='Murder Arrests', freq=F,
    border=NA, ylim=ylim)
title('Urban Pop <= Mean', font.main=1)

hist(murder_highpop,
    breaks=xbks, col=cols[2],
    main=NA,
    xlab='Murder Arrests', freq=F,
    border=NA, ylim=ylim)
title('Urban Pop > Mean', font.main=1)
```

It is sometimes preferable to show the [ECDF](https://jadamso.github.io/Rbooks/01_02_Data.html#empirical-cumulative-distribution-function) instead. And you can glue various combinations together to convey more information all at once

```{r}
layout( t(c(1, 2, 2)))
# Full Sample Density
hist(USArrests[, 'Murder'],
    main=NA,
    xlab='Murder Arrests',
    breaks=xbks, freq=F, border=NA)
title('Full Sample Density', font.main=1)

# Split Sample Distribution Comparison
F_lowpop <- ecdf(murder_lowpop)
plot(F_lowpop, col=cols[1],
    pch=16, xlab='Murder Arrests',
    main=NA, bty='n')
title('Split Sample Distributions', font.main=1)
F_highpop <- ecdf(murder_highpop)
plot(F_highpop, add=T, col=cols[2], pch=16)

legend('bottomright', col=cols,
    pch=16, bty='n', inset=c(0, .1),
    title='% Urban Pop.',
    legend=c('Low (<= Mean)', 'High (> Mean)'))
```

```{r, message=F}
# Simple Interactive Scatter Plot
# plot(Assault ~ UrbanPop, USArrests, col=grey(0, .5), pch=16,
#    cex=USArrests[, 'Murder']/diff(range(USArrests[, 'Murder']))*2,
#    main='US Murder arrests (per 100, 000)')
```

You can also split data into more than two groups. For more than three groups, boxplots are often more effective than histograms or ECDF's.
```{r}
# K Groups with even spacing (not even counts)
K <- 4
USArrests[, 'UrbanPop_Kcut'] <- cut(USArrests[, 'UrbanPop'], K)
table(USArrests[, 'UrbanPop_Kcut'] )

# Full sample
#boxplot(USArrests[, 'Murder'], main='',
#    xlab='All Data', ylab='Murder Arrests')

# Boxplots for each group
Kcols <- hcl.colors(K, alpha=.5)
boxplot(Murder ~ UrbanPop_Kcut, USArrests,
    main=NA, col=Kcols, border=NA,
    varwidth=T, #show number of obs. per group
    xlab='Urban Population', ylab='Murder Arrests')

# 4 Groups with equal numbers of observations
#Qcuts <- c(
#    '0%'=min(USArrests[, 'UrbanPop'])-10*.Machine[['double.eps']],
#    quantile(USArrests[, 'UrbanPop'], probs=c(.25, .5, .75, 1)))
#USArrests[, 'UrbanPop_cut'] <- cut(USArrests[, 'UrbanPop'], Qcuts)
#boxplot(Murder ~ UrbanPop_cut, USArrests, col=hcl.colors(4, alpha=.5))
```


## Simpson's Paradox

Conditional distributions can be tricky, so take your time making and interpreting them.

Consider the following example, which shows that $400$ men and $400$ women applied to university and that only a share were accepted in each department (English or Engineering). If you computed percents for each sex within each cell, you would find men have a higher overall acceptance rate, but women have higher acceptance rates within each department! This counterintuitive phenomena is known as *Simpson's Paradox*, and the example is part of a real debate about discrimination (<http://homepage.stat.uiowa.edu/~mbognar/1030/Bickel-Berkeley.pdf>).

:::{.callout-important title="Definition" collapse="false"}
*Simpson's paradox* arises when a relationship that holds inside every subgroup reverses when the data are pooled. The within-group conditional distributions $\hat{p}_{y\mid x, g}$ point one way; the overall $\hat{p}_{y\mid x}$ points the other.
:::

Even if women have higher admission rates within both departments, women can still have a lower overall admission rate if women disproportionately apply to the more selective department (English) and men disproportionately apply to the less selective department (Engineering). The same issue is relevant for a variety of labor market issues, such as the gender pay gap, and a great many other social issues, such as why are some countries rich and others poor.^[In general, one should also remember that a ratio can change due changes in either the numerator or the denominator.]


| Department   | Men                  | Women                 | Total                 |
|-------------|-----------------------|-----------------------|-----------------------|
| English     | 100 (40, **$40\%$**)      | 350 (150, **$43\%$**)     | 450 (190, **$42\%$**)     |
| Engineering | 300 (160, **$53\%$**)     | 50 (30, **$60\%$**)       | 350 (190, **$54\%$**)     |
| Total       | 400 (200, **$50\%$**)     | 400 (180, **$45\%$**)     | 800 (380, **$48\%$**)     |

: School Applicants (Admitted), by Sex and Department


:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
Explore this issue further with real data
```{r}
UCBAdmissions
```
:::

The same issue shows up in continuous data. The figure below shows three groups. Within each group the relationship between $X1$ and $X2$ is negative, but the group centers rise together from lower-left to upper-right.
```{r, echo=F}
set.seed(5)
library(mvtnorm)
N <- 400
### Group 1
mu1    <- c(0, 0)
sigma1 <- rbind(c(4, -0.9), c(-0.9, 2) )
g1     <- data.frame(rmvnorm(n=N, mu1, sigma1), Group=1)
### Group 2
mu2    <- c(3, 3)
sigma2 <- rbind(c(2, -0.9), c(-0.9, 1) )
g2     <- data.frame(rmvnorm(n=N, mu2, sigma2), Group=2)
### Group 3
mu3    <- c(6, 6)
sigma3 <- rbind(c(3, -0.9), c(-0.9, 3) )
g3     <- data.frame(rmvnorm(n=N, mu3, sigma3), Group=3)

## Combined Data (Initially Observed)
df_agg <- rbind(g1, g2, g3)
df_agg <- df_agg[order(df_agg$X1), ]

## Scatterplot, colored by group
pt_cols <- c(rgb(.8, 0, 0, .2), rgb(0, .8, 0, .2), rgb(0, 0, .8, .2))
ln_cols <- c(rgb(.8, 0, 0, .8), rgb(0, .8, 0, .8), rgb(0, 0, .8, .8))
plot(X2 ~ X1, df_agg, pch=16, col=pt_cols[df_agg$Group], cex=.8,
    main=NA, xlab='X1', ylab='X2')

## Within-group regression lines (each negative)
for(g in 1:3){
    dg <- df_agg[df_agg$Group==g, ]
    abline(lm(X2 ~ X1, data=dg), col=ln_cols[g], lwd=2)
}
## Pooled regression line (positive)
abline(lm(X2 ~ X1, data=df_agg), lwd=3, lty=2)
legend('topleft', bty='n', cex=.8,
    legend=c('within group', 'pooled'),
    lwd=c(2, 3), lty=c(1, 2), col=c(grey(0, .7), 'black'))
```
Each within-group regression line (solid) slopes downward, yet the pooled regression line (dashed) slopes upward. A researcher who ignored the groups would report a positive relationship that holds within no single group. Fitting lines like these is the subject of [Simple Regression](https://jadamso.github.io/Rbooks/02_13_SimpleRegression.html).

## Exercises

1. Explain the difference between a joint distribution, a marginal distribution, and a conditional distribution. Using the education-and-sex example from this chapter, describe in words what $\hat{p}_{y=1 \mid x=14}$ tells you and how it differs from $\hat{p}_{x=14 \mid y=1}$.

2. Suppose you observe a sample of $n=20$ people with two discrete variables: $X_{i}$ takes values in $\{A, B\}$ and $Y_{i}$ takes values in $\{1, 2\}$. The counts are: $(A,1)=6$, $(A,2)=4$, $(B,1)=3$, $(B,2)=7$. Compute the joint distribution $\hat{p}_{xy}$, both marginal distributions, and the conditional distribution $\hat{p}_{y \mid x}$.

3. Using the `USArrests` dataset, split the variable `Assault` into two groups based on whether `UrbanPop` is above or below its median. Plot overlapping histograms of `Assault` for both groups with semi-transparent colors. Then plot the ECDF for each group on the same axes with a legend.

#### **Further Reading**. {-}

* <https://www.econometrics-with-r.org/2-1-random-variables-and-probability-distributions.html> -- joint, marginal, and conditional distributions in an econometrics context.
* <https://r-graph-gallery.com/histogram.html> -- gallery of histogram styles in R, useful for building marginal-distribution plots.
* <https://www.khanacademy.org/math/statistics-probability/random-variables-stats-library/random-variables-discrete/v/valid-discrete-probability-distribution-examples> -- short video introduction to discrete probability distributions and joint tables.

:::{.callout-note title="Recall" collapse="false" icon=false}
This chapter introduced joint, marginal, and conditional distributions for two variables, and showed how mixing up conditional directions produces the base-rate fallacy and Simpson's paradox. The next chapter uses conditioning on a factor variable to compare two groups, and tests the differences with the resampling tools from Part 1.
:::

