
# Inference
***

Often, we are interested in *gradients*: how $Y$ changes with $X$. The linear model from [Simple Regression](https://jadamso.github.io/Rbooks/02_13_SimpleRegression.html) depicts this as a single constant, $\hat{b}_{1}$, but the local-regression models from [Local Regression](https://jadamso.github.io/Rbooks/02_14_LocalRegression.html) do not. A great first step to assess gradients is to plot the predicted values over the explanatory values. A great second step is to compute gradients and summarize them. In any case, we compute confidence intervals to account for variability across samples.

## Applied Methods

#### **LOESS** {-}

When $\hat{X}_{i}$ is unevenly distributed across its range, a fixed bandwidth $h$ leaves some local fits with many neighbors and others with very few.

:::{.callout-important icon=false title="Key Definition" collapse="false"}
*LOESS* (locally estimated scatterplot smoothing) fits a local linear (or quadratic) regression at each design point using an *adaptive bandwidth*: instead of fixing $h$, it includes a fixed *share* of the data — set by the parameter `span` — at each $x$.
:::

LOESS is useful for smoothing data with uneven $X$: because the *share* of points in each neighborhood is fixed, the local fit has roughly the same effective sample size everywhere, even where the data are sparse. The `span` parameter plays the role $h$ does for fixed-bandwidth methods, so smaller spans mean narrower neighborhoods, more flexibility, and noisier curves, while larger spans pool more data and smooth more aggressively.

```{r}
## Data
library(Ecdat)
dat <- Wages1[, c('wage', 'school')]
dat <- dat[order(Wages1[, 'school']), ]

# Adaptive bandwidth with non-uniform weights
dat <- Wages1[order(Wages1[, 'school']), c('wage', 'school')]
plot(wage ~ school, pch=16, col=grey(0, .1), data=dat,
    main=NA, xlab='School', ylab='Wage')

reg_lo4 <- loess(wage ~ school, data=dat, span=.4)
reg_lo8 <- loess(wage ~ school, data=dat, span=.8)

cols <- hcl.colors(3, alpha=.75)[-3]
lines(dat[, 'school'], predict(reg_lo4),
    col=cols[1], type='o', pch=2)
lines(dat[, 'school'], predict(reg_lo8),
    col=cols[2], type='o', pch=2)

legend('topleft', title='Loess',
    legend=c('span=.4 ', 'span=.8'),
    lty=1, col=cols, cex=.8)
```


#### **Confidence Bands**. {-}

A single fitted curve does not show how much that curve would change in another sample, so we draw a region around it.

:::{.callout-important icon=false title="Key Definition" collapse="false"}
A *confidence band* is a confidence interval drawn at every design point: a region around the fitted curve that captures sample-to-sample variability. A bootstrap confidence band collects the predictions from many resamples and reads off pointwise quantiles at each $x$.
:::

Confidence bands are useful for showing where a fitted curve is well-supported by the data and where another sample could move it. The width is typically wider at the edges of $X$ (few neighbors) and narrower in the dense middle, which is itself a quick visual diagnostic. The same bootstrap that produces the band also produces standard errors for any single design point and for the gradient summaries below.

```{r}
# Loess
reg_lo <- loess(wage ~ school, data=dat, span=.6)

# Predictions at design points
pred_design <- data.frame(school=unique(dat[, 'school']))
preds_lo <- predict(reg_lo, newdata=pred_design)

# Plot
lo_col <- hcl.colors(3, alpha=.5)[2]
plot(wage ~ school, pch=16, col=grey(0, .1), data=dat,
    main=NA, xlab='School', ylab='Wage')
lines(pred_design[, 'school'], preds_lo,
    col=lo_col,
    type='o', pch=2)

# Boot CI
boot_lo <- matrix(NA, nrow=nrow(pred_design), ncol=399)
for (b in 1:399) {
    # xy_i <- xy[-i, ] #jackknife for i in 1:nrow(xy)
    xy_b <- dat[sample(nrow(dat), replace=T), ]
    reg_b <- loess(wage ~ school, data=xy_b, span=.8)
    boot_lo[, b] <- predict(reg_b, newdata=pred_design)
}
boot_cb <- apply(boot_lo, 1, quantile,
    probs=c(.025, .975), na.rm=T)

# Plot CI
polygon(
    c(pred_design[[1]], rev(pred_design[[1]])),
    c(boot_cb[1, ], rev(boot_cb[2, ])),
    col=lo_col,
    border=NA)
```


:::{.callout-note icon=false collapse="true" title="Must Know"}
Construct a bootstrap confidence band for the following loess regression
```{r}
# Adaptive-width subsamples with non-uniform weights
xy <- USArrests[, c('UrbanPop', 'Murder')]
xy0 <- xy[order(xy[, 'UrbanPop']), ]
names(xy0) <- c('x', 'y')

plot(y ~ x, pch=16, col=grey(0, .5), data=xy0,
    main=NA, xlab='Urban Population', ylab='Murder Arrests')
reg_lo2 <- loess(y ~ x, data=xy0, span=.6)

red_col <- rgb(1, 0, 0, .5)
lines(xy0[, 'x'], predict(reg_lo2),
    col=red_col, type='o', pch=2)
```
:::


#### **Gradients**.{-}

Once the fitted curve can bend, "the slope" is no longer a single number; we want the slope locally at each $x$.

:::{.callout-important icon=false title="Key Definition" collapse="false"}
A *gradient* $\hat{b}_{1}(x)$ at the design point $x$ is the rate of change of the fitted value $\hat{y}(x)$ with respect to $\hat{X}_{i}$ in a neighborhood of $x$.
:::

The gradient is useful as the local analog of the simple-regression slope: for a globally linear model it equals the single coefficient $\hat{b}_{1}$ everywhere; for a local model it varies with $x$ and traces out how the marginal relationship changes across the range of $X$. Three common ways to summarize gradients are:

1. For all methods, including regressograms, you can approximate gradients with small finite differences. For some small difference $d$, we can manually compute
\begin{eqnarray}
\hat{b}_{1}(x) &=& \frac{ \hat{y}(x+\frac{d}{2}) - \hat{y}(x-\frac{d}{2})}{d},
\end{eqnarray}

2. When using split-sample regressions or local linear regressions, you can use the estimated slope coefficients $\hat{b}_{1}(x)$ as a gradient estimate.^[One benefit of LLLS is that it is theoretically motivated: assuming that $Y_{i}=m(X_{i}) + \epsilon_{i}$, we can then take a Taylor approximation: $m(X_{i}) + \epsilon_{i} \approx m(x) + m'(x)[X_{i}-x] + \epsilon_{i} = [m(x) - m'(x)x ] + m'(x)X_{i} + \epsilon_{i} = b_{0}(x,h) + b_{1}(x,h) X_{i}$.]

3. More sophisticated methods that are beyond the scope of this class.

:::{.callout-note icon=false collapse="true" title="Must Know"}
Suppose a local model predicts wages near $x=12$ years of schooling: $\hat{y}(11.5)=9.2$ and $\hat{y}(12.5)=10.4$ (dollars per hour).
With step $d=1$, the finite-difference gradient at $x=12$ is
\begin{eqnarray}
\hat{b}_{1}(12) = \frac{\hat{y}(12.5) - \hat{y}(11.5)}{d} = \frac{10.4 - 9.2}{1} = 1.2.
\end{eqnarray}
Near $12$ years of schooling, each extra year is associated with about $\$1.20$ more per hour.
```{r}
d <- 1
y_hi <- 10.4  # prediction at x + d/2
y_lo <- 9.2   # prediction at x - d/2
(y_hi - y_lo) / d
```
:::

After computing gradients, you can summarize them in various plots: Histograms and Scatterplots. The confidence band only shows variability across samples, whereas these plots show variability within-sample. You can also plot all Gradients with their CI's (Chaudhuri 1999; Henderson et al. 2012).

```{r}
## Gradients
pred_lo <- predict(reg_lo)
grad_dx <- diff(dat[, 'school'])
grad_dy <- diff(pred_lo)
grad_lo <-grad_dy/grad_dx

## Visual Summary
par(mfrow=c(1, 2))
hist(grad_lo,  breaks=20,
    border=NA, freq=F,
    col=lo_col,
    xlab=expression(d~hat(y)/dx),
    main=NA) ## Distributional Summary
  
## Visual Summary 2
grad_x  <- dat[, 'school'][-1]
plot(grad_x+grad_dx, grad_lo,
    xlab='x', ylab=expression(d~hat(y)/dx),
    col=lo_col, pch=16, main=NA) ## Diminishing Returns?
```

A different kind of summary collapses the whole gradient curve into a single number for tabular reporting.

:::{.callout-important icon=false title="Key Definition" collapse="false"}
The *marginal effect at the mean* evaluates the gradient at a single design point (usually $\hat{M}_{X}$):
$$\hat{b}_{1}(x=\hat{M}_{X}).$$
The *mean of the gradients* (also called the *average effect*) averages $\hat{b}_{1}(x)$ across every observation in the dataset:
$$\frac{1}{n}\sum_{i=1}^{n} \hat{b}_{1}(x=\hat{X}_{i}).$$
:::

These two summaries are useful for different questions: the *marginal effect at the mean* answers "what is the slope at a typical $X$?", while the *mean of the gradients* answers "what is the average slope across the sample?". The two coincide for a globally linear model but can differ for any other, sometimes substantially when the relationship bends. You may also be interested in the *median* of the gradients, or in measures of *effect heterogeneity* like the interquartile range or standard deviation of the gradients. Such statistics can be presented in tabular form: "mean gradient (sd gradient)" or "mean gradient (estimated SE), sd gradient (estimated SE)".

:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
These two summaries usually differ.
Suppose a local model has gradients $\hat{b}_{1}(x)$ at schooling levels $x=(8,10,12,14,16)$ equal to $(0.4, 0.8, 1.2, 0.9, 0.3)$, and the sample mean schooling is $\hat{M}_{X}=12$.

* The *marginal effect at the mean* is the single gradient at $x=\hat{M}_{X}$: $\hat{b}_{1}(12) = 1.2$.
* The *mean of the gradients* averages over all five points: $\frac{0.4+0.8+1.2+0.9+0.3}{5} = \frac{3.6}{5} = 0.72$.

They differ because the gradient is not constant: here the relationship is steepest near the mean and flatter at the extremes.
```{r}
grads <- c(0.4, 0.8, 1.2, 0.9, 0.3)
grads[3]      # marginal effect at the mean (x = 12)
mean(grads)   # mean of the gradients
```
:::

```{r}
## Tabular Summary
tab_stats <- c(
    Mean=mean(grad_lo, na.rm=T),
    SD=sd(grad_lo, na.rm=T))

## Use bootstrap to approximate sampling dist
boot_stats <- matrix(NA, nrow=299, ncol=2)
colnames(boot_stats) <- c('Mean SE', 'SD SE')
for(b in 1:nrow(boot_stats)){
    xy_b <- dat[sample(1:nrow(dat), replace=T), ]
    reg_lo <- loess(wage ~ school, data=xy_b, span=.6)
    pred_lo <- predict(reg_lo)
    grad_lo <- diff(pred_lo)/diff(xy_b[, 'school'])
    dydx_mean <- mean(grad_lo, na.rm=T)
    dydx_sd <- sd(grad_lo, na.rm=T)
    boot_stats[b, 1] <- dydx_mean
    boot_stats[b, 2] <- dydx_sd
}
## SEs and CIs
boot_se <- apply(boot_stats, 2, sd)
boot_quants <- apply(boot_stats, 2, quantile, probs=c(0.025, 0.975))
boot_quants <- apply( round(boot_quants, 3), 2, paste0, collapse=', ')

## Printing
tab_regstyle <- data.frame(
  Estimate  = round(tab_stats, 3),
  SE = paste0('(', round(boot_se, 3), ')'),
  CI_95 = paste0('[', boot_quants, ']')
)
tab_regstyle
```


```{r}
#| echo: false
#| message: false
#| warning: false
#| results: asis

library(knitr)
tab <- data.frame(
  'Estimate' = round(tab_stats, 2),
  'Bootstrap SE'  =  paste0('(', round(boot_se, 3), ')'),
  'Bootstrap 95 CI' = paste0('[', boot_quants, ']')
 )
kable(
  tab,
  format='html',
  digits = 3,
  align = 'l',
  caption = 'Summary of Local Gradients')
```


#### **Hypothesis Testing**.{-}

We can test whether the mean gradient is statistically different from zero. We can compute the p-value directly from the bootstrap distribution. Under $H_0$, the mean gradient equals zero, so we center the bootstrap distribution at zero and ask how often it produces values as extreme as the observed mean gradient.

```{r}
## P-value via null bootstrap ECDF
## Center bootstrap means at zero (impose H0)
boot_centered <- boot_stats[, 'Mean SE'] - mean(boot_stats[, 'Mean SE'])

## ECDF of centered bootstrap distribution
boot_ecdf <- ecdf(boot_centered)

## Two-sided p-value: P(|centered mean| >= |observed mean|)
p_boot <- 1 - boot_ecdf(abs(tab_stats['Mean'])) +
              boot_ecdf(-abs(tab_stats['Mean']))

cat('Bootstrap p-value (ECDF):', format.pval(p_boot, digits=3), '\n')
```

The small p-value indicates that the mean gradient is statistically distinguishable from zero: on average, wages increase with schooling.

We can also use a normal approximation. Under the null hypothesis $H_0$: the mean gradient equals zero, we use the test statistic
$$t = \frac{\text{mean gradient}}{\text{SE(mean gradient)}}$$
and the p-value comes from the standard normal distribution.

```{r}
## P-value for mean gradient
## Standard Normal Approximation
t_stat <- tab_stats['Mean'] / boot_se['Mean SE']
p_val  <- 2 * pnorm(-abs(t_stat))

cat('Mean gradient:', round(tab_stats['Mean'], 3), '\n')
cat('Bootstrap SE: ', round(boot_se['Mean SE'], 3), '\n')
cat('t-statistic:  ', round(t_stat, 3), '\n')
cat('p-value:      ', format.pval(p_val, digits=3), '\n')
```

::: {.callout-tip icon=false collapse="true" title="Test Yourself"}
Compute the t-statistic and p-value using jackknife standard errors instead of the bootstrap.

```{r}
## Jackknife SE for mean gradient
n <- nrow(dat)
jack_means <- numeric(n)
for (i in 1:n) {
    dat_i  <- dat[-i, ]
    reg_i  <- loess(wage ~ school, data=dat_i, span=.6)
    pred_i <- predict(reg_i)
    grad_i <- diff(pred_i) / diff(dat_i[, 'school'])
    jack_means[i] <- mean(grad_i, na.rm=T)
}

## Jackknife SE
jack_se <- sqrt((n - 1) / n * sum((jack_means - mean(jack_means))^2))

## t-statistic and p-value
t_jack <- tab_stats['Mean'] / jack_se
p_jack <- 2 * pnorm(-abs(t_jack))

cat('Jackknife SE: ', round(jack_se, 3), '\n')
cat('t-statistic:  ', round(t_jack, 3), '\n')
cat('p-value:      ', format.pval(p_jack, digits=3), '\n')
```
:::


## Theory

We use simulations where the true data-generating process is known to better understand different models.

#### **Sampling Distributions**.{-}

Bootstrap confidence bands are approximations: they estimate how much predicted values vary from sample to sample. The simulation below compares a bootstrap CI constructed from one sample to the true sampling variation observed across many independent samples.

```{r}
## Ages
Xmx <- 70
Xmn <- 15

##Generate Sample Data
dat_sim <- function(n=1000){
    X <- seq(Xmn, Xmx, length.out=n)
    ## Random Productivity
    e <- runif(n, 0, 1E6)
    beta <-  1E-10*exp(1.4*X -.015*X^2)
    Y    <-  (beta*X + e)/10
    return(data.frame(Y, X))
}
dat0 <- dat_sim(1000)
dat0 <- dat0[order(dat0[, 'X']), ]

## Data from one sample
plot(Y ~ X, data=dat0, pch=16, col=grey(0, .05),
    main=NA, ylab='Yearly Productivity ($)', xlab='Age' )
reg_lo <- loess(Y ~ X, data=dat0, span=.8)

## Plot Bootstrap CI for Single Sample
pred_design <- data.frame(X=seq(Xmn, Xmx))
preds_lo <- predict(reg_lo, newdata=pred_design)
boot_lo <- matrix(NA, nrow=nrow(pred_design), ncol=399)
for (b in 1:399) {
    dat0_i <- dat0[sample(nrow(dat0), replace=T), ]
    reg_i <- loess(Y ~ X, data=dat0_i, span=.8)
    boot_lo[, b] <- predict(reg_i, newdata=pred_design)
}
boot_cb <- apply(boot_lo, 1, quantile,
    probs=c(.025, .975), na.rm=T)
polygon(
    c(pred_design[[1]], rev(pred_design[[1]])),
    c(boot_cb[1, ], rev(boot_cb[2, ])),
    col=hcl.colors(3, alpha=.25)[2],
    border=NA)

# Construct CI across Multiple Samples
sample_lo <- matrix(NA, nrow=nrow(pred_design), ncol=399)
for (b in 1:399) {
    xy_b <- dat_sim(1000) #Entirely new sample
    reg_b <- loess(Y ~ X, data=xy_b, span=.8)
    sample_lo[, b] <- predict(reg_b, newdata=pred_design)
}
ci_lo <- apply(sample_lo, 1, quantile,
    probs=c(.025, .975), na.rm=T)
polygon(
    c(pred_design[[1]], rev(pred_design[[1]])),
    c(ci_lo[1, ], rev(ci_lo[2, ])),
    col=grey(0, alpha=.25),
    border=NA)
```

#### **Bias**.{-}

The plot below shows the average prediction across $B$ samples against the true mean. Notice that LOESS tracks the true mean closely. Notice that a linear model does not.


```{r}
set.seed(123)

## Shared parameters used by Bias and Variance
B        <- 300
n_bias   <- 200
Nseq     <- seq(50, 500, by=50)
span_lo  <- 0.5
true_m   <- function(x) (1E-10*exp(1.4*x - .015*x^2)*x + 5E5) / 10
x0       <- seq(Xmn, Xmx, length.out=120)
m0       <- true_m(x0)
pred_design <- data.frame(X=x0)

## Storage for Bias predictions (filled via <<- when n == n_bias)
pred_lm_bias <- matrix(NA_real_, nrow=length(x0), ncol=B)
pred_lo_bias <- matrix(NA_real_, nrow=length(x0), ncol=B)

## One sweep: gradient SEs across all n, plus bias predictions at n_bias
SE <- matrix(NA, nrow=2, ncol=length(Nseq))
for (ni in seq_along(Nseq)) {
    n <- Nseq[ni]
    stats <- matrix(NA, nrow=2, ncol=B)
    for (b in 1:B) {
        dat_b  <- dat_sim(n)
        fit_lm <- lm(Y ~ X, data=dat_b)
        fit_lo <- loess(Y ~ X, data=dat_b, span=0.8)
        grad_lo <- diff(predict(fit_lo))/diff(dat_b[, 'X'])
        if(n == n_bias){
            pred_lm_bias[, b] <- predict(fit_lm, newdata=pred_design)
            fit_lo_b <- loess(Y ~ X, data=dat_b, span=span_lo)
            pred_lo_bias[, b] <- predict(fit_lo_b, newdata=pred_design)
        }
        stats[, b] <- c(coef(fit_lm)[2], mean(grad_lo, na.rm=T))
    }
    SE[, ni] <- apply(stats, 1, sd)
}
```

```{r}
m_lm <- rowMeans(pred_lm_bias, na.rm=TRUE)
m_lo <- rowMeans(pred_lo_bias, na.rm=TRUE)

y_rng <- range(c(m0, m_lm, m_lo), na.rm=TRUE)
plot(x0, m0, type='l', lwd=2, col=rgb(0, 0, 0, .8), ylim=y_rng,
  xlab='Age', ylab=' Mean Productivity ($)', main=NA)
lines(x0[is.finite(m_lm)], m_lm[is.finite(m_lm)], lwd=2, col=rgb(1, 0, 0, .8))
lines(x0[is.finite(m_lo)], m_lo[is.finite(m_lo)], lwd=2, col=rgb(0, 0, 1, .8))
legend('topright', lty=1, lwd=2, col=c(rgb(0, 0, 0, .8), rgb(1, 0, 0, .8), rgb(0, 0, 1, .8)),
  legend=c('True mean',
    'Linear',
    paste0('Loess(', span_lo, ')')
))
```


#### **Variance**.{-}

The model estimates also vary from sample to sample. Notice that the OLS slope coefficient generally varies less than the LOESS mean gradient, even though it is biased.

Also notice that there are diminishing returns to larger sample sizes. Both the OLS slope coefficient and the loess mean gradient vary less from sample to sample as $n$ grows, making hypothesis tests more accurate.

```{r}
cols <- c(rgb(1, 0, 0, .8), rgb(0, 0, 1, .8))
matplot(Nseq, t(SE), type='b', pch=16,
    lty=1, lwd=2, col=cols,
    ylab='standard error', xlab='sample size',
    main=NA)
legend('topright',
    lty=1, lwd=2, col=cols,
    legend=c('OLS slope', 'Loess mean gradient'))
```




#### **Bias-Variance Tradeoff**.{-}

The previous chapter introduced the bias-variance tradeoff intuitively with the bandwidth knob; here we restate it more formally and check it with simulations.

:::{.callout-important icon=false title="Key Definition" collapse="false"}
The expected prediction error of an estimator decomposes into two pieces. Squared *bias* is how far its average prediction sits from the truth. *Variance* is how much its predictions wobble around their own average.
:::

The bias-variance decomposition is useful as the formal expression of why we cannot simply minimize one source of error and ignore the other: cutting one usually raises the other, so there is a tradeoff to navigate. Concretely:

* High-bias / low-variance models. E.g., strict linear models are stable across samples but can miss curvature.
* Low-bias / high-variance models. E.g., very flexible local linear models can capture curvature but may overfit noise.

:::{.callout-note icon=false collapse="true" title="Must Know"}
```{r}
# Bias functions
mean_pred <- function(P) rowMeans(P, na.rm=TRUE)
avg_bias2 <- function(P) mean( (mean_pred(P) - m0)^2, na.rm=TRUE)

# Variance
var_pred <- function(P) apply(P, 1, var, na.rm=TRUE)
avg_var   <- function(P) mean(var_pred(P), na.rm=TRUE)

metrics <- rbind(
  c(avg_bias2(pred_lm_bias),  avg_var(pred_lm_bias)),
  c(avg_bias2(pred_lo_bias),  avg_var(pred_lo_bias))
)
colnames(metrics) <- c('Avg Bias^2', 'Avg Variance')
rownames(metrics) <- c('Linear', 'Loess')
round(metrics, 4)
```
:::

#### **Consistency**.{-}

The bias-variance tradeoff raises a question: can both be reduced simultaneously as $n$ grows? For local regression, the answer is yes — by shrinking the bandwidth with $n$. The neighborhood shrinks (reducing bias) while the local sample size still grows (reducing variance). This is often written as bandwidth conditions:
\begin{eqnarray}
h_n \to 0
\quad\text{and}\quad
n h_n \to \infty.
\end{eqnarray}
In LOESS language, this corresponds to the span shrinking with $n$, but not too fast.

:::{.callout-note icon=false collapse="true" title="Must Know"}
A common bandwidth rule shrinks $h$ with the sample size, for example $h_{n}=n^{-1/5}$. Check the two conditions:

* $h_{n}=n^{-1/5}\to 0$: at $n=100$ it is about $0.40$; at $n=10{,}000$ it is about $0.16$. The neighborhood shrinks.
* $n h_{n}=n^{4/5}\to\infty$: at $n=100$ it is about $40$; at $n=10{,}000$ it is about $1585$. The local sample size still grows.

So bias falls (the window narrows) while variance falls (each local fit uses more data).
```{r}
n <- c(100, 10000)
h_n <- n^(-1/5)
rbind(h_n=round(h_n, 3), n_times_h=round(n*h_n, 1))
```
:::

The simulation below illustrates this at one target point $x_0$: absolute error in estimating $m(x_0)$ tends to decrease with larger $n$.

```{r}
set.seed(42)

x0_target <- 40  # mid-career age
n_grid <- c(60, 120, 240, 480)
R <- 120

avg_abs_err <- numeric(length(n_grid))
for (ni in seq_along(n_grid)) {
  n <- n_grid[ni]
  span_n <- min(0.9, 1.8*n^(-1/4)) # shrinks with n
  errs <- replicate(R, {
    dat_b <- dat_sim(n)
    fit <- loess(Y ~ X, data=dat_b, span=span_n, degree=1)
    mhat <- predict(fit, newdata=data.frame(X=x0_target))
    abs(mhat - true_m(x0_target))
  })
  avg_abs_err[ni] <- mean(errs, na.rm=TRUE)
}

plot(n_grid, avg_abs_err, type='b', pch=16,
     xlab='Sample size (n)',
     ylab='Average Bias',
     main=NA)
```

A similar result holds for OLS when the true relationship is linear. Even with unlimited data, however, a misspecified model cannot recover the true conditional mean.


## Exercises

1. A linear regression has low variance but can have high bias when the true relationship is nonlinear. A loess regression with a small span has low bias but high variance. Explain why increasing the sample size $n$ helps reduce the standard error for both models, but only helps reduce bias for loess (with an appropriate bandwidth rule).

2. Using the `USArrests` dataset, fit a loess regression of `Murder` on `UrbanPop` with `span = 0.6`. Compute the finite-difference gradients $\hat{b}_{1}(x)$ and report their mean and standard deviation. Compare these to the OLS slope $\hat{b}_{1}$ from a simple linear regression.

3. Using the `USArrests` dataset and the loess fit from the previous question, write R code to construct a bootstrap confidence band with $B = 399$ replicates. Plot the scatterplot, the loess curve, and the shaded 95% confidence band.

#### **Further Reading**. {-}

* <https://www.econometrics-with-r.org/5-htaciam.html> -- hypothesis tests and confidence intervals for regression coefficients, with worked R examples.
* <https://learningstatisticswithr.com/book/regression.html#hypothesistestsforregressionmodels> -- testing regression slopes, interpreting p-values, and understanding standard errors.
* <https://www.khanacademy.org/math/statistics-probability/significance-tests-one-sample/more-significance-testing-videos/v/hypothesis-testing-and-p-values> -- video introduction to hypothesis testing logic and p-values.

#### **Recall** {-}

This chapter put inference around the local fits from the previous chapter on the `Wages1` data: bootstrap confidence bands for a LOESS fit, gradient summaries (marginal effect at the mean vs mean of the gradients), hypothesis tests, and a simulation walk through bias, variance, and consistency. The small worked example with $\hat{y}(11.5)=9.2$ and $\hat{y}(12.5)=10.4$ gave a finite-difference gradient $\hat{b}_{1}(12)=1.2$ dollars per year of schooling, and the gradient-vector example $(0.4, 0.8, 1.2, 0.9, 0.3)$ at $x=(8, 10, 12, 14, 16)$ showed the marginal-effect-at-the-mean ($1.2$) and the mean-of-the-gradients ($0.72$) diverging when the relationship bends. The next chapter pulls back from samples and develops the bivariate probability machinery (joint distributions, conditional expectations, and the population coefficient $\beta_{1}$) that says what we are estimating.
