# Comparing Multiple Groups
***

[Comparing Two Groups](02_12_TwoGroups.qmd) split a cardinal outcome by a two-level factor and tested whether the two groups differed.
This chapter extends that idea to $G$ groups: how to describe several groups at once, how to compare their medians with bootstrap intervals, and then three ways to test whether they differ, one built on means, one built on ranks, and one built on category proportions.

## Describing Groups

The [census regions](01_03_SummaryStatistics.qmd#unordered-factors) are an unordered factor with $G=4$ levels.
The built-in `state.region` vector lists them in the same state order as the rows of `USArrests`, so it can be paired directly with the murder rates.

```{r}
# Outcome: murder arrests per 100K, one value per state
y_hat <- USArrests[, 'Murder']
# Group: the four-level region factor, in the same state order as USArrests.
# Its levels are the values the categorical variable can take.
region <- state.region
levels(region)
table(region)
```

#### **Group Summaries** {-}

A factor with $G$ levels splits a cardinal variable into $G$ conditional distributions, one per group.
The group summaries from [Bivariate Distributions](02_11_BivariateDistributions.qmd#conditional-distributions) apply directly.
The [ECDF](01_02_Data.qmd#empirical-cumulative-distribution-function) shows every observation, and drawing one per group on the same axes compares the whole distributions at once: a curve further to the right has higher values at every quantile.
The regions are ordered by median murder rate from lowest to highest, so South is last in the legend, and each line's thickness scales with its group size.

```{r, fig.height=5.2}
# Order regions by their median murder rate, lowest to highest
region_order <- names(sort(tapply(y_hat, region, median)))
region <- factor(region, levels=region_order)

# One ECDF per region, on the same axes
line_cols <- hcl.colors(4)
n_g <- table(region)
lwd_g <- 1.5 + 2.5 * (n_g - min(n_g)) / (max(n_g) - min(n_g))

par(mar=c(4, 4, 4.5, 1))
plot(NULL, xlim=range(y_hat), ylim=c(0, 1),
    xlab='Murder arrests (per 100,000)', ylab='Proportion of states', main=NA)
for (g in seq_along(levels(region))) {
    F_g_hat <- ecdf(y_hat[region == levels(region)[g]])
    plot(F_g_hat, add=TRUE, col=line_cols[g], lwd=lwd_g[g],
        verticals=TRUE, do.points=FALSE, col.01line=NA)
}
legend('topleft', inset=c(0, -.2), xpd=TRUE, ncol=2, bty='n',
    legend=paste0(levels(region), ' (n=', n_g, ')'), col=line_cols, lty=1, lwd=lwd_g)
```

The South's curve sits to the right of the others over its whole range, so its higher values are not driven by a few states.

A table of group sizes and centers puts numbers on the figure.
The *conditional mean* $\hat{m}_{Y}(g)$ is the average of $\hat{y}_{i}$ among the $n_g$ observations in group $g$, and `aggregate()` computes it for every group at once.

```{r}
# Group size, mean, and median of the murder rate by region
summary_fun <- function(y_hat) {
    out <- c(n=length(y_hat), mean=mean(y_hat), median=median(y_hat))
    return(out)
}
aggregate(y_hat ~ region, FUN=summary_fun)
```

The South has the highest center and the Northeast the lowest.
Whether such differences are larger than chance alone would produce is the question the rest of this chapter takes up.

## Equal Medians

In [Comparing Two Groups](02_12_TwoGroups.qmd#quantile-differences), we bootstrapped a confidence interval for the difference between two medians; with $G$ groups, we can instead bootstrap one interval per group and compare them all on one figure.

A *notched boxplot* cuts a notch into the side of each group's box, showing a confidence interval for that group's median.
These are useful for comparing medians across many groups at a glance: if two notches do not overlap, that is strong evidence the two medians differ.
Overlapping notches do not show that the medians are equal, because comparing two separate $95\%$ intervals is stricter than a $95\%$ test of their difference.
Here we use a bootstrap percentile interval for the median in each region, which needs no such formula.^[
R's built-in `notch=TRUE` draws $\pm 1.58 \hat{IQR}/\sqrt{n_g}$ around each median, a rule of thumb from normal theory (see `?boxplot.stats`).]


```{r, warning=FALSE}
# Bootstrap the median within each region, keeping each region's size
n_boot <- 999
boot_medians <- matrix(NA, n_boot, nlevels(region))
colnames(boot_medians) <- levels(region)
for (b in seq(n_boot)) {
    for (g in levels(region)) {
        y_boot <- sample(y_hat[region == g], replace=TRUE)
        boot_medians[b, g] <- median(y_boot)
    }
}

# 95% bootstrap percentile intervals: one column per region
med_ci <- apply(boot_medians, 2, quantile, probs=c(.025, .975))
round(med_ci, 2)

# Boxplot statistics, with the bootstrap intervals as the notches
bx <- boxplot(y_hat ~ region, plot=FALSE)
bx[['conf']] <- med_ci

# One notched box per region, box width scaled to group size
bxp(bx, notch=TRUE, varwidth=TRUE,
    boxfill=hcl.colors(4, alpha=.45),
    whisklty=0, staplelty=0,
    ylab='Murder arrests (per 100,000)', xlab='Region', main=NA)
title('Murder rates by region', font.main=1)
# Raw Observations, in place of whiskers and outliers
stripchart(y_hat ~ region,
    pch='-', col=grey(0, .5), cex=2,
    vert=TRUE, add=TRUE)
```

Each box shows the median and the two quartiles, and each dash is one state's murder rate.
`varwidth=TRUE` also scales each box's width to $\sqrt{n_g}$, so a region with more states draws a wider box.
With only $9$ to $16$ states per region, each median's interval is nearly as wide as the box itself.
When an interval reaches past a quartile, the notch folds back outside the box.
The South's notch lies above the Northeast's, so the South's median is higher.
Which other pairs of regions have notches that do not overlap?

## Equal Means (ANOVA)

In [Comparing Groups](02_12_TwoGroups.qmd), we tested whether two groups had the same mean.
We now extend this to $G$ groups.
The *Analysis of Variance* (ANOVA) approach asks: are the observed differences in group means larger than we would expect from random variation alone?

#### **Decomposition** {-}

If group means differ, some of the spread in $Y$ must come from differences *between* groups rather than *within* them.
We can split each piece out.

:::{.callout-important icon=false title="Key Definition" collapse="false"}
Label each observation $\hat{y}_{ig}$ for individual $i$ in group $g$, where group $g=1,\dots,G$ has $n_g$ observations with group mean $\hat{m}_{Y}(g)$, and the grand mean across all $n=\sum_g n_g$ observations is $\hat{m}_{Y}$.
The *ANOVA decomposition* splits the total sum of squares into between-group and within-group pieces:
$$\begin{aligned}
\underbrace{\sum_{g=1}^{G}\sum_{i=1}^{n_g}(\hat{y}_{ig}-\hat{m}_{Y})^2}_{\hat{TSS}}
&=
\underbrace{\sum_{g=1}^{G} n_g (\hat{m}_{Y}(g)-\hat{m}_{Y})^2}_{\hat{BSS}}
\;+\;
\underbrace{\sum_{g=1}^{G}\sum_{i=1}^{n_g}(\hat{y}_{ig}-\hat{m}_{Y}(g))^2}_{\hat{WSS}},
\end{aligned}$$
where $\hat{BSS}$ measures variation in group means around the grand mean and $\hat{WSS}$ measures variation of observations around their own group mean.
:::

The decomposition is useful as the foundation for any test of "do group means differ?": if they do not, $\hat{BSS}$ should be small relative to $\hat{WSS}$.
The equality holds exactly because the cross-term $\sum_{g,i}(\hat{y}_{ig}-\hat{m}_{Y}(g))(\hat{m}_{Y}(g)-\hat{m}_{Y})$ vanishes, since the within-group deviations sum to zero around their own group mean.

In R, `aov()` fits the ANOVA model and `summary()` returns the standard ANOVA table.
The ANOVA table reports $\hat{BSS}$ ("Sum Sq" for `region`), $\hat{WSS}$ ("Sum Sq" for `Residuals`).
```{r}
# Fit ANOVA
fit_aov <- aov(y_hat ~ region)
summary(fit_aov)
```

:::{.callout-note icon=false collapse="true" title="Must Know"}
For example, suppose we have $G=3$ groups with the following data:

| Group A | Group B | Group C |
|:---:|:---:|:---:|
| 4 | 8 | 5 |
| 5 | 7 | 6 |
| 6 | 9 | 7 |

Group means: $\hat{m}_{Y}(A) = 5$, $\hat{m}_{Y}(B) = 8$, $\hat{m}_{Y}(C) = 6$.
Grand mean: $\hat{m}_Y = (5+8+6)/3 = 19/3 \approx 6.33$.
Each group has $n_g = 3$.

Between-group sum of squares:
$$\begin{aligned}
\hat{BSS} &= 3(5-19/3)^2 + 3(8-19/3)^2 + 3(6-19/3)^2 \\
&= 3(16/9) + 3(25/9) + 3(1/9) = 48/9 + 75/9 + 3/9 = 14
\end{aligned}$$
Within-group sum of squares:
$$\begin{aligned}
\hat{WSS} &= [(4-5)^2+(5-5)^2+(6-5)^2] + [(8-8)^2+(7-8)^2+(9-8)^2] \\
& + [(5-6)^2+(6-6)^2+(7-6)^2] = 2+2+2 = 6
\end{aligned}$$

```{r}
# Data
y0_hat <- c(4, 5, 6, 8, 7, 9, 5, 6, 7)
g0 <- factor(c('A', 'A', 'A', 'B', 'B', 'B', 'C', 'C', 'C'))

# Verify by hand
m0_hat <- mean(y0_hat)
m0_g_hat <- aggregate(y0_hat, list(g0), mean)$x
n0_g <- aggregate(y0_hat, list(g0), length)$x

BSS <- sum(n0_g * (m0_g_hat - m0_hat)^2)
WSS <- sum(aggregate(y0_hat, list(g0), function(y_hat) sum((y_hat - mean(y_hat))^2))$x)
cbind(BSS, WSS)
```
:::

#### **The F-statistic** {-}

We have the decomposition; now we need a single number that tells us when the between-group piece is "large" relative to the within-group piece.

:::{.callout-important icon=false title="Key Definition" collapse="false"}
The *ANOVA $F$-statistic* is the ratio of average between-group variation to average within-group variation:
$$\hat{F} = \frac{\hat{BSS}/(G-1)}{\hat{WSS}/(n-G)}.$$
Writing $Y_{ig}$ for the random outcome of individual $i$ in group $g$, the null is $H_0: \mathbb{E}[Y_{i1}]=\mathbb{E}[Y_{i2}]=\dots=\mathbb{E}[Y_{iG}]$.
Under the null (with the usual parametric assumptions) it follows an $F_{G-1, n-G}$ distribution; we can also get a $p$-value nonparametrically by bootstrapping the labels.
:::

The ANOVA F-statistic is useful as the global "do any means differ?" test: one number and one $p$-value, regardless of how many groups you have.
The numerator is the average between-group variation per degree of freedom, the denominator is the average within-group variation, and a large $\hat{F}$ means group means are more spread out than individual-level noise would predict.

```{r}
F_fun <- function(y_hat,g){
    m_y_hat <- mean(y_hat)
    m_g_hat <- aggregate(y_hat, list(g), mean)$x
    n_g <- aggregate(y_hat, list(g), length)$x

    BSS <- sum(n_g * (m_g_hat - m_y_hat)^2)
    WSS <- sum(aggregate(y_hat, list(g), function(y_hat) sum((y_hat - mean(y_hat))^2))$x)

    n <- length(y_hat)
    G <- nlevels(g)
    F_obs <- (BSS/(G-1)) / (WSS/(n-G))
    return(F_obs)
}
F_obs <- F_fun(y_hat=y_hat, g=region)
F_obs
```


To compute a $p$-value, we can use permutation or bootstrap methods to build a null distribution, just as in previous chapters.

```{r}
# Pairs bootstrap: resample group labels to break the Y--group pairing
n <- length(y_hat)

F_nullboot <- vector(length=999)
for( b in seq_along(F_nullboot)){
    idx <- sample(n, replace = TRUE)
    g_boot <- region[idx]
    F_b <- F_fun(y_hat=y_hat, g=g_boot)
    F_nullboot[b] <- F_b
}

# Bootstrap p-value
p_boot <- mean(F_nullboot >= F_obs)
p_boot

hist(F_nullboot, breaks = 40,
    col = grey(0.5, 0.5), border = NA,
    freq = FALSE, main = NA, xlab = 'Null Bootstrap F-statistic',
    xlim = range(c(0, F_nullboot, F_obs)))
title('Null Bootstrap F-test', font.main = 1)
abline(v = F_obs, col = rgb(1, 0, 0, .8), lwd = 2)
legend('topright', legend = 'Observed F',
    col = rgb(1, 0, 0, .8), lwd = 2, bty='n')
```

:::{.callout-note icon=false collapse="true" title="Must Know"}
Pairing the observed murder rates `y_hat` with a *resampled* `region` vector breaks the link between outcome and group: under the null, group labels are exchangeable, so this approximates the distribution of $\hat{F}$ when no group effect exists.
:::

Under additional parametric assumptions (independent observations, equal variances, normal errors), $\hat{F}$ (under the null) follows an $F_{G-1,\;n-G}$ distribution.
From that theoretical null distribution, we can then compute p-values (which is the default report method in R).
```{r}
# Degrees of freedom for the regional data
n <- length(y_hat)
G <- nlevels(region)
p_param <- 1 - pf(F_obs, G-1, n-G)

 # Compare parametric and bootstrap p-values
cbind(F_obs, p_param, p_boot)
```

:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
The parametric F-test assumes (i) independent observations, (ii) normal errors within each group, and (iii) equal variances across groups (*homoscedasticity*).
If these assumptions are doubtful, then the nonparametric Kruskal-Wallis test (below) is a safer alternative.
As a quick diagnostic, visually compare side-by-side boxplots: if the spread differs across groups or the distribution has excessive skew/kurtosis, be cautious with the parametric test.

With $G-1=2$ and $n-G=6$ degrees of freedom, we compare to an $F_{2,6}$ distribution: $p = 1 - F_{2,6}(7) \approx 0.027$.
```{r}
# Data (the three-group example from the decomposition above)
y0_hat <- c(4, 5, 6, 8, 7, 9, 5, 6, 7)
g0 <- factor(c('A', 'A', 'A', 'B', 'B', 'B', 'C', 'C', 'C'))

# Computation
summary(aov(y0_hat ~ g0))

# Verify by hand
n0 <- length(y0_hat)
G0 <- nlevels(g0)
m0_hat <- mean(y0_hat)
m0_g_hat <- aggregate(y0_hat, list(g0), mean)$x
n0_g <- aggregate(y0_hat, list(g0), length)$x
BSS <- sum(n0_g * (m0_g_hat - m0_hat)^2)
WSS <- sum(aggregate(y0_hat, list(g0), function(y_hat) sum((y_hat - mean(y_hat))^2))$x)
Fstat <- (BSS/(G0-1)) / (WSS/(n0-G0))
pval  <- 1 - pf(Fstat, G0-1, n0-G0)
cbind(BSS, WSS, Fstat, pval)
```
:::

## Equal Distributions

ANOVA tests whether group *means* differ; sometimes the entire *distributions* differ in ways the mean would miss.

#### **Kruskal-Wallis Test** {-}

When data are skewed, heavy-tailed, or only ordinal, a test built on means may overlook differences that show up in the rank order.

:::{.callout-important icon=false title="Key Definition" collapse="false"}
The *Kruskal-Wallis test* is a nonparametric test for equal distributions across $G$ groups, testing $H_0: F_1 = F_2 = \dots = F_G$ against $H_A:$ at least one $F_g$ differs (where $F_g$ is the continuous distribution of group $g=1,...G$).
It uses the average ranks within each group rather than the raw observations, so it only requires the data to be ordinal.
With two groups it reduces to the *Mann-Whitney U* (equivalently *Wilcoxon rank-sum*) test.
:::

The Kruskal-Wallis test is useful when the equal-variance normal-error assumption is implausible, as with income, response times, or other heavy-tailed outcomes.
Rank-based tests need only the orderings of the data.
A significant result tells us *that* some distribution differs, but not *which* one; the global rejection has to be followed up with pairwise tests.

To conduct the test, first denote individuals $i=1,...n$ with overall ranks $\hat{r}_1,....\hat{r}_{n}$.
Each individual belongs to group $g=1,...G$, and each group $g$ has $n_{g}$ individuals with average rank $\bar{r}_{g} = \sum_{i} \hat{r}_{i} /n_{g}$.
The Kruskal Wallis statistic is
$$\begin{aligned}
\hat{KW} &= (n-1) \frac{\sum_{g=1}^{G} n_{g}( \bar{r}_{g} - \bar{r}  )^2  }{\sum_{i=1}^{n} ( \hat{r}_{i} - \bar{r}  )^2},
\end{aligned}$$
where  $\bar{r} = \frac{n+1}{2}$ is the grand mean rank.

In the special case with only two groups, the Kruskal Wallis test reduces to the *Mann–Whitney U* test (also known as the *Wilcoxon rank-sum* test).
In this case, we can write the hypotheses in terms of individual outcomes in each group, $Y_i$ in one group $Y_j$ in the other; $H_0: Prob(Y_i > Y_j)=Prob(Y_j > Y_i)$  versus $H_A: Prob(Y_i > Y_j) \neq Prob(Y_j > Y_i)$.
The corresponding test statistic is
$$\begin{aligned}
\hat{U}   &= \min(\hat{U}_1, \hat{U}_2) \\
\hat{U}_g &= \sum_{i~\text{in}~g}\sum_{j~\text{in}~-g}
           \Bigl[\mathbf 1( \hat{y}_{i} > \hat{y}_{j}) + \tfrac12\mathbf 1(\hat{y}_{i} = \hat{y}_{j})\Bigr].
\end{aligned}$$

Interpretation template:

* If the Kruskal-Wallis p-value is large, keep the global null and stop.
* If the p-value is small, report that at least one group differs.
* Report both statistical and practical importance (distribution plots + effect size summaries), not only p-values.

```{r}
kw <- kruskal.test(y_hat ~ region)
kw
```

A significant Kruskal-Wallis result tells us *that* some group differs, not *which*.
To answer *which*, use a post-hoc pairwise test together with a multiple-comparisons correction.
Here, `pairwise.wilcox.test()` runs a Mann-Whitney test on each pair of regions, and the Holm correction adjusts the $p$-values for the number of pairs tested.

```{r}
pairwise.wilcox.test(y_hat, region, p.adjust.method='holm')
```

Looking back at the boxplot in [Equal Medians](#equal-medians), which region stands out?

:::{.callout-note icon=false collapse="true" title="Must Know"}
Worked example: $G=3$ groups with values $A=\{1,2,3\}$, $B=\{4,5,6\}$, $C=\{7,8,9\}$.

Pooling and ranking gives the ranks $1,\ldots,9$ in order.
Group rank averages: $\bar{r}_A=2$, $\bar{r}_B=5$, $\bar{r}_C=8$.
Grand mean rank: $\bar{r}=(n+1)/2=5$.
Plug into $\hat{KW}$ with $n=9$, $n_g=3$:
$$
\hat{KW} = (n-1) \cdot \frac{3(2-5)^2 + 3(5-5)^2 + 3(8-5)^2}{\sum_{i=1}^{9}(\hat{r}_i - 5)^2}
        = 8 \cdot \frac{27 + 0 + 27}{60} = 7.2.
$$

```{r}
y0_hat <- c(1,2,3, 4,5,6, 7,8,9)
g0 <- factor(rep(c('A','B','C'), each=3))
kruskal.test(y0_hat ~ g0)
```

For the two-group case, the Mann-Whitney statistic counts pairwise wins.
Compare $A=\{1,3,5\}$ to $B=\{2,4,6\}$: out of $3\times 3 = 9$ pairs, $A$ wins three ($(3,2), (5,2), (5,4)$), so $\hat{U}_A=3$, $\hat{U}_B=6$, giving $\hat{U}=\min(3,6)=3$.

```{r}
a0_hat <- c(1,3,5); b0_hat <- c(2,4,6)
wilcox.test(a0_hat, b0_hat)
```
:::

#### **Categorical ANOVA** {-}

When the outcome is a factor, we can still ask whether the groups spread their observations across the categories in the same proportions.

:::{.callout-important icon=false title="Key Definition" collapse="false"}
Let $\hat{p}(y)$ be the proportion of all $n$ observations in category $y$, and $\hat{p}(y \mid g)$ the proportion of the $n_g$ observations in group $g$.
The *categorical ANOVA* measures variation with one minus the [Herfindahl index](01_03_SummaryStatistics.qmd#unordered-factors), and splits it into between-group and within-group pieces:
$$\begin{aligned}
\underbrace{\frac{n}{2}\left(1-\hat{H}\right)}_{\hat{TSS}}
&=
\underbrace{\frac{1}{2}\sum_{g=1}^{G} n_g \sum_{y}\left(\hat{p}(y \mid g)-\hat{p}(y)\right)^2}_{\hat{BSS}}
\;+\;
\underbrace{\sum_{g=1}^{G} \frac{n_g}{2}\left(1-\hat{H}(g)\right)}_{\hat{WSS}},
\end{aligned}$$
where $\hat{H}=\sum_{y}\hat{p}(y)^2$ is the Herfindahl index of the pooled data and $\hat{H}(g)=\sum_{y}\hat{p}(y \mid g)^2$ is the Herfindahl index within group $g$.
:::

The categorical ANOVA is useful when the outcome is an unordered factor, such as occupation, brand, or vote choice, where there are no means to compare and no ranks to sort.
Recall that $1-\hat{H}$ is zero when every observation falls in one category and grows as the categories become equally common, so it plays the role of the variance.
The pieces are also ordinary ANOVA sums of squares.
Make one 0/1 indicator for each category and compute its $\hat{TSS}$, $\hat{BSS}$, and $\hat{WSS}$ as in the [decomposition](#decomposition) above.
Adding each piece up across the categories and halving it gives the three pieces here.
With a binary outcome, the two indicators have the same sums of squares, so the categorical ANOVA is exactly the ANOVA of a single 0/1 variable.
The share $\hat{BSS}/\hat{TSS}$ reports the fraction of the variation in $Y$ that lies between groups.^[See Light, R. J., and B. H. Margolin (1971), "An Analysis of Variance for Categorical Data," *Journal of the American Statistical Association*, 66(335), 534--544, <https://doi.org/10.1080/01621459.1971.10482297>.]

For example, the urban population bands from [Ordered Factors](01_03_SummaryStatistics.qmd#ordered-factors) sort each region's states into mostly rural, mixed, and mostly urban.
The bands are ordered, but the categorical ANOVA ignores that order, so it treats them as it would any unordered factor.

```{r, fig.height=5.2}
# Outcome: urban population in three bands, an ordered factor
urban_band <- cut(USArrests[, 'UrbanPop'], breaks=c(0, 50, 75, 100),
    labels=c('mostly rural', 'mixed', 'mostly urban'), ordered_result=TRUE)
tab <- table(region, urban_band)
tab

# Proportions within each region: each bar sums to 1
p_band_region <- prop.table(tab, 1)
cols <- hcl.colors(3, alpha=.45)
par(mar=c(4, 4, 3.5, 1))
barplot(t(p_band_region), col=cols, border=NA,
    xlab='Region', ylab='Proportion of states', main=NA)
legend('top', inset=c(0, -.15), xpd=TRUE, ncol=3, bty='n',
    fill=cols, border=NA, legend=levels(urban_band))
```

The South and North Central lean rural, while the Northeast and West lean urban.
The function below computes the three pieces and the between-group share.

```{r}
# Categorical ANOVA: variation is measured by one minus the Herfindahl index
cat_anova <- function(y_hat, g) {
    # Total variation, from the pooled proportions
    n <- length(y_hat)
    p_hat <- table(y_hat) / n
    TSS <- n / 2 * (1 - sum(p_hat^2))
    # Within-group variation, from each group's own proportions
    WSS_g <- tapply(y_hat, g, function(y_g) {
        p_g_hat <- table(y_g) / length(y_g)
        return(length(y_g) / 2 * (1 - sum(p_g_hat^2)))
    })
    WSS <- sum(WSS_g)
    # Between-group variation, from the decomposition
    BSS <- TSS - WSS
    return(c(BSS=BSS, WSS=WSS, TSS=TSS, share=BSS / TSS))
}
cat_anova(urban_band, region)
```

About $11\%$ of the variation in urban bands lies between regions.

:::{.callout-note icon=false collapse="true" title="Must Know"}
Suppose we record how $n=8$ workers commute, $n_g=4$ from each of two neighborhoods.
Neighborhood $A$ has car, car, car, bus, and neighborhood $B$ has car, bus, bike, bike.

Pooled, the proportions are $\hat{p}(\text{car})=4/8$, $\hat{p}(\text{bus})=2/8$, and $\hat{p}(\text{bike})=2/8$, so $\hat{H}=1/4+1/16+1/16=3/8$ and $\hat{TSS}=\frac{8}{2}(1-3/8)=5/2$.
Within $A$, $\hat{H}(A)=(3/4)^2+(1/4)^2=5/8$, and within $B$, $\hat{H}(B)=(1/4)^2+(1/4)^2+(2/4)^2=3/8$.
So $\hat{WSS}=\frac{4}{2}(1-5/8)+\frac{4}{2}(1-3/8)=3/4+5/4=2$.
For $\hat{BSS}$, each neighborhood's proportions differ from the pooled ones by $1/4$ in two categories and by $0$ in the third, so $\hat{BSS}=\frac{1}{2}[4(1/16+0+1/16)+4(1/16+0+1/16)]=1/2$.
The pieces add up, $5/2=1/2+2$, and the between-group share is $\hat{BSS}/\hat{TSS}=1/5$.
```{r}
# Commute mode of 8 workers, 4 from each neighborhood
y0_hat <- factor(c('car', 'car', 'car', 'bus', 'car', 'bus', 'bike', 'bike'))
g0 <- factor(c('A', 'A', 'A', 'A', 'B', 'B', 'B', 'B'))
table(g0, y0_hat)
cat_anova(y0_hat, g0)
```
:::

To compute a $p$-value, we use a [permutation test](02_12_TwoGroups.qmd#permutation-tests): shuffle the region labels, which keeps each region's number of states fixed, and recompute the share each time.

```{r}
# Observed between-group share
share_obs <- cat_anova(urban_band, region)[['share']]

# Null distribution: shuffle region labels to break the outcome--group pairing
share_perm <- rep(NA, 999)
for (b in seq_along(share_perm)) {
    g_perm <- sample(region, replace=FALSE)
    share_perm[b] <- cat_anova(urban_band, g_perm)[['share']]
}

# Permutation p-value
p_perm <- mean(share_perm >= share_obs)
p_perm

hist(share_perm, breaks=40,
    col=grey(0.5, 0.5), border=NA,
    freq=FALSE, main=NA, xlab='Null Permutation Between-Group Share',
    xlim=range(c(0, share_perm, share_obs)))
title('Null Permutation Categorical ANOVA', font.main=1)
abline(v=share_obs, col=rgb(1, 0, 0, .8), lwd=2)
legend('topright', legend='Observed share',
    col=rgb(1, 0, 0, .8), lwd=2, bty='n')
```

Because the bands are ordered, the Kruskal-Wallis test also applies.

```{r}
# Kruskal-Wallis ranks the bands, so it uses their order
kruskal.test(urban_band ~ region)
```

The Kruskal-Wallis $p$-value is smaller.
The regions differ mostly along one direction, from rural to urban, and ranks concentrate on that direction.
The categorical ANOVA spreads its attention over every way the proportions could differ, so a rank test is usually more powerful when the categories are ordered.
The categorical ANOVA can still detect differences that ranks miss, such as one group concentrated in the middle band and another split between the two extremes.

:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
Make a binary outcome that marks the mostly urban states.
Check that the categorical ANOVA of this outcome gives the same $\hat{BSS}$ and $\hat{WSS}$ as the ordinary ANOVA of the 0/1 variable.
```{r}
# 1 for mostly urban states, 0 otherwise
d_hat <- as.numeric(urban_band == 'mostly urban')
cat_anova(factor(d_hat), region)
summary(aov(d_hat ~ region))
```
:::

## Exercises

1. Comment the script you wrote for this chapter, then restart R and check that it runs from a clean session, then check the script with AI as explained in [Working with AI](04_A_WorkingWithAI.qmd).
   Write three sentences from memory on the main statistical idea of this chapter, and ask the assistant what is wrong, vague, or missing.
   Finish with your own questions about whatever you found hardest.

2. Load `Wages1` from the `Ecdat` package.
   Draw a boxplot of `wage` by `sex`, then add an ECDF of `wage` for each sex on the same axes.
   Compute the mean and median wage for each sex with `aggregate()`.
   Does one sex's ECDF sit to the right of the other's over the whole range, or do the curves cross?

3. An ANOVA $F$-test rejects the null that all group means are equal.
   Does this tell you *which* group is different, or only that at least one differs?

4. Using the `USArrests` dataset with `state.region` as the grouping variable, compute the between-group sum of squares ($\hat{BSS}$) and within-group sum of squares ($\hat{WSS}$) for `Assault` by hand (using `aggregate()`).
   Verify your $\hat{F}$ statistic matches the output of `summary(aov(Assault ~ state.region, data = ...))`.

#### **Further Reading** {-}

* <https://www.econometrics-with-r.org/7-htaciimr.html> -- hypothesis tests for comparing groups in a regression framework, including F-tests and joint restrictions.
* <https://online.stat.psu.edu/stat462/node/137/> -- ANOVA as a special case of regression, with examples of multiple group comparisons.


#### **Recall** {-}

This chapter first described several groups with ECDFs and a table of group means, compared their medians with bootstrap notches on boxplots, and then tested whether the description reflected more than chance.
ANOVA's decomposition of variation between and within groups produced an $F$-statistic; the worked three-group example with $A=\{4,5,6\}, B=\{8,7,9\}, C=\{5,6,7\}$ made the arithmetic concrete, giving $\hat{BSS}=14$ and $\hat{WSS}=6$ by hand, which `aov(y0_hat ~ g0)` reproduces exactly.
The Kruskal-Wallis test replaced outcomes with ranks when the equal-variance normal-error assumptions were doubtful.
The categorical ANOVA measured the variation of a factor outcome with one minus the Herfindahl index; for the commutes in two neighborhoods it gave $\hat{TSS}=5/2$, $\hat{BSS}=1/2$, and $\hat{WSS}=2$ by hand.
