# Association is not Causation
***

## Introduction

Earlier chapters described how two variables move together, fitted predictive models, and measured uncertainty about those summaries.
None of those tools alone identifies what would happen to $Y$ if we intervened and changed $X$.
This chapter brings together three warnings: Simpson's paradox shows that pooling groups can reverse a relationship, association can appear without causation or causation without linear correlation, and predictive models trade bias against variance without answering a causal question.

## Simpson's Paradox

A relationship visible inside every subgroup can flip direction once those subgroups are pooled.
The [conditional distributions](02_11_BivariateDistributions.qmd#conditional-distributions) from Chapter 11 compare outcomes within values of another variable; adding a subgroup $g$ lets us compare $\hat{P}(y\mid x,g)$ within each group to the pooled $\hat{P}(y\mid x)$.

:::{.callout-important icon=false title="Key Definition" collapse="false"}
*Simpson's paradox* arises when a relationship that holds inside every subgroup reverses when the data are pooled.
The within-group conditional distributions $\hat{P}(y\mid x, g)$ point one way; the overall $\hat{P}(y\mid x)$ points the other.
:::

Simpson's paradox is useful as a warning whenever you are tempted to pool subgroups: the pooled relationship reflects both the within-group pattern and how the subgroups differ in size.
The classic example involves university admissions, where $400$ men and $400$ women apply to one of two departments (English or Engineering).
Women can have higher admission rates *within* both departments yet a lower *overall* admission rate, if women disproportionately apply to the more selective department (English) and men to the less selective one (Engineering).
The example is part of a real debate about discrimination (<http://homepage.stat.uiowa.edu/~mbognar/1030/Bickel-Berkeley.pdf>) and the same logic applies to the gender pay gap, cross-country growth comparisons, and many other social issues.^[A ratio can also change due to changes in either the numerator or the denominator.]


| Department   | Men                  | Women                 | Total                 |
|-------------|-----------------------|-----------------------|-----------------------|
| English     | 100 (40, **$40\%$**)      | 350 (150, **$43\%$**)     | 450 (190, **$42\%$**)     |
| Engineering | 300 (160, **$53\%$**)     | 50 (30, **$60\%$**)       | 350 (190, **$54\%$**)     |
| Total       | 400 (200, **$50\%$**)     | 400 (180, **$45\%$**)     | 800 (380, **$48\%$**)     |

: School Applicants (Admitted), by Sex and Department


:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
Explore this issue further with real data.
```{r}
UCBAdmissions
```
:::

The same issue shows up in continuous data.
The figure below shows three groups.
Within each group the relationship between $X_1$ and $X_2$ is negative, but the group centers rise together from lower-left to upper-right.
```{r, echo=FALSE}
set.seed(5)
library(mvtnorm)
N <- 400
### Group 1
mu1    <- c(0, 0)
sigma1 <- rbind(c(4, -0.9), c(-0.9, 2) )
g1     <- data.frame(rmvnorm(n=N, mu1, sigma1), Group=1)
### Group 2
mu2    <- c(3, 3)
sigma2 <- rbind(c(2, -0.9), c(-0.9, 1) )
g2     <- data.frame(rmvnorm(n=N, mu2, sigma2), Group=2)
### Group 3
mu3    <- c(6, 6)
sigma3 <- rbind(c(3, -0.9), c(-0.9, 3) )
g3     <- data.frame(rmvnorm(n=N, mu3, sigma3), Group=3)

## Combined Data (Initially Observed)
df_agg <- rbind(g1, g2, g3)
df_agg <- df_agg[order(df_agg$X1), ]

## Scatterplot, colored by group
pt_cols <- c(rgb(.8, 0, 0, .2), rgb(0, .8, 0, .2), rgb(0, 0, .8, .2))
plot(X2 ~ X1, df_agg, pch=16, col=pt_cols[df_agg$Group], cex=.8,
    main=NA, xlab='X1', ylab='X2')

# Within-group regression lines are negative.
ln_cols <- c(rgb(.8, 0, 0, .8), rgb(0, .8, 0, .8), rgb(0, 0, .8, .8))
for (g in seq_len(3)) {
    abline(
        lm(X2 ~ X1, data=df_agg[df_agg$Group == g, ]),
        col=ln_cols[g], lwd=2)
}

# The pooled regression line is positive.
abline(lm(X2 ~ X1, data=df_agg), lwd=3, lty=2)
legend('topleft', bty='n', cex=.8,
    legend=c('within group', 'pooled'),
    lwd=c(2, 3), lty=c(1, 2), col=c(grey(0, .7), 'black'))
```
Each within-group regression line (solid) slopes downward, yet the pooled regression line (dashed) slopes upward.
A researcher who ignored the groups would report a positive relationship that holds within no single group.
The [Simple Regression](02_14_SimpleRegression.qmd) chapter explains how these lines are fit.
Neither the pooled line nor the within-group lines alone identify what would happen under an intervention.

## Association Is Not Causation


Sampling variability is one problem: relationships in the population may not appear in a sample, and apparent sample relationships may not exist in the population.
[Inference](02_16_Inference.qmd) helps quantify that uncertainty, but statistical significance does not distinguish association from causation.
Two other problems remain: real relationships can average out, and random data can acquire mechanically induced relationships.
To make both problems concrete, we focus on the Pearson correlation from [Statistics of Association](02_13_Associations.qmd#cardinal-data) and examine

* Causation without correlation
* Correlation without causation


#### **Causation without correlation** {-}

Examples of the first problem include *nonlinear effects* and *heterogeneous effects* that average out.

```{r}
set.seed(123)
n <- 10000

# X causes Y via Y = X^2 + noise
X <- runif(n, min = -1, max = 1)
epsilon <- rnorm(n, mean = 0, sd = 0.1)
Y <- X^2 + epsilon  # clear causal effect of X on Y
plot(X, Y, pch=16, col=grey(0, .05),
    main=NA, xlab='X', ylab='Y')

# Correlation over the entire range
title( paste0('Cor: ', round( cor(X, Y), 1) ) ,
    font.main=1)
```

```{r}
# Heterogeneous Effects
X <- rnorm(n) # Randomized 'treatment'

# Heterogeneous effects based on group
group <- rbinom(n, size = 1, prob = 0.5)
epsilon <- rnorm(n, mean = 0, sd = 1)
Y <- ifelse(group == 1,
            X + epsilon,   # positive effect
            -X + epsilon)  # negative effect
plot(X, Y, pch=16, col=grey(0, .05),
    main=NA, xlab='X', ylab='Y')

# Correlation in the pooled sample
title( paste0('Cor: ', round( cor(X, Y), 1) ),
    font.main=1 )
```


#### **Correlation without causation** {-}
Examples of the second problem include *shared denominators* and *selection bias* that induce correlations.

Consider three completely random variables.
We can induce a mechanical relationship between the first two variables by dividing them both by the third variable.

```{r}
set.seed(123)
n <- 20000

# Independent components
A <- runif(n)
B <- runif(n)
C <- runif(n)
par(mfrow=c(1, 2))
plot(A, B, pch=16, col=grey(0, .05),
    main=NA, xlab='A', ylab='B')
title('Independent Variables', font.main=1)

# Ratios with a shared denominator
X <- A / C
Y <- B / C
plot(X, Y, pch=16, col=grey(0, .05),
    main=NA, xlab='X', ylab='Y')
title('With Common Divisor', font.main=1)

# Correlation
cor(X, Y)
```

Consider an admissions rule into university: applicants are accepted if they have either high test scores or strong extracurriculars.
Even if there is no general relationship between test scores and extracurriculars, you will see one amongst university students.

```{r}

# Independent traits in the population
test_score        <- rnorm(n, mean = 0, sd = 1)
extracurriculars  <- rnorm(n, mean = 0, sd = 1)

# Selection above thresholds
threshold <- 1.0
admitted <- (test_score > threshold) | (extracurriculars > threshold)
mean(admitted)  # admission rate

par(mfrow = c(1, 2))
# Full population
plot(test_score, extracurriculars,
     pch=16, col=grey(0, .05),
     main=NA, xlab='Test Score', ylab='Extracurriculars')
title('General Sample', font.main=1)
# Admitted only
plot(test_score[admitted], extracurriculars[admitted],
     pch=16, col=grey(0, .05),
     main=NA, xlab='Test Score', ylab='Extracurriculars')
title('University Sample', font.main=1)
# Correlation among admitted applicants only
cor(test_score[admitted], extracurriculars[admitted])
```


:::{.callout-note icon=false collapse="true" title="Must Know"}
In the dataset below, compute the correlation statistic and comment on whether any relationship is causal.
```{r}
plot(Murder ~ Assault, USArrests, pch=16, col=grey(0, .5),
    main=NA, xlab='Assault Arrests', ylab='Murder Arrests')
```
:::

Note that the examples above are not the only examples of "correlation does not mean causation".
Many real datasets have temporal and spatial interdependence that create additional issues.
Many real datasets also have economic interdependence, which also creates additional issues.
The [Observational Data](03_26_ObservationalData.qmd) and [Experimental Data](03_27_ExperimentalData.qmd) chapters return to the designs and assumptions needed for causal claims.

## Bias-Variance Tradeoff

The applied [Bias-Variance Tradeoff](02_16_Inference.qmd#bias-variance-tradeoff) section compared narrow and wide LOESS spans.
We now use simulations where the true data-generating process is known to separate bias from variance and study how both change across samples.

#### **Sampling Distributions**{-}

Bootstrap confidence bands are approximations: they estimate how much predicted values vary from sample to sample.
The simulation below compares a bootstrap CI constructed from one sample to the true sampling variation observed across many independent samples.

```{r}
## Ages
Xmx <- 70
Xmn <- 15

##Generate Sample Data
dat_sim <- function(n=1000){
    X <- seq(Xmn, Xmx, length.out=n)
    ## Random Productivity
    e <- runif(n, 0, 1E6)
    beta <-  1E-10*exp(1.4*X -.015*X^2)
    Y    <-  (beta*X + e)/10
    return(data.frame(Y, X))
}
dat0 <- dat_sim(1000)
dat0 <- dat0[order(dat0[, 'X']), ]

## Data from one sample
plot(Y ~ X, data=dat0, pch=16, col=grey(0, .05),
    main=NA, ylab='Yearly Productivity ($)', xlab='Age' )
reg_lo <- loess(Y ~ X, data=dat0, span=.8)

## Plot Bootstrap CI for Single Sample
pred_design <- data.frame(X=seq(Xmn, Xmx))
preds_lo <- predict(reg_lo, newdata=pred_design)
boot_lo <- matrix(NA, nrow=nrow(pred_design), ncol=399)
for (b in 1:399) {
    dat0_i <- dat0[sample(nrow(dat0), replace=TRUE), ]
    reg_i <- loess(Y ~ X, data=dat0_i, span=.8)
    boot_lo[, b] <- predict(reg_i, newdata=pred_design)
}
boot_cb <- apply(boot_lo, 1, quantile,
    probs=c(.025, .975), na.rm=TRUE)
polygon(
    c(pred_design[[1]], rev(pred_design[[1]])),
    c(boot_cb[1, ], rev(boot_cb[2, ])),
    col=hcl.colors(3, alpha=.25)[2],
    border=NA)

# Construct CI across Multiple Samples
sample_lo <- matrix(NA, nrow=nrow(pred_design), ncol=399)
for (b in 1:399) {
    xy_b <- dat_sim(1000) #Entirely new sample
    reg_b <- loess(Y ~ X, data=xy_b, span=.8)
    sample_lo[, b] <- predict(reg_b, newdata=pred_design)
}
ci_lo <- apply(sample_lo, 1, quantile,
    probs=c(.025, .975), na.rm=TRUE)
polygon(
    c(pred_design[[1]], rev(pred_design[[1]])),
    c(ci_lo[1, ], rev(ci_lo[2, ])),
    col=grey(0, alpha=.25),
    border=NA)
```

#### **Bias**{-}

The plot below shows the average prediction across $B$ samples against the true mean.
Notice that LOESS tracks the true mean closely.
Notice that a linear model does not.


```{r}
set.seed(123)

## Shared parameters used by Bias and Variance
B        <- 300
n_bias   <- 200
Nseq     <- seq(50, 500, by=50)
span_lo  <- 0.5
true_m   <- function(x) (1E-10*exp(1.4*x - .015*x^2)*x + 5E5) / 10
x0       <- seq(Xmn, Xmx, length.out=120)
m0       <- true_m(x0)
pred_design <- data.frame(X=x0)

## Storage for Bias predictions (filled via <<- when n == n_bias)
pred_lm_bias <- matrix(NA_real_, nrow=length(x0), ncol=B)
pred_lo_bias <- matrix(NA_real_, nrow=length(x0), ncol=B)

## One sweep: gradient SEs across all n, plus bias predictions at n_bias
SE <- matrix(NA, nrow=2, ncol=length(Nseq))
for (ni in seq_along(Nseq)) {
    n <- Nseq[ni]
    stats <- matrix(NA, nrow=2, ncol=B)
    for (b in 1:B) {
        dat_b  <- dat_sim(n)
        fit_lm <- lm(Y ~ X, data=dat_b)
        fit_lo <- loess(Y ~ X, data=dat_b, span=0.8)
        grad_lo <- diff(predict(fit_lo))/diff(dat_b[, 'X'])
        if(n == n_bias){
            pred_lm_bias[, b] <- predict(fit_lm, newdata=pred_design)
            fit_lo_b <- loess(Y ~ X, data=dat_b, span=span_lo)
            pred_lo_bias[, b] <- predict(fit_lo_b, newdata=pred_design)
        }
        stats[, b] <- c(coef(fit_lm)[2], mean(grad_lo, na.rm=TRUE))
    }
    SE[, ni] <- apply(stats, 1, sd)
}
```

```{r}
m_lm <- rowMeans(pred_lm_bias, na.rm=TRUE)
m_lo <- rowMeans(pred_lo_bias, na.rm=TRUE)

y_rng <- range(c(m0, m_lm, m_lo), na.rm=TRUE)
plot(x0, m0, type='l', lwd=2, col=rgb(0, 0, 0, .8), ylim=y_rng,
  xlab='Age', ylab=' Mean Productivity ($)', main=NA)
lines(x0[is.finite(m_lm)], m_lm[is.finite(m_lm)], lwd=2, col=rgb(1, 0, 0, .8))
lines(x0[is.finite(m_lo)], m_lo[is.finite(m_lo)], lwd=2, col=rgb(0, 0, 1, .8))
legend('topright', lty=1, lwd=2, col=c(rgb(0, 0, 0, .8), rgb(1, 0, 0, .8), rgb(0, 0, 1, .8)),
  legend=c('True mean',
    'Linear',
    paste0('Loess(', span_lo, ')')
))
```


#### **Variance**{-}

The model estimates also vary from sample to sample.
Notice that the OLS slope coefficient generally varies less than the LOESS mean gradient, even though it is biased.

Also notice that there are diminishing returns to larger sample sizes.
Both the OLS slope coefficient and the loess mean gradient vary less from sample to sample as $n$ grows, making hypothesis tests more accurate.

```{r}
cols <- c(rgb(1, 0, 0, .8), rgb(0, 0, 1, .8))
matplot(Nseq, t(SE), type='b', pch=16,
    lty=1, lwd=2, col=cols,
    ylab='standard error', xlab='sample size',
    main=NA)
legend('topright',
    lty=1, lwd=2, col=cols,
    legend=c('OLS slope', 'Loess mean gradient'))
```




#### **Bias-Variance Decomposition**{-}

The applied section in [Inference](02_16_Inference.qmd#bias-variance-tradeoff) introduced the bias-variance tradeoff intuitively with the bandwidth knob; here we decompose it formally and check both pieces with simulations.

:::{.callout-important icon=false title="Key Definition" collapse="false"}
The expected prediction error of an estimator decomposes into two pieces.
Squared *bias* is how far its average prediction sits from the truth.
*Variance* is how much its predictions wobble around their own average.
:::

The bias-variance decomposition is useful as the formal expression of why we cannot simply minimize one source of error and ignore the other: cutting one usually raises the other, so there is a tradeoff to navigate.
Concretely:

* High-bias / low-variance models.
  E.g., strict linear models are stable across samples but can miss curvature.
* Low-bias / high-variance models.
  E.g., very flexible local linear models can capture curvature but may overfit noise.

:::{.callout-note icon=false collapse="true" title="Must Know"}
```{r}
# Bias functions
mean_pred <- function(P) rowMeans(P, na.rm=TRUE)
avg_bias2 <- function(P) mean( (mean_pred(P) - m0)^2, na.rm=TRUE)

# Variance
var_pred <- function(P) apply(P, 1, var, na.rm=TRUE)
avg_var   <- function(P) mean(var_pred(P), na.rm=TRUE)

metrics <- rbind(
  c(avg_bias2(pred_lm_bias),  avg_var(pred_lm_bias)),
  c(avg_bias2(pred_lo_bias),  avg_var(pred_lo_bias))
)
colnames(metrics) <- c('Avg Bias^2', 'Avg Variance')
rownames(metrics) <- c('Linear', 'Loess')
round(metrics, 4)
```
:::

#### **Consistency**{-}

The bias-variance tradeoff raises a question: can both be reduced simultaneously as $n$ grows?
For local regression, the answer is yes, provided we shrink the bandwidth as $n$ grows.
The neighborhood shrinks (reducing bias) while the local sample size still grows (reducing variance).
This is often written as bandwidth conditions:
$$
h_n \to 0
\quad\text{and}\quad
n h_n \to \infty.
$$
In LOESS language, this corresponds to the span shrinking with $n$, but not too fast.

:::{.callout-note icon=false collapse="true" title="Must Know"}
A common bandwidth rule shrinks $h$ with the sample size, for example $h_{n}=n^{-1/5}$.
Check the two conditions:

* $h_{n}=n^{-1/5}\to 0$: at $n=100$ it is about $0.40$; at $n=10{,}000$ it is about $0.16$.
  The neighborhood shrinks.
* $n h_{n}=n^{4/5}\to\infty$: at $n=100$ it is about $40$; at $n=10{,}000$ it is about $1585$.
  The local sample size still grows.

So bias falls (the window narrows) while variance falls (each local fit uses more data).
```{r}
n <- c(100, 10000)
h_n <- n^(-1/5)
rbind(h_n=round(h_n, 3), n_times_h=round(n*h_n, 1))
```
:::

The simulation below illustrates this at one target point $x_0$: absolute error in estimating $M(x_0)$ tends to decrease with larger $n$.

```{r}
set.seed(42)

x0_target <- 40  # mid-career age
n_grid <- c(60, 120, 240, 480)
R <- 120

avg_abs_err <- numeric(length(n_grid))
for (ni in seq_along(n_grid)) {
  n <- n_grid[ni]
  span_n <- min(0.9, 1.8*n^(-1/4)) # shrinks with n
  errs <- replicate(R, {
    dat_b <- dat_sim(n)
    fit <- loess(Y ~ X, data=dat_b, span=span_n, degree=1)
    mhat <- predict(fit, newdata=data.frame(X=x0_target))
    abs(mhat - true_m(x0_target))
  })
  avg_abs_err[ni] <- mean(errs, na.rm=TRUE)
}

plot(n_grid, avg_abs_err, type='b', pch=16,
     xlab='Sample size (n)',
     ylab='Average Bias',
     main=NA)
```

A similar result holds for OLS when the true relationship is linear.
Even with unlimited data, however, a misspecified model cannot recover the true conditional mean.
These results concern prediction across samples; they do not establish that changing $X$ would cause $Y$ to change.

## Exercises

1. Comment the script you wrote for this chapter, then restart R and check that it runs from a clean session, then check the script with AI as explained in [Working with AI](04_A_WorkingWithAI.qmd).
   Write three sentences from memory on the main statistical idea of this chapter, and ask the assistant what is wrong, vague, or missing.
   Finish with your own questions about whatever you found hardest.

2. Use `UCBAdmissions` to compute the overall admission rate for men and women and the admission rate for each group within each department.
   Identify any reversal between the pooled and within-department relationships, then explain why that reversal does not by itself establish whether admissions were discriminatory.

3. Give an example of a dataset where Pearson's correlation $\hat{R}_{XY}$ is close to zero but there is a clear causal relationship between $X$ and $Y$.
   What feature of the relationship causes the correlation to miss it, and which alternative statistic from [Statistics of Association](02_13_Associations.qmd) might detect it?

4. Explain why a model can have low prediction error but still fail to estimate the causal effect of $X$ on $Y$.
   In your answer, distinguish the bias-variance tradeoff from confounding and selection bias.

#### **Further Reading** {-}

* <http://homepage.stat.uiowa.edu/~mbognar/1030/Bickel-Berkeley.pdf> -- the Berkeley admissions study behind the classic Simpson's-paradox example.
* <https://www.khanacademy.org/math/statistics-probability/describing-relationships-quantitative-data/scatterplots-and-correlation/v/correlation-and-causality> -- short video on why correlation does not imply causation.
* <https://www.khanacademy.org/computing/computer-science/informationtheory/moderninfotheory/v/bias-variance> -- visual introduction to the bias-variance tradeoff.

#### **Recall** {-}

Simpson's paradox showed that a pooled relationship can reverse every within-group relationship.
The correlation examples showed both causation without linear correlation and correlation without causation.
The bias-variance analysis showed how model flexibility affects prediction across samples, but predictive performance alone does not identify a causal effect.
The next chapter pulls back from samples and develops the [Bivariate Probability](02_18_BivariateProbability.qmd) machinery that says what the book's statistics estimate.
