#  Testing Theory
***

## Theoretical Distributions

Just as with one sample tests, we can compute standardized differences. The observed difference is converted into a $t$-statistic by dividing by an estimated standard error. Note, however, that we have to compute the standard error for the difference statistic, which is a bit more complicated than for a single sample. Under the assumption that both populations are independently distributed, we can analytically derive the sampling distribution for the differences between two groups.

#### **Differences in Means**. {-}

In particular, the $t$-statistic is used to compare two groups.
\begin{eqnarray}
\hat{t} = \frac{
    \hat{M}_{Y1} - \hat{M}_{Y2}
}{
    \sqrt{\hat{S}_{Y1}+\hat{S}_{Y2}}/\sqrt{n}
},
\end{eqnarray}
With normally distributed means, this statistic follows Student's [t-distribution](https://en.wikipedia.org/wiki/Student%27s_t-distribution).
Welch's [$t$-statistic](https://en.wikipedia.org/wiki/Welch%27s_t-test) is an adjustment for two normally distributed populations with potentially unequal variances or sample sizes. With the above assumptions, one can conduct hypothesis tests entirely using math.

```{r}
# Sample 1 (e.g., males)
n1 <- 100
Y1 <- rnorm(n1, 0, 2)
#hist(Y1, freq=F, main='Sample 1')

# Sample 2 (e.g., females)
n2 <- 80
Y2 <- rnorm(n2, 1, 1)
#hist(Y2, freq=F, main='Sample 2')

t.test(Y1, Y2, var.equal=F)
```


```{r, eval=F, echo=F}
## Manual T-stat
n1  <- length(Y1)
n2  <- length(Y2)
d <- mean(Y1)-mean(Y2)
se_hat <- sqrt(var(Y1)/n1 + var(Y2)/n2);
t_obs <- d/se_hat

# Alt estimate
s1  <- var(Y1)
s2  <- var(Y2)
s   <- ((n1-1)*s1 + (n2-1)*s2)/(n1+n2-2)
d_se <- sqrt(s*(1/n1+1/n2))
# t stat
t_stat <- d/d_se
t_stat
```

#### **Differences in Medians**. {-}

If we want to test for the differences in medians across groups with independent observations, we can also use notches in the boxplot. If the notches of two boxes do not overlap, then there is rough evidence that the difference in medians is statistically significant. The square root of the sample size is also shown as the bin width in each boxplot.^[Let each group $g$ have median $\tilde{M}_{g}$, interquartile range $\hat{IQR}_{g}$, observations $n_{g}$. We can compute standard deviation of the median as $\tilde{S}_{g}= \frac{1.25 \hat{IQR}_{g}}{1.35 \sqrt{n_{g}}}$. As a rough guess, the interval $\tilde{M}_{g} \pm 1.7 \tilde{S}_{g}$ is the historical default and displayed as a *notch* in the boxplot. See also <https://www.tandfonline.com/doi/abs/10.1080/00031305.1978.10479236>.]

```{r}
Y3 <- rnorm(n1, 3, 3)

boxplot(Y1, Y2, Y3,
    col=c(
        rgb(1,0,0,.5),
        rgb(0,1,0,.5),
        rgb(0,0,1,.5)),
    notch=T,
    varwidth=T)
```

There are also non-parametric alternatives for distributional comparisons. The [Kruskal-Wallis test](https://en.wikipedia.org/wiki/Kruskal%E2%80%93Wallis_test) compares medians across multiple groups using ranks rather than assuming normality.
```{r, eval=F}
library(Ecdat)
data(Caschool)
Caschool[,'stratio'] <- Caschool[,'enrltot']/Caschool[,'teachers']
kruskal.test(Caschool[,'stratio'], Caschool[,'county'])

# Multiple pairwise tests
# pairwise.wilcox.test(Caschool[,'stratio'], Caschool[,'county'])
```

#### **Regression Slope**. {-}

The same ideas apply to regression coefficients. Recall that the OLS slope $\hat{b}_1 = \hat{C}_{XY}/\hat{V}_X$ is a statistic computed from the data. In the Simple Regression chapter, we estimated variability using data-driven methods (jackknife and bootstrap). Under standard assumptions, we can instead derive the sampling distribution analytically: the $t$-statistic
\begin{eqnarray}
\hat{t} = \frac{\hat{b}_1}{\hat{s}_{\hat{b}_1}}
\end{eqnarray}
follows a $t$-distribution with $n-2$ degrees of freedom, where $\hat{s}_{\hat{b}_1}$ is the standard error of the slope. This allows us to test $H_0: \beta_1 = 0$ using math alone.

```{r}
# USArrests data (same as Simple Regression chapter)
xy <- USArrests[, c('Murder', 'UrbanPop')]
colnames(xy) <- c('y', 'x')
reg <- lm(y ~ x, dat = xy)

# The summary reports theoretical t-statistics and p-values
summary(reg)$coefficients
```

```{r}
# Manual verification of the theoretical t-statistic
n <- nrow(xy)
b1 <- coef(reg)[2]
e <- resid(reg)
s_e <- sqrt(sum(e^2) / (n - 2))               # residual standard error
s_b1 <- s_e / sqrt(sum((xy[,'x'] - mean(xy[,'x']))^2)) # SE of slope
t_stat <- b1 / s_b1

# Compare with summary output
c(manual = t_stat, summary = summary(reg)$coefficients[2, 3])

# Two-sided p-value from t-distribution
2 * pt(-abs(t_stat), df = n - 2)
```

We can compare the bootstrap sampling distribution of $\hat{t}$ (histogram) with the theoretical $t_{n-2}$ distribution (curve). They should be similar, but are not identical because the theoretical result assumes normally distributed errors.

```{r}
# Bootstrap distribution of the t-statistic
B <- 399
boot_t <- vector(length = B)
for (b in seq(B)) {
    b_id <- sample(nrow(xy), replace = TRUE)
    xy_b <- xy[b_id, ]
    reg_b <- lm(y ~ x, dat = xy_b)
    # Use theoretical SE from summary
    boot_t[b] <- summary(reg_b)$coefficients[2, 3]
}

# Overlay: bootstrap histogram vs. theoretical t-distribution
t_range <- range(c(boot_t, -4, 4))
hist(boot_t, breaks = 30, freq = FALSE, border = NA,
    col = grey(0.5, 0.5),
    xlim = t_range,
    xlab = expression(hat(t)),
    main = 'Bootstrap vs. Theoretical Distribution',
    font.main = 1)
# Theoretical t(n-2) density
t_grid <- seq(t_range[1], t_range[2], length = 200)
lines(t_grid, dt(t_grid, df = n - 2), lwd = 2)
legend('topright', legend = c('Bootstrap', 't(n-2)'),
    fill = c(grey(0.5, 0.5), NA), border = NA,
    lwd = c(NA, 2), lty = c(NA, 1), bty = 'n')
```

## Type II Errors

When we test a hypothesis, we start with a claim called the null hypothesis $H_0$ and an alternative claim $H_A$. Because we base conclusions on sample data, which has variability, mistakes are possible. There are two types of errors:

* *Type I Error*: Rejecting a true null hypothesis. (False Positive).
* *Type II Error*: Failing to reject a false null hypothesis (False Negative).

| True Situation | Decision: Fail to Reject $H_0$ | Decision: Reject $H_0$ |
|---|---|---|
| $H_0$ is True  |  Correct (no detection)  |  Type I Error (False Positive) |
| $H_0$ is False |  Type II Error (False Negative; missed detection) | Correct (effect detected) |

:::{.callout-tip icon=false collapse="true"}
Here is a Courtroom example: Someone suspected of committing a crime is at trial, and they are either guilty or not (a Bernoulli random variable). You hypothesize that the suspect is innocent, and a jury can either convict them (decide guilty) or free them (decide not-guilty). Recall that fail-to-reject a hypothesis does not mean accepting it, so deciding not-guilty does not necessarily mean innocent.

| True Situation | Decision: Free | Decision: Convict |
|---|---|---|
| Suspect Innocent |  Correctly Freed  | Falsely Convicted |
| Suspect Guilty   |  Falsely Freed    | Correctly Convicted |
:::

#### **Statistical Power**. {-}

The probability of Type I Error is called *significance level* and denoted by $Prob(\text{Type I Error}) = \alpha$. The probability of correctly rejecting a false null is called *power* and denoted by $\text{Power} = 1 - \beta = 1 -  Prob(\text{Type II Error})$.

Significance is often chosen by statistical analysts to be $\alpha=0.05$. Power is less often chosen, instead following from a decision about power.


:::{.callout-tip icon=false collapse="true"}
The code below runs a small simulation using a shifted, nonparametric bootstrap. Two-sided test; studentized statistic, for $H0: \mu = 0$

```{r power-sim, eval=F}
# Power for Two-sided test;
# nonparametric bootstrap, studentized statistic
n <- 25
mu <- 0
alpha <- 0.05
B <- 299

sim_reps <- 100

p_values <- vector(length=sim_reps)
for (i in seq(p_values)) {
    # Generate data
    X <- rnorm(n, mean=0.2, sd=1)
    # Observed statistic
    X_bar <- mean(X)
    T_obs <-  (X_bar - mu) / (sd(X)/ sqrt(n)) ##studentized
    # Bootstrap null distribution of the statistic
    T_boot <- vector(length=B)
    X_null <- X - X_bar + mu # Impose the null by recentering
    for (b in seq(T_boot)) {
      X_b <- sample(X_null, size = n, replace = TRUE)
      T_b <- (mean(X_b) - mu) / (sd(X_b)/sqrt(n))
      T_boot[b] <- T_b
    }
    # Two-sided bootstrap p-value
    pval <- mean(abs(T_boot) >= abs(T_obs))
    p_values[i] <- pval
    }
power <- mean(p_values < alpha)
power
```
:::

There is an important Trade-off for fixed sample sizes: Increasing significance (fewer false positive) often lowers power (more false negatives). Generally, power depends on the effect size and sample size: bigger true effects and larger $n$ make it easier to detect real differences (higher power, lower $\beta$).

#### **Regression Slope Example**. {-}

Continuing the USArrests regression, consider testing $H_0: \beta_1 = 0$. We can simulate Type I and Type II error rates by generating data that resemble the original. When the true slope is $0$ (null is true), rejecting is a Type I Error. When the true slope is nonzero (null is false), failing to reject is a Type II Error.

```{r}
# DGP calibrated to USArrests regression
beta0 <- coef(reg)[1]
beta1_true <- coef(reg)[2]
sigma_e <- summary(reg)$sigma
x_mean <- mean(xy[, 'x'])
x_sd <- sd(xy[, 'x'])

# Function: simulate data, return theoretical p-value for slope
sim_pval <- function(n, beta1) {
    X <- rnorm(n, mean = x_mean, sd = x_sd)
    Y <- beta0 + beta1 * X + rnorm(n, 0, sigma_e)
    summary(lm(Y ~ X))$coefficients[2, 4]
}

sim_reps <- 200
alpha <- 0.05

# Type I Error rate: true slope = 0
pvals_null <- replicate(sim_reps, sim_pval(n = 50, beta1 = 0))
type1_rate <- mean(pvals_null < alpha)
type1_rate  # should be close to alpha

# Power: true slope = beta1_true
pvals_alt <- replicate(sim_reps, sim_pval(n = 50, beta1 = beta1_true))
power_est <- mean(pvals_alt < alpha)
power_est
```

```{r}
# Power increases with sample size
n_grid <- c(10, 25, 50, 100, 200)
power_by_n <- sapply(n_grid, function(ni) {
    pvals <- replicate(sim_reps, sim_pval(ni, beta1 = beta1_true))
    mean(pvals < alpha)
})

plot(n_grid, power_by_n, type = 'b', pch = 16,
    xlab = 'Sample size (n)', ylab = 'Power',
    ylim = c(0, 1),
    main = 'Power for Regression Slope', font.main = 1)
abline(h = alpha, lty = 2, col = 'red')
```

## Bayesian Updating

#### **Bayes' Theorem**. {-}

Bayes' theorem maps predictive statements into inferential statements. In bivariate form,
\begin{eqnarray}
Prob(X_i=x \mid Y_i=y)
&=&
\frac{Prob(Y_i=y \mid X_i=x)Prob(X_i=x)}{Prob(Y_i=y)}.
\end{eqnarray}

Interpretation:

* $Prob(X_i=x)$ is the prior probability for $X_i=x$.
* $Prob(Y_i=y|X_i=x)$ is the likelihood of seeing $Y_i=y$ if $X_i=x$ is true.
* $Prob(X_i=x|Y_i=y)$ is the posterior, your updated probability after seeing $Y_i=y$.

A useful way to remember this is
\begin{eqnarray}
\text{Posterior} \propto \text{Likelihood} \times \text{Prior}.
\end{eqnarray}
For two states, posterior odds are prior odds times a likelihood ratio.

Building on the coin-flip examples from the previous chapter, suppose we observe the outcome of the second coin ($Y_i$) and want to update our belief about the first coin ($X_i$).

**Fair coins.** With $Prob(X_i=x,Y_i=y)=1/4$ for all $x,y\in\{0,1\}$, the prior is $Prob(X_i=1)=1/2$ and the likelihoods are
\begin{eqnarray}
Prob(Y_i=1|X_i=1)=1/2,\quad Prob(Y_i=1|X_i=0)=1/2.
\end{eqnarray}
Applying Bayes' theorem,
\begin{eqnarray}
Prob(X_i=1|Y_i=1)
&=& \frac{Prob(Y_i=1|X_i=1)Prob(X_i=1)}{Prob(Y_i=1)}
= \frac{(1/2)(1/2)}{1/2} = \frac{1}{2}.
\end{eqnarray}
The posterior equals the prior because the coins are independent --- seeing one coin tells you nothing about the other.

```{r}
# Fair coins
probs <- c(0.25, 0.25, 0.25, 0.25)
P_xy <- matrix(probs, nrow = 2,
  dimnames = list(c("Y=0","Y=1"), c("X=0","X=1")))

# Marginals
P_x <- colSums(P_xy)
P_y <- rowSums(P_xy)

# Likelihood: Prob(Y=1 | X=x)
PY1_x <- P_xy["Y=1", ] / P_x
PY1_x

# Posterior: Prob(X=1 | Y=1)
posterior_x1 <- (PY1_x["X=1"] * P_x["X=1"]) / P_y["Y=1"]
posterior_x1  # equals the prior
```

**Unfair coins.** Now consider the joint distribution from the previous chapter with `probs <- c(0.4, 0.2, 0.1, 0.3)`:

|    | $x=0$  | $x=1$   | Marginal    |
|:------:|:---:|:---:|:---:|
| $y=0$        |$0.4$|$0.1$|$0.5$|
| $y=1$        |$0.2$|$0.3$|$0.5$|
| **Marginal** |$0.6$|$0.4$| $1$ |

The prior is $Prob(X_i=1)=0.4$. The likelihoods are
\begin{eqnarray}
Prob(Y_i=1|X_i=1) = \frac{0.3}{0.4} = 0.75,\quad
Prob(Y_i=1|X_i=0) = \frac{0.2}{0.6} = \frac{1}{3}.
\end{eqnarray}
Applying Bayes' theorem,
\begin{eqnarray}
Prob(X_i=1|Y_i=1)
&=& \frac{0.75 \times 0.4}{0.75 \times 0.4 + \frac{1}{3}\times 0.6}
= \frac{0.3}{0.3+0.2} = 0.6.
\end{eqnarray}
The prior probability of heads on the first coin was $0.4$, but after seeing heads on the second coin, it increases to $0.6$. The coins are *not* independent: when the second coin shows heads, it is more likely that the first coin also shows heads.

```{r}
# Unfair coins
probs <- c(0.4, 0.2, 0.1, 0.3)
P_xy <- matrix(probs, nrow = 2,
  dimnames = list(c("Y=0","Y=1"), c("X=0","X=1")))
P_xy

# Marginals
P_x <- colSums(P_xy)
P_y <- rowSums(P_xy)

# Likelihood: Prob(Y=1 | X=x)
PY1_x <- P_xy["Y=1", ] / P_x
PY1_x

# Posterior: Prob(X=1 | Y=1)
posterior_x1 <- (PY1_x["X=1"] * P_x["X=1"]) / P_y["Y=1"]
posterior_x1  # prior was 0.4, posterior is 0.6

# Also compute Prob(X=1 | Y=0)
PY0_x <- P_xy["Y=0", ] / P_x
posterior_x1_y0 <- (PY0_x["X=1"] * P_x["X=1"]) / P_y["Y=0"]
posterior_x1_y0  # seeing tails lowers belief to 0.2
```

#### **Bayes' Theorem Applied**. {-}

Suppose a screening test has sensitivity 0.90 and false positive rate 0.08, while prevalence is 0.12:
\begin{eqnarray}
Prob(Y_i=1|X_i=1)=0.90,\quad
Prob(Y_i=1|X_i=0)=0.08,\quad
Prob(X_i=1)=0.12,
\end{eqnarray}
where $X_i=1$ means "condition present" and $Y_i=1$ means "test positive".

Then
\begin{eqnarray}
Prob(X_i=1|Y_i=1)
&=&
\frac{0.90\times0.12}{0.90\times0.12 + 0.08\times0.88}
\approx 0.605.
\end{eqnarray}
Even with a good test, posterior probability depends strongly on prevalence.

```{r}
# States: X in {0,1}, signal Y in {0,1}
# Prior Prob(X_i=1)
p_x1 <- 0.12

# Test characteristics
p_y1_x1 <- 0.90
p_y1_x0 <- 0.08

# Law of total probability for Prob(Y_i=1)
p_y1 <- p_y1_x1 * p_x1 + p_y1_x0 * (1 - p_x1)

# Bayes posterior Prob(X_i=1 | Y_i=1)
p_x1_y1 <- (p_y1_x1 * p_x1) / p_y1
p_x1_y1

# Also compute Prob(X_i=1 | Y_i=0)
p_y0_x1 <- 1 - p_y1_x1
p_y0_x0 <- 1 - p_y1_x0
p_y0 <- p_y0_x1 * p_x1 + p_y0_x0 * (1 - p_x1)
p_x1_y0 <- (p_y0_x1 * p_x1) / p_y0
p_x1_y0
```

#### **Regression Slope Example**. {-}

The screening test has the same structure as testing a regression slope. Let $X_i=1$ mean "slope is truly nonzero" and $Y_i=1$ mean "the test rejects $H_0$." Then sensitivity equals power, the false positive rate equals $\alpha$, and prevalence is your prior belief that the slope is nonzero. Using the simulation results from the previous section:

```{r}
# Prior: 50-50 chance the slope is nonzero
prior_nonzero <- 0.5

# "Sensitivity" = power, "False positive rate" = alpha
p_reject <- power_est * prior_nonzero + alpha * (1 - prior_nonzero)

# Posterior: Prob(slope nonzero | test rejects)
post_nonzero <- (power_est * prior_nonzero) / p_reject
post_nonzero
```

Even with moderate power, a rejection substantially updates the posterior probability that the slope is nonzero. A weaker prior (e.g., $Prob(\beta_1 \neq 0) = 0.1$) would require stronger evidence to reach the same posterior.

```{r}
# How the posterior changes with different priors
prior_grid <- seq(0.01, 0.99, by = 0.01)
post_grid <- (power_est * prior_grid) /
    (power_est * prior_grid + alpha * (1 - prior_grid))

plot(prior_grid, post_grid, type = 'l', lwd = 2,
    xlab = 'Prior Prob(slope nonzero)',
    ylab = 'Posterior Prob(slope nonzero | reject)',
    main = 'Bayesian Interpretation of Rejection',
    font.main = 1)
abline(0, 1, lty = 2, col = grey(0.5))
```

## Further Reading

Many introductory econometrics textbooks have a good appendix on probability and statistics. See also the further reading about Probability Theory in the Statistics chapter.

* <https://www.r-bloggers.com/2024/03/calculating-conditional-probability-in-r/>
