# Multivariate Data I
***

First, note that you can summarize a dataset with multiple variables using the previous tools.

```{r}
# Inspect Dataset on police arrests for the USA in 1973
head(USArrests)

library(psych)
pairs.panels( USArrests[,c('Murder','Assault','UrbanPop')],
    hist.col=grey(0,.25), breaks=30, density=F, hist.border=NA, # Diagonal
    ellipses=F, rug=F, smoother=F, pch=16, col='red' # Lower Triangle
    )
```

You can also use size, color, and shape to distinguish conditional relationships.
```{r}
# High Assault Areas
assault_high <- USArrests$Assault > median(USArrests$Assault)
col_high <- rgb(1,0,0,.5)
col_low <- rgb(0,0,1,.5)
cols <- ifelse(assault_high, col_high, col_low)

# Scatterplot
# Show High Assault Areas via 'cex=' or 'pch='
# Could further add regression lines for each data split
plot(Murder~UrbanPop, USArrests, pch=16, col=cols)

outer_legend <- function(...) {
  opar <- par(fig=c(0, 1, 0, 1), oma=c(0, 0, 0, 0), 
    mar=c(0, 0, 0, 0), new=TRUE)
  on.exit(par(opar))
  plot(0, 0, type='n', bty='n', xaxt='n', yaxt='n')
  legend(...)
}
outer_legend('topright',
    legend=c('many assualts', 'few assaults'),
    pch=16, col=c(col_high, col_low),
    horiz=T, cex=1, bty='n')
```



See also <https://plotly.com/r/bubble-charts/>
```{r, message=F}
library(plotly)
# Scatter Plot
USArrests$ID <- rownames(USArrests)
fig <- plot_ly(
    USArrests, x = ~UrbanPop, y = ~Assault,
    mode='markers',
    type='scatter',
    hoverinfo='text',
    text = ~paste('<b>', ID, '</b>',
        "<br>Urban  :", UrbanPop,
        "<br>Assault:", Assault,
        "<br>Murder :", Murder),
    color=~Murder,
    marker=list(
        size=~Murder,
        opacity=0.5,
        showscale=T,  
        colorbar = list(title='Murder Arrests (per 100,000)')))
fig <- layout(fig,
    showlegend=F,
    title='Crime and Urbanization in America 1975',
    xaxis = list(title = 'Percent of People in an Urban Area'),
    yaxis = list(title = 'Assault Arrests per 100,000 People'))
fig
```



## Multiple Linear Regression

With $K$ variables, the linear model is
\begin{eqnarray}
\hat{Y}_{i}=b_0 + b_1 \hat{X}_{i1}+ b_2 \hat{X}_{i2}+\ldots+b_K \hat{X}_{iK}+e_i
\end{eqnarray}
Our objective is
\begin{eqnarray}
\min_{b_0 ~~ b_1 ~~... ~~ b_{K}} \sum_{i=1}^{n} e_{i}^2, 
\end{eqnarray}
which yields the best fitting coefficients $b_0^{*} ~~ b_1^{*} ~~... ~~ b_{K}^{*}$.


```{r}
# Check
reg <- lm(Murder~Assault+UrbanPop, data=USArrests)
coef(reg)
```

To measure the ``Goodness of fit'' of the model, we can again plot our predictions.
```{r}
plot(USArrests$Murder, predict(reg), pch=16, col=grey(0,.5))
abline(a=0,b=1, lty=2)
```

We can also again compute sums of squared errors. Adding random data may sometimes improve the fit, however, so we adjust the $\hat{R}^2$ by the number of covariates $K$.
\begin{eqnarray}
\hat{R}_{yY}^2 = \frac{\hat{ESS}}{\hat{TSS}}=1-\frac{\hat{RSS}}{\hat{TSS}}\\
\hat{R}^2_{\text{adj.}} = 1-\frac{n-1}{n-K}(1-\hat{R}_{yY}^2)
\end{eqnarray}

```{r}
ksims <- 1:30
for(k in ksims){ 
    USArrests[,paste0('R',k)] <- runif(nrow(USArrests),0,20)
}
reg_sim <- lapply(ksims, function(k){
    rvars <- c('Assault','UrbanPop', paste0('R',1:k))
    rvars2 <- paste0(rvars, collapse='+')
    reg_k <- lm( paste0('Murder~',rvars2), data=USArrests)
})
R2_sim <- sapply(reg_sim, function(reg_k){  summary(reg_k)$r.squared })
R2adj_sim <- sapply(reg_sim, function(reg_k){  summary(reg_k)$adj.r.squared })

plot.new()
plot.window(xlim=c(0,30), ylim=c(0,1))
points(ksims, R2_sim)
points(ksims, R2adj_sim, pch=16)
axis(1)
axis(2)
mtext(expression(R^2),2, line=3)
mtext('Additional Random Covariates', 1, line=3)
legend('topleft', horiz=T,
    legend=c('Undjusted', 'Adjusted'), pch=c(1,16))
```


#### **Factor Variables**. {-}

So far, we have discussed cardinal data where the difference between units always means the same thing: e.g., $4-3=2-1$. There are also factor variables

* Ordered: refers to Ordinal data. The difference between units means something, but not always the same thing. For example, $4th - 3rd \neq 2nd - 1st$.
* Unordered: refers to Categorical data. The difference between units is meaningless. For example, $D-C=?$

To analyze either factor, we often convert them into indicator variables or dummies; $\hat{D}_{c}=\mathbf{1}( \text{Factor} = c)$. One common case is if you have observations of individuals over time periods, then you may have two factor variables. An unordered factor that indicates who an individual is; for example $\hat{D}_{i}=\mathbf{1}( \text{Individual} = i)$, and an order factor that indicates the time period; for example $\hat{D}_{t}=\mathbf{1}( \text{Time} \in [\text{month}~ t, \text{month}~ t+1) )$. There are many other cases you see factor variables, including spatial ID's in purely cross sectional data.

Be careful not to handle categorical data as if they were cardinal. E.g., generate city data with Leipzig=1, Lausanne=2, LosAngeles=3, ... and then include city as if it were a cardinal number (that's a big no-no). The same applied to ordinal data; PopulationLeipzig=2, PopulationLausanne=3, PopulationLosAngeles=1.

```{r}
N <- 1000
x <- runif(N,3,8)
e <- rnorm(N,0,0.4)
fo <- factor(rbinom(N,4,.5), ordered=T)
fu <- factor(rep(c('A','B'),N/2), ordered=F)
dA <- 1*(fu=='A')
y <- (2^as.integer(fo)*dA )*sqrt(x)+ 2*as.integer(fo)*e
dat_f <- data.frame(y,x,fo,fu)
```

With factors, you can still include them in the design matrix of an OLS regression. For example, 
\begin{eqnarray}
\hat{Y}_{i} = b_{0} + \hat{X}_{i} b_{1} + \sum_{t}\hat{D}_{t}b_{t}
\end{eqnarray}
When, as commonly done, the factors are modeled as being additively seperable, they are modeled "fixed effects".^[There are also *random effects*: the factor variable comes from a distribution that is uncorrelated with the regressors. This is rarely used in economics today, however, and are mostly included for historical reasons and special cases where fixed effects cannot be estimated due to data limitations.] Simply including the factors into the OLS regression yields a "dummy variable" fixed effects estimator. The fixed effects and dummy variable approach are algebraically equal: same coefficients and same residuals. (See Hansen Econometrics, Theorem 17.1)
<!--
In fact, if the fixed effect is ``fully unstructured then the only way to consistently estimate the coefficient $\beta$ is by an estimator which is invariant'' (Hansen Econometrics, p). 
-->
```{r}
library(fixest)
fe_reg1 <- feols(y~x|fo+fu, dat_f)
coef(fe_reg1)
fixef(fe_reg1)[1:2]

# Compare Coefficients
fe_reg0 <- lm(y~-1+x+fo+fu, dat_f)
coef( fe_reg0 )
```

With fixed effects, we can also compute averages for each group: $\hat{M}_{Yt}=\sum_{i}^{n_{t}} \hat{Y}_{it}/n_{t}$, where each period $t$ has $n_{t}$ observations denoted $\hat{Y}_{it}$. We can then construct a *between estimator*: $\hat{M}_{Yt} = b_{0} + \hat{M}_{Xt} b_{1}$. Or we can subtract the average from each group to construct a *within estimator*: $(\hat{Y}_{it} - \hat{M}_{Yt}) = (\hat{X}_{it}-\hat{M}_{Xt})b_{1}$. 

<!-- 
> The labels "random effects" and "fixed effects" are misleading. These are labels which arose in the early literature and we are stuck with these labels today. In a previous era regressors were viewed as "fixed". Viewing the individual effect as an unobserved regressor leads to the label of the individual effect as "fixed". Today, we rarely refer to regressors as "fixed" when dealing with observational data. We view all variables as random. Consequently describing u i as "fixed" does not make much sense and it is hardly a contrast with the "random effect" label since under either assumption u i is treated as random. Once again, the labels are unfortunate but the key difference is whether u i is correlated with the regressors.
-->



## Variability Estimates

Above, we computed the coefficient $\hat{B}$ for a particular sample: $\hat{X}_{1i}, \hat{X}_{2i}, ..., \hat{Y}_{i}$. We now seek to know how much the best-fitting coefficients $B^{*}$ varies from sample to sample. I.e., $\hat{B}$ is our estimate and we want to know the variance of our estimator: $\mathbb{V}(B^{*})$. To estimate this variability, we can use the same *data-driven* methods introduced previously. 

As before, we can also conduct independent hypothesis tests using t-values. However, we can also conduct *joint tests* that account for interdependancies in our estimates. For example, to test whether two coefficients both equal $0$, we bootstrap the joint distribution of coefficients.

```{r}
# Bootstrap SE's
boots <- 1:399
boot_regs <- lapply(boots, function(b){
    b_id <- sample( nrow(USArrests), replace=T)
    xy_b <- USArrests[b_id,]
    reg_b <- lm(Murder~Assault+UrbanPop, dat=xy_b)
})
boot_coefs <- sapply(boot_regs, coef)

# Recenter at 0 to impose the null
#boot_means <- rowMeans(boot_coefs)
#boot_coefs0 <- sweep(boot_coefs, MARGIN=1, STATS=boot_means)
```

```{r}
boot_coef_df <- as.data.frame(cbind(ID=boots, t(boot_coefs)))
fig <- plotly::plot_ly(boot_coef_df,
    type = 'scatter', mode = 'markers',
    x = ~UrbanPop, y = ~Assault,
    text = ~paste('<b> bootstrap dataset: ', ID, '</b>',
            '<br>Coef. Urban  :', round(UrbanPop,3),
            '<br>Coef. Murder :', round(Assault,3),
            '<br>Coef. Intercept :', round(`(Intercept)`,3)),
    hoverinfo='text',
    showlegend=F,
    marker=list( color='rgba(0, 0, 0, 0.5)'))
fig <- plotly::layout(fig,
    showlegend=F,
    title='Joint Distribution of Coefficients (under the null)',
    xaxis = list(title='UrbanPop Coefficient'),
    yaxis = list(title='Assualt Coefficient'))
fig
```


## Hypothesis Tests

#### **F-statistic**. {-}

We can also use an $F$ test for any $q$ hypotheses. Specifically, when $q$ hypotheses *restrict* a model, the degrees of freedom drop from $k_{u}$ to $k_{r}$ and the residual sum of squares $\hat{RSS}=\sum_{i}(\hat{Y}_{i}-\hat{y}_{i})^2$ typically increases. We compute the statistic
\begin{eqnarray}
\hat{F}_{q} = \frac{(\hat{RSS}_{r}-\hat{RSS}_{u})/(k_{u}-k_{r})}{\hat{RSS}_{u}/(n-k_{u})}
\end{eqnarray}

If you test whether all $K$ variables are significant, the restricted model is a simple intercept and $\hat{RSS}_{r}=\hat{TSS}$, and $\hat{F}_{q}$ can be written in terms of $\hat{R}^2$: $\hat{F}_{K} = \frac{\hat{R}^2}{1-\hat{R}^2} \frac{n-K}{K-1}$. The first fraction is the relative goodness of fit, and the second fraction is an adjustment for degrees of freedom (similar to how we adjusted the $\hat{R}^2$ term before). 

To conduct a hypothesis test, first compute a null distribution by randomly reshuffling the outcomes and recompute the F statistic, and then compare how often random data give something as extreme as your initial statistic. For some intuition on this F test, examine how the adjusted $\hat{R}^2$ statistic varies with bootstrap samples. 
```{r}
# Bootstrap under the null
boots <- 1:399
boot_regs0 <- lapply(boots, function(b){
  # Generate bootstrap sample
  xy_b <- USArrests
  b_id <- sample( nrow(USArrests), replace=T)
  # Impose the null
  xy_b$Murder <-  xy_b$Murder[b_id]
  # Run regression
  reg_b <- lm(Murder~Assault+UrbanPop, dat=xy_b)
})
# Get null distribution for adjusted R2
R2adj_sim0 <- sapply(boot_regs0, function(reg_k){
    summary(reg_k)$adj.r.squared })
hist(R2adj_sim0, xlim=c(-.1,1), breaks=25, border=NA,
    main='', xlab=expression('adj.'~R[b]^2))

# Compare to initial statistic
abline(v=summary(reg)$adj.r.squared, lwd=2, col=2)
```

Note that *hypothesis testing is not to be done routinely*, as additional complications arise when testing multiple hypothesis sequentially.

Under some additional assumptions $F_{q}$ follows an F-distribution. For more about F-testing, see <https://online.stat.psu.edu/stat501/lesson/6/6.2> and <https://www.econometrics.blog/post/understanding-the-f-statistic/>

#### **ANOVA** {-}

Under some additional parametric assumptions about the data generating process, the F-statistic follows an $F$ distribution. This case is well-studied historically, often under the title Analysis of Variance (ANOVA). An important case corresponds to the restricted model having no explanatory variables (i.e., our model is $\hat{Y}_{i}=b_{0}$ and our predictions are $\hat{y}_{i}=\hat{M}_{Y}$).

```{r}
reg_full <- lm(Murder ~ Assault + UrbanPop + Rape, data = USArrests)
reg_none <- lm(Murder ~ 1, dat=USArrests)
anova(reg_none, reg_full)

# Manual F-test
rss0 <- sum(resid(reg_none)^2) # restricted
rss1 <- sum(resid(reg_full)^2) # unrestricted
df0  <- df.residual(reg_none)
df1  <- df.residual(reg_full)
F    <- ((rss0 - rss1)/(df0-df1)) / (rss1/df1) # observed F stat
p    <- 1-pf(F, df0-df1, df1) # where F falls in the F-distribution
cbind(F, p)
```

Whether you take a parametric or nonparametric approach to hypothesis testing, you can easily test whether variables are additively separable with an F test.
```{r}
# Empirical Example
reg1 <- lm(Murder~Assault+UrbanPop, USArrests)
reg2 <- lm(Murder~Assault*UrbanPop, USArrests)
anova(reg1, reg2)
```

```{r, eval=F}
# Simulation Example
N <- 1000
x <- runif(N,3,8)
e <- rnorm(N,0,0.4)
fo <- factor(rbinom(N,4,.5), ordered=T)
fu <- factor(rep(c('A','B'),N/2), ordered=F)
dA <- 1*(fu=='A')
y <- (2^as.integer(fo)*dA )*sqrt(x)+ 2*as.integer(fo)*e
dat_f <- data.frame(y,x,fo,fu)

reg0 <- lm(y~-1+x+fo+fu, dat_f)
reg1 <- lm(y~-1+x+fo*fu, dat_f)
reg2 <- lm(y~-1+x*fo*fu, dat_f)

anova(reg0, reg2)
anova(reg0, reg1, reg2)
```


## Further Reading

For OLS, see

* <https://bookdown.org/josiesmith/qrmbook/linear-estimation-and-minimizing-error.html>
* <https://www.econometrics-with-r.org/4-lrwor.html>
* <https://www.econometrics-with-r.org/6-rmwmr.html>
* <https://www.econometrics-with-r.org/7-htaciimr.html>
* <https://bookdown.org/ripberjt/labbook/bivariate-linear-regression.html>
* <https://bookdown.org/ripberjt/labbook/multivariable-linear-regression.html>
* <https://online.stat.psu.edu/stat462/node/137/>
* <https://book.stat420.org/>
* Hill, Griffiths & Lim (2007), Principles of Econometrics, 3rd ed., Wiley, S. 86f.
* Verbeek (2004), A Guide to Modern Econometrics, 2nd ed., Wiley, S. 51ff.
* Asteriou & Hall (2011), Applied Econometrics, 2nd ed., Palgrave MacMillan, S. 177ff.
* <https://online.stat.psu.edu/stat485/lesson/11/>

To derive OLS coefficients in Matrix form, see

* <https://jrnold.github.io/intro-methods-notes/ols-in-matrix-form.html>
* <https://www.fsb.miamioh.edu/lij14/411_note_matrix.pdf>
* <https://web.stanford.edu/~mrosenfe/soc_meth_proj3/matrix_OLS_NYU_notes.pdf>

For fixed effects, see

* <https://www.econometrics-with-r.org/10-rwpd.html>
* <https://bookdown.org/josiesmith/qrmbook/topics-in-multiple-regression.html>
* <https://bookdown.org/ripberjt/labbook/multivariable-linear-regression.html>
* <https://www.princeton.edu/~otorres/Panel101.pdf>
* <https://www.stata.com/manuals13/xtxtreg.pdf>

