# Multiple Linear Regression
***

We often want to know the relationship between one outcome variable $\hat{Y}_{i}$ and multiple explanatory/predictor variables. For historical reasons, we start by fitting a linear model to data.  

## The Linear Model

#### **Model and Objective**. {-}
Let there be $K$ cardinal explanatory variables: $\hat{X}_{1i},  \hat{X}_{2i},... \hat{X}_{Ki}$. Our model is 
\begin{eqnarray}
\hat{Y}_{i} &=& b_0 + b_1 \hat{X}_{i1}+ b_2 \hat{X}_{i2}+\ldots+b_K \hat{X}_{iK} + e_i \\
 &=& \sum_{k=0}^{K} b_k \hat{X}_{ik} + e_i,
\end{eqnarray}
where $\hat{X}_{i0}=1$ and $e_{i}$ is an error term. 

Our objective is to minimize the sum of squared errors:
\begin{eqnarray}
\min_{b_0 ~~ b_1 ~~... ~~ b_{K}} \sum_{i=1}^{n} e_{i}^2,
\end{eqnarray}
hence the common term *Ordinary Least Squares* regression.

We can analytically solve this problem using [matrix-calculations](https://jadamso.github.io/Rbooks/03_26_MiscTopics.html#matrix-calculations), although we do not do that here (for a detailed derivation, see [these notes](https://jrnold.github.io/intro-methods-notes/ols-in-matrix-form.html)). Solving yields the best fitting model with coefficients $B^{*}=(b_0^{*} ~~ b_1^{*} ~~... ~~ b_{K}^{*})$.

```{r}
# Check
reg <- lm(Murder ~ Assault+UrbanPop, data=USArrests)
coef(reg)
```

:::{.callout-note icon=false collapse="true" title="Must Know"}
Add `Rape` as a third explanatory variable: run `lm(Murder ~ Assault + UrbanPop + Rape, data=USArrests)` and compare the coefficients to those above. Which coefficients change the most? Does the sign of any coefficient change? What does this tell you about the relationship between the explanatory variables?
:::

#### **Goodness of fit**. {-}

To measure how well the model fits the data, we can again plot our predictions.
```{r}
plot(USArrests$Murder, predict(reg), pch=16, col=grey(0, .5),
    main=NA, xlab='Observed Murder', ylab='Predicted Murder')
abline(a=0, b=1, lty=2)
```

We can also again compute sums of squared errors. Adding random data may sometimes improve the fit, however, so we adjust the $\hat{R}^2$ by the number of covariates $K$.
\begin{eqnarray}
\hat{R}_{yY}^2 = \frac{\hat{ESS}}{\hat{TSS}}=1-\frac{\hat{RSS}}{\hat{TSS}}\\
\hat{R}^2_{\text{adj.}} = 1-\frac{n-1}{n-K}(1-\hat{R}_{yY}^2)
\end{eqnarray}

:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
Suppose $n=50$, $K=3$, and $\hat{R}^2=0.70$. Then
$$\hat{R}^2_{\text{adj.}} = 1 - \frac{49}{47}(1-0.70) = 1 - 1.043 \times 0.30 \approx 0.687.$$
The adjustment reduces $\hat{R}^2$ slightly. With more covariates the penalty grows: if instead $K=10$, then $\hat{R}^2_{\text{adj.}} = 1 - \frac{49}{40}(0.30) = 1 - 0.3675 = 0.6325$.
:::

```{r}
ksims <- 1:30
for(k in ksims){ 
    USArrests[, paste0('R', k)] <- runif(nrow(USArrests), 0, 20)
}
R2_sim <- rep(NA, length(ksims))
R2adj_sim <- rep(NA, length(ksims))
for(k in ksims){
    rvars <- c('Assault', 'UrbanPop', paste0('R', 1:k))
    rvars2 <- paste0(rvars, collapse='+')
    reg_k <- lm( paste0('Murder ~ ', rvars2), data=USArrests)
    R2_sim[k] <- summary(reg_k)$r.squared
    R2adj_sim[k] <- summary(reg_k)$adj.r.squared
}

plot.new()
plot.window(xlim=c(0, 30), ylim=c(0, 1))
points(ksims, R2_sim)
points(ksims, R2adj_sim, pch=16)
axis(1)
axis(2)
mtext(expression(R^2), 2, line=3)
mtext('Additional Random Covariates', 1, line=3)
legend('topleft', horiz=T,
    legend=c('Unadjusted', 'Adjusted'), pch=c(1, 16))
```




## Variability Estimates and Hypothesis Tests

Above, we computed the coefficient $\hat{B}$ for a particular sample: $\hat{X}_{1i}, \hat{X}_{2i}, ..., \hat{Y}_{i}$. We now seek to know how much the best-fitting coefficients $B^{*}$ varies from sample to sample. I.e., $\hat{B}$ is our estimate and we want to know the standard error of our estimator: $SE(B^{*})$. To estimate this variability, we can use the same data-driven methods introduced previously. 

As before, we can also conduct independent hypothesis tests using t-values. However, we can also conduct *joint tests* that account for interdependancies in our estimates. For example, to test whether two coefficients both equal $0$, we bootstrap the joint distribution of coefficients.

```{r}
# Bootstrap SE's
boot_coefs <- matrix(NA, nrow=399, ncol=3)
for(b in seq(nrow(boot_coefs))){
    b_id <- sample( nrow(USArrests), replace=T)
    dat_b <- USArrests[b_id, ]
    reg_b <- lm(Murder ~ Assault+UrbanPop, data=dat_b)
    boot_coefs[b, ] <- coef(reg_b)
}
colnames(boot_coefs) <- names(coef(reg))

# Recenter at 0 to impose the null
#boot_means <- rowMeans(boot_coefs)
#boot_coefs0 <- sweep(boot_coefs, MARGIN=1, STATS=boot_means)
```

```{r}
boot_coef_df <- as.data.frame(cbind(
    ID=seq(nrow(boot_coefs)),
    boot_coefs))
fig <- plotly::plot_ly(boot_coef_df,
    type = 'scatter', mode = 'markers',
    x = ~ UrbanPop, y = ~ Assault,
    text = ~ paste('<b> bootstrap dataset: ', ID, '</b>',
            '<br>Coef. Urban  :', round(UrbanPop, 3),
            '<br>Coef. Murder :', round(Assault, 3),
            '<br>Coef. Intercept :', round(`(Intercept)`, 3)),
    hoverinfo='text',
    showlegend=F,
    marker=list( color='rgba(0, 0, 0, 0.5)'))
fig <- plotly::layout(fig,
    showlegend=F,
    title='Joint Distribution of Coefficients (under the null)',
    xaxis = list(title='UrbanPop Coefficient'),
    yaxis = list(title='Assualt Coefficient'))
fig
```


#### **F-statistic**. {-}

We can also use an $F$ test for any $q$ hypotheses. Specifically, when $q$ hypotheses *restrict* a model, the degrees of freedom drop from $k_{u}$ to $k_{r}$ and the residual sum of squares $\hat{RSS}=\sum_{i}(\hat{Y}_{i}-\hat{y}_{i})^2$ typically increases. We compute the statistic
\begin{eqnarray}
\hat{F}_{q} = \frac{(\hat{RSS}_{r}-\hat{RSS}_{u})/(k_{u}-k_{r})}{\hat{RSS}_{u}/(n-k_{u})}
\end{eqnarray}

If you test whether all $K$ variables are jointly significant, the restricted model is a simple intercept and $\hat{RSS}_{r}=\hat{TSS}$, and $\hat{F}_{q}$ can be written in terms of $\hat{R}^2$: $\hat{F}_{K} = \frac{\hat{R}^2}{1-\hat{R}^2} \frac{n-K}{K-1}$. The first fraction is the relative goodness of fit, and the second fraction is an adjustment for degrees of freedom (similar to how we adjusted the $\hat{R}^2$ term before). We can also write this test statistic for whether all variables are statistically significant in terms of the adjusted $\hat{R}^2$:
\begin{eqnarray}
\hat{F}_{K} = \frac{(K-1) + (n-K)\hat{R}^2_{\text{adj.}}}{(K-1)(1 - \hat{R}^2_{\text{adj.}})}
\end{eqnarray}
Notice that $\hat{F}_{K}=1$ when $\hat{R}^2_{\text{adj.}}=0$, meaning the model does not improve fit beyond what random covariates would achieve. The F-statistic exceeds $1$ when $\hat{R}^2_{\text{adj.}}>0$ and falls below $1$ when $\hat{R}^2_{\text{adj.}}<0$.

:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
Suppose $\hat{R}^2_{\text{adj.}}=0.50$, $K=3$, $n=50$. Then
$$\hat{F}_K = \frac{(K-1) + (n-K)\hat{R}^2_{\text{adj.}}}{(K-1)(1 - \hat{R}^2_{\text{adj.}})} = \frac{(3-1) + (50-3)(0.50)}{(3-1)(1-0.50)} = \frac{2+23.5}{2 \times 0.5} = \frac{25.5}{1} = 25.5.$$
This is well above $1$, indicating that the model explains substantially more variation than random covariates. You can verify: `qf(0.95, 2, 47)` $\approx 3.20$, so $\hat{F}_K = 25.5$ is highly significant.
:::

To conduct a hypothesis test, first compute a null distribution by randomly reshuffling the outcomes and recompute the F statistic, and then compare how often random data give something as extreme as your initial statistic. For some intuition on this F test, examine how the adjusted $\hat{R}^2$ statistic varies with bootstrap samples. 
```{r}
# Bootstrap under the null
boots <- 1:399
R2adj_sim0 <- rep(NA, length(boots))
for(b in boots){
  # Generate bootstrap sample
  xy_b <- USArrests
  b_id <- sample( nrow(USArrests), replace=T)
  # Impose the null
  xy_b$Murder <- xy_b$Murder[b_id]
  # Run regression
  reg_b <- lm(Murder ~ Assault+UrbanPop, data=xy_b)
  R2adj_sim0[b] <- summary(reg_b)$adj.r.squared
}
hist(R2adj_sim0, xlim=c(-.1, 1), breaks=25, border=NA,
    freq=F, main=NA, xlab=expression('adj.'~R[b]^2))
title('Bootstrap F-test', font.main=1)

# Compare to initial statistic
abline(v=summary(reg)$adj.r.squared, lwd=2, col=rgb(1, 0, 0, .8))
```

Note that *hypothesis testing is not to be done routinely*, as additional complications arise when testing multiple hypothesis sequentially.

Under some additional assumptions $F_{q}$ follows an F-distribution. For more about F-testing, see <https://online.stat.psu.edu/stat501/lesson/6/6.2> and <https://www.econometrics.blog/post/understanding-the-f-statistic/>



## Coefficient Interpretation

Notice that we have gotten pretty far without actually trying to meaningfully interpret regression coefficients. That is because the above procedure will always give us number, regardless as to whether the true data generating process is linear or not. So, to be cautious, we have been interpreting the regression outputs while being agnostic as to how the data are generated. We now consider a special situation where we know the data are generated according to a linear process and are only uncertain about the parameter values.

If the data generating process is truly linear then we have a famous result that lets us attach a simple interpretation of OLS coefficients as unbiased estimates of the effect of $X$. For example, generate a simulated dataset with $30$ observations and two exogenous variables. Assume the following relationship: $Y_{i} = \beta_0 + \beta_1 X_{i1} + \beta_2 X_{i2} + \epsilon_{i}$ where the variables and the error term are realizations of the following data generating processes:
```{r}
N <- 30
B <- c(10, 2, -1)

x1 <- runif(N, 0, 5)
x2 <- rbinom(N, 1, .7)
X <- cbind(1, x1, x2)
e <- rnorm(N, 0, 3)
Y <- X%*%B + e
dat <- data.frame(Y, X)
coef(lm(Y ~ x1+x2, data=dat))
```

Simulate the distribution of coefficients under a correctly specified model. Interpret the average.
```{r}
N <- 30
B <- c(10, 2, -1)

Coefs <- matrix(NA, nrow=3, ncol=400)
for(sim in 1:400){
    x1 <- runif(N, 0, 5)
    x2 <- rbinom(N, 1, .7)
    X <- cbind(1, x1, x2)
    e <- rnorm(N, 0, 3)
    Y <- X%*%B + e
    dat <- data.frame(Y, x1, x2)
    Coefs[, sim] <- coef(lm(Y ~ x1+x2, data=dat))
}

par(mfrow=c(1, 2))
for(i in 2:3){
    hist(Coefs[i, ], xlab=bquote(beta[.(i)]), main=NA, border=NA, freq=F)
    abline(v=mean(Coefs[i, ]), lwd=2, col=rgb(1, 0, 0, .8))
    abline(v=B[i], col=rgb(0, 0, 1, .8), lty=2)
}
```


Many economic phenomena are nonlinear, even when including potential transforms of $Y$ and $X$. Sometimes the linear model may still be a good or even great approximation. But sometimes not, and it is hard to know ex-ante. Examine the distribution of coefficients under this mispecified model and try to interpret the average.
```{r}
N <- 30

Coefs <- matrix(NA, nrow=3, ncol=600)
for(sim in 1:600){
    x2 <- runif(N, 0, 5)
    x3 <- rbinom(N, 1, .7)
    e <- rnorm(N, 0, 3)
    Y <- 10*x3 + 2*log(x2)^x3 + e
    dat <- data.frame(Y, x2, x3)
    Coefs[, sim] <- coef(lm(Y ~ x2+x3, data=dat))
}

par(mfrow=c(1, 2))
for(i in 2:3){
    hist(Coefs[i, ],  xlab=bquote(beta[.(i)]), main=NA, border=NA, freq=F)
    abline(v=mean(Coefs[i, ]), col=rgb(1, 0, 0, .8), lwd=2)
}
```

:::{.callout-note icon=false collapse="true" title="Must Know"}
Compare the two simulation results. In the correctly specified case, the estimated coefficients center on the true values. What happens in the misspecified case? Why might OLS still be useful even when the model is misspecified?
:::

In general, you can interpret your regression coefficients as "adjusted correlations". There are (many) tests for whether the relationships in your dataset are actually additively separable and linear.

#### **$\hat{R}^2$ vs. $\hat{\beta}$**. {-}

A small $\hat{R}^2$ does not mean your coefficients are useless. The two quantities answer different questions: $\hat{\beta}_{k}$ tells us how $\hat{Y}$ moves with $\hat{X}_{k}$, holding other regressors fixed, while $\hat{R}^2$ tells us how much of the variation in $\hat{Y}$ the model captures. If most of the variation in $\hat{Y}$ comes from things outside the model, $\hat{R}^2$ will be small even when $\hat{\beta}_{k}$ is precisely estimated. For example, wage regressions on years of schooling often report $\hat{R}^2$ around $0.1$, yet the schooling coefficient is one of the most studied quantities in labor economics.

That said, we still care about $\hat{R}^2$. A very small $\hat{R}^2$ can signal that the model is misspecified (missing important variables or interactions, nonlinear relationships). In this case, you are not actually estimating a theoretical parameter.^[Even if the model is correctly specified, we will care about $R^2$.  The unexplained variance affects standard errors, so smaller $\hat{R}^2$ typically widens confidence intervals. Alternatively, if you want to forecast $\hat{Y}$, a small $\hat{R}^2$ means wide prediction intervals.]

## Exercises

1. Explain in your own words why $\hat{R}^2$ can increase when you add a random noise variable to a regression, but $\hat{R}^2_{\text{adj.}}$ generally does not. What does this imply about using $\hat{R}^2$ alone to choose between models with different numbers of covariates?

2. Using the `USArrests` dataset, regress `Murder` on `Assault`, `UrbanPop`, and `Rape`. Report the $\hat{R}^2_{\text{adj.}}$ and the $F$-statistic. Then test whether `Rape` is jointly significant with the other variables by comparing the full model to a restricted model that excludes `Rape`.

3. Using `lm()`, fit the model `Murder ~ Assault + UrbanPop` on the `USArrests` data. Write a bootstrap loop (at least 200 replications) to estimate the standard error of the `Assault` coefficient. Compare your bootstrap standard error to the one reported by `summary()`.

#### **Further Reading**. {-}

* <https://www.econometrics-with-r.org/6-rmwmr.html> -- multiple regression in R with worked examples, hypothesis tests, and diagnostics.
* <https://book.stat420.org/> -- a full applied-statistics textbook covering OLS theory, matrix derivations, and model selection in R.
* <https://www.econometrics-with-r.org/10-rwpd.html> -- panel data and fixed-effects estimation with practical R code.

