# Mixed-Data Regression
***

## Introduction

The previous chapter compared means and distributions across several groups.
This chapter rewrites those group comparisons as regressions, first with treatment coding and then with effect coding.
It then develops two resampling methods with different purposes: permutation tests impose a null relationship by shuffling a factor variable, while subsampling keeps each observation intact and checks how estimates vary when some observations are omitted.
These methods sharpen what a regression can establish, but none turns an observational relationship into a causal effect.
The word *effect* in *effect coding* names a parameterization, not a causal claim.

We continue with `USArrests` and the four-level `state.region` factor from [Comparing Multiple Groups](03_22_MultipleGroups.qmd).

```{r}
# Recreate the regional data so this chapter runs independently.
dat <- USArrests
dat$Region <- state.region
table(dat$Region)
```

## Regression

[ANOVA](03_22_MultipleGroups.qmd#equal-means-anova) is a special case of linear regression where the only explanatory variables are group indicators.

#### **ANOVA** {-}

The model $\hat{Y}_{ig} = b_0 + \sum_{g=2}^{G} b_g \hat{D}_{ig} + \hat{E}_{ig}$ produces the same $F$-statistic as the ANOVA model in the previous chapter.
After fitting, $\hat{B}_0$ estimates the mean of the reference group and each $\hat{B}_g$ estimates the difference between group $g$ and the reference.
This default representation is called *treatment coding*.

```{r}
# Compare the regional means to treatment-coded coefficients.
M_g <- aggregate(Murder ~ Region, data=dat, FUN=mean)[,'Murder']
M_g

# The intercept is the reference mean.
M_g - M_g[1]

# Each remaining coefficient is a difference from the reference.
fit_lm <- lm(Murder ~ Region, data=dat)
coef(fit_lm)
```

The ANOVA table for a regression with a single factor is therefore the same as the table from `aov()`.

```{r}
fit_aov <- aov(Murder ~ Region, data=dat)
anova(fit_aov)
anova(fit_lm)
```

:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
Suppose there are $G=3$ groups with means $\hat{M}_1=4$, $\hat{M}_2=7$, and $\hat{M}_3=5$.
Using group 1 as the reference, the regression coefficients are $\hat{B}_0=4$, $\hat{B}_2=7-4=3$, and $\hat{B}_3=5-4=1$.
The intercept equals the reference-group mean, and each slope equals a difference from that reference.

```{r}
# Verify the treatment-coded coefficients.
Y <- c(3, 4, 5, 6, 7, 8, 4, 5, 6)
g <- factor(c(1, 1, 1, 2, 2, 2, 3, 3, 3))
coef(lm(Y ~ g))
```
:::

#### **Effect Coding** {-}

Treatment coding makes every coefficient depend on which group R selects as the reference.
Effect coding instead compares each displayed group with the mean across all group means.

:::{.callout-important icon=false title="Key Definition" collapse="false"}
With $G$ factor levels, *effect coding* creates $G-1$ columns.
For column $g$, observations in level $g$ receive $1$, observations in the omitted level receive $-1$, and observations in the other levels receive $0$.
In a factor-only regression, $\hat{B}_0$ is the unweighted mean of the $G$ group means, each displayed $\hat{B}_g$ is that group's deviation from the unweighted mean, and the omitted group's effect is $-\sum_{g=1}^{G-1}\hat{B}_g$.
:::

Effect coding is useful when no group is a natural reference and the deviations from a common center are easier to discuss than several pairwise differences.
Because the four regions contain different numbers of states, the unweighted mean of their four means is not the observation-weighted mean across all 50 states.

```{r}
# Display the sum-to-zero contrast matrix used for effect coding.
effect_matrix <- contr.sum(nlevels(dat$Region))
rownames(effect_matrix) <- levels(dat$Region)
effect_matrix

# Apply contrasts to this model only, without changing global options.
fit_effect <- lm(
    Murder ~ Region,
    data=dat,
    contrasts=list(Region='contr.sum')
)
coef(fit_effect)

# Compare each regional mean with the unweighted mean of group means.
region_means <- aggregate(Murder ~ Region, data=dat, FUN=mean)
effect_mean <- mean(region_means[,'Murder'])
region_effects <- region_means[,'Murder'] - effect_mean
data.frame(
    region=region_means[,'Region'],
    group_mean=region_means[,'Murder'],
    coded_effect=region_effects
)

# Recover the omitted West effect from the sum-to-zero restriction.
-sum(coef(fit_effect)[-1])
```

Changing the contrasts changes the meanings of the coefficients, but not the fitted values, residuals, $\hat{R}^2$, or global $F$-statistic.

```{r}
treatment_F <- anova(fit_lm)[['F value']][1]
effect_F <- anova(fit_effect)[['F value']][1]

c(
    maximum_fitted_difference=max(abs(fitted(fit_lm) - fitted(fit_effect))),
    maximum_residual_difference=max(abs(resid(fit_lm) - resid(fit_effect))),
    treatment_F=treatment_F,
    effect_F=effect_F
)
```

:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
The effect-coded intercept is about $7.28$, while the overall mean of `Murder` across all states is about $7.79$.
Use `table(dat$Region)` and the regional means to explain why these numbers differ.
Then change the order of the factor levels and verify that the fitted values remain unchanged.
:::

#### **Regression with Factors** {-}

We can extend the group-comparison model by adding continuous predictors alongside the factor.
For example, we might ask whether regional differences in murder arrest rates persist after holding `Assault` fixed.

```{r}
# Add a continuous predictor to the regional model.
reg_combined <- lm(Murder ~ Region + Assault, data=dat)
summary(reg_combined)
```

Holding `Assault` fixed changes the conditional association being summarized.
It does not by itself remove confounding, reverse causation, measurement error, or other reasons that a regression coefficient may not be causal.

#### **Fixed Effects** {-}

The dummy-variable model treats each group as having its own intercept.
This framing is useful when group identity captures features we cannot otherwise measure.

:::{.callout-important icon=false title="Key Definition" collapse="false"}
*Fixed effects* are group-specific intercepts that absorb every time-invariant feature of the group, entering the regression as
$$
\hat{Y}_{i} = \sum_{k=0}^{K} b_k \hat{X}_{ik} + \sum_{g}\hat{D}_{g}b_{g}.
$$
Including the dummy variables directly in OLS or using a dedicated routine such as `fixest::feols()` gives the same coefficients on the remaining regressors and the same residuals.
:::

Fixed effects are useful when a factor predicts the outcome but the coefficients on its individual levels are not the main object of interest.
The coefficient on another predictor then describes its partial relationship with the outcome after removing level differences across the factor.
The dummy-variable approach and a dedicated fixed-effects estimator are algebraically equal, giving the same coefficient and residuals.

```{r}
# Estimate the Assault coefficient with explicit region indicators.
reg_dv <- lm(Murder ~ -1 + Assault + Region, data=dat)
coef(reg_dv)['Assault']

# Estimate the same coefficient by absorbing the region effects.
reg_fe <- fixest::feols(Murder ~ Assault | Region, data=dat)
coef(reg_fe)
```

With fixed effects, we can also compute averages for each group: $\hat{M}_{Yg}=\sum_{i}^{n_{g}}\hat{Y}_{ig}/n_{g}$, where group $g$ has $n_g$ observations denoted $\hat{Y}_{ig}$.
We can construct a *between estimator* from $\hat{M}_{Yg}=b_0+\hat{M}_{Xg}b_1$.
Alternatively, we can subtract each group average to construct a *within estimator*: $(\hat{Y}_{ig}-\hat{M}_{Yg})=(\hat{X}_{ig}-\hat{M}_{Xg})b_1$.
Absorbing fixed group differences does not remove confounders that vary within a group.

#### **Comparing Models** {-}

The [ANOVA $F$-statistic](03_22_MultipleGroups.qmd#the-f-statistic) also compares nested regression models.
Under additional parametric assumptions, the statistic follows an $F$ distribution under the null.
We can build a sequence of nested models, each adding one layer of complexity, to see which predictors contribute.
The first asks whether group means differ, the second whether a continuous predictor adds explanatory power, and the third whether that predictor's slope varies across groups.

```{r}
# Order the nested models from simplest to most complex.
reg0 <- lm(Murder ~ 1, data=dat)
reg1 <- lm(Murder ~ Region, data=dat)
reg2 <- lm(Murder ~ Region + Assault, data=dat)
reg3 <- lm(Murder ~ Region * Assault, data=dat)

# Test whether Region adds information.
anova(reg0, reg1)

# Test whether Assault adds information beyond Region.
anova(reg1, reg2)

# Test whether the Assault slope differs by Region.
anova(reg2, reg3)
```

```{r}
# Reproduce the first nested-model F-test by hand.
rss0 <- sum(resid(reg0)^2)
rss1 <- sum(resid(reg1)^2)
df0 <- df.residual(reg0)
df1 <- df.residual(reg1)
F <- ((rss0 - rss1)/(df0 - df1))/(rss1/df1)
p <- 1 - pf(F, df0 - df1, df1)
cbind(F, p)
```

:::{.callout-note icon=false collapse="true" title="Must Know"}
The sequence of $F$-tests builds from simple to complex.
The first test asks whether Region adds information at all.
The second asks whether Assault adds information beyond Region.
The third asks whether the Assault-Murder relationship varies across regions.
A small $p$-value rejects the simpler model, but it does not explain why the relationship exists.
:::

## Permutation Tests

The theoretical $F$-test uses assumptions about independent Normal errors with equal variance.
A permutation test instead constructs a null distribution by rearranging the observed factor labels.

:::{.callout-important icon=false title="Key Definition" collapse="false"}
A *permutation test for a factor variable* repeatedly shuffles the factor labels without replacement, recomputes a global statistic such as $\hat{F}$, and compares the observed statistic with this permutation distribution.
Under the null, the labels must be exchangeable across observations.
:::

Permutation tests are useful when we want a reference distribution that preserves the observed outcomes and the exact number of observations in each group.
They differ from the label bootstrap in the previous chapter because sampling without replacement keeps all four regional group counts fixed.

#### **A Global Factor Test** {-}

The global $F$-statistic tests the factor as one object.
This is preferable to separately permuting its $G-1$ coded columns, which would destroy the rule that each observation belongs to exactly one group.
It is also coding-invariant, so treatment and effect coding produce the same observed and permuted statistics.

```{r}
#| fig-cap: "Permutation distribution for the regional F-statistic."
#| fig-alt: "Histogram of F-statistics after shuffling regional labels. The observed F-statistic is marked by a red vertical line to the right of the simulated null distribution."

set.seed(23)

F_stat <- function(Y, group) {
    fit <- lm(Y ~ group)
    F_value <- anova(fit)[['F value']][1]
    return(F_value)
}

F_obs <- F_stat(Y=dat[,'Murder'], group=dat[,'Region'])
B <- 9999
F_perm <- rep(NA, B)

for (b in seq(B)) {
    group_b <- sample(dat[,'Region'], replace=FALSE)
    F_perm[b] <- F_stat(Y=dat[,'Murder'], group=group_b)
}

# Add one simulated result to avoid reporting a zero Monte Carlo p-value.
p_perm <- (sum(F_perm >= F_obs) + 1)/(B + 1)
p_param <- anova(fit_lm)[['Pr(>F)']][1]
c(F_obs=F_obs, p_permutation=p_perm, p_parametric=p_param)

hist(F_perm, breaks=40, border=NA, freq=FALSE, main=NA,
    xlim=c(0, F_obs*1.05),
    xlab='F-statistic under permuted region labels')
abline(v=F_obs, col=rgb(1, 0, 0, .8), lwd=2)
title(paste0('permutation p-value = ',
    formatC(p_perm, format='f', digits=4)), font.main=1)
```

The observed regional $F$-statistic lies in the far-right tail of the permutation distribution.
The test therefore provides evidence that murder arrest rates and region are associated, under the assumption that the labels are exchangeable under the null.

#### **What Can Be Permuted** {-}

The simple shuffle above is appropriate for the factor-only model because the null removes the only relationship in the model.
In a regression with controls, naively shuffling `Region` would also break its relationship with those controls.
Adjusted permutation tests therefore require a design-specific restricted permutation or a method that permutes residuals under a reduced model.

Exchangeability is also a substantive assumption rather than a computational fact.
States may be spatially related, and region was not randomly assigned.
The small permutation $p$-value does not estimate the effect of moving a state to another region.

:::{.callout-note icon=false collapse="true" title="Must Know"}
A permutation test changes the pairing between `Murder` and `Region` while preserving the four group counts.
If `Region` had been randomly assigned in an experiment, the shuffle could reproduce the assignment mechanism.
For these observational data, it instead tests a no-association null under exchangeability.
:::

## Subsampling

Permutation and subsampling both rerun a statistic on altered data, but they answer different questions.
Permutation breaks a relationship to impose a null.
Subsampling preserves each row and asks how much an estimate changes when some rows are left out.

:::{.callout-important icon=false title="Key Definition" collapse="false"}
*Random subsampling* draws $B$ samples of size $m<n$ without replacement and recomputes the statistic in each.
When $m=n-d$, the procedure is a random delete-$d$ jackknife.
:::

Random subsampling is useful for checking whether a regression result is stable across overlapping subsets of the observed data.
The table compares it with the [bootstrap](01_05_Sampling.qmd#bootstrap-distribution) and [jackknife](01_05_Sampling.qmd#jackknife-distribution).

| Method | Sample Size per Iteration | Number of Iterations | Resample |
|---|---:|---:|---|
| Bootstrap | $n$ | $B$ | With replacement |
| Jackknife | $n-1$ | $n$ | Without replacement |
| Random delete-$d$ subsample | $m=n-d$ | $B$ | Without replacement |

#### **Random Delete-$d$ Samples** {-}

The raw spread of estimates across overlapping subsamples is not the standard error of the full-sample estimator.
For the random delete-$d$ jackknife used here, the variance estimate applies the correction [@PolitisEtAl1999]
$$
\hat{V}^{\text{sub}}
=
\frac{m}{dB}
\sum_{b=1}^{B}
\left(\hat{B}^{\text{sub}}_{b}-\bar{\hat{B}}^{\text{sub}}\right)^2,
$$
and $\hat{SE}^{\text{sub}}=\sqrt{\hat{V}^{\text{sub}}}$.
The example uses a Normal approximation around the full-sample coefficient rather than taking raw percentiles of the overlapping subsample estimates.

```{r}
#| fig-cap: "Slope estimates from random delete-10 subsamples."
#| fig-alt: "Histogram of regression slopes across delete-10 subsamples. The full-sample slope is marked in red and the corrected Normal confidence limits are marked with blue dashed lines."

xy <- USArrests[,c('Murder', 'UrbanPop')]
colnames(xy) <- c('y', 'x')
reg <- lm(y ~ x, data=xy)
b_full <- coef(reg)['x']

# Draw random delete-10 samples without replacement.
n <- nrow(xy)
d <- 10
m <- n - d
B_sub <- 999
rs_coefs <- rep(NA, B_sub)

for (b in seq(B_sub)) {
    b_id <- sample(n, m, replace=FALSE)
    xy_b <- xy[b_id,]
    reg_b <- lm(y ~ x, data=xy_b)
    rs_coefs[b] <- coef(reg_b)['x']
}

# Correct the spread for overlapping delete-d samples.
rs_mean <- mean(rs_coefs)
rs_var <- (m/d)*mean((rs_coefs - rs_mean)^2)
rs_se <- sqrt(rs_var)
rs_ci <- b_full + qnorm(c(.025, .975))*rs_se
c(b_full=b_full, subsample_SE=rs_se, lower=rs_ci[1], upper=rs_ci[2])

hist(rs_coefs, breaks=25, border=NA, freq=FALSE, main=NA,
    xlim=range(c(rs_coefs, rs_ci)),
    xlab='Subsample slope on UrbanPop')
abline(v=b_full, lwd=2, col=rgb(1, 0, 0, .8))
abline(v=rs_ci, lty=2, col=rgb(0, 0, 1, .8))
title(paste0('delete-', d, ' SE = ', round(rs_se, 3)), font.main=1)
```

:::{.callout-note icon=false collapse="true" title="Must Know"}
The example deletes $d=10$ observations and retains $m=40$ per iteration.
Increasing $d$ makes individual subsets less similar to the full dataset, while increasing $B$ only reduces Monte Carlo noise from which subsets happened to be drawn.
Neither change creates new information beyond the original 50 states.
:::

#### **Permutation Versus Subsampling** {-}

A permutation iteration keeps all $n$ observations but changes which outcome is paired with each factor label.
A subsampling iteration keeps the original pairings but uses only $m<n$ observations.
The permutation distribution describes a test statistic under a null relationship.
The subsampling distribution describes sensitivity to omitting observations and requires a scaling correction before it estimates full-sample uncertainty.

## What the Results Establish

Treatment coding and effect coding are two coordinate systems for the same fitted group means.
The word *effect* does not change the statistical model or supply a causal interpretation.

The permutation test provides evidence against a no-association null under exchangeability.
It identifies a surprising arrangement of outcomes and labels, not what would happen under an intervention.

The subsampling result shows how a coefficient responds to deleting observations.
A stable coefficient can still be confounded, and an unstable coefficient is not necessarily biased.

These limits extend the distinction introduced in [Association and Prediction Are Not Causation](02_17_NotCausation.qmd).
The later chapters on [Observational Data](03_26_ObservationalData.qmd) and [Experimental Data](03_27_ExperimentalData.qmd) develop the designs and assumptions required for causal claims.

## Exercises

1. Comment the script you wrote for this chapter, then restart R and check that it runs from a clean session, then check the script with AI as explained in [Working with AI](04_A_WorkingWithAI.qmd).
   Write three sentences from memory on the main statistical idea of this chapter, and ask the assistant what is wrong, vague, or missing.
   Finish with your own questions about whatever you found hardest.

2. Suppose three groups have means $4$, $7$, and $5$.
   Compute the treatment-coded coefficients using the first group as the reference.
   Then compute the effect-coded intercept and all three group effects, checking that the effects sum to zero.

3. Using `USArrests` and `state.region`, construct a permutation test of whether mean `Assault` arrest rates differ across regions.
   Use $B=9999$ permutations, report the global $F$-statistic and permutation $p$-value, and compare the result with `anova(lm(Assault ~ Region, data=dat))`.

4. Extend the running example by fitting `Murder ~ Region + Assault + UrbanPop` and `Murder ~ Region * Assault + UrbanPop`.
   Use `anova()` to test whether the Region-Assault interaction terms are jointly significant after controlling for `UrbanPop`.
   Explain why this test does not establish that assault arrests cause murder arrests.

5. Repeat the subsampling example with $d=5$, $d=10$, and $d=20$.
   Apply the delete-$d$ correction in each case and compare the standard errors.
   Explain why increasing $B$ makes the simulation more stable without increasing the information in the original sample.

#### **Further Reading** {-}

* <https://www.econometrics-with-r.org/7-htaciimr.html> -- hypothesis tests for comparing groups in a regression framework, including $F$-tests and joint restrictions.
* <https://online.stat.psu.edu/stat462/node/137/> -- ANOVA as a special case of regression, with examples of multiple-group comparisons.
* @PolitisEtAl1999 -- a systematic treatment of subsampling methods and their statistical foundations.
* @BinEtAl2015 -- a comparison of subsampling and bootstrapping for the stability of multivariable regression models.

#### **Recall** {-}

This chapter rewrote ANOVA as a regression with a factor variable and showed that treatment and effect coding change coefficient meanings without changing the fitted model.
The regional example gave an effect-coded intercept of about $7.28$, the unweighted mean of the four regional means, while its global $F$-statistic was identical under both codings.
The permutation test shuffled the factor labels to test a no-association null, whereas random delete-$d$ subsampling preserved each row and measured sensitivity to omitted observations.
None of these operations made the regional regression causal.
The next chapter audits fitted regressions for outliers, leverage, collinearity, and misspecification.
