# Misc. Multivariate Topics
***



## Prediction

#### **Describe vs. Explain vs. Predict**.{-}

Understanding whether we aim to *describe*, *explain*, or *predict* is central to empirical analysis. The three distinct purposes are to accurately describe empirical patterns, explain why the empirical patterns exist, predict what empirical patterns will exist.


| Objective | Core Question | Goal | Methods |
|---|---|---|---|
| Describe | What is happening? | Characterize patterns & facts | Summary stats, visualizations, correlations |
| Explain | Why is it happening? | Establish causal relationships & mechanisms | Theoretical models; experimentation; counterfactual reasoning |
| Predict | What will happen? | Anticipate outcomes; support decisions that affect the future | Machine learning, treatment effect forecasting, policy simulations



Here is an example for minimum wages.

* Describe: Studies document that following minimum-wage increases, overall low-wage employment may look roughly stable in the short run, but disaggregated data often show larger employment declines over longer horizons, especially among youths and racial minorities.
* Explain: Studies investigate economic mechanisms such as (i) whether lower productivity populations are more vulnerable and employers adjust along hiring margins (fewer openings, higher required skills), and (ii) whether effects are larger in sectors like retail/food service where minimum wages bite hardest.
* Predict: A policy simulation that raises the wage floor incorporates subgroup-specific elasticities to forecast different employment losses  (e.g., unemployment for teenage or black workers and income gains for stayers)


The distinctions matter, as they help you recognize that predictive accuracy or good data visualization does not equal causal insight. You can then better align statistical tools with questions. Try remembering this: Describe reality, Explain causes, Predict futures

:::{.callout-note icon=false collapse="true" title="Must Know"}
Consider the relationship between class size and student test scores. Write one sentence for each: (a) a descriptive finding, (b) an explanatory hypothesis, and (c) a predictive statement. Which of the three requires the strongest assumptions?
:::


#### **Prediction Intervals**. {-}

Here, we consider a range for $Y_{i}$ rather than for $m(X_{i})$. These intervals also take into account the residuals, the variability of individuals rather than the variability of their mean.
```{r}
# Bivariate Data from USArrests
xy <- USArrests[, c('Murder', 'UrbanPop')]
colnames(xy) <- c('y', 'x')
xy0 <- xy[order(xy[, 'x']), ]
```

For a nice overview of different types of intervals, see <https://www.jstor.org/stable/2685212>. For an in-depth view, see "Statistical Intervals: A Guide for Practitioners and Researchers" or "Statistical Tolerance Regions: Theory, Applications, and Computation". See <https://robjhyndman.com/hyndsight/intervals/> for constructing intervals for future observations in a time-series context. See Davison and Hinkley, chapters 5 and 6 (also Efron and Tibshirani, or Wehrens et al.)


```{r}

## Data
library('wooldridge')
dat <- wage1[, c('wage', 'educ')]
dat <- dat[order(wage1[, 'educ']), ]


# Design Points
X0 <- unique(dat[, 'educ'])
# Regression
reg_lo <- loess(wage ~ educ, data=dat, span=.8)
preds_lo <- predict(reg_lo, newdata=data.frame(educ=X0))

# Bootstrap Residuals
n <- nrow(dat)
res_lo <- lapply(1:399, function(i){
    dat_b <- dat[sample(n, replace=T), ]
    reg_b <- loess(wage ~ educ, data=dat_b, span=.8)
    resids_b <- resid(reg_b)
    cbind(e=resids_b, educ=dat_b[, 'educ'])
})
res_lo <- do.call(rbind, res_lo)

## Smooth residuals
res_fun <- function(x0, h){
	# Include only residuals within h distance to x0
    ki0 <- abs(res_lo[, 'educ']-x0)/h
    ki0 <- dunif(ki0)/h
    ei0 <- res_lo[ki0 != 0, 'e'] #subset estimates   
    ## Local quantiles
    res_i <- quantile(ei0, probs=c(.025, .975), na.rm=T)
    res_i
}
res_ci <- sapply(X0, res_fun, h=5)

## Prediction Interval
pi_lo1 <- preds_lo + res_ci[1, ]
pi_lo2 <- preds_lo + res_ci[2, ]

#Plot
col <- hcl.colors(3, alpha=.5)[2]
plot(wage ~ educ, pch=16, col=grey(0, .25),
    data=dat, ylim=c(0, 20), main=NA)
lines(X0, preds_lo,
    col=col, type='o', pch=2)
polygon( c(X0, rev(X0)),
	c(pi_lo1, rev(pi_lo2)),
    col=col, density=12, angle=90)
```


:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
Consider also this example below with a linear model. Expand on it by using the Normal Approximation for the residuals (hint: use `qnorm`).

```{r}
reg <- lm(wage ~ educ, data=dat)

# Bootstrap Prediction Interval
boot_resids <- lapply(1:399, function(b){
    b_id <- sample( nrow(dat), replace=T)
    dat_b <- dat[b_id, ]
    reg_b <- lm(wage ~ educ, data=dat_b)
    e_b <- resid(reg_b)
    x_b <- reg_b[['model']][['educ']]
    res_b <- cbind(e_b, x_b)
})
boot_resids <- as.data.frame(do.call(rbind, boot_resids))

# Errors do not vary with x
ehat <- quantile(boot_resids[, 'e_b'], probs=c(.025, .975))

# PI
phat <- predict(reg)
boot_pi <- cbind(phat + ehat[1], phat + ehat[2])

# Plot Bootstrap PI
plot(wage ~ educ, data=dat, pch=16, col=grey(0, .25), main=NA,
    ylim=c(-5, 20))
title('Prediction Intervals', font.main=1)
x <- dat[, 'educ']
lines(x, phat, type='o', pch=2)
polygon( c(x, rev(x)),
    c(boot_pi[, 1], rev(boot_pi[, 2])),
    col=grey(0, .2), border=NA)

# Normal PI (Make by hand)
#pi <- predict(reg, interval='prediction')
#lines( x, pi[, 'lwr'], lty=2)
#lines( x, pi[, 'upr'], lty=2)
```
:::

#### **Cross Validation**.{-}

Perhaps the most common approach to selecting a bandwidth is to minimize *prediction* error. *Leave-one-out cross-validation* minimizes the average "leave-one-out" mean square prediction errors:
\begin{eqnarray}
\min_{h} \quad \frac{1}{n} \sum_{i=1}^{n} \left[ \hat{Y}_{i} - \hat{y_{[i]}}(X,h) \right]^2,
\end{eqnarray}
where $\hat{y}_{[i]}(X,h)$ is the model predicted value at $X_{i}$ based on a dataset that excludes $X_{i}$, and $h$ is the bandwidth (e.g., bin size in a regressogram).

Minimizing out-sample prediction error is perhaps the simplest computational approach to choose bandwidths, and it also addresses an issue that plagues observational studies in the social sciences: your model explains everything and predicts nothing. Note, however, minimizing prediction error is not objectively "best". 

In practice, compare models with cross-validation (below). The preferred model is the one with best out-of-sample performance, not the one with the most flexible in-sample fit. The same idea selects bin counts and bandwidths in [Local Relationships](https://jadamso.github.io/Rbooks/03_22_Regressograms.html#model-selection).

```{r}
library(wooldridge)
# Crossvalidated bandwidth for regression
xy_mat <- data.frame(x=wage1[, 'educ'], y=wage1[, 'wage'])

## Grid Search
BWS <- c(0.5, 0.6, 0.7, 0.8)
BWS_CV <- sapply(BWS, function(h){
    E_bw <- sapply(1:nrow(xy_mat), function(i){
        llls <- loess(y ~ x, data=xy_mat[-i, ], span=h,
            degree=1, surface='direct')
        pred_i <- predict(llls, newdata=xy_mat[i, ])
        e_i <-  (pred_i- xy_mat[i, 'y'])
        return(e_i)
    })
    return( mean(E_bw^2) )
})

## Plot MSE
par(mfrow=c(1, 2))
plot(BWS, BWS_CV, ylab='CV error', pch=3,
    xlab='bandwidth (h)')

## Plot Optimal Predictions
h_star <- BWS[which.min(BWS_CV)]
llls <- loess(y ~ x, data=xy_mat, span=h_star,
    degree=1, surface='direct')
plot(xy_mat, pch=16, col=grey(0, .25),
    xlab='X', ylab='Predictions', main=NA)

pred_id <- order(xy_mat[, 'x'])
lines(xy_mat[pred_id, 'x'], predict(llls)[pred_id],
    col=rgb(1, 0, 0, .8), lwd=2)
```

:::{.callout-note icon=false collapse="true" title="Must Know"}
Leave-one-out cross-validation is just an average of out-of-sample errors.
Suppose $n=4$ and, for one bandwidth $h$, leaving out each point in turn gives the predictions $\hat{y}_{[i]}$ below.

| $i$ | $\hat{Y}_i$ | $\hat{y}_{[i]}$ | $\hat{Y}_i - \hat{y}_{[i]}$ |
|:---:|:---:|:---:|:---:|
| 1 | 10 | 9 | 1 |
| 2 | 12 | 14 | -2 |
| 3 | 15 | 14 | 1 |
| 4 | 11 | 11 | 0 |

The cross-validation score is the mean squared error: $(1^2 + (-2)^2 + 1^2 + 0^2)/4 = 6/4 = 1.5$.
Repeating this for several bandwidths, the one with the smallest score is preferred.
:::

## Filtering

#### **Filtering**.{-}
In some cases, we may want to smooth time series data instead of predict into the future. We can distinguish two types of smoothing, incorporating future observations or not. When only weighting two other observations, the differences can be expressed as trying to estimate the average with different available data:

* Filtering: $\mathbb{E}[Y_{t} | X_{t-1}, X_{t-2}]$
* Smoothing: $\mathbb{E}[Y_{t} | X_{t-1}, X_{t+1}]$

One example of *filtering* is *Exponential Filtering* (sometimes confusingly referred to as "Exponential Smoothing") which weights only previous observations using an exponential kernel.
```{r}
# Time series data
set.seed(1)
## Underlying Trend
x0 <- cumsum(rnorm(500, 0, 1))
## Observed Datapoints
x <- x0 + runif(length(x0), -10, 10)
dat <- data.frame(t=seq_along(x), x0=x0, x=x)

## Asymmetric kernel (filtering: uses current and past only)
bw1 <- c(2/3, 1/3)
s1 <- filter(x, bw1/sum(bw1), sides=1)

## Symmetric kernel (smoothing: also uses future)
bw2 <- c(1/6, 2/3, 1/6)
s2 <- filter(x, bw2/sum(bw2), sides=2)

## Plot the series with both smoothers
plot(x ~ t, data=dat, pch=16, col=grey(0, .25),
    main=NA, xlab='Time', ylab='x')
lines(dat$t, s1, col=rgb(1, 0, 0, .8), lwd=2)
lines(dat$t, s2, col=rgb(0, 0, 1, .8), lwd=2)
legend('topleft', lwd=2, bty='n',
    col=c(rgb(1, 0, 0, .8), rgb(0, 0, 1, .8)),
    legend=c('Asymmetric (filtering)', 'Symmetric (smoothing)'))
```

There are several cross-validation procedures for filtering time series data (Opsomer et al., 2001).
One is called *time series cross-validation* (TSCV), which is useful for temporally dependent data (Hart, 1991; Härdle and Philippe, 1992; Bergmeir et al., 2018).
Unlike ordinary cross-validation, TSCV only ever predicts a point from its past, never its future, which respects the time ordering of the data.
We use it below to choose the window of a one-sided moving average.
```{r}
# Time series cross-validation for a one-sided moving average
# For each window h, predict each point from the previous h points
x <- dat$x
ma_windows <- 2:20
tscv_mse <- sapply(ma_windows, function(h){
    ## Weight the previous h points equally, 0 weight on the current point
    ma_coefs <- c(0, rep(1/h, h))
    x_pred <- filter(x, ma_coefs, sides=1)
    mean((x - x_pred)^2, na.rm=T)
})

## CV-optimal window
h_star <- ma_windows[which.min(tscv_mse)]

par(mfrow=c(1, 2))
## Plot CV error against window size
plot(ma_windows, tscv_mse, type='o', pch=16, main=NA,
    xlab='Moving-average window (h)', ylab='TSCV mean squared error')
abline(v=h_star, col=rgb(1, 0, 0, .8), lwd=2)

## Plot the series with the CV-optimal filter
x_smooth <- filter(x, c(0, rep(1/h_star, h_star)), sides=1)
plot(x ~ t, data=dat, pch=16, col=grey(0, .25), main=NA,
    xlab='Time', ylab='x')
lines(dat$t, x_smooth, col=rgb(1, 0, 0, .8), lwd=2)
title(paste0('One-sided MA, h = ', h_star), font.main=1)

## See also https://robjhyndman.com/hyndsight/tscvexample/
```

#### **Bayes' Theorem**. {-}

Bayes' theorem maps predictive statements into inferential statements. In bivariate form,
\begin{eqnarray}
Prob(X_i=x \mid Y_i=y)
&=&
\frac{Prob(Y_i=y \mid X_i=x)Prob(X_i=x)}{Prob(Y_i=y)}.
\end{eqnarray}

Interpretation:

* $Prob(X_i=x)$ is the prior probability for $X_i=x$.
* $Prob(Y_i=y|X_i=x)$ is the likelihood of seeing $Y_i=y$ if $X_i=x$ is true.
* $Prob(X_i=x|Y_i=y)$ is the posterior, your updated probability after seeing $Y_i=y$.

A useful way to remember this is
\begin{eqnarray}
\text{Posterior} \propto \text{Likelihood} \times \text{Prior}.
\end{eqnarray}
For two states, posterior odds are prior odds times a likelihood ratio.

For a concrete example, suppose a screening test has sensitivity 0.90 and false positive rate 0.08, while prevalence is 0.12:
\begin{eqnarray}
Prob(Y_i=1|X_i=1)=0.90,\quad
Prob(Y_i=1|X_i=0)=0.08,\quad
Prob(X_i=1)=0.12,
\end{eqnarray}
where $X_i=1$ means "condition present" and $Y_i=1$ means "test positive".

Then
\begin{eqnarray}
Prob(X_i=1|Y_i=1)
&=&
\frac{0.90\times0.12}{0.90\times0.12 + 0.08\times0.88}
\approx 0.605.
\end{eqnarray}
Even with a good test, posterior probability depends strongly on prevalence.

```{r}
# States: X in {0, 1}, signal Y in {0, 1}
# Prior Prob(X_i=1)
p_x1 <- 0.12

# Test characteristics
p_y1_x1 <- 0.90
p_y1_x0 <- 0.08

# Law of total probability for Prob(Y_i=1)
p_y1 <- p_y1_x1 * p_x1 + p_y1_x0 * (1 - p_x1)

# Bayes posterior Prob(X_i=1 | Y_i=1)
p_x1_y1 <- (p_y1_x1 * p_x1) / p_y1
p_x1_y1

# Also compute Prob(X_i=1 | Y_i=0)
p_y0_x1 <- 1 - p_y1_x1
p_y0_x0 <- 1 - p_y1_x0
p_y0 <- p_y0_x1 * p_x1 + p_y0_x0 * (1 - p_x1)
p_x1_y0 <- (p_y0_x1 * p_x1) / p_y0
p_x1_y0
```

:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
Notice that a positive test result raises the probability of having the condition from $12\%$ (prior) to about $60\%$ (posterior). A negative test result lowers it from $12\%$ to about $1.4\%$. If prevalence were higher, say $50\%$, the posterior after a positive test would be $\frac{0.90 \times 0.50}{0.90 \times 0.50 + 0.08 \times 0.50} = \frac{0.45}{0.49} \approx 0.92$. The same test is far more informative when prevalence is higher.
:::

## Subsampling
Random *subsampling* is one of many hybrid approaches that tries to combine the best of the core methods: Bootstrap and Jacknife.

| | Sample Size per Iteration | Number of Iterations | Resample |
| -------- | ------- | ------- | ------- |
Bootstrap | $n$     | $B$  | With Replacement |
Jackknife | $n-1$   | $n$  | Without Replacement |
Random Subsample | $m < n$ | $B$  | Without Replacement |

```{r}
xy <- USArrests[, c('Murder', 'UrbanPop')]
colnames(xy) <- c('y', 'x')
reg <-  lm(y ~ x, data=xy)
coef(reg)

# Random Subsamples
rs_regs <- lapply(1:399, function(b){
    b_id <- sample( nrow(xy), nrow(xy)-10, replace=F)
    xy_b <- xy[b_id, ]
    reg_b <- lm(y ~ x, data=xy_b)
})
rs_coefs <- sapply(rs_regs, coef)['x', ]
rs_se <- sd(rs_coefs)

hist(rs_coefs, breaks=25,
    main=NA, border=NA, freq=F,
    xlab=expression(hat(b)[b]))
title(paste0('SE est. = ', round(rs_se, 4)), font.main=1)
abline(v=coef(reg)['x'], lwd=2, col=rgb(1, 0, 0, .8))
rs_ci_percentile <- quantile(rs_coefs, probs=c(.025, .975))
abline(v=rs_ci_percentile, lty=2, col=rgb(0, 0, 1, .8))
```

:::{.callout-note icon=false collapse="true" title="Must Know"}
The random subsample uses $n-10$ observations per iteration. How would you expect the standard error estimate to change if you used $n-20$ instead? What is the trade-off between subsample size $m$ and the number of iterations $B$?
:::


## Inference.

#### **Matrix Calculations**. {-}

First, note that you can use matrix algebra in R
```{r}
x_mat1 <- matrix( seq(2, 7), 2, 3)
x_mat1

x_mat2 <- matrix( seq(4, -1), 2, 3)
x_mat2

x_mat1*x_mat2 #NOT classical matrix multiplication
x_mat1^x_mat2

tcrossprod(x_mat1, x_mat2) #x_mat1 %*% t(x_mat2)

crossprod(x_mat1, x_mat2)
```

:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
Matrix multiplication `%*%` follows the rule that element $(i,j)$ of the product equals the dot product of row $i$ of the first matrix and column $j$ of the second. For `crossprod(A, B)`, R computes $A'B$ efficiently without explicitly transposing. This is useful in OLS where we need $(\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'\mathbf{Y}$ repeatedly.
:::

Second, note that you can apply functions to matrices
```{r}
sum_squared <- function(x1, x2) {
    y <- (x1 + x2)^2
    return(y)
}
sum_squared(x_mat1, x_mat2)


solve(x_mat1[1:2, 1:2])
```


Third, note that we can conduct OLS regressions efficiently using matrix algebra. Grouping the coefficients as a vector $B=(b_0 ~~ b_1 ~~... ~~ b_{K})$, we want to find coefficients that minimize the sum of squared errors. The linear model has a fairly simple solution for $\hat{B}^{*}$ if you know linear algebra. Denoting $\hat{\mathbf{X}}_{i} = [1~~  \hat{X}_{i1} ~~...~~ \hat{X}_{iK}]$ as a row vector, we can write the model as $\hat{Y}_{i} = \hat{\mathbf{X}}_{i}B + e_{i}$. We can then write the model in matrix form
\begin{eqnarray}
\hat{Y} &=& \hat{\textbf{X}}B + E \\
\hat{Y} &=& \begin{pmatrix} 
\hat{Y}_{1} \\ \vdots \\ \hat{Y}_{N}
\end{pmatrix} \quad
\hat{\textbf{X}} = \begin{pmatrix} 
1 & \hat{X}_{11} & ... & \hat{X}_{1K} \\
& \vdots & & \\
1 & \hat{X}_{n1} & ... & \hat{X}_{nK} 
\end{pmatrix}
\end{eqnarray}
Minimizing the squared errors: $\min_{B} e_{i}^2 = \min_{B} (E' E)$, yields coefficient estimates and predictions 
\begin{eqnarray}
\hat{B^{*}} &=& (\hat{\textbf{X}}'\hat{\textbf{X}})^{-1}\hat{\textbf{X}}'\hat{Y}\\
\hat{y} &=& \hat{\textbf{X}} \hat{B^{*}} \\
\hat{E} &=& \hat{Y} - \hat{y} \\
\end{eqnarray}
```{r}
# Manually Compute
Y <- USArrests[, 'Murder']
X <- USArrests[, c('Assault', 'UrbanPop')]
X <- as.matrix(cbind(1, X))

XtXi <- solve(t(X)%*%X)
Bhat <- XtXi %*% (t(X)%*%Y)
c(Bhat)
```

#### **Bias**. {-}

With matrix notation, we can established that OLS is an unbiased estimator for additively separable and linear relationships. Assume
\begin{eqnarray}
Y_{i} = X_{i} \beta + \epsilon_{i} &\quad& \mathbb{E}[\epsilon_{i} | X_{i} ]=0,
\end{eqnarray}
then notice that
\begin{eqnarray}
B^{*} = (\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'Y = (\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'(\mathbf{X}\beta + \epsilon) = \beta + (X'X)^{-1}X'\epsilon\\
\mathbb{E}\left[ B^{*} \right] = \mathbb{E}\left[ (\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'Y \right] = \beta + (\mathbf{X}'\mathbf{X})^{-1}\mathbb{E}\left[ \mathbf{X}'\epsilon \right] = \beta
\end{eqnarray}

If we do not know that the data generating process is additively separable and linear, then we do not know how to interpret $\hat{B}^{*}$ (outside of a very mathematical expression, which I do not cover here). If we run a local linear regression analysis, on the other hand, we get the estimator
\begin{eqnarray}
\beta(\mathbf{x}) &=& [\mathbf{X}'\mathbf{K}(\mathbf{x})\mathbf{X}]^{-1} \mathbf{X}'\mathbf{K}(\mathbf{x})Y \\
\mathbf{K}(\mathbf{x}) &=& \begin{pmatrix}
K\left(\frac{\mathbf{X}_{1}-\mathbf{x}}{h_{1}}\right) & ... & 0\\
\vdots & & \\
0 & ... & K\left(\frac{\mathbf{X}_{P}-\mathbf{x}}{h_{P}}\right)
\end{pmatrix},
\end{eqnarray}
which is unbiased far more often.


## Multiple Testing

When you test one hypothesis at the $5\%$ level, there is a $5\%$ chance of a false positive.
When you test many hypotheses, those chances accumulate.
This is the statistical engine behind much of [Data Scientism](https://jadamso.github.io/Rbooks/03_25_DataScientism.html): search long enough and something will look significant.

#### **Family-Wise Error Rate**. {-}

The *family-wise error rate* (FWER) is the probability of at least one false positive across a family of $K$ tests.
If the tests are independent and each has false-positive rate $\alpha$, then
\begin{eqnarray}
\text{FWER} &=& 1-(1-\alpha)^{K}.
\end{eqnarray}
With $\alpha=0.05$, one test has FWER $0.05$, but $K=10$ tests give $1-0.95^{10}\approx 0.40$ and $K=100$ tests give $1-0.95^{100}\approx 0.99$.
Almost any dataset will yield a "significant" result if you test it enough ways.

```{r}
# Family-wise error rate grows with the number of tests
alpha <- 0.05
K <- c(1, 5, 10, 20, 50, 100)
fwer <- 1 - (1-alpha)^K
cbind(K, fwer=round(fwer, 3))
```

#### **Bonferroni and Holm Corrections**. {-}

To control the FWER, we make each individual test stricter.
The *Bonferroni correction* tests each hypothesis at level $\alpha/K$ instead of $\alpha$.
Equivalently, it multiplies each $p$-value by $K$ and keeps the $\alpha$ threshold.
This guarantees FWER $\leq \alpha$, but it is conservative: with many tests, it can miss real effects.
The *Holm correction* is a less conservative stepwise version that still controls the FWER, and is generally preferred.

```{r}
# Five p-values from five separate tests
pvals <- c(0.001, 0.013, 0.021, 0.04, 0.21)

# Adjust for multiple testing
cbind(
    raw        = pvals,
    bonferroni = p.adjust(pvals, method='bonferroni'),
    holm       = p.adjust(pvals, method='holm'))
```

:::{.callout-note icon=false collapse="true" title="Must Know"}
A researcher runs $K=5$ tests and reports a $p$-value of $0.013$.
On its own, $0.013 < 0.05$ looks significant.
The Bonferroni-adjusted value is $0.013 \times 5 = 0.065 > 0.05$, so it is *not* significant once we account for the five tests.
The threshold view is equivalent: compare the raw $p$-value to $\alpha/K = 0.05/5 = 0.01$.
:::

#### **Post-Hoc Pairwise Tests**. {-}

A global test, such as the ANOVA $F$-test in [Comparing Multiple Groups](https://jadamso.github.io/Rbooks/03_20_MultipleGroups.html), tells you that *some* group differs, but not *which* one.
The natural follow-up is to compare every pair of groups.
With $G$ groups there are $G(G-1)/2$ pairs, so these *post-hoc* comparisons are themselves a multiple-testing problem and need a correction.

```{r}
# Global test: do murder rates differ by region?
dat <- USArrests
dat$Region <- state.region
summary(aov(Murder ~ Region, data=dat))
```

```{r}
# Post-hoc: which regions differ? (pairwise t-tests, Holm-adjusted)
pairwise.t.test(dat$Murder, dat$Region, p.adjust.method='holm')
```

```{r}
# Tukey's Honest Significant Differences:
# pairwise confidence intervals that already account for all comparisons
TukeyHSD(aov(Murder ~ Region, data=dat))
```

:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
Rerun `pairwise.t.test` on the region data with `p.adjust.method='none'` and again with `'bonferroni'`.
How many pairs are significant at the $5\%$ level under each?
The uncorrected version flags more differences; explain why, and state which version you should report.
:::

## Exercises

1. Explain the distinction between description, explanation, and prediction using a concrete example other than minimum wages. For each objective, state what statistical tools you would use and what kind of evidence would be needed.

2. Using the `USArrests` dataset, regress `Murder` on `Assault` and `UrbanPop`. Compute $\hat{B}^{*}$ both with `lm()` and manually via $(\hat{\mathbf{X}}'\hat{\mathbf{X}})^{-1}\hat{\mathbf{X}}'\hat{Y}$ using matrix operations in R. Verify that the two approaches give the same coefficients.

3. Using the `wage1` dataset from the `wooldridge` package, fit a `loess` regression of `wage` on `educ` for three different span values (e.g., 0.4, 0.6, 0.8). Use leave-one-out cross-validation to compute the mean squared prediction error for each span. Which span minimizes prediction error, and does a more flexible fit always predict better?

4. A researcher runs $K=20$ independent tests, each at the $5\%$ level. Compute the family-wise error rate $1-(1-0.05)^{20}$, and find the per-test Bonferroni threshold $\alpha/K$ that holds the family-wise rate at $5\%$. Then create a vector of 20 $p$-values, apply `p.adjust()` with the `'bonferroni'` and `'holm'` methods, and report how many results remain significant under each.

#### **Further Reading**. {-}

* <https://otexts.com/fpp3/> -- Hyndman and Athanasopoulos, *Forecasting: Principles and Practice* (3rd ed.), covering time-series smoothing, cross-validation, and prediction in R.
* <https://cran.r-project.org/web/packages/np/vignettes/np.pdf> -- the `np` package vignette for nonparametric kernel estimation and bandwidth selection.
* Shmueli, G. (2010). To explain or to predict? *Statistical Science*, 25(3), 289--310. <https://doi.org/10.1214/10-STS330> -- a clear discussion of the describe/explain/predict distinction in statistical modeling.
* <https://cran.r-project.org/web/packages/multcomp/vignettes/multcomp-examples.pdf> -- worked examples of multiple-comparison procedures and post-hoc tests in R.
