# Data Scientism
***

In practice, it is hard to find a good natural experiment. For example, suppose we asked "what is the effect of wages on police demanded?" and examined a policy which lowered the educational requirements from 4 years to 2 to become an officer. This increases the labour supply, but it also affects the demand curve through "general equilibrium": as some of the new officers were potentially criminals and, with fewer criminals, the demand for police shifts down.

In practice, it is also easy to find a bad instrument. Paradoxically, natural experiments are something you are supposed to find but never search for. As you search for good instruments, for example, sometimes random noise will appear like a good instrument (*spurious instruments*). In this age of big data, we are getting increasingly more data and, perhaps surprisingly, this makes it easier to make false discoveries.

We will consider three classical ways for false discoveries to arise. After that, there are examples with the latest and greatest empirical recipes---we don't have so many theoretical results yet but I think you can understand the issue with the numerical example. Although it is difficult to express numerically, you must also know that if you search for a good natural experiment for too long, you can also be led astray from important questions. There are good reasons to be excited about empirical social science, but we would be wise to recall some earlier wisdom from economists on the matter.

> The most reckless and treacherous of all theorists is he who professes to let facts and figures speak for themselves, who keeps in the background the part he has played, perhaps unconsciously, in selecting and grouping them
>
> ---  Alfred Marshall, 1885 


> The blind transfer of the striving for quantitative measurements to a field where the specific conditions are not present which give it its basic importance in the natural sciences is the result of an entirely unfounded prejudice. It is probably responsible for the worst aberrations and absurdities produced by scientism in the social sciences. It not only leads frequently to the selection for study of the most irrelevant aspects of the phenomena because they happen to be measurable, but also to "measurements" and assignments of numerical values which are absolutely meaningless. What a distinguished philosopher recently wrote about psychology is at least equally true of the social sciences, namely that it is only too easy "to rush off to measure something without considering what it is we are measuring, or what measurement means. In this respect some recent measurements are of the same logical type as Plato's determination that a just ruler is 729 times as happy as an unjust one."
>
> --- F.A. Hayek, 1943

> if you torture the data long enough, it will confess
>
> --- R. Coase (Source Unknown)

<!---
''torture the data (to) confess'' (Coase,  Essays on economics and economists.  1995, p. 27)
--->

> the definition of a causal parameter is not always clearly stated, and formal statements of identifying conditions in terms of well-specified economic models are rarely presented. Moreover, the absence of explicit structural frameworks makes it difficult to cumulate knowledge across studies conducted within this framework. Many studies produced by this research program have a `stand alone' feature and neither inform nor are influenced by the general body of empirical knowledge in economics.
>
> --- J.J. Heckman, 2000


> without explicit prior consideration of the effect of the instrument choice on the parameter being estimated, such a procedure is effectively the opposite of standard statistical practice in which a parameter of interest is defined first, followed by an estimator that delivers that parameter. Instead, we have a procedure in which the choice of the instrument, which is guided by criteria designed for a situation in which there is no heterogeneity, is implicitly allowed to determine the parameter of interest. This goes beyond the old story of looking for an object where the light is strong enough to see; rather, we have at least some control over the light but choose to let it fall where it may and then proclaim that whatever it illuminates is what we were looking for all along.
>
> --- A. Deaton, 2010


## False Positives

#### **Data Errors**. {-}
A huge amount of data normally means a huge amount of data cleaning/merging/aggregating. This avoids many copy-paste errors, which are a recipe for [disaster](https://blog.hurree.co/8-of-the-biggest-excel-mistakes-of-all-time), but may also introduce other types of errors. Some spurious results are driven by honest errors in data cleaning. According to one [estimate](https://www.pnas.org/doi/10.1073/pnas.1212247109), this is responsible for around one fifth of all medical science retractions (there is even a whole [book](https://www.amazon.de/Much-Cost-Coding-Errors-Implementation/dp/1543772994) about this!). Although there are not similar meta-analysis in economics, there are some high-profile examples. This includes papers that are highly influential, like [Lott, Levitt](https://scienceblogs.com/deltoid/2005/12/02/lott-levitt-and-coding-errors) and [Reinhart and Rogoff](https://blogs.lse.ac.uk/impactofsocialsciences/2013/04/24/reinhart-rogoff-revisited-why-we-need-open-data-in-economics/) as well as others the top economics journals, like the [RESTUD](https://academic.oup.com/restud/article/90/2/1009/6982752) and [AER](https://www.aeaweb.org/articles?id=10.1257/aer.113.7.2053). There are some reasons to think such errors are more widespread across the social sciences; e.g., in [Census data](https://www2.census.gov/ces/tp/tp-2002-17.pdf) and [Aid data](https://www.sciencedirect.com/science/article/abs/pii/S0305750X11001951). So be careful!

Note: one reason to plot your data is to help spot such errors. Data errors are only one route to a spurious finding. The next one is searching the same dataset for any specification that "works".

#### **P-Hacking**. {-}
Another class of errors pertains to *p-hacking* (and its various synonyms: *data drudging*, *star mining*,...). While there are cases of fraudulent data manipulation (which can be considered as a dishonest data error), p-hacking need not even be intentional. You can simply be trying different variable transformations to uncover patterns in the data, for example, without accounting for how easy it is to find patterns when transforming completely random data. P-hacking is [pernicious](https://elephantinthelab.org/a-replication-crisis-in-the-making/) and [widespread](https://www.americanscientist.org/article/the-statistical-crisis-in-science).

:::{.callout-important title="Definition" collapse="false"}
*P-hacking* is the practice (often unintentional) of trying many models, transformations, or subsamples until one returns a small $p$-value, then reporting only that one. The reported $p$-value no longer reflects the true probability of seeing the result under the null, because the search itself was not accounted for.
:::

```{r}
# P-hacking OLS with different explanatory vars
set.seed(123)
n <- 50
X1 <- runif(n)

# Regression Machine:
# repeatedly finds covariate, runs regression
# stops when statistically significant at .1%
p <- 1
i <- 0
while(p >= .001){ 
    # Get Random Covariate
    X2 <-  runif(n)
    # Merge and `Analyze'
    dat_i <- data.frame(X1, X2)
    reg_i <- lm(X1 ~ X2, data=dat_i)
    # update results in global environment
    p <- summary(reg_i)$coefficients[2, 4]
    i <- i+1
}
#summary(reg_i)

plot(X1 ~ X2, data=dat_i,
    pch=16, col=grey(0, .5), main=NA)
title(paste0('Random Dataset ', i, ':   p=',
    formatC(p, digits=2, format='fg')), font.main=1)
abline(reg_i, col=rgb(1, 0, 0, .8))
```

:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
The loop ran through `i` random datasets before finding one with $p < 0.001$. With truly random data, each regression has a $0.1\%$ chance of producing $p < 0.001$. After $k$ independent tests, the probability of at least one false positive is $1-(1-\alpha)^k$. The probability climbs fast in $k$:

```{r}
alpha <- 0.001
k <- c(1, 10, 100, 700, 5000)
round(1 - (1 - alpha)^k, 3)
```

So at $k=700$ tries the chance of at least one false discovery exceeds 50%, and by $k=5000$ it is essentially 1. The more you search, the more likely you are to find a spurious result.
See [Multiple Testing](https://jadamso.github.io/Rbooks/03_26_MiscTopics.html#multiple-testing) in the next chapter for corrections that control this family-wise error rate.
:::

```{r}
# P-hacking 2SLS with different explanatory vars
# and different instrumental vars
library(fixest)
p <- 1
ii <- 0
set.seed(123)
while(p >= .05){
    # Get Random Covariates
    X2 <-  runif(n)    
    X3 <-  runif(n)
    # Create Treatment Variable based on Cutoff
    cutoffs <- seq(0, 1, length.out=11)[-c(1, 11)]
    for(tau in cutoffs){
        T3 <- 1*(X3 > tau)
        # Merge and `Analyze'
        dat_i <- data.frame(X1, X2, T3)
        ivreg_i <- feols(X1 ~ 1|X2 ~ T3, data=dat_i)
        # Update results in global environment
        ptab <- summary(ivreg_i)$coeftable
        if( nrow(ptab)==2){
            p <- ptab[2, 4]
            ii <- ii+1
        }
    }
}
summary(ivreg_i)
```

:::{.callout-note icon=false collapse="true" title="Must Know"}
The p-hacking example found a statistically significant IV result using random data. What safeguards could a researcher adopt to avoid this problem? Consider: (a) pre-registration of hypotheses, (b) adjusting for multiple testing, and (c) out-of-sample validation.
:::


## Spurious Regression

Even without any coding errors or p-hacking, you can sometimes make a false discovery. We begin with a motivating empirical example of "US Gov't Spending on Science".

:::{.callout-important title="Definition" collapse="false"}
*Spurious regression* arises when two unrelated time series share a common trend and so produce a large $\hat{R}^2$ and a small $p$-value when one is regressed on the other. The OLS standard errors assume independent observations; with trending series this assumption fails badly, and significance becomes a function of sample length rather than any underlying relationship.
:::


First, get and inspect some data from https://tylervigen.com/spurious-correlations
```{r}
# Your data is not made up in the computer (hopefully!)
vigen_csv <- read.csv(
    'https://raw.githubusercontent.com/the-mad-statter/whysospurious/master/data-raw/tylervigen.csv'
) 
class(vigen_csv)
names(vigen_csv)
vigen_csv[1:5, 1:5]
```


Examine some data
```{r}
par(mfrow=c(1, 2), mar=c(2, 2, 2, 1))
plot.new()
plot.window(xlim=c(1999, 2009), ylim=c(5, 9)*1000)
lines(science_spending/3 ~ year, data=vigen_csv, lty=1, col=rgb(1, 0, 0, .8), pch=16)
text(2003, 8200, 'US spending on science, space, technology (USD/3)', col=rgb(1, 0, 0, .8), cex=.6, srt=30)
lines(hanging_suicides ~ year, data=vigen_csv, lty=1, col=rgb(0, 0, 1, .8), pch=16)
text(2004, 6500, 'US Suicides by hanging, strangulation, suffocation (Deaths)', col=rgb(0, 0, 1, .8), cex=.6, srt=30)
axis(1)
axis(2)


plot.new()
plot.window(xlim=c(2002, 2009), ylim=c(0, 5))
lines(cage_films ~ year, data=vigen_csv[vigen_csv$year>=2002, ], lty=1, col=rgb(1, 0, 0, .8), pch=16)
text(2006, 0.5, 'Number of films with Nicolas Cage (Films)', col=rgb(1, 0, 0, .8), cex=.6, srt=0)
lines(pool_fall_drownings/25 ~ year, data=vigen_csv[vigen_csv$year>=2002, ], lty=1, col=rgb(0, 0, 1, .8), pch=16)
text(2006, 4.5, 'Number of drownings by falling into pool (US Deaths/25)', col=rgb(0, 0, 1, .8), cex=.6, srt=0)
axis(1)
axis(2)
```

:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
Both plots show variables with high correlations ($r > 0.9$) that have no causal connection. The correlations arise because both series trend upward over time. When two time series share a common trend, their correlation can be very high even if they are completely unrelated. This is why regression with trending time series requires special care (e.g., detrending or differencing).
:::

#### **Another Example**. {-}
The US government spending on science is ruining cinema
(p<.001)!?
```{r}
# Drop Data before 1999
vigen_csv <- vigen_csv[vigen_csv$year >= 1999, ]

# Run OLS Regression of Cage's film count on science spending.
# The p-value on `science_spending` will be tiny, despite zero plausible causal link.
reg1 <-  lm(cage_films ~ -1 + science_spending, data=vigen_csv)
summary(reg1)
```
It's not all bad, because people in Maine stay married longer?
```{r}
plot.new()
plot.window(xlim=c(1999, 2009), ylim=c(7, 9))
lines(log(maine_divorce_rate*1000) ~ year, data=vigen_csv)
lines(log(science_spending/10) ~ year, data=vigen_csv, lty=2)
axis(1)
axis(2)
legend('topright', lty=c(1, 2), bty='n', legend=c(
    'log(maine_divorce_rate*1000)',
    'log(science_spending/10)'))
```


For more intuition on spurious correlations, try http://shiny.calpoly.sh/Corr_Reg_Game/
The same principles apply to more sophisticated methods.

## Spurious Causal Impacts

As discussed earlier, good natural experiments are hard to find and bad ones are easy to find. The remainder of this section illustrates the second point: applying canonical causal-inference recipes (IV, RDD, DID) to data that is purely noise.

:::{.callout-important title="Definition" collapse="false"}
A *spurious instrument* is a variable that satisfies the formal statistical criteria for an instrumental variable (relevance, exogeneity) by chance, despite having no plausible causal connection to the endogenous regressor. Spurious instruments become easier to find as the number of candidate variables grows — the same multiple-testing problem as p-hacking, but applied to instrument selection.
:::

#### **Example: Vigen IV's**. {-}
We now run IV regressions for different variable combinations in the dataset of spurious relationships
```{r}
knames <- names(vigen_csv)[2:11] # First 10 Variables
#knames <- names(vigen_csv)[-1] # Try All Variables
p <- 1
ii <- 1
ivreg_list <- vector('list', factorial(length(knames))/factorial(length(knames)-3))

# Choose 3 variable
for( k1 in knames){
for( k2 in setdiff(knames, k1)){
for( k3 in setdiff(knames, c(k1, k2)) ){   
    X1 <- vigen_csv[, k1]
    X2 <- vigen_csv[, k2]
    X3 <- vigen_csv[, k3]
    # Merge and `Analyze'        
    dat_i <- na.omit(data.frame(X1, X2, X3))
    ivreg_i <- feols(X1 ~ 1|X2 ~ X3, data=dat_i)
    ivreg_list[[ii]] <- list(ivreg_i, c(k1, k2, k3))
    ii <- ii+1
}}}
pvals <- sapply(ivreg_list, function(ivreg_i){ivreg_i[[1]]$coeftable[2, 4]})

plot(ecdf(pvals), xlab='p-value', ylab='CDF', main=NA)
title('Frequency IV is Statistically Significant', font.main=1)
abline(v=c(.01, .05), col=c(rgb(1, 0, 0, .8), rgb(0, 0, 1, .8)))

# Most Significant Spurious Combinations
pvars <- sapply(ivreg_list, function(ivreg_i){ivreg_i[[2]]})
pdat <- data.frame(t(pvars), pvals)
pdat <- pdat[order(pdat$pvals), ]
head(pdat)

# Fraction of all combinations significant at conventional levels
c(at_0.05 = mean(pvals < 0.05, na.rm=TRUE),
  at_0.01 = mean(pvals < 0.01, na.rm=TRUE))
```
A noticeable share of the random variable triples returns "significant" results at both the 5% and 1% levels — far more than the 5% and 1% rates a single hypothesis test would imply.

#### **Simulation Study**. {-}
We apply the three major credible methods (IV, RDD, DID) to random walks. Each time, we find a result that fits mold and add various extensions that make it appear robust. One could tell a story about how $X_{2}$ affects $X_{1}$ but $X_{1}$ might also affect $X_{2}$, and how they discovered an instrument $X_{3}$ to provide the first causal estimate of $X_{2}$ on $X_{1}$. The analysis looks scientific and the story sounds plausible, so you could probably be convinced *if it were not just random noise.*


```{r, message=F, warning=F}
n <- 1000
n_index <- seq(n)

set.seed(1)
random_walk1 <- cumsum(runif(n, -1, 1))

set.seed(2)
random_walk2 <- cumsum(runif(n, -1, 1))

par(mfrow=c(1, 2))
plot(random_walk1, pch=16, col=rgb(1, 0, 0, .25),
    xlab='Time', ylab='Random Value')
plot(random_walk2, pch=16, col=rgb(0, 0, 1, .25),
    xlab='Time', ylab='Random Value')
```

**IV**. First, find an instrument that satisfy various statistical criterion to provide a causal estimate of $X_{2}$ on $X_{1}$.
```{r}
# 'Find' 'valid' ingredients
library(fixest)
random_walk3 <- cumsum(runif(n, -1, 1))
dat_i <- data.frame(
    X1=random_walk1,
    X2=random_walk2,
    X3=random_walk3)
ivreg_i <- feols(X1 ~ 1|X2 ~ X3, data=dat_i)
summary(ivreg_i)

# After experimenting with different instruments
# you can find even stronger results!
```

**RDD**. Second, find a large discrete change in the data that you can associate with a policy. You can use this as an instrument too, also providing a causal estimate of $X_{2}$ on $X_{1}$.
```{r, message=F, warning=F}
# Let the data take shape
# (around the large differences before and after)
n1 <- 290
wind1 <- c(n1-300, n1+300)
dat1 <- data.frame(t=n_index, y=random_walk1, d=1*(n_index > n1))
dat1_sub <- dat1[ n_index>wind1[1] & n_index < wind1[2], ]

# Then find your big break
reg0 <- lm(y ~ t, data=dat1_sub[dat1_sub$d==0, ])
reg1 <- lm(y ~ t, data=dat1_sub[dat1_sub$d==1, ])

# The evidence should show openly (it's just science)
plot(random_walk1, pch=16, col=rgb(0, 0, 1, .25),
    xlim=wind1, xlab='Time', ylab='Random Value')
abline(v=n1, lty=2)
lines(reg0$model$t, reg0$fitted.values, col=rgb(0, 0, 0, .8))
lines(reg1$model$t, reg1$fitted.values, col=rgb(0, 0, 0, .8))
```

```{r, message=F, warning=F, results='asis'}
# Dress with some statistics for added credibility
rdd_sub <- lm(y ~ d+t+d*t, data=dat1_sub)
rdd_full <- lm(y ~ d+t+d*t, data=dat1)
stargazer::stargazer(rdd_sub, rdd_full, 
    type='html',
    title='Recipe RDD',
    header=F,
    omit=c('Constant'),
    notes=c('First column uses a dataset around the discontinuity.',
    'Smaller windows are more causal, and where the effect is bigger.'))
```

**DID**. Third, find a change in the data that you can associate with a policy where the control group has parallel trends. This also provides a causal estimate of $X_{2}$ on $X_{1}$.
```{r, message=F, warning=F}
# Find a reversal of fortune
# (A good story always goes well with a nice pre-trend)
n2 <- 318
wind2 <- c(n2-20, n2+20)
plot(random_walk2, pch=16, col=rgb(0, 0, 1, .5),
    xlim=wind2, ylim=c(-15, 15), xlab='Time', ylab='Random Value')
points(random_walk1, pch=16, col=rgb(1, 0, 0, .5))
abline(v=n2, lty=2)
```

```{r, message=F, warning=F, results='asis'}
# Knead out any effects that are non-causal (aka correlation)
dat2A <- data.frame(t=n_index, y=random_walk1, d=1*(n_index > n2), RWid=1)
dat2B <- data.frame(t=n_index, y=random_walk2, d=0, RWid=2)
dat2  <- rbind(dat2A, dat2B)
dat2$RWid <- as.factor(dat2$RWid)
dat2$tid <- as.factor(dat2$t)
dat2_sub <- dat2[ dat2$t>wind2[1] & dat2$t < wind2[2], ]

# Report the stars for all to enjoy
# (what about the intercept?)
# (stable coefficients are the good ones?)
did_fe1 <- lm(y ~ d+tid, data=dat2_sub)
did_fe2 <- lm(y ~ d+RWid, data=dat2_sub)
did_fe3 <- lm(y ~ d*RWid+tid, data=dat2_sub)
stargazer::stargazer(did_fe1, did_fe2, did_fe3,
    type='html',
    title='Recipe DID',
    header=F,
    omit=c('tid', 'RWid', 'Constant'),
    notes=c(
     'Fixed effects for time in column 1, for id in column 2, and both in column 3.',
     'Fixed effects control for most of your concerns.',
     'Anything else creates a bias in the opposite direction.'))
```

:::{.callout-note icon=false collapse="true" title="Must Know"}
All three empirical recipes (IV, RDD, DID) produced statistically significant results from random walks with no true relationship. What is the common thread? In your own empirical work, how would you distinguish a genuine causal finding from a spurious one?
:::


## Exercises

1. Explain the difference between p-hacking and spurious regression. Both can produce statistically significant results from unrelated data, but the mechanisms differ. Give a concrete example of each using concepts from this chapter.

2. Using the `tylervigen.csv` dataset from the chapter, pick any two variables and run an OLS regression. Then pick a third variable as an instrument and run a 2SLS regression with `fixest::feols`. Report the p-values for both. Do these significant results reflect a real causal relationship? Why or why not?

3. Simulate two independent random walks of length 500 in R (each starting at 0, with increments from `runif(1, -1, 1)`). Regress one on the other using `lm()` and record the p-value on the slope. Repeat this 200 times and plot the distribution of p-values. What fraction are below 0.05, and how does this compare to what you would expect under independence?

#### **Further Reading**. {-}

* <https://www.econometrics-with-r.org/9-1-testing-joint-hypotheses.html> -- multiple testing and joint hypothesis testing in a regression context.
* Athey, S. and Imbens, G. W. (2019). Machine learning methods that economists should know about. *Annual Review of Economics*, 11, 685--725. <https://doi.org/10.1146/annurev-economics-080217-053433> -- a survey of when and how machine learning tools complement econometric inference.


:::{.callout-note title="Recall" collapse="false" icon=false}
This chapter cataloged three ways significant results arise from noise — data errors, $p$-hacking, and spurious regression — and showed that all three causal-inference recipes (IV, RDD, DID) can produce statistically significant findings from purely random data. In the next chapter we collect the remaining multivariate topics: prediction, multiple testing corrections, and the matrix-form OLS theory that underlies everything above.
:::
