# Observational Data
***

Most economic data are not generated by an experiment. They arrive with dependence across time, across space, and across the variables themselves — through market equilibrium, optimization, and selection. This chapter walks through those three forms of dependence in turn, each of which violates the independence assumption of standard regression and complicates causal interpretation.

## Temporal Interdependence

Many observational datasets have *temporal dependence*, meaning that values at one point in time are related to past values. This violates the standard assumption of independence used in many statistical methods.

Stock prices are classic examples of temporally dependent processes. If Apple's stock was high yesterday, it is more likely (but not guaranteed) to be high today.
```{r}
# Load a CSV of daily Apple stock prices and plot the daily high as a line.
library(plotly)
stock <- read.csv('https://raw.githubusercontent.com/plotly/datasets/master/finance-charts-apple.csv')
fig <- plot_ly(stock, type = 'scatter', mode = 'lines')%>%
  add_trace(x = ~ Date, y = ~ AAPL.High) %>%
  layout(showlegend = F)
fig
```

A *random walk* is the simplest mathematical model of temporal dependence. Each new value is just the previous value plus a random shock (white noise).
```{r}
# Generate Random Walk
tN <- 200
y <- numeric(tN)
y[1] <- stock$AAPL.High[1]
for (ti in 2:tN) {
    y[ti] <- y[ti-1] + runif(1, -10, 10)
}
#x <- runif(tN, -1, 1) White Noise

y_dat <- data.frame(Date=1:tN, RandomWalk=y)
fig <- plot_ly(y_dat, type = 'scatter', mode = 'lines') %>%
  add_trace(x= ~ Date, y= ~ RandomWalk) %>%
  layout(showlegend = F)
fig
```

:::{.callout-note icon=false collapse="true" title="Must Know"}
Modify the random walk above by changing `runif(1, -10, 10)` to `rnorm(1, 0, 5)`. How does the choice of shock distribution affect the path? Does it change whether the series is stationary or nonstationary?

*Hint.* Both choices produce zero-mean, finite-variance shocks, so each simulated path is still a random walk: the variance of $Y_t$ grows linearly in $t$, so the series is nonstationary regardless of whether the shocks are uniform or normal. The shape of individual paths looks slightly different but the long-run behavior is the same.
:::

In both plots, we see that today's value is not independent of past values. In contrast to *cross-sectional* data (e.g. individual incomes), *time series* often require special methods to account for memory and nonstationarity.


It often helps to see the marginal distribution alongside the time path.
A nonstationary series can disguise itself in a histogram because the histogram pools across $t$.
```{r}
# Time path beside the marginal distribution
layout(matrix(c(1, 2), nrow = 1), widths = c(4, 1))

par(mar=c(4, 4, 1, 1))
plot(1:tN, y, type='l', lwd=2,
    xlab='t', ylab=expression(y[t]))
title('Random walk: time path', font.main=1, adj=0)

par(mar=c(4, 1, 1, 2))
y_hist <- hist(y, breaks=20, plot=F)
barplot(y_hist$counts, horiz=TRUE, space=0, border=NA,
    col=grey(.5, .5), axes=FALSE)
axis(1)
title('Marginal', font.main=1, adj=0)
```

#### **Stationary**. {-}

The simplest time-series methods implicitly assume the data have a stable distribution over time; this assumption fails often enough that we need a name for it.

:::{.callout-important title="Key Definition" collapse="false"}
A time series is *stationary* if both its mean and variance do not depend on $t$:

* Stationary means: $\mathbb{E}[Y_{t}]=\mathbb{E}[Y_{t'}]$ for all time periods $t, t'$.
* Stationary variances: $\mathbb{V}[Y_{t}]=\mathbb{V}[Y_{t'}]$ for all time periods $t, t'$.
:::

Stationarity is useful as a precondition for most classical time-series tools, and violations come in distinct flavors: a series with a deterministic trend is mean-nonstationary, while a random walk is variance-nonstationary because $\mathbb{V}[Y_t]$ grows linearly in $t$.

For example, consider the data generating process $Y_t = \beta t + \epsilon_t$, with $\epsilon_t \sim \text{N}(0, \sigma + \alpha t)$, with parameters $\beta$ affecting mean stationarity and $\alpha$ affecting variance stationarity.

```{r}
tN <- 200
simulate_series <- function(beta, alpha, sigma=.2){
    y <- numeric(tN)
    for (ti in 1:tN) {
        mean_ti <- beta*ti
        sd_ti <- (.2 + alpha*ti)
        y[ti] <- mean_ti + rnorm(1, sd=sd_ti)
    }
    return(y)
}

# Plotting Functions
plot_setup <- function(alpha, beta){
    plot.new()
    plot.window(xlim=c(1, tN), ylim=c(-5, 20))
    axis(1)
    axis(2)
    mtext(expression(y[t]), 2, line=2.5)
    mtext('Time (t)', 1, line=2.5)
}
plot_title <- function(alpha, beta){
    beta_name <- ifelse(beta==0, 'Mean Stationary', 'Mean Nonstationary')
    alpha_name <- ifelse(alpha==0, 'Var Stationary', 'Var Nonstationary')
    title(paste0(beta_name, ', ', alpha_name), font.main=1, adj=0)
}

par(mfrow = c(2, 2))
for(alpha in c(0, .015)){
for(beta in c(0, .05)){
    plot_setup(alpha=alpha, beta=beta)
    sim_cols <- c(rgb(.8, 0, 0, .5), rgb(0, 0, .8, .5))
    for( sim in 1:2){
        y_sim <- simulate_series(beta=beta, alpha=alpha)
        lines(y_sim, col=sim_cols[sim], lwd=2)
    }
    plot_title(alpha=alpha, beta=beta)
}}
```

:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
In the top-left panel ($\beta=0, \alpha=0$), both the mean and variance are constant over time -- this is stationary. In the top-right panel ($\beta>0, \alpha=0$), the mean drifts upward but the variance stays constant -- mean nonstationary. In the bottom-left ($\beta=0, \alpha>0$), the mean is constant but the spread grows -- variance nonstationary. The bottom-right has both problems.
:::

#### **Measures of temporal association**. {-}

When values at one time depend on values at other times, we need correlation-based diagnostics across lags to characterize the dependence.

:::{.callout-important title="Key Definition" collapse="false"}
The *autocorrelation function* of $Y$ at lag $k$ is the correlation of $Y_t$ with its own lagged value $Y_{t-k}$,
$$ACF_{Y}(k) = \frac{\mathbb{C}(Y_{t},Y_{t-k})}{ \sqrt{\mathbb{V}(Y_{t})\mathbb{V}(Y_{t-k})}}.$$
The *cross-correlation function* between $Y$ and $X$ at lag $k$ is the corresponding correlation of $Y_t$ with $X_{t-k}$,
$$CCF_{YX}(k) = \frac{\mathbb{C}(Y_{t},X_{t-k})}{ \sqrt{\mathbb{V}(Y_t)\mathbb{V}(X_{t-k})}}.$$
:::

The ACF is useful for detecting temporal persistence (memory) within one series: for stationary processes it typically decays quickly, while nonstationary processes show ACFs that decay slowly or persist.
The CCF is useful for detecting lagged relationships between two series, such as leading indicators or external drivers; if $X$ is white noise, any visible structure in the CCF likely reflects nonstationarity in $Y$.

:::{.callout-note icon=false collapse="true" title="Must Know"}
Worked example: $y=(2,4,6,8,10)$, with $n=5$ and mean $\bar{y}=6$.
R's `acf()` uses the biased estimator
$$
\widehat{ACF}(1) = \frac{\frac{1}{n}\sum_{t=2}^{n}(y_t - \bar{y})(y_{t-1}-\bar{y})}{\frac{1}{n}\sum_{t=1}^{n}(y_t - \bar{y})^2}.
$$
Deviations are $(-4,-2,0,2,4)$; lag-1 cross-products are $(-4)(-2)=8$, $(-2)(0)=0$, $(0)(2)=0$, $(2)(4)=8$, summing to $16$.
Sum of squared deviations is $16+4+0+4+16=40$.
So $\widehat{ACF}(1) = 16/40 = 0.4$.

```{r}
y_acf <- c(2,4,6,8,10)
acf(y_acf, plot=FALSE)$acf[2]
```
:::

```{r}
par(mfrow = c(2, 2))
for(alpha in c(0, .015)){
for(beta in c(0, .05)){
    y_sim <- simulate_series(beta=beta, alpha=alpha)
    acf(y_sim, main=NA)
    plot_title(alpha=alpha, beta=beta)
}}
```

A *cross-correlation* plot diagnoses lagged dependence between two series at once:

:::{.callout-note icon=false collapse="true" title="Must Know"}
Worked CCF example. Let $y=(1,2,3,4,5)$ and $x=(2,3,4,5,6)$, with $\bar{y}=3, \bar{x}=4$.
The lag-1 cross-correlation pairs $y_t$ with $x_{t-1}$, giving three cross-products $(2-3)(2-4) = 2$, $(3-3)(3-4) = 0$, $(4-3)(4-4) = 0$, $(5-3)(5-4) = 2$.
With $n=5$, the (biased) denominator is $\sqrt{\sum_t(y_t-\bar y)^2 \sum_t(x_t-\bar x)^2}/n = \sqrt{10 \cdot 10}/5 = 2$, so $\widehat{CCF}(1) = (2+0+0+2)/5 / 2 = 0.4$.
```{r}
y_ccf <- c(1,2,3,4,5)
x_ccf <- c(2,3,4,5,6)
ccf(y_ccf, x_ccf, lag.max=1, plot=FALSE)$acf
```
:::

```{r}
x_sim <- runif(tN, -1, 1) # White Noise
par(mfrow = c(2, 2))
for(alpha in c(0, .015)){
for(beta in c(0, .05)){
    y_sim <- simulate_series(beta=beta, alpha=alpha)
    ccf(y_sim, x_sim, main=NA)
    plot_title(alpha=alpha, beta=beta)
}}
```

## Spatial Interdependence

Many observational datasets exhibit *spatial dependence*, meaning that values at one location tend to be related to values at nearby locations. This violates the standard assumption of independent observations used in many classical statistical methods.

For example, elevation is spatially dependent: if one location is at high elevation, nearby locations are also likely (though not guaranteed) to be high. Similarly, socioeconomic outcomes like disease rates or income often cluster geographically due to shared environmental or social factors.

Just as stock prices today depend on yesterday, spatial variables often depend on neighboring regions, creating a need for specialized statistical methods that account for spatial autocorrelation.

#### **Raster vs. Vector Data**. {-}
Spatial data typically comes in two formats, each suited to different types of information:

* *Vector data* uses geometric shapes (points, lines, polygons) to store data. E.g., a census tract map that stores data on population demographics.
* *Raster data* uses grid cells (typically squares, but sometimes hexagons) to store data. E.g., an image that stores data on elevation above seawater.
 

```{r}
# Vector Data
library(sf)
northcarolina_vector <- st_read(system.file('shape/nc.shp', package='sf'))
plot(northcarolina_vector['BIR74'], main=NA)
title('Number of Live Births in 1974', font.main=1)
# https://r-spatial.github.io/spdep/articles/sids.html
```

```{r}
# Raster Data
library(terra)
luxembourg_elevation_raster <- rast(system.file('ex/elev.tif', package='terra'))
plot(luxembourg_elevation_raster)
```

#### **Stationary.** {-}
Just as with temporal data, stationarity in spatial data means that the statistical properties (like mean, variance, or spatial correlation) are roughly the same across space.

* Stationary Means: $\mathbb{E}[Y(s)]=\mathbb{E}[Y(s')]$ for all spatial locations $s,s'$
* Stationary Vars: $\mathbb{V}[Y(s)]=\mathbb{V}[Y(s')]$ for all spatial locations $s,s'$

```{r}
# Simulated 2D spatial fields
set.seed(1)
n <- 20
x <- y <- seq(0, 1, length.out = n)
grid <- expand.grid(x = x, y = y)

# 1. Stationary: Gaussian with constant mean and var
z_stationary <- matrix(rnorm(n^2, 0, 1), n, n)

# 2. Nonstationary: Mean increases with x and y
z_nonstationary <- outer(x, y, function(x, y) 3*x*y) + rnorm(n^2, 0, 1)

par(mfrow = c(1, 2))
# Stationary field
image(x, y, z_stationary,
      main = NA,
      col = terrain.colors(100),
      xlab = 'x', ylab = 'y')
title('Stationary Field', font.main = 1)
# Nonstationary field
image(x, y, z_nonstationary,
      main = NA,
      col = terrain.colors(100),
      xlab = 'x', ylab = 'y')
title('Nonstationary Field', font.main = 1)
```

#### **Measures of spatial association**. {-}

Just as temporal data can exhibit autocorrelation, spatial data can too — nearby locations tend to be more (or less) similar than spatial independence would predict.

:::{.callout-important title="Key Definition" collapse="false"}
*Moran's I* is an index of spatial autocorrelation ranging from $-1$ to $1$. Values near $1$ indicate that nearby locations have similar values, $-1$ that they have dissimilar values, and $0$ that they are spatially independent.
:::

Moran's I is useful as the spatial analog of the ACF: it summarizes overall spatial dependence in one number, and large positive values signal that observations near each other share information we cannot treat as independent.
A two-variable analog measures *spatial cross-correlation* — computed similarly to the CCF but with spatial rather than temporal lags.

```{r}
# Raster Data Example
autocor(luxembourg_elevation_raster, method='moran', global=T)
```

:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
Moran's I ranges from $-1$ to $1$. A value near $1$ indicates strong positive spatial autocorrelation (nearby locations have similar values), near $-1$ indicates negative spatial autocorrelation (nearby locations have dissimilar values), and near $0$ indicates spatial independence. For elevation data, we expect strong positive autocorrelation because terrain changes gradually.
:::

Cross-Correlation. We can also assesses the relationship between two variables at varying distances. 
```{r}
# Vector Data Example
dat <- as.data.frame(northcarolina_vector)[, c('BIR74', 'SID74')]
mu <- colMeans(dat)

# Format Distances
dmat <- st_distance( st_centroid(northcarolina_vector) )
dmat <- units::set_units(dmat, 'km')

# At Which Distances to Compute CCF
# summary(dmat[, 1])
rdists <- c(-1, seq(0, 100, by=25)) # includes 0
rdists <- units::set_units(rdists , 'km')

# Compute Cross-Covariances
varXY <- prod( apply(dat, 2, sd) )
CCF <- lapply( seq(2, length(rdists)), function(ri){
    # Which Observations are within (rmin, rmax] distance
    dmat_r <- dmat
    d_id <- (dmat_r > rdists[ri-1] & dmat_r <= rdists[ri]) 
    dmat_r[!d_id]  <- NA
    # Compute All Covariances (Stationary)
    covs_r <- lapply(1:nrow(dmat_r), function(i){
        pairsi <- which(!is.na(dmat_r[i, ]))        
        covXiYj <- sapply(pairsi, function(j) {
            dXi <- dat[i, 1] - mu[1]
            dYj <- dat[j, 2] - mu[2]
            return(dXi*dYj)
        })
        return(covXiYj)
    })
    corXY <- unlist(covs_r)/varXY
    return(corXY)
} )
```

```{r}
# Plot Cross-Covariance Function
x <- as.numeric(rdists[-1])

par(mfrow=c(1, 2))

# Distributional Summary
boxplot(CCF,
    outline=F, whisklty=0, staplelty=0,
    ylim=c(-1, 1), #quantile(unlist(CCF), probs=c(.05, .95)),
    names=x,
    main=NA,
    xlab='Distance [km]',
    ylab='Cross-Correlation of BIR74 and SID74')
title('Binned Medians and IQRs', font.main=1, adj=0)
abline(h=0, lty=2)

# Inferential Summary
CCF_means <- sapply(CCF, mean)
plot(x, CCF_means,
    ylim=c(-1, 1),
    type='o', pch=16,
    main=NA,
    xlab='Distance [km]',
    ylab='Cross-Correlation of BIR74 and SID74')
title('Binned Means + 95% Confidence Band', font.main=1, adj=0)
abline(h=0, lty=2)    
# Quick and Dirty Subsampling CI
CCF_meanCI <- sapply(CCF, function(corXY){
    ss_size <- floor(length(corXY)*3/4)
    corXY_boot <- sapply(1:200, function(b){
        corXY_b <- sample(corXY, ss_size, replace=F)
        mean(corXY_b, na.rm=T)
    })
    quantile(corXY_boot,  probs=c(.025, .975), na.rm=T)
})
polygon( c(x, rev(x)), 
    c(CCF_meanCI[1, ], rev(CCF_meanCI[2, ])), 
    col=grey(0, .25), border=NA)
```

## Economic Interdependence

In addition to spatial and temporal dependence, many observational datasets exhibit interdependence between variables for economic reasons.

#### **Endogeneity**. {-}

Economic outcomes are not free variables we can move at will — many are jointly determined, affected by selection, or driven by variables we have not measured.

:::{.callout-important title="Key Definition" collapse="false"}
A regressor is *endogenous* if it is correlated with the error term in a regression. The three classic sources are

 * *reverse causality*: $Y \to X$.
 * *simultaneity*: $Y \to X$ and $X \to Y$.
 * *omitted variables*: $Z\to Y$ and either $Z\to X$ or $X \to Z$.
:::

Endogeneity is the central reason observational data resist clean causal interpretation: in the linear model $Y=X\beta + \epsilon$, endogeneity means $\mathbb{E}[X'\epsilon] \neq 0$, so OLS is biased away from the structural causal parameter.^[$X$ and $\epsilon$ may be correlated for other reasons too, such as when $X$ is measured with error.]
The next chapter introduces experimental and quasi-experimental designs that supply exogenous variation in $X$ and recover what OLS alone cannot.

:::{.callout-note icon=false collapse="true" title="Must Know"}
Consider the relationship between education and wages. Identify a potential source of (a) omitted variable bias, (b) reverse causality, and (c) measurement error. For each, explain the direction in which you expect the OLS estimate to be biased.
:::

```{r}
# Simulate data with an endogeneity issue
n <- 300
z <- rbinom(n, 1, .5)
xy <- sapply(z, function(zi){
    y <- rnorm(1, zi, 1)
    x <- rnorm(1, zi*2, 1)
    c(x, y)
})
xy <- data.frame(x=xy[1, ], y=xy[2, ])
plot(y ~ x, data=xy, pch=16, col=grey(0, .25), main=NA)
abline(lm(y ~ x, data=xy), col=rgb(1, 0, 0, .8))
```

I will focus on the seminal economic example to provide some intuition.


#### **Competitive Market Equilibrium**. {-}
This model has three structural relationships: (1) market supply is the sum of quantities supplied by individual firms at a given price, (2) market demand is the sum of quantities demanded by individual people at a given price, and (3) market supply equals market demand in equilibrium. Assuming market supply and demand are linear, we can write these three relationships as
\begin{eqnarray}
\label{eqn:market_supply}
Q_{S}(P) &=& \alpha_{S} + \beta_{S} P + E_{S},\\
\label{eqn:market_demand}
Q_{D}(P) &=& \alpha_{D} - \beta_{D} P + E_{D},\\
\label{eqn:market_eq}
Q_{D} &=& Q_{S} = Q.
%%  $Q_{D}(P) = \sum_{i} q_{D}_{i}(P)$,
\end{eqnarray}

#### **Reduced Form**. {-}

When two variables are jointly determined in equilibrium, neither one is "exogenous" — we can only observe their joint outcomes.

:::{.callout-important title="Key Definition" collapse="false"}
The *reduced form* of a structural system writes each endogenous outcome (here $P^{*}$ and $Q^{*}$) as a function of the exogenous variables (here the intercepts and shocks) only.
:::

The reduced form is useful as a clean separation between what observational data can identify and what they cannot: the reduced-form intercepts and shocks combine in known ways, but the structural slopes $\beta_S, \beta_D$ require additional identifying variation (introduced in the next chapter) to recover.
The equilibrium condition $Q_D = Q_S$ in the structural equations above is exactly what triggers this simultaneity, leaving both price and quantity as outcomes.
```{r}
# Demand Curve Simulator
qd_fun <- function(p, Ad=8, Bd=-.8, Ed_sigma=.25){
    Qd <- Ad + Bd*p + rnorm(1, 0, Ed_sigma)
    return(Qd)
}

# Supply Curve Simulator
qs_fun <- function(p, As=-8, Bs=1, Es_sigma=.25){
    Qs <- As + Bs*p + rnorm(1, 0, Es_sigma)
    return(Qs)
}

# Quantity Supplied and Demanded at 3 Prices
cbind(P=8:10, D=qd_fun(8:10), S=qs_fun(8:10))

# Market Equilibrium Finder
eq_fun <- function(demand, supply, P){
    # Compute EQ (what we observe)
    eq_id <- which.min( abs(demand-supply) )
    eq <- c(P=P[eq_id], Q=demand[eq_id]) 
    return(eq)
}
```

```{r}
# Simulations Parameters
N <- 300 # Number of Market Interactions
P <- seq(5, 10, by=.01) # Price Range to Consider

# Generate Data from Competitive Market  
# Plot Underlying Process
plot.new()
plot.window(xlim=c(0, 2), ylim=range(P))
EQ1 <- sapply(1:N, function(n){
    # Market Data Generating Process
    demand <- qd_fun(P)
    supply <- qs_fun(P)
    eq <- eq_fun(demand, supply, P)    
    # Plot Theoretical Supply and Demand
    lines(demand, P, col=grey(0, .01))
    lines(supply, P, col=grey(0, .01))
    points(eq[2], eq[1], col=grey(0, .05), pch=16)
    # Save Data
    return(eq)
})
axis(1)
axis(2)
mtext('Quantity', 1, line=2)
mtext('Price', 2, line=2)
```

Suppose we ask "what is the effect of price on quantity?" You can simply run a regression of quantity, $Y$, on price, $X$, and get the slope coefficient $\hat{b}_{1} = \hat{C}_{Q^{*} P^{*}} / \hat{V}_{P^{*}}$. You always get a number back, but it is hard to interpret meaningfully. 
```{r}
# Analyze Market Data
dat1 <- data.frame(t(EQ1), cost='1', T=1:N)
reg1 <- lm(Q ~ P, data=dat1)
summary(reg1)
```
This simple example has a profound insight: price-quantity data does not generally tell you how price affects quantity (or vice-versa). The reason is simultaneity: price and quantity mutually cause one another in markets.^[Although there are many ways this simultaneity can happen, economic theorists have made great strides in analyzing the simultaneity problem as it arises from equilibrium market relationships. In fact, 2SLS arose to understand agricultural markets. With a linear structure to supply and demand, we can even use algebra to solve for the equilibrium price and quantity analytically as
\begin{eqnarray}
P^{*} &=& \frac{\alpha_{D}-\alpha_{S}}{\beta_{D}+\beta_{S}} + \frac{E_{D} - E_{S}}{\beta_{D}+\beta_{S}}, \\
Q^{*} &=& \frac{\alpha_{S}\beta_{D}+ \alpha_{D}\beta_{S}}{\beta_{D}+\beta_{S}} + \frac{E_{S}\beta_{D}+ E_{D}\beta_{S}}{\beta_{D}+\beta_{S}}.
\end{eqnarray}
]

Moreover, this example also clarifies that our initial question "what is the effect of price on quantity?" is misguided. We could more sensibly ask  "what is the effect of price on quantity supplied?" or "what is the effect of price on quantity demanded?"


#### **Contamination**. {-}

With multiple linear regression, endogeneity biases are not just a problem for your main variable of interest. Suppose you are interested in how $X_{1}$ affects $Y$, conditioning on $X_{2}$, and that the data generating process is linear: $Y=\beta_{0}+\beta_{1}X_{1}+\beta_{2}X_{2}+\epsilon$. You paid special attention in your research design to find a case where $X_{1}$ is truly exogenous. Unfortunately, if $X_{2}$ is correlated with the error term, then there is also a bias for $X_{1}$. The magnitude of the bias for $X_{1}$ depends on the correlations between $X_{1}$ and $X_{2}$ as well as $X_{2}$ and $\epsilon$.^[Denoting $X=[X_{1}, X_{2}]$, you estimate the OLS coefficients $\hat{B}^{*}$. After algebraic work, you can find the bias $\mathbb{E}[ \hat{B}^{*} - \beta]$ suffers from a contamination effect.
\begin{eqnarray}
\hat{B^{*}} &=& [\hat{X}'\hat{X}]^{-1}X'y \\
\mathbb{E}[X'\epsilon] &=& 
\begin{bmatrix}
0 \\ \rho
\end{bmatrix}\\
\mathbb{E}[ \hat{B}^{*} - \beta] &=& [\hat{X}'\hat{X}]^{-1} \begin{bmatrix}
0 \\ \rho
\end{bmatrix} \neq
\begin{bmatrix}
0 \\ \rho
\end{bmatrix}
\end{eqnarray}
In words: when $X_2$'s row of $X'\epsilon$ has expectation $\rho$, the bias *spreads* through the inverse $(X'X)^{-1}$ onto every coefficient — including the one you cared about, $X_1$.
]

In the next chapter we look at the standard remedy: experimental and quasi-experimental designs that supply exogenous variation in a regressor and let us recover structural parameters that observational data alone cannot identify.


## Exercises

1. Explain why simultaneity in a competitive market makes it difficult to interpret the OLS slope $\hat{b}_1$ from a regression of quantity on price. In your answer, distinguish between the structural demand slope $\beta_D$ and what OLS actually estimates.

2. Using the North Carolina births data (`sf::st_read(system.file("shape/nc.shp", package="sf"))`), compute Moran's I for the variable `SID74` (sudden infant deaths in 1974). Is there evidence of spatial autocorrelation? Briefly explain what a high value of Moran's I would imply for standard OLS inference.

3. Simulate a random walk of length 200 in R (starting at 0, with increments drawn from `runif(1, -1, 1)`). Plot the series, compute the ACF using `acf()`, and describe how the ACF pattern differs from what you would expect for white noise.

#### **Further Reading**. {-}

* <https://theeffectbook.net/ch-CausalPaths.html> -- causal diagrams, confounding, and omitted variable bias explained with examples.
* <https://www.mostlyharmlesseconometrics.com/> -- companion site for Angrist and Pischke's textbook on credible research designs for observational data.
* <https://mixtape.scunning.com/03-directed_acyclical_graphs> -- DAGs and the logic of identification in observational studies.


#### Recall {-}

This chapter introduced three forms of observational dependence — temporal, spatial, and economic — and the diagnostic tools (ACF/CCF, Moran's I, endogeneity taxonomy) for each. The supply-demand simulation made the endogeneity bite concrete: we ran 300 market periods through `qd_fun`/`qs_fun`, fit `lm(Q ~ P, data=dat1)`, and got a slope that confuses the supply and demand sides because price and quantity are jointly determined. In the next chapter we turn to experimental and quasi-experimental designs that supply the exogenous variation needed to identify causal effects: RDD, DID, and 2SLS.

