# Local Relationships
***

In [Local Regression](https://jadamso.github.io/Rbooks/02_14_LocalRegression.html), we estimated nonparametric models with a single explanatory variable.
We now extend those tools to multiple explanatory variables.
The ideas carry over directly: instead of binning one variable, we bin each variable, and instead of fitting a line in pieces, we fit a plane in pieces.
This chapter covers three nonparametric tools for multivariate data: regressograms, piecewise regressions, and cross-validated model selection.
It then shows how to summarize the gradients these models produce.

## Regressograms

#### **Setup**. {-}

We want a flexible estimate of $\mathbb{E}[Y \mid X_1, X_2]$ that does not commit to any functional form for either explanatory variable.

:::{.callout-important title="Key Definition" collapse="false"}
A *multivariate regressogram* cuts each explanatory variable into exclusive bins (each encoded as a dummy variable), then predicts $\mathbb{E}[\hat{Y}]$ within each bin combination using OLS on the full set of dummies. It is the multivariate generalization of the univariate regressogram from [Local Regression](https://jadamso.github.io/Rbooks/02_14_LocalRegression.html#regressograms).
:::

The multivariate regressogram is useful for estimating a conditional mean without committing to linearity or any particular interaction structure — every cell gets its own prediction.
This flexibility is also its main drawback: as the number of explanatory variables grows, the cells empty out and predictions in sparse regions become noisy; the simulated data below uses two predictors to keep things visible.

```{r}
## Simulate data: y is a nonlinear function of x1 and x2 plus noise.
N <- 10000
e <- rnorm(N)
x1 <- seq(.1, 20, length.out=N)
x2 <- runif(N, 0, 1)
y  <- 3*exp(-2*x2 + 1.5*x1 - .1*x1^2)*x1 + e
dat <- data.frame(x1, x2, y)

## Create color palette (reused in later examples)
col_scale <- seq(min(y)*1.1, max(y)*1.1, length.out=401)
ycol_pal <- hcl.colors(length(col_scale), alpha=.5)
names(ycol_pal) <- sort(col_scale)

## Add legend (reused in later examples)
add_legend <- function(col_scale,
    yl=11,
    colfun=function(x){ hcl.colors(x, alpha=.5) },
    ...) {
  opar <- par(fig=c(0, 1, 0, 1), oma=c(0, 0, 0, 0), 
              mar=c(0, 0, 0, 0), new=TRUE)
  on.exit(par(opar))
  h <- hist(col_scale, plot=F, breaks=yl-1)$mids
  plot(0, 0, type='n', bty='n', xaxt='n', yaxt='n')
  legend(...,
    legend=h,
    fill=colfun(length(h)),
    border=NA,
    bty='n')
}


## Plot Data
par(oma=c(0, 0, 0, 2))
plot(x1 ~ x2, dat,
    col=ycol_pal[cut(y, col_scale)],
    pch=16, cex=.5,
    main=NA)
title('Raw Data', font.main=1)
add_legend(x='topright', col_scale=col_scale,
    yl=6, inset=c(0, .05), title='y')
```

#### **OLS Baseline**. {-}

We start with a global linear baseline, fitted without an interaction.
```{r}
## OLS (without interaction; the additive baseline)
reg <- lm(y ~ x1+x2, data=dat)

## Grid Points for Prediction
# X1 bins
l1 <- 11
bks1 <- seq(0, 20, length.out=l1)
h1 <- diff(bks1)[1]/2
mids1 <- bks1[-1]-h1
# X2 bins
l2 <- 11
bks2 <- seq(0, 1, length.out=l2)
h2 <- diff(bks2)[1]/2
mids2 <- bks2[-1]-h2
# Grid
pred_x <- expand.grid(x1=mids1, x2=mids2)

## OLS Predictions
pred_ols <- predict(reg, newdata=pred_x)
pred_df_ols  <- cbind(pred_ols, pred_x)

## Plot Predictions
par(oma=c(0, 0, 0, 2))
plot(x1 ~ x2, pred_df_ols,
    col=ycol_pal[cut(pred_ols, col_scale)],
    pch=15, cex=2,
    main=NA)
title('OLS Predictions', font.main=1)
add_legend(x='topright', col_scale=col_scale,
    yl=6, inset=c(0, .05), title='y')
```

#### **Multivariate Bins**. {-}

```{r}
##################
# Multivariate Regressogram
##################

## Cut each variable into bins, then interact them: each (x1c, x2c) cell
## gets its own mean. With the `*` interaction the model is fully saturated.
dat$x1c <- cut(dat$x1, bks1)
dat$x2c <- cut(dat$x2, bks2)

## Regressogram (interacted bins -> one mean per (x1c, x2c) cell)
reg <- lm(y ~ x1c*x2c, data=dat)

## Predicted Values
## For Points in Middle of Each Bin
pred_df_rgrm <- expand.grid(
    x1c=levels(dat$x1c),
    x2c=levels(dat$x2c))
pred_df_rgrm$yhat <- predict(reg, newdata=pred_df_rgrm)
pred_df_rgrm <- cbind(pred_df_rgrm, pred_x)

## Plot Predictions
par(oma=c(0, 0, 0, 2))
plot(x1 ~ x2, pred_df_rgrm,
    col=ycol_pal[cut(pred_df_rgrm$yhat, col_scale)],
    pch=15, cex=2,
    main=NA)
title('Regressogram Predictions', font.main=1)
add_legend(x='topright', col_scale=col_scale,
    yl=6, inset=c(0, .05), title='y')
```

:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
The regressogram divides the $(x_1, x_2)$ space into a grid of bins and estimates the mean of $y$ within each bin. Compared to OLS, the regressogram makes no assumption about functional form: it can capture nonlinear and interaction effects. The trade-off is that it requires more data to fill each bin, especially as the number of variables grows.
:::

:::{.callout-note icon=false collapse="true" title="Must Know"}
A regressogram prediction is just the average of $\hat{Y}_i$ within a bin.
Consider six observations, with $x_1$ and $x_2$ each split into "low" and "high".

```{r}
ex <- data.frame(
    y  = c(2, 4, 5, 9, 8, 12),
    x1 = c('low', 'low', 'low', 'high', 'high', 'high'),
    x2 = c('low', 'low', 'high', 'low', 'high', 'high'))

# Regressogram prediction = mean of y within each (x1, x2) cell
aggregate(y ~ x1 + x2, data=ex, FUN=mean)
```

The fitted value for any point is its cell mean: $(2+4)/2 = 3$ for the (low, low) cell, $5$ for (low, high), $9$ for (high, low), and $(8+12)/2 = 10$ for (high, high).
:::


## Local Regressions

#### **Piecewise Model**. {-}

A regressogram is *locally constant* — every point in a bin gets the same prediction — and we sometimes want slopes within each bin instead of a flat shelf.

:::{.callout-important title="Key Definition" collapse="false"}
A multivariate *piecewise regression* bins one explanatory variable and fits a separate OLS line on the others within each bin.
:::

A *piecewise regression* is useful as a middle ground between the flat regressogram and the global OLS plane: it tracks gradients within each bin so predictions vary smoothly, and jumps only at the bin edges.
We bin one variable, then fit a full linear model of $\hat{Y}$ on the explanatory variables within each bin.

```{r}
##################
# Multivariate Piecewise Regression
##################

## Bin x1, then fit a linear model of y on (x1, x2) within each bin
dat$x1c <- cut(dat$x1, bks1)
preg <- lm(y ~ x1c/(x1 + x2), data=dat)

## Predicted Values
## For the same grid of points used above
pred_df_preg <- pred_x
pred_df_preg$x1c <- cut(pred_df_preg$x1, bks1)
pred_df_preg$yhat <- predict(preg, newdata=pred_df_preg)

## Plot Predictions
par(oma=c(0, 0, 0, 2))
plot(x1 ~ x2, pred_df_preg,
    col=ycol_pal[cut(pred_df_preg$yhat, col_scale)],
    pch=15, cex=2,
    main=NA)
title('Piecewise Regression Predictions', font.main=1)
add_legend(x='topright', col_scale=col_scale,
    yl=6, inset=c(0, .05), title='y')
```

The formula `y ~ x1c/(x1 + x2)` gives each `x1` bin its own intercept and its own slopes on `x1` and `x2`.
Running one regression per data subset with `split()` gives identical predictions, as shown for the bivariate case in [Local Regression](https://jadamso.github.io/Rbooks/02_14_LocalRegression.html#piecewise-regression).
A fully *local linear regression* generalizes this further, replacing hard bins with kernel weights and a separate bandwidth for each variable; see [Local Regression](https://jadamso.github.io/Rbooks/02_14_LocalRegression.html#local-linear-regression) for the bivariate version.

:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
Compare the three sets of predictions in this chapter (OLS, regressogram, and piecewise).
Which captures the curvature in `x1`?
Which is smoothest?
Refit the piecewise model with finer bins (`bks1 <- seq(0, 20, length.out=21)`) and describe how the prediction plot changes.
:::


#### **Break Points**. {-}
Kinks and discontinuities in the relationship between $Y$ and $X$ can be modeled with factor variables, just like any other transformation.
We can then use $F$-tests to examine whether a break is statistically significant.
Both the interaction-model $F$-test (`anova`) and the *Chow test* below test the same null of "no break", and give the same answer.
```{r}
library(Ecdat)
reg <- lm(wage ~ school, data=Wages1)

# F-test for a break at school = 12
reg2 <- lm(wage ~ school*I(school>12), data=Wages1)
anova(reg, reg2)

# Chow test for the same break
data_splits <- split(Wages1, Wages1$school <= 12)
resids <- sapply(data_splits, function(subdat){
    reg_sub <- lm(wage ~ school, data=subdat)
    sum(resid(reg_sub)^2)
})
Ns <- sapply(data_splits, function(subdat){ nrow(subdat) })

# Chow F-statistic: does splitting the sample reduce RSS enough?
Rt <- (sum(resid(reg)^2) - sum(resids))/sum(resids)
Rb <- (sum(Ns) - 2*reg$rank)/reg$rank
Ft <- Rt*Rb
pf(Ft, reg$rank, sum(Ns)-2*reg$rank, lower.tail=F)

# See also
# strucchange::sctest(wage ~ school, data=Wages1, type='Chow', point=.5)
# segmented::segmented(reg)
```
A small $p$-value is evidence that the wage-schooling slope changes at 12 years (the end of secondary school).

:::{.callout-note icon=false collapse="true" title="Must Know"}
The Chow test checks whether a structural break exists at a specified point. What happens if you do not know where the break is? How does searching over many potential break points affect the validity of your test? (Hint: think about multiple testing.)
:::



## Model Selection

#### **Leave-One-Out and K-Fold CV**. {-}

A flexible model can fit the training data perfectly while predicting new observations badly, so we need a way to measure performance on data the model has not seen.

:::{.callout-important title="Key Definition" collapse="false"}
*Cross-validation (CV)* picks a model by minimizing out-of-sample prediction error. *Leave-one-out CV* drops one observation at a time, refits, and averages the squared prediction errors:
\begin{eqnarray}
\min_{\mathbf{H}} \quad \frac{1}{n} \sum_{i=1}^{n} \left[ \hat{Y}_{i} - \hat{y_{[i]}}(\mathbf{X},\mathbf{H}) \right]^2,
\end{eqnarray}
where $\hat{y}_{[i]}(\mathbf{X},\mathbf{H})$ is the model's prediction at $\mathbf{X}_{i}$ from a fit that excluded $\mathbf{X}_{i}$, and $\mathbf{H}$ is the bandwidth matrix. *$k$-fold CV* drops one of $k$ random groups at a time instead of one observation.
:::

Cross-validation is useful because it directly penalizes overfitting: a too-flexible model predicts held-out observations poorly even though it fits in-sample data tightly.
With a weighted least squares regression on three explanatory variables, the bandwidth matrix has the form
\begin{eqnarray}
\mathbf{H}=\begin{pmatrix}
h_{1} & 0 & 0  \\
 0    & h_{2} & 0 \\
 0    & 0 & h_{3} \\
\end{pmatrix},
\end{eqnarray}
where each $h_{k}$ is the bandwidth for variable $X_{k}$. In the discrete-bin setting below, the role of the bandwidth is played by the *number of bins $L$*: more bins means finer resolution (a smaller effective bandwidth) and fewer observations per bin.

There are many types of cross-validation (Arlot and Celisse, 2010; Bates et al., 2023). *Generalized cross-validation* adjusts for the degrees of freedom, whereas the `npreg` function in R uses *least-squares cross-validation* (Racine, 2019, p. 74) by default. You can refer to extensions on a case by case basis.

:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
Cross-validation works by repeatedly holding out one observation, fitting the model on the rest, and measuring how well the model predicts the held-out point. A model that overfits (e.g., too many bins) will have low in-sample error but high cross-validation error. The bandwidth $\mathbf{H}$ that minimizes cross-validation error balances flexibility against overfitting.
:::

To make this concrete, we use cross-validation to choose the number of `x1` bins for the piecewise regression above.
We split the data into 5 folds, fit the model on 4 of them, and measure the squared prediction error on the held-out fold.

```{r}
## 5-fold cross-validation for the number of x1 bins
set.seed(1)
n_folds <- 5
fold_id <- sample(rep(1:n_folds, length.out=nrow(dat)))
bin_counts <- c(2, 4, 6, 8, 12, 16)

cv_mse <- sapply(bin_counts, function(L){
    bks <- seq(0, 20, length.out=L+1)
    fold_mse <- sapply(1:n_folds, function(k){
        train <- dat[fold_id != k, ]
        test  <- dat[fold_id == k, ]
        train$x1c <- cut(train$x1, bks)
        test$x1c  <- cut(test$x1, bks)
        reg_k <- lm(y ~ x1c/(x1 + x2), data=train)
        pred_k <- predict(reg_k, newdata=test)
        mean((test$y - pred_k)^2, na.rm=TRUE)
    })
    mean(fold_mse)
})

## Plot CV error against model complexity
plot(bin_counts, cv_mse, type='o', pch=16,
    xlab='Number of x1 bins', ylab='5-fold CV mean squared error',
    main=NA)
L_star <- bin_counts[which.min(cv_mse)]
abline(v=L_star, col=rgb(1, 0, 0, .8), lwd=2)
title(paste0('CV-optimal: ', L_star, ' bins'), font.main=1)
```

Too few bins cannot capture the curvature in `x1` (high bias); too many bins split the data so finely that each local regression is noisy (high variance).
The cross-validated error is smallest in between.

:::{.callout-tip icon=false collapse="true" title="Test Yourself"}
The cross-validation above selects bins for the piecewise model.
Adapt the code to instead select bins for the *regressogram* `lm(y ~ x1c*x2c)`, cutting both `x1` and `x2` into `L` bins.
What happens to the cross-validated error once `L` is large enough that some grid cells are empty? (Hint: empty cells give `NA` predictions.)
:::




## Hypothesis Testing

#### **Finite Differences**. {-}

We first summarize gradients — how $\hat{y}$ changes with each $x_p$ at a point — before testing them.

:::{.callout-important title="Key Definition" collapse="false"}
A *gradient* of a fitted multivariate model is the partial derivative $\partial \hat{y}/\partial x_p$ evaluated at a point. For a regressogram with no slope coefficients, it must be approximated by a finite difference between adjacent bins,
\begin{eqnarray}
\hat{\beta}_{p}(\mathbf{x}) &=& \frac{ \hat{y}(x_{1},...,x_{p}+ \frac{h_{p}}{2}...,x_{P})-\hat{y}(x_{1},...,x_{p}-\frac{h_{p}}{2}...,x_{P})}{h_{p}},
\end{eqnarray}
while a piecewise or local linear regression has within-bin slopes that are themselves gradient estimates.
:::

The gradient is useful because, unlike the single OLS coefficient, it varies with $\mathbf{x}$ — letting us see whether the marginal effect of $X_p$ is the same at small, medium, and large values of the other regressors.
We can compute it two ways:

1. For regressograms, approximate the gradient with a small finite difference between adjacent bins (formula above).

2. For split-sample regressions or local linear regressions, read the within-bin slope coefficients $\hat{\beta}_{p}(\mathbf{x})$ directly as gradient estimates.

After computing gradients, you can summarize them in various plots: a histogram of $\hat{\beta}_{p}(\mathbf{x})$, a scatterplot of $\hat{\beta}_{p}(\mathbf{x})$ against $X_{p}$, or the sorted gradients shown with their confidence intervals (Chaudhuri, 1999; Henderson et al., 2012).


You may also be interested in a particular gradient or a single summary statistic.
For example, a bivariate regressogram can estimate the marginal effect of $X_{1}$ at the means; $\hat{\beta_{1}}(\bar{\mathbf{x}}=[\bar{x_{1}}, \bar{x_{2}}])$.
You may also be interested in the mean of the marginal effects (sometimes said simply as "average effect"), which averages the marginal effect over all datapoints in the dataset: $1/n \sum_{i}^{n} \hat{\beta_{1}}(\mathbf{X}_{i})$, or the median marginal effect.
Such statistics are single numbers that can be presented similar to an OLS regression table, where each row corresponds to a variable and each cell has two elements: "mean gradient (sd gradient)".

We illustrate both methods on the models fit earlier in this chapter.
Method 1 differences the regressogram predictions; method 2 reads the slopes directly off the piecewise regression.

```{r}
##################
# Method 1: finite-difference gradients from the regressogram
##################

## Arrange the regressogram predictions into an x1-by-x2 grid
grid_yhat <- matrix(pred_df_rgrm$yhat,
    nrow=length(mids1), ncol=length(mids2))

## Approximate the x1-gradient by differencing adjacent x1 bins
bin_width1 <- diff(bks1)[1]
grad_x1 <- apply(grid_yhat, 2, diff)/bin_width1

## Summarize the gradient distribution
par(mfrow=c(1, 2))
hist(grad_x1, breaks=20, border=NA, freq=F,
    main=NA, xlab=expression(hat(beta)[1](x)))
title('Distribution of x1-gradients', font.main=1)
plot(rep(mids1[-1], ncol(grad_x1)), grad_x1,
    pch=16, col=grey(0, .5),
    xlab=expression(x[1]), ylab=expression(hat(beta)[1](x)),
    main=NA)
title('Gradient vs x1', font.main=1)
```

#### **Slope Coefficients**. {-}

```{r}
##################
# Method 2: slope coefficients from the piecewise regression
##################

## The within-bin slope on x1 is itself a gradient estimate
coef_preg <- coef(preg)
slope_x1 <- coef_preg[grep(':x1$', names(coef_preg))]

## Mean and standard deviation of the x1-gradient
c(mean=mean(slope_x1, na.rm=TRUE), sd=sd(slope_x1, na.rm=TRUE))
```

The mean gradient is a single-number summary, much like an OLS slope.
The standard deviation across bins measures how much the gradient changes over the data, which a single OLS coefficient cannot show.
To test whether a mean gradient differs from zero, bootstrap it exactly as in [Inference](https://jadamso.github.io/Rbooks/02_15_Inference.html#hypothesis-testing).

:::{.callout-note icon=false collapse="true" title="Must Know"}
A regressogram has no slope coefficients, so its gradient must be approximated by finite differences: the change in the predicted value between two adjacent bins, divided by the distance between them.
Suppose the regressogram predicts $\hat{y} = 8$ in one $x_1$-bin and $\hat{y} = 14$ in the next, and the bin midpoints are $2$ units apart.
The estimated $x_1$-gradient is $(14 - 8)/2 = 3$: moving one unit in $x_1$ raises the prediction by about $3$.
A piecewise or local linear regression avoids this approximation, because its within-bin slope coefficient *is* the gradient estimate.
:::

## Exercises

1. A regressogram bins each explanatory variable and estimates a separate mean within each bin combination. How does this differ from a standard OLS regression with the same variables? What advantage does it have when the true relationship is nonlinear, and what is its main drawback as the number of explanatory variables grows?

2. Using the simulated data from this chapter (`x1 <- seq(.1, 20, length.out = 1000)`, `x2 <- runif(1000, 0, 1)`, `y <- 3*exp(-2*x2 + 1.5*x1 - .1*x1^2)*x1 + rnorm(1000)`), build a regressogram by cutting `x1` into 5 bins and `x2` into 5 bins. Compute the mean of `y` within each of the 25 bin combinations. Which bin combination has the highest predicted value?

3. Using `lm()`, fit two models to the `mtcars` dataset with `mpg` as the outcome: (a) a standard linear model `mpg ~ wt + hp`, and (b) a regressogram model where `wt` and `hp` are each cut into 4 bins using `cut()` and interacted (`mpg ~ cut(wt,4) * cut(hp,4)`). Compare the $\hat{R}^2$ of each model. Plot the predicted values from both models against the observed `mpg`.

#### **Further Reading**. {-}

* <https://cran.r-project.org/web/packages/np/vignettes/np.pdf> -- the `np` package vignette for kernel and local-linear nonparametric regression in R.
* <https://www.econometrics-with-r.org/11-rwmad.html> -- nonparametric methods and their connection to regression, including bin-smoothers.


#### Recall {-}

This chapter generalized regressograms, piecewise regression, and cross-validation to the multivariate setting, and showed two ways to summarize gradients: finite differences from the regressogram, and within-bin slopes from the piecewise model. The 5-fold CV exercise on `y ~ x1c/(x1 + x2)` made the bias-variance trade-off concrete: we swept the number of `x1` bins through $\{2, 4, 6, 8, 12, 16\}$, watched the cross-validated MSE bottom out at the CV-optimal $L^{*}$, and saw it rise again as we kept adding bins. In the next chapter we step back from estimation and ask what happens when the data themselves are dependent over time, across space, or through market equilibrium — observational complications that violate the independence assumption used here.
